### 2022-08-25 example.Rmd
example Rmd

---
title: "Untitled"
author: "username"
date: '2022-08-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-02-28 rpkgs.R
update

@@ -1,12 +1,0 @@
-install.packages("systemfonts")
-install.packages("gplots")
-install.packages("ggplot2")
-install.packages("rmarkdown")
-install.packages("knitr")
-install.packages("kableExtra")
-
-packageurl <- "https://cran.r-project.org/src/contrib/Archive/rvcheck/rvcheck_0.1.8.tar.gz"
-install.packages(packageurl, repos=NULL, type="source")
-
-install.packages("BiocManager")
-BiocManager::install("edgeR")

### 2023-03-01 example.Rmd
Dockerfile done, now working on Rmd

@@
 ---
-title: "Untitled"
-author: "username"
-date: '2022-08-25'
-output: html_document
+title: "Enrichment analysis with gold standard reproducibility."
+author: "Mark Ziemann & Anusuiya Bora"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    toc: true
+    toc_float: true
+    code_folding: hide
+    fig_width: 7
+    fig_height: 7
+theme: cosmo
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+Source: https://github.com/markziemann/udocker_r_example
+
+Alt titles:
+
+* "Recipes for extremely reproducible enrichment analysis"
+
+## Intro
+
+Previously my team have written about methodological problems in enrichment analysis.
+These include statistical problems, like lack of FDR correction, lack of correct background
+ and poor reporting of methodological details (Wijesooriya et al, 2022).
+
+Therefore there is a need for community-agreed guidelines for conducting and presenting
+enrichment analysis work, which is something that has been noted by others
+(Timmons et al, 2015; Reimand et al 2019), but has only recently been addressed (Zhao & Rhee 2023).
+
+The purpose of this work is to offer a postive examplar of an enrichment analysis
+that fills all criteria:
+
+1. Statistically rigourous.
+
+2. Well documented in the manuscript and public code.
+
+3. Maximally reproducible.
+
+While rigour and good documentation are features of enrichment analyses in high quality specialist
+genomics journals, reproducibility remains a distinct weakness.
+
+When trying to reproduce or replicate the work of others we are often frustrated by
+the unavailability of input data, of functional annotations, of code, and full descriptions
+of results.
+
+For example, while conducting a replication study of 20 journal articles, we found that
+one of the most frequently used web-tools, DAVID bioinformatics, was going to move from
+version 6.7 to version 6.8, with no option to continue using v6.7.
+
+Changes such as these mean that the results of many journal articles can no longer be
+independently replicated, only a couple of years after original publication.
+
+As time passes and software versions update, it will become ever more difficult to replicate
+previous analyses.
+This means that it will be impossible to audit many research articles if those findings come
+under question.
+
+Although it doesn't need to be this way due to the precise nature of modern computers and software.
+In order for bioinformatics analysis to be 100% reproducible requires a few key elements.
+These are:
+
+* Input data
+
+* Set of instructions (analysis code).
+
+* Operating system
+
+With these things working together, it is possible for a data analysis workflow to be 100%
+replicable and auditable for many years into the future.
+
+To achieve this requires some planning:
+
+1. Input data needs to be deposited to a public and persistent location.
+
+2. Analysis code packaged in a working Docker container.
+
+A Docker container is a minimal operating system that has the necessary software to complete a task.
+Docker is used extensively in web development, to rapidly set up many modular services into
+full stacks of functionality.
+For example using Docker containers, a website front end and back end can be set up once, and then
+deployed at huge scale (eg thousands of instances) with minimal work involved.
+
+In addition, Docker containers ensure consistent behaviour across different types of computing
+infrastructure, so it doesn't matter whether the container is run on a linux workstation,
+desktop windows PC or cloud server, the end results should be identical.
+
+While there have been many general guides to computatonal reproducibility in the literature,
+there is a need for boilerplate code templates that researchers can quickly adapt/tweak for their own
+projects.
+
+The demand for this is expected to be huge, as there are massive numbers of published articles featuring
+such analyses.
+Just in 2022, there were 9,917 PubMed articles with the keywords pathway/enrichment/ontology analysis
+in the title or abstract alone.
+
+The more rigourous, well documented and reproducible these studies are, we will find less wasted
+research resources and greater trust/support from the general community.
+
+## Data set being analysed
+
+For this guide I will be using bulk RNA-seq data from a previous study, which is
+deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
+The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.
+
+```{r,libs}
+
+suppressPackageStartupMessages({
+  library("getDEE2")
+  library("DESeq2")
+  library("kableExtra")
+  library("vioplot")
+  library("clusterProfiler")
+  library("eulerr")
+})
+
 ```
 
-## R Markdown
+The raw data have been processed by the DEE2 project, and the summary gene expression counts are
+available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
+(Ziemann et al, 2019).
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+```{r,fetch}
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+mdat <- getDEE2Metadata("hsapiens")
 
-```{r cars}
-summary(cars)
+# get sample sheet
+ss <- subset(mdat,SRP_accession=="SRP038101")
+
+# fetch the whole set of RNA-seq data
+x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
+mx <- x$GeneCounts
+rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
+dim(mx)
+
+# aza no filtering
+ss$trt <- grepl("Treated",ss$Experiment_title)
+
+ss %>%
+  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
+  kable_paper("hover", full_width = F)
+
 ```
 
-## Including Plots
+## Analysis of read counts
 
-You can also embed plots, for example:
+Although I know this data set is good as I was involved in the project,
+it is a good idea to show at least the number of reads for each sample.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r,counts1}
+
+par(mar=c(5,7,5,1))
+barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")
+
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+Now we can have a look at the read count distribution for one of the samples.
+
+```{r,dist1}
+
+# look at the distribution of counts
+l <- lapply(1:ncol(mx),function(i) log10( mx[,i] +1 ))
+names(l) <- colnames(mx)
+l <- rev(l)
+
+hist(l[[6]],main="Distribution of read counts in SRR1171523",
+  xlab="log10(readcounts+1)",breaks=30)
+
+abline(v=1,col="red",lty=2)
+
+```
+
+We can see that most genes have a very low read count, consistent
+with most genes being undetectable in any sample.
+
+Based on this typical distribution of read counts, I would consider
+any gene with fewer than 10 reads as not detected.
+This could be considered a strict detection threshold.
+Setting a strict threshold like this could be helpful in selecting
+genes that are likely to validate with qRT-PCR.
+
+I will run DESeq2 with and without this threshold to see how
+it impacts the results.
+
+## Differential expression analysis with and without filtering
+
+To see whether this has any effect on downstream analysis, DESeq2 will
+be run on the same dataset without any filtering, with light filtering and with
+strict filtering.
+
+Here I define light filter threshold as the lowest baseline expression level that
+DE genes start to appear.
+I know this contrast will yield thousands of DE genes, so this is a reasonable
+way to set the detection threshold.
+The DE genes with lowest baseline expression showed mean read counts of about 2.
+
+```{r,de1}
+
+# DESeq2 without any fitering
+dds <- DESeqDataSetFromMatrix(countData = mx , colData = ss, design = ~ trt )
+res <- DESeq(dds)
+z <- results(res)
+vsd <- vst(dds, blind=FALSE)
+zz <- cbind(as.data.frame(z),mx)
+de <- as.data.frame(zz[order(zz$pvalue),])
+
+sig <- subset(de,padj<0.05)
+
+head(de,10) %>%
+  kbl(caption="Top DE genes for Aza treatment (unfiltered)") %>%
+  kable_paper("hover", full_width = F)
+
+DET=nrow(mx)
+NSIG=nrow(sig)
+NUP=nrow(subset(sig,log2FoldChange>0))
+NDN=nrow(subset(sig,log2FoldChange<0))
+
+HEADER=paste(DET,"genes included;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
+
+plot(log10(de$baseMean) ,de$log2FoldChange,
+  cex=0.6,pch=19,col="darkgray",
+  main="no detection filter",
+  xlab="log10(basemean)",ylab="log2(fold change)")
+
+points(log10(sig$baseMean) ,sig$log2FoldChange,
+  cex=0.6,pch=19,col="red")
+
+mtext(HEADER)
+
+abline(v=1,lty=2)
+
+deup <- rownames(subset(de,padj<=0.05 & log2FoldChange>0))
+deup <- unique(sapply(strsplit(deup," "),"[[",2))
+
+dedn <- rownames(subset(de,padj<=0.05 & log2FoldChange<0))
+dedn <- unique(sapply(strsplit(dedn," "),"[[",2))
+
+all <- rownames(de)
+all <- unique(sapply(strsplit(all," "),"[[",2))
+
+```
+
+look at the lowest basemean genes
+
+```{r,low}
+
+sig <- sig[order(sig$baseMean),]
+
+head(sig,10) %>%
+  kbl(caption="lowest basemean genes") %>%
+  kable_paper("hover", full_width = F)
+
+```
+
+The counts look quite low.
+Not sure that these DEGs could be validated with qRT-PCR.
+
+## Now run DESeq2 with filtering
+
+```{r,de2}
+
+mxf <- mx[which(rowMeans(mx)>=10),]
+dim(mxf)
+dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
+res <- DESeq(dds)
+z <- results(res)
+vsd <- vst(dds, blind=FALSE)
+zz <-cbind(as.data.frame(z),mxf)
+def <-as.data.frame(zz[order(zz$pvalue),])
+
+head(def,10) %>%
+  kbl(caption="Top DE genes for Aza treatment (filtered)") %>%
+  kable_paper("hover", full_width = F)
+
+sigf <- subset(def,padj<=0.05)
+
+DET=nrow(mxf)
+NSIG=nrow(sigf)
+NUP=nrow(subset(sigf,log2FoldChange>0))
+NDN=nrow(subset(sigf,log2FoldChange<0))
+
+HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
+
+plot(log10(def$baseMean) ,def$log2FoldChange,
+  cex=0.6,pch=19,col="darkgray",
+  main="strict filter",
+  xlab="log10(basemean)",ylab="log2(fold change)")
+
+points(log10(sigf$baseMean) ,sigf$log2FoldChange,
+  cex=0.6,pch=19,col="red")
+
+mtext(HEADER)
+
+defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
+defup <- unique(sapply(strsplit(defup," "),"[[",2))
+
+defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
+defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
+
+bg <- rownames(def)
+bg <- unique(sapply(strsplit(bg," "),"[[",2))
+
+```
+
+## Now run ORA based enrichment analysis
+
+I will run enrichment analysis using the correct correct background list,
+and compare the results to the incorrect whole genome background.
+
+```{r,gs}
+
+if (! file.exists("ReactomePathways.gmt")) {
+  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
+    destfile="ReactomePathways.gmt.zip")
+  unzip("ReactomePathways.gmt.zip")
+}
+
+genesets2 <- read.gmt("ReactomePathways.gmt")
+
+```
+
+### ORA with correct background list.
+
+Firstly with the upregulated genes
+
+```{r,ora2up}
+
+ora2_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora2_up$geneID <- NULL
+ora2_up <- subset(ora2_up,p.adjust<0.05 & Count >=10)
+ora2_ups <- rownames(ora2_up)
+
+gr <- as.numeric(sapply(strsplit(ora2_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora2_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora2_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora2_up$BgRatio,"/"),"[[",2))
+
+ora2_up$es <- gr/br
+ora2_up <- ora2_up[order(-ora2_up$es),]
+ora2_up$Description=NULL
+
+head(ora2_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (filtered)") %>%
+  kable_paper("hover", full_width = F)
+
+topup2 <- rev(head(ora2_up$es,20))
+names(topup2) <- rev(head(ora2_up$ID,20))
+
+```
+
+Now with the downregulated genes
+
+```{r,ora2dn}
+
+ora2_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora2_dn$geneID <- NULL
+ora2_dn <- subset(ora2_dn,p.adjust<0.05 & Count >=10)
+ora2_dns <- rownames(ora2_dn)
+
+gr <- as.numeric(sapply(strsplit(ora2_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora2_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora2_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora2_dn$BgRatio,"/"),"[[",2))
+
+ora2_dn$es <- gr/br
+ora2_dn <- ora2_dn[order(-ora2_dn$es),]
+ora2_dn$Description=NULL
+
+head(ora2_dn) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (filtered)") %>%
+  kable_paper("hover", full_width = F)
+
+topdn2 <- head(ora2_dn$es,20)
+names(topdn2) <- head(ora2_dn$ID,20)
+
+```
+
+Make a barplot
+
+```{r,topbarplot2}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",20),rep("red",20))
+
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES")
+
+mtext("correct background")
+
+```
+
+### ORA with whole genome background list.
+
+Firstly with the upregulated genes
+
+```{r,ora1up}
+
+ora1_up <- as.data.frame(enricher(gene = defup ,
+  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora1_up$geneID <- NULL
+ora1_up <- subset(ora1_up,p.adjust<0.05 & Count >=10)
+ora1_ups <- rownames(ora1_up)
+
+gr <- as.numeric(sapply(strsplit(ora1_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora1_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora1_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora1_up$BgRatio,"/"),"[[",2))
+
+ora1_up$es <- gr/br
+ora1_up <- ora1_up[order(-ora1_up$es),]
+ora1_up$Description=NULL
+
+head(ora1_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (unfiltered)") %>%
+  kable_paper("hover", full_width = F)
+
+topup1 <- rev(head(ora1_up$es,20))
+names(topup1) <- rev(head(ora1_up$ID,20))
+
+```
+
+Now with the downregulated genes
+
+```{r,ora1dn}
+
+ora1_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora1_dn$geneID <- NULL
+ora1_dn <- subset(ora1_dn,p.adjust<0.05 & Count >=10)
+ora1_dns <- rownames(ora1_dn)
+
+gr <- as.numeric(sapply(strsplit(ora1_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora1_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora1_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora1_dn$BgRatio,"/"),"[[",2))
+
+ora1_dn$es <- gr/br
+ora1_dn <- ora1_dn[order(-ora1_dn$es),]
+ora1_dn$Description=NULL
+
+head(ora1_dn) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (unfiltered)") %>%
+  kable_paper("hover", full_width = F)
+
+topdn1 <- head(ora1_dn$es,20)
+names(topdn1) <- head(ora1_dn$ID,20)
+
+```
+
+Make a barplot
+
+```{r,topbarplot1}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",20),rep("red",20))
+
+barplot(c(topdn1,topup1),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES")
+
+mtext("whole genome background")
+
+```
+
+## Compare pathway results
+
+```{r,venn1,fig.width=6,fig.height=6}
+
+v1 <- list("correct up"=ora2_ups, "wg up"=ora1_ups)
+
+v2 <- list("correct dn"=ora2_dns, "wg dn"=ora1_dns)
+
+par(mar=c(10,10,10,10))
+
+par(mfrow=c(2,1))
+plot(euler(v1),quantities = TRUE)
+plot(euler(v2),quantities = TRUE)
+
+```
+
+Jaccard index
+
+```{r,jaccard}
+
+jaccard <- function(a, b) {
+    intersection = length(intersect(a, b))
+    union = length(a) + length(b) - intersection
+    return (intersection/union)
+}
+
+correct <- c(ora2_ups,ora2_dns)
+incorrect <- c(ora1_ups,ora1_dns)
+
+jaccard(correct,incorrect)
+
+```
+
+## Session information
+
+For reproducibility
+
+<br><details><summary><b>Click HERE to show session info</b></summary><br><p>
+
+```{r,session}
+
+sessionInfo()
+
+```
+
+</details>
+
+## References

### 2023-03-03 manuscript.Rmd
ms

--- 2022-08-25 example.Rmd
@@
 ---
-title: "Untitled"
-author: "username"
-date: '2022-08-25'
-output: html_document
+title: "Enrichment analysis with gold standard reproducibility"
+author: "Mark Ziemann and Anusuiya Bora"
+date: "`r Sys.Date()`"
+output:
+  rmdformats::robobook:
+  html_document:
+    toc: true
+    toc_float: true
+    fig_width: 7
+    fig_height: 5
+theme: cosmo
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
+<b>Alt titles:<b>
 
-## R Markdown
+* "Recipes for extremely reproducible enrichment analysis"
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+* "A guide to completely reproducible enrichment analysis"
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+Mark Ziemann<sup>1</sup>*, Anusuiya Bora<sup>2</sup>
 
-```{r cars}
-summary(cars)
-```
+**Affiliations**
 
-## Including Plots
+1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
 
-You can also embed plots, for example:
+2. Vellore Institute of Technology, Vellore, India.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
+(*) Corresponding author: m.ziemann@deakin.edu.au
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+## Abstract
+
+asddsadasd
+
+## Introduction
+
+Enrichment analysis is one of the most used methods in computational biology,
+and involves the summarisation of omics data to reflect biological changes, such as
+the detection of pathway activation in development or disease [1,2](Slonim 2002;Khatri et al, 2012). 
+While enrichment analysis is undoubtedly a key method for the interpretation of large
+data sets, there are concerns that inadvertant misuse of these methods leads to 
+unreliable results.
+These potential problems include (i) the lack of correct background for over-representation tests,
+(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
+[3-6](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
+
+Another emerging problem is that while computational research is theoretically amenable to complete
+reproducibility, in reality this is rarely achieved in computation-intensive research due to lack
+of detail provided in methods sections, lack of shared data and lack of code [7](Peng 2011).
+
+The lack of reproducibility is particularly acute for enrichment analyses, as web-based and
+point-and-click tools appear to be far more popular than computer script based methods 
+[5](Wijesooriya et al, 2022).
+Web-based tools are a particular concern, as it is known that algorithms and functional annotation
+sets undergo regular updates, but the versions are not regularly reported nor captured, 
+making later replication difficult if not impossible [5](Wijesooriya et al, 2022).
+
+A lack of reproducibility does not mean a study is invalid, however it cannot be ruled out that
+mistakes have been made in the enrichment analysis and that the results are biased or false.
+As omics data are becoming more routine in the drug development process and are used to
+inform clinical practice, reproducibility should become a minimum requirement for 
+bioinformatics analysis of omics data.
+
+Although it doesn't need to be this way due to the precise nature of modern computers and software.
+In order for bioinformatics analysis to be 100% reproducible requires a few key elements.
+These are:
+
+* Input data
+
+* Set of instructions (analysis code).
+
+* Operating system
+
+With these things working together, it is possible for a data analysis workflow to be 100%
+replicable and auditable for many years into the future.
+
+To achieve this requires some planning:
+
+1. Input data needs to be deposited to a public and persistent location.
+
+2. Analysis code packaged in a working Docker container.
+
+A Docker container is a minimal operating system that has the necessary software to complete a task.
+Docker is used extensively in web development, to rapidly set up many modular services into
+full stacks of functionality.
+For example using Docker containers, a website front end and back end can be set up once, and then
+deployed at huge scale (eg thousands of instances) with minimal work involved.
+
+In addition, Docker containers ensure consistent behaviour across different types of computing
+infrastructure, so it doesn't matter whether the container is run on a linux workstation,
+desktop windows PC or cloud server, the end results should be identical.
+
+While there have been many general guides to computatonal reproducibility in the literature,
+there is a need for boilerplate code templates that researchers can quickly adapt/tweak for their own
+projects.
+
+The demand for this is expected to be huge, as there are massive numbers of published articles featuring
+such analyses.
+Just in 2022, there were 9,917 PubMed articles with the keywords pathway/enrichment/ontology analysis
+in the title or abstract alone.
+
+The more rigourous, well documented and reproducible these studies are, we will find less wasted
+research resources and greater trust/support from the general community.
+
+## Source
+
+Source: https://github.com/markziemann/udocker_r_example
+
+Alt titles:
+
+* "Recipes for extremely reproducible enrichment analysis"

### 2023-03-05 example.Rmd
big update

@@
 ---
-title: "Enrichment analysis with gold standard reproducibility."
+title: "An example of extremely reproducible enrichment analysis."
 author: "Mark Ziemann & Anusuiya Bora"
@@
 Alt titles:
-
-* "Recipes for extremely reproducible enrichment analysis"
 
@@
   library("kableExtra")
-  library("vioplot")
   library("clusterProfiler")

### 2023-03-05 manuscript.Rmd
big update

@@
 ---
-title: "Enrichment analysis with gold standard reproducibility"
+title: "A recipe for extremely reproducible enrichment analysis"
+Enrichment analysis with gold standard reproducibility"
 author: "Mark Ziemann and Anusuiya Bora"
@@
 
-* "Recipes for extremely reproducible enrichment analysis"
+* "Enrichment analysis with gold standard reproducibility"
 

### 2023-03-06 example.Rmd
progress

--- 2022-08-25 example.Rmd
@@
 ---
-title: "Untitled"
-author: "username"
-date: '2022-08-25'
-output: html_document
+title: "An example of extremely reproducible enrichment analysis."
+author: "Mark Ziemann & Anusuiya Bora"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    toc: true
+    toc_float: true
+    code_folding: hide
+    fig_width: 7
+    fig_height: 7
+theme: cosmo
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+Source: https://github.com/markziemann/enrichment_recipe
+
+## Introduction
+
+This guide is a Rmarkdown script that conducts differential expression and enrichment analysis,
+which are very popular workflows for transcriptome data.
+
+This is meant to be a boilerplate template, which you can remix and modify to suit your analytical
+needs.
+
+In the code chunk below called `libs`, you can add and remove required R library dependancies.
+Check that the libraries listed here match the Dockerfile, otherwise you might get errors.
+
+```{r,libs}
+
+suppressPackageStartupMessages({
+  library("getDEE2")
+  library("DESeq2")
+  library("kableExtra")
+  library("clusterProfiler")
+  library("fgsea")
+  library("eulerr")
+})
+
 ```
 
-## R Markdown
+For this guide I will be using bulk RNA-seq data from a previous study, which is
+deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
+The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+The raw data have been processed by the DEE2 project, and the summary gene expression counts are
+available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
+(Ziemann et al, 2019).
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
+storage drive.
 
-```{r cars}
-summary(cars)
+```{r,fetch}
+
+mdat <- getDEE2Metadata("hsapiens")
+
+# get sample sheet
+ss <- subset(mdat,SRP_accession=="SRP038101")
+
+# fetch the whole set of RNA-seq data
+x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
+mx <- x$GeneCounts
+rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
+dim(mx)
+
+# aza no filtering
+ss$trt <- grepl("Treated",ss$Experiment_title)
+
+ss %>%
+  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
+  kable_paper("hover", full_width = F)
+
 ```
 
-## Including Plots
+## Data quality control
 
-You can also embed plots, for example:
+QC is important, even if you are using public transcriptome data.
+For RNA-seq it is a good idea to show the number of reads for each sample.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+```{r,counts1}
+
+par(mar=c(5,7,5,1))
+barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")
+
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+## Differential expression analysis
+
+I will be using DESeq2 for DE analysis, the count matrix is prefiltered using
+a detection threshold of 10 reads per sample across all samples.
+All genes that meet the detection threshold will comprise the background list.
+
+```{r,de1}
+
+mxf <- mx[which(rowMeans(mx)>=10),]
+dim(mxf)
+dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
+res <- DESeq(dds)
+z <- results(res)
+vsd <- vst(dds, blind=FALSE)
+zz <-cbind(as.data.frame(z),mxf)
+def <-as.data.frame(zz[order(zz$pvalue),])
+
+head(def,10) %>%
+  kbl(caption="Top DE genes for Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+sigf <- subset(def,padj<=0.05)
+
+DET=nrow(mxf)
+NSIG=nrow(sigf)
+NUP=nrow(subset(sigf,log2FoldChange>0))
+NDN=nrow(subset(sigf,log2FoldChange<0))
+
+HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")
+
+plot(log10(def$baseMean) ,def$log2FoldChange,
+  cex=0.6,pch=19,col="darkgray",
+  main="strict filter",
+  xlab="log10(basemean)",ylab="log2(fold change)")
+
+points(log10(sigf$baseMean) ,sigf$log2FoldChange,
+  cex=0.6,pch=19,col="red")
+
+mtext(HEADER)
+
+defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
+defup <- unique(sapply(strsplit(defup," "),"[[",2))
+
+defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
+defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
+
+bg <- rownames(def)
+bg <- unique(sapply(strsplit(bg," "),"[[",2))
+
+```
+
+In the next sections I will run enrichment analysis with over-representation test
+and compare it to functional class scoring.
+
+## Enrichment using over-representation test
+
+I've compiled a reporting checklist:
+
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | Reactome (current date) |
+| Tool used | ClusterProfiler (check version at foot of report) |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+
+```{r,gs}
+
+if (! file.exists("ReactomePathways.gmt")) {
+  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
+    destfile="ReactomePathways.gmt.zip")
+  unzip("ReactomePathways.gmt.zip")
+}
+
+genesets2 <- read.gmt("ReactomePathways.gmt")
+gsets <- gmtPathways("ReactomePathways.gmt")
+
+```
+
+Firstly with the upregulated genes
+
+```{r,oraup}
+
+ora_up <- as.data.frame(enricher(gene = defup ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_up$geneID <- NULL
+ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
+ora_ups <- rownames(ora_up)
+
+gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
+
+ora_up$es <- gr/br
+ora_up <- ora_up[order(-ora_up$es),]
+ora_up$Description=NULL
+
+head(ora_up) %>%
+  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topup2 <- rev(head(ora_up$es,20))
+names(topup2) <- rev(head(ora_up$ID,20))
+
+```
+
+Now with the downregulated genes
+
+```{r,oradn}
+
+ora_dn <- as.data.frame(enricher(gene = defdn ,
+  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
+
+ora_dn$geneID <- NULL
+ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
+ora_dns <- rownames(ora_dn)
+
+gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))
+
+br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
+  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))
+
+ora_dn$es <- gr/br
+ora_dn <- ora_dn[order(-ora_dn$es),]
+ora_dn$Description=NULL
+
+head(ora_dn) %>%
+  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+topdn2 <- head(ora_dn$es,20)
+names(topdn2) <- head(ora_dn$ID,20)
+
+```
+
+Make a barplot
+
+```{r,topbarplot2}
+
+par(mar=c(5,20,5,1))
+
+cols <- c(rep("blue",20),rep("red",20))
+
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES")
+
+mtext("ORA test")
+
+```
+
+## Enrichment analysis with functional class scoring
+
+```{r,fgsea}
+
+def2 <- def
+def2$genename <- sapply(strsplit(rownames(def2)," "),"[[",2)
+def2 <- def2[,c("stat","genename")]
+def2 <- aggregate(. ~ genename, def2, sum)
+stat <- def2$stat
+names(stat) <- def2$genename
+fgseaRes <- fgsea(pathways=gsets, stats=stat, minSize=10, nPermSimple = 10000)
+fgseaRes <- fgseaRes[order(fgseaRes$pval),]
+fgsea_up <- subset(fgseaRes,padj<0.05 & ES>0)
+fgsea_ups <- fgsea_up$pathway
+fgsea_dn <- subset(fgseaRes,padj<0.05 & ES<0)
+fgsea_dns <- fgsea_dn$pathway
+
+```
+
+## Compare pathway results
+
+```{r,venn1,fig.width=6,fig.height=6}
+
+v1 <- list("ORA up"=ora_ups, "FCS up"=fgsea_ups)
+v2 <- list("ORA dn"=ora_dns, "FCS dn"=fgsea_dns)
+par(mar=c(10,10,10,10))
+par(mfrow=c(2,1))
+plot(euler(v1),quantities = TRUE)
+plot(euler(v2),quantities = TRUE)
+
+```
+
+Jaccard index
+
+```{r,jaccard}
+
+jaccard <- function(a, b) {
+    intersection = length(intersect(a, b))
+    union = length(a) + length(b) - intersection
+    return (intersection/union)
+}
+
+ora <- c(ora_ups,ora_dns)
+fcs <- c(fgsea_ups,fgsea_dns)
+
+jaccard(ora,fcs)
+
+```
+
+## Scatterplot
+
+## Session information
+
+For reproducibility
+
+<br><details><summary><b>Click HERE to show session info</b></summary><br><p>
+
+```{r,session}
+
+sessionInfo()
+
+```
+
+</details>
+
+## References

### 2023-03-08 example.Rmd
work

@@
 | --- | --- |
-| Origin of gene sets | Reactome (current date) |
+| Origin of gene sets | Reactome (2023-03-06) |
 | Tool used | ClusterProfiler (check version at foot of report) |
@@
 | Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 10 |
 
@@
 
-if (! file.exists("ReactomePathways.gmt")) {
-  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
-    destfile="ReactomePathways.gmt.zip")
-  unzip("ReactomePathways.gmt.zip")
-}
-
-genesets2 <- read.gmt("ReactomePathways.gmt")
-gsets <- gmtPathways("ReactomePathways.gmt")
+genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")
+gsets <- gmtPathways("ReactomePathways_2023-03-06.gmt")
 
@@
 
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | Reactome (2023-03-06) |
+| Tool used | FGSEA (check version at foot of report) |
+| Statistical test used | Kolmogorov-Smirnov test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 10 |
+
 ```{r,fgsea}
@@
 </details>
-
-## References

### 2023-03-09 example.Rmd
working on docs

@@
 ```{r,gs}
-
+# from https://reactome.org/download/current/ReactomePathways.gmt.zip
 genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")

### 2023-03-09 manuscript.Rmd
mswork

@@
 title: "A recipe for extremely reproducible enrichment analysis"
-Enrichment analysis with gold standard reproducibility"
 author: "Mark Ziemann and Anusuiya Bora"
@@
 [5](Wijesooriya et al, 2022).
-Web-based tools are a particular concern, as it is known that algorithms and functional annotation
-sets undergo regular updates, but the versions are not regularly reported nor captured, 
-making later replication difficult if not impossible [5](Wijesooriya et al, 2022).
+Web-based tools are a potential concern, as it is known that algorithms and functional annotation
+sets undergo regular updates, but the historical versions are typically not archived, nor
+are versions regularly reported in publications, making later replication difficult if not
+impossible in most cases [5](Wijesooriya et al, 2022).
+Indeed, in some cases the entire web resource can become unavailable, and hence irreproducible due to
+a phenomenon known as link decay [8](Hennessey & Ge 2013).
 
-A lack of reproducibility does not mean a study is invalid, however it cannot be ruled out that
-mistakes have been made in the enrichment analysis and that the results are biased or false.
+A lack of reproducibility does not mean a study is invalid, however a lack of auditability means
+it cannot be ruled out that mistakes have been made in the enrichment analysis and that the results
+are biased or false.
 As omics data are becoming more routine in the drug development process and are used to
@@
 
+If research data and code are shared, reproducibility still cannot be guaranteed due to the
+differences in various computing environments such as Windows, Linux, Mac, which provide varying
+sets of libraries for programming languages to use.
+Even if the operating system is specified in the research methods, reproducibility can be difficult if the
+operating system is old or obscure, and where dependancies are unavailable.
+To address this, many articles have been written to endorse the use of "virtual machines" or "Docker containers"
+to encapsulate the operating system and dependancies along with the research code to facilitate
+a high degree of reproducibility [9-20](Stein 2010; Dudley & Butte 2010; Sandve et al, 2013;
+Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016; da Veiga Leprevost et al, 2017;
+Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
+
+Despite the availability of such tools for over a decade and the outsized impact such analysis has
+on the conclusions of an omics study, enrichment analyses using fully reproducible virtual machines
+or containers remains a miniscule fraction of the overall share of enrichment analyses.
+A small number of enrichment analysis applications have been developed to have beneficial properties
+of being "containerised" and with archived hostorical versions [21,22](Ge et al, 2020; Perampalam & Dick 2020).
+This helps reproducibility greatly, but is insufficient as it lacks information about how the input data was
+used and a record of any parameter selections.
+
+Therefore the most complete reproducibility solution is to use containerisation to provide a scripted
+workflow with prescribed links to research data, external databases and complete record of parameter selection.
+Designed correctly, these Docker images could encapsulate bioinformatics workflows and ensure their
+computational reproducibility for decades to come, with just one or a few simple commands.
+
+The barriers to entry include:
+
+* The fact that conventional web based and GUI tools are easier to use
+
+* It adds an overhead of effort to the project
+
+* There's no time to learn new stuff
+
+* It's not a requirement of the journal
+
+We address this by:
+
+* providing working templates for docker image and Rmarkdown scripts for common transcriptome and enrichment analysis
+routines.
+
+* Providing step-by-step guides to help users customise these templates for their own work, deploy the analysis,
+and verify the results.
+
+To highlight the need for extremely reproducible enrichment analysis, we first investigate the computational
+reproducibility of enrichment analyses shown in a sample of 20 journal articles published in 2019.
+
+## Methods
+
+### A systematic assessment of reproducibility in enrichment analysis
+
+<Anusuiya to populate>
+
+### A container based approach to addressing reproducibility of enrichment analysis
+
+Here describe the set up of the system.
+The following text is a bit random and needs to be culled.
 Although it doesn't need to be this way due to the precise nature of modern computers and software.
@@
 
+## Concluding remarks
+
 While there have been many general guides to computatonal reproducibility in the literature,
@@
 
-## Source
+## Bibliography
 
-Source: https://github.com/markziemann/udocker_r_example
-
-Alt titles:
-
-* "Recipes for extremely reproducible enrichment analysis"
+1.
+Slonim DK.
+From patterns to pathways: gene expression data analysis comes of age.
+Nat Genet [Internet].
+2002 [cited 2023 Mar 3];32 Suppl(S4):502–8.
+Available from: https://www.nature.com/articles/ng1033z
+2.
+Khatri P, Sirota M, Butte AJ.
+Ten years of pathway analysis: current approaches and outstanding challenges.
+PLoS Comput Biol [Internet].
+2012 [cited 2023 Mar 3];8(2):e1002375.
+Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
+3.
+Timmons JA, Szkop KJ, Gallagher IJ.
+Multiple sources of bias confound functional enrichment analysis of global -omics data.
+Genome Biol [Internet].
+2015;16(1):186.
+Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
+4.
+Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al.
+Pathway enrichment analysis and visualization of omics data using g:Profiler, GSEA, Cytoscape and EnrichmentMap.
+Nat Protoc [Internet].
+2019;14(2):482–517.
+Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
+5.
+Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M.
+Urgent need for consistent standards in functional enrichment analysis.
+PLoS Comput Biol [Internet].
+2022;18(3):e1009935.
+Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
+6.
+Zhao K, Rhee SY.
+Interpreting omics data with pathway enrichment analysis.
+Trends Genet [Internet].
+2023 [cited 2023 Mar 1];0(0).
+Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0168952523000185%3Fshowall%3Dtrue
+7.  Peng RD. Reproducible research in computational science. Science [Internet]. 2011;334(6060):1226–7. Available from: http://dx.doi.org/10.1126/science.1213847
+8.
+Hennessey J, Ge S.
+A cross disciplinary study of link decay and the effectiveness of mitigation techniques.
+BMC Bioinformatics [Internet].
+2013;14 Suppl 14(S14):S5.
+Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
+9.  Stein LD. The case for cloud computing in genome informatics. Genome Biol [Internet]. 2010;11(5):207. Available from: http://dx.doi.org/10.1186/gb-2010-11-5-207
+10.  Dudley JT, Butte AJ. In silico research in the era of cloud computing. Nat Biotechnol [Internet]. 2010;28(11):1181–5. Available from: http://dx.doi.org/10.1038/nbt1110-1181
+11.
+Sandve GK, Nekrutenko A, Taylor J, Hovig E.
+Ten simple rules for reproducible computational research.
+PLoS Comput Biol [Internet].
+2013;9(10):e1003285.
+Available from: http://dx.doi.org/10.1371/journal.pcbi.1003285
+12.  Piccolo SR, Frampton MB. Tools and techniques for computational reproducibility. Gigascience [Internet]. 2016;5(1). Available from: http://dx.doi.org/10.1186/s13742-016-0135-4
+13.
+Grüning B, Chilton J, Köster J, Dale R, Soranzo N, van den Beek M, et al.
+Practical computational reproducibility in the life sciences.
+Cell Syst [Internet].
+2018;6(6):631–5.
+Available from: http://dx.doi.org/10.1016/j.cels.2018.03.014
+14.
+Lewis J, Breeze CE, Charlesworth J, Maclaren OJ, Cooper J.
+Where next for the reproducibility agenda in computational biology?
+BMC Syst Biol [Internet].
+2016;10(1).
+Available from: http://dx.doi.org/10.1186/s12918-016-0288-x
+15.
+da Veiga Leprevost F, Grüning BA, Alves Aflitos S, Röst HL, Uszkoreit J, Barsnes H, et al.
+BioContainers: an open-source and community-driven framework for software standardization.
+Bioinformatics [Internet].
+2017;33(16):2580–2.
+Available from: http://dx.doi.org/10.1093/bioinformatics/btx192
+16.
+Pasquier T, Lau MK, Han X, Fong E, Lerner BS, Boose ER, et al.
+Sharing and Preserving Computational Analyses for Posterity with encapsulator.
+Comput Sci Eng [Internet].
+2018;20(4):111–24.
+Available from: https://ieeexplore.ieee.org/abstract/document/8409369
+17.
+Kulkarni N, Alessandrì L, Panero R, Arigoni M, Olivero M, Ferrero G, et al.
+Reproducible bioinformatics project: a community for reproducible bioinformatics analysis pipelines.
+BMC Bioinformatics [Internet].
+2018;19(Suppl 10):349.
+Available from: http://dx.doi.org/10.1186/s12859-018-2296-x
+18.
+Hung LH, Hu J, Meiss T, Ingersoll A, Lloyd W, Kristiyanto D, et al.
+Building containerized workflows using the BioDepot-workflow-builder.
+Cell Syst [Internet].
+2019;9(5):508-514.e3.
+Available from: http://dx.doi.org/10.1016/j.cels.2019.08.007
+19.
+Brito JJ, Li J, Moore JH, Greene CS, Nogoy NA, Garmire LX, et al.
+Recommendations to enhance rigor and reproducibility in biomedical research.
+Gigascience [Internet].
+2020 [cited 2023 Mar 9];9(6).
+Available from: https://pubmed.ncbi.nlm.nih.gov/32479592/
+20.
+Piccolo SR, Ence ZE, Anderson EC, Chang JT, Bild AH.
+Simplifying the development of portable, scalable, and reproducible workflows.
+Elife [Internet].
+2021;10.
+Available from: http://dx.doi.org/10.7554/eLife.71069
+21.
+Ge SX, Jung D, Yao R.
+ShinyGO: a graphical gene-set enrichment tool for animals and plants.
+Bioinformatics [Internet].
+2020 [cited 2023 Mar 9];36(8):2628–9.
+Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
+22.
+Perampalam P, Dick FA.
+BEAVR: a browser-based tool for the exploration and visualization of RNA-seq data.
+BMC Bioinformatics [Internet].
+2020 [cited 2023 Mar 9];21(1):221.
+Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
+23.
+Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
+DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
+Genome Biol [Internet].
+2014;15(8):406.
+Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
+24.
+Ziemann M, Kaspi A, El-Osta A.
+Digital expression explorer 2: a repository of uniformly processed RNA sequencing data.
+Gigascience [Internet].
+2019;8(4).
+Available from: http://dx.doi.org/10.1093/gigascience/giz022

### 2023-03-11 example.Rmd
progress

@@
 fgseaRes <- fgseaRes[order(fgseaRes$pval),]
+
+```
+
+```{r,fgsea_sig}
+
 fgsea_up <- subset(fgseaRes,padj<0.05 & ES>0)
@@
 fgsea_dns <- fgsea_dn$pathway
+
+fgsea_up <-  data.frame(fgsea_up[order(-fgsea_up$ES),])
+
+fgsea_dn <-  data.frame(fgsea_dn[order(fgsea_dn$ES),])
+
+head(fgsea_up,10) %>%
+  kbl(row.names = FALSE, caption="FGSEA:Top upregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
+
+head(fgsea_dn,10) %>%
+  kbl(row.names = FALSE, caption="FGSEA:Top downregulated pathways in Aza treatment") %>%
+  kable_paper("hover", full_width = F)
 

### 2023-03-20 manuscript.Rmd
intro

@@
 
-Enrichment analysis is one of the most used methods in computational biology,
-and involves the summarisation of omics data to reflect biological changes, such as
-the detection of pathway activation in development or disease [1,2](Slonim 2002;Khatri et al, 2012).
+Reproducibility is essential to the scientific enterprise by confirming the validity of new discoveries
+[1](National Academies of Sciences 2019, p1).
+According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was "significant
+crisis" of reproducibility [2](Baker 2016).
+While bioinformatics and other branches of computational research are theoretically amenable to
+complete reproducibility, in reality this is rarely achieved due to lack of detail provided in methods
+sections, lack of shared data and lack of code [3](Peng 2011).
+Indeed even if data and code are shared, reproducibility still cannot be guaranteed due to the
+differences in computing environments over time, and the availability of dependancies [4](Perkel 2020).
+
+To address this, many articles have been written to endorse the use of "virtual machines" or
+"Docker containers" to encapsulate the operating system and dependancies along with the research code
+to facilitate a high degree of reproducibility [4-16](Perkel 2020; Stein 2010; Dudley & Butte 2010;
+Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016;
+da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019;
+Brito et al, 2020; Piccolo et al, 2021).
+Despite the availability of such tools for over a decade, bioinformatics analyses using fully
+reproducible virtual machines or containers remains a miniscule fraction.
+
+In this article, we will focus on enrichment analysis, which is one of the most used methods in
+computational biology, and involves the summarisation of omics data to reflect biological changes,
+such as the detection of pathway activation in development or disease
+[17,18](Slonim 2002; Khatri et al, 2012).
 While enrichment analysis is undoubtedly a key method for the interpretation of large
@@
 (ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
-[3-6](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
+[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
+Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
+the validity of the methods and results.
+We are concerned about the reliability of published enrichment analyses, as web-based and
+point-and-click tools appear to be far more popular than computer script based methods for this type
+of work [21](Wijesooriya et al, 2022).
+Web-based tools are a potential concern, algorithms and functional annotation sets undergo regular
+updates, but the historical versions are typically not archived, nor are versions regularly reported
+in publications, which hampers later reproducibility.
+Indeed, in some cases the entire web resource can become unavailable, and hence irreproducible due to
+a phenomenon known as link decay [23](Hennessey & Ge 2013).
 
-Another emerging problem is that while computational research is theoretically amenable to complete
-reproducibility, in reality this is rarely achieved in computation-intensive research due to lack
-of detail provided in methods sections, lack of shared data and lack of code [7](Peng 2011).
-
-The lack of reproducibility is particularly acute for enrichment analyses, as web-based and
-point-and-click tools appear to be far more popular than computer script based methods
-[5](Wijesooriya et al, 2022).
-Web-based tools are a potential concern, as it is known that algorithms and functional annotation
-sets undergo regular updates, but the historical versions are typically not archived, nor
-are versions regularly reported in publications, making later replication difficult if not
-impossible in most cases [5](Wijesooriya et al, 2022).
-Indeed, in some cases the entire web resource can become unavailable, and hence irreproducible due to
-a phenomenon known as link decay [8](Hennessey & Ge 2013).
-
-A lack of reproducibility does not mean a study is invalid, however a lack of auditability means
-it cannot be ruled out that mistakes have been made in the enrichment analysis and that the results
-are biased or false.
-As omics data are becoming more routine in the drug development process and are used to
-inform clinical practice, reproducibility should become a minimum requirement for
-bioinformatics analysis of omics data.
-
-If research data and code are shared, reproducibility still cannot be guaranteed due to the
-differences in various computing environments such as Windows, Linux, Mac, which provide varying
-sets of libraries for programming languages to use.
-Even if the operating system is specified in the research methods, reproducibility can be difficult if the
-operating system is old or obscure, and where dependancies are unavailable.
-To address this, many articles have been written to endorse the use of "virtual machines" or "Docker containers"
-to encapsulate the operating system and dependancies along with the research code to facilitate
-a high degree of reproducibility [9-20](Stein 2010; Dudley & Butte 2010; Sandve et al, 2013;
-Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016; da Veiga Leprevost et al, 2017;
-Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
-
-Despite the availability of such tools for over a decade and the outsized impact such analysis has
-on the conclusions of an omics study, enrichment analyses using fully reproducible virtual machines
-or containers remains a miniscule fraction of the overall share of enrichment analyses.
 A small number of enrichment analysis applications have been developed to have beneficial properties
-of being "containerised" and with archived hostorical versions [21,22](Ge et al, 2020; Perampalam & Dick 2020).
-This helps reproducibility greatly, but is insufficient as it lacks information about how the input data was
-used and a record of any parameter selections.
+of being "containerised" and with archived historical versions [24,25](Ge et al, 2020;
+Perampalam & Dick 2020).
+This helps reproducibility greatly, but is insufficient as it lacks information about how the input
+data was used and a record of any parameter selections.
 
 Therefore the most complete reproducibility solution is to use containerisation to provide a scripted
-workflow with prescribed links to research data, external databases and complete record of parameter selection.
-Designed correctly, these Docker images could encapsulate bioinformatics workflows and ensure their
+workflow with prescribed links to research data, external databases and complete record of parameter
+selection.
+Designed correctly, these containers could encapsulate bioinformatics workflows and ensure their
 computational reproducibility for decades to come, with just one or a few simple commands.
+There are multiple barriers to entry for such workflows including (i) the added difficulty compares to
+existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
+there is a lack of step-by-step guides, and few existing templates which can be rapidly adapted.
 
-The barriers to entry include:
+We address this firstly by providing working templates for docker image and Rmarkdown scripts for
+common transcriptome and enrichment analysis routines.
+Secondly, we provide step-by-step written and video guides to help users customise these templates for
+their own work, deploy the analysis, verify the results, and archive the Docker image.
 
-* The fact that conventional web based and GUI tools are easier to use
-
-* It adds an overhead of effort to the project
-
-* There's no time to learn new stuff
-
-* It's not a requirement of the journal
-
-We address this by:
-
-* providing working templates for docker image and Rmarkdown scripts for common transcriptome and enrichment analysis
-routines.
-
-* Providing step-by-step guides to help users customise these templates for their own work, deploy the analysis,
-and verify the results.
-
-To highlight the need for extremely reproducible enrichment analysis, we first investigate the computational
-reproducibility of enrichment analyses shown in a sample of 20 journal articles published in 2019.
+But is this extra work really necessary?
+Although our previous work suggests that enrichment analyses on the whole lack reproducibility
+[5](Wijesooriya et al, 2022), we lack the empirical data confirming this directly.
+Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
+to assess directly whether current standards are appropriate, by trying to reproduce the enrichment
+analyses of previously published studies.
+In the following section we investigate the computational reproducibility of enrichment analyses
+shown in a sample of 20 journal articles published in 2019.
+This should provide a sound basis for recommendations.
 
@@
 1.
+National Academies of Sciences, Engineering, Medicine, Policy, Global Affairs, Committee on Science, Engineering, Medicine, Public Policy, et al.
+Reproducibility and Replicability in Science.
+Washington, D.C., DC: National Academies Press; 2019.
+2.  Baker M. 1,500 scientists lift the lid on reproducibility. Nature [Internet]. 2016 [cited 2023 Mar 20];533(7604):452–4. Available from: https://www.nature.com/articles/533452a
+3.  Peng RD. Reproducible research in computational science. Science [Internet]. 2011;334(6060):1226–7. Available from: http://dx.doi.org/10.1126/science.1213847
+4.
+Perkel JM.
+Challenge to scientists: does your ten-year-old code still run?
+Nature [Internet].
+2020 [cited 2023 Mar 20];584(7822):656–8.
+Available from: https://www.nature.com/articles/d41586-020-02462-7
+5.  Stein LD. The case for cloud computing in genome informatics. Genome Biol [Internet]. 2010;11(5):207. Available from: http://dx.doi.org/10.1186/gb-2010-11-5-207
+6.  Dudley JT, Butte AJ. In silico research in the era of cloud computing. Nat Biotechnol [Internet]. 2010;28(11):1181–5. Available from: http://dx.doi.org/10.1038/nbt1110-1181
+7.
+Sandve GK, Nekrutenko A, Taylor J, Hovig E.
+Ten simple rules for reproducible computational research.
+PLoS Comput Biol [Internet].
+2013;9(10):e1003285.
+Available from: http://dx.doi.org/10.1371/journal.pcbi.1003285
+8.  Piccolo SR, Frampton MB. Tools and techniques for computational reproducibility. Gigascience [Internet]. 2016;5(1). Available from: http://dx.doi.org/10.1186/s13742-016-0135-4
+9.
+Grüning B, Chilton J, Köster J, Dale R, Soranzo N, van den Beek M, et al.
+Practical computational reproducibility in the life sciences.
+Cell Syst [Internet].
+2018;6(6):631–5.
+Available from: http://dx.doi.org/10.1016/j.cels.2018.03.014
+10.
+Lewis J, Breeze CE, Charlesworth J, Maclaren OJ, Cooper J.
+Where next for the reproducibility agenda in computational biology?
+BMC Syst Biol [Internet].
+2016;10(1).
+Available from: http://dx.doi.org/10.1186/s12918-016-0288-x
+11.
+da Veiga Leprevost F, Grüning BA, Alves Aflitos S, Röst HL, Uszkoreit J, Barsnes H, et al.
+BioContainers: an open-source and community-driven framework for software standardization.
+Bioinformatics [Internet].
+2017;33(16):2580–2.
+Available from: http://dx.doi.org/10.1093/bioinformatics/btx192
+12.
+Pasquier T, Lau MK, Han X, Fong E, Lerner BS, Boose ER, et al.
+Sharing and Preserving Computational Analyses for Posterity with encapsulator.
+Comput Sci Eng [Internet].
+2018;20(4):111–24.
+Available from: https://ieeexplore.ieee.org/abstract/document/8409369
+13.
+Kulkarni N, Alessandrì L, Panero R, Arigoni M, Olivero M, Ferrero G, et al.
+Reproducible bioinformatics project: a community for reproducible bioinformatics analysis pipelines.
+BMC Bioinformatics [Internet].
+2018;19(Suppl 10):349.
+Available from: http://dx.doi.org/10.1186/s12859-018-2296-x
+14.
+Hung LH, Hu J, Meiss T, Ingersoll A, Lloyd W, Kristiyanto D, et al.
+Building containerized workflows using the BioDepot-workflow-builder.
+Cell Syst [Internet].
+2019;9(5):508-514.e3.
+Available from: http://dx.doi.org/10.1016/j.cels.2019.08.007
+15.
+Brito JJ, Li J, Moore JH, Greene CS, Nogoy NA, Garmire LX, et al.
+Recommendations to enhance rigor and reproducibility in biomedical research.
+Gigascience [Internet].
+2020 [cited 2023 Mar 9];9(6).
+Available from: https://pubmed.ncbi.nlm.nih.gov/32479592/
+16.
+Piccolo SR, Ence ZE, Anderson EC, Chang JT, Bild AH.
+Simplifying the development of portable, scalable, and reproducible workflows.
+Elife [Internet].
+2021;10.
+Available from: http://dx.doi.org/10.7554/eLife.71069
+17.
 Slonim DK.
@@
 Available from: https://www.nature.com/articles/ng1033z
-2.
+18.
 Khatri P, Sirota M, Butte AJ.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
-3.
+19.
 Timmons JA, Szkop KJ, Gallagher IJ.
@@
 Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
-4.
+20.
 Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al.
@@
 Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
-5.
+21.
 Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M.
@@
 Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
-6.
+22.
 Zhao K, Rhee SY.
@@
 Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0168952523000185%3Fshowall%3Dtrue
-7.  Peng RD. Reproducible research in computational science. Science [Internet]. 2011;334(6060):1226–7. Available from: http://dx.doi.org/10.1126/science.1213847
-8.
+23.
 Hennessey J, Ge S.
@@
 Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
-9.  Stein LD. The case for cloud computing in genome informatics. Genome Biol [Internet]. 2010;11(5):207. Available from: http://dx.doi.org/10.1186/gb-2010-11-5-207
-10.  Dudley JT, Butte AJ. In silico research in the era of cloud computing. Nat Biotechnol [Internet]. 2010;28(11):1181–5. Available from: http://dx.doi.org/10.1038/nbt1110-1181
-11.
-Sandve GK, Nekrutenko A, Taylor J, Hovig E.
-Ten simple rules for reproducible computational research.
-PLoS Comput Biol [Internet].
-2013;9(10):e1003285.
-Available from: http://dx.doi.org/10.1371/journal.pcbi.1003285
-12.  Piccolo SR, Frampton MB. Tools and techniques for computational reproducibility. Gigascience [Internet]. 2016;5(1). Available from: http://dx.doi.org/10.1186/s13742-016-0135-4
-13.
-Grüning B, Chilton J, Köster J, Dale R, Soranzo N, van den Beek M, et al.
-Practical computational reproducibility in the life sciences.
-Cell Syst [Internet].
-2018;6(6):631–5.
-Available from: http://dx.doi.org/10.1016/j.cels.2018.03.014
-14.
-Lewis J, Breeze CE, Charlesworth J, Maclaren OJ, Cooper J.
-Where next for the reproducibility agenda in computational biology?
-BMC Syst Biol [Internet].
-2016;10(1).
-Available from: http://dx.doi.org/10.1186/s12918-016-0288-x
-15.
-da Veiga Leprevost F, Grüning BA, Alves Aflitos S, Röst HL, Uszkoreit J, Barsnes H, et al.
-BioContainers: an open-source and community-driven framework for software standardization.
-Bioinformatics [Internet].
-2017;33(16):2580–2.
-Available from: http://dx.doi.org/10.1093/bioinformatics/btx192
-16.
-Pasquier T, Lau MK, Han X, Fong E, Lerner BS, Boose ER, et al.
-Sharing and Preserving Computational Analyses for Posterity with encapsulator.
-Comput Sci Eng [Internet].
-2018;20(4):111–24.
-Available from: https://ieeexplore.ieee.org/abstract/document/8409369
-17.
-Kulkarni N, Alessandrì L, Panero R, Arigoni M, Olivero M, Ferrero G, et al.
-Reproducible bioinformatics project: a community for reproducible bioinformatics analysis pipelines.
-BMC Bioinformatics [Internet].
-2018;19(Suppl 10):349.
-Available from: http://dx.doi.org/10.1186/s12859-018-2296-x
-18.
-Hung LH, Hu J, Meiss T, Ingersoll A, Lloyd W, Kristiyanto D, et al.
-Building containerized workflows using the BioDepot-workflow-builder.
-Cell Syst [Internet].
-2019;9(5):508-514.e3.
-Available from: http://dx.doi.org/10.1016/j.cels.2019.08.007
-19.
-Brito JJ, Li J, Moore JH, Greene CS, Nogoy NA, Garmire LX, et al.
-Recommendations to enhance rigor and reproducibility in biomedical research.
-Gigascience [Internet].
-2020 [cited 2023 Mar 9];9(6).
-Available from: https://pubmed.ncbi.nlm.nih.gov/32479592/
-20.
-Piccolo SR, Ence ZE, Anderson EC, Chang JT, Bild AH.
-Simplifying the development of portable, scalable, and reproducible workflows.
-Elife [Internet].
-2021;10.
-Available from: http://dx.doi.org/10.7554/eLife.71069
-21.
+24.
 Ge SX, Jung D, Yao R.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
-22.
+25.
 Perampalam P, Dick FA.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
+
 23.

### 2023-03-23 manuscript.Rmd
update ms

@@
 Despite the availability of such tools for over a decade, bioinformatics analyses using fully
-reproducible virtual machines or containers remains a miniscule fraction.
+reproducible virtual machines or containers remains only a small fraction.
 
@@
 of work [21](Wijesooriya et al, 2022).
-Web-based tools are a potential concern, algorithms and functional annotation sets undergo regular
-updates, but the historical versions are typically not archived, nor are versions regularly reported
-in publications, which hampers later reproducibility.
-Indeed, in some cases the entire web resource can become unavailable, and hence irreproducible due to
+Web-based tools are a potential concern, firstly because sometimes methods are not completely described,
+and secondly, algorithms and functional annotation sets undergo regular updates.
+These historical versions are typically not archived, which hampers later reproducibility.
+In some cases the entire web resource can become unavailable, and hence irreproducible due to
 a phenomenon known as link decay [23](Hennessey & Ge 2013).
@@
 computational reproducibility for decades to come, with just one or a few simple commands.
-There are multiple barriers to entry for such workflows including (i) the added difficulty compares to
+There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
 existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
-there is a lack of step-by-step guides, and few existing templates which can be rapidly adapted.
+a lack of step-by-step guides and templates aimed at beginner users.
 
@@
 to assess directly whether current standards are appropriate, by trying to reproduce the enrichment
-analyses of previously published studies.
+analyses of previously published studies, and examining whether the conclusions of those articles
+are supported or not.
 In the following section we investigate the computational reproducibility of enrichment analyses

### 2023-03-24 example.Rmd
add stuff

@@
 
-par(mar=c(5,20,5,1))
+par(mar=c(5,25,5,1))
 
@@
 
+fgsea_up <- head(fgsea_up,20)
+fgsea_up_vec <- fgsea_up$ES
+names(fgsea_up_vec) <- fgsea_up$pathway
+fgsea_up_vec <- rev(fgsea_up_vec)
+
+fgsea_dn <- head(fgsea_dn,20)
+fgsea_dn_vec <- fgsea_dn$ES * -1
+names(fgsea_dn_vec) <- fgsea_dn$pathway
+
+```
+
+Make a barplot
+
+```{r,topbarplot3}
+
+par(mar=c(5,25,5,1))
+
+cols <- c(rep("blue",20),rep("red",20))
+
+barplot(c(fgsea_dn_vec,fgsea_up_vec),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES")
+
+mtext("FCS test")
+
 ```

### 2023-03-24 manuscript.Rmd
add stuff

@@
 ## Introduction
+
+```{r,libs}
+
+library("DiagrammeR")
+
+```
 
@@
 Designed correctly, these containers could encapsulate bioinformatics workflows and ensure their
-computational reproducibility for decades to come, with just one or a few simple commands.
+computational reproducibility for decades to come with just one or a few simple commands.
 There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
@@
 <Anusuiya to populate>
+<references to start at [101](Author Year)>
 
@@
 
-Here describe the set up of the system.
-The following text is a bit random and needs to be culled.
-Although it doesn't need to be this way due to the precise nature of modern computers and software.
-In order for bioinformatics analysis to be 100% reproducible requires a few key elements.
-These are:
+Infographic:
 
-* Input data
+* source control
 
-* Set of instructions (analysis code).
+* containerisation
 
-* Operating system
+* literate programming
 
-With these things working together, it is possible for a data analysis workflow to be 100%
-replicable and auditable for many years into the future.
+* documentation
 
-To achieve this requires some planning:
+The templates provided include the Dockerfile and an Rmarkdown-based workflow script.
+These templates are intended to be altered by the end user, but here is an overview of the contents of
+each.
 
-1. Input data needs to be deposited to a public and persistent location.
+The Dockerfile uses a base image from Bioconductor which should correspond to the latest available
+stable release version (currently 3.16), running on R v4.2.2 and Ubuntu 22.04.
+The image contains instruction for installing a few useful utilities including `nano` and `git` for
+modifying scripts, and `magic-wormhole` for transfering data between computers.
+It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
+analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` [201](Ziemann et al, 2019),
+differential expression with `DESeq2` [202](Love & Huber, 2014),
+enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
+[203-205](Yu et al, 2012; Korotkevich et al, 2016; Kaspi & Ziemann 2020).
+It also contains instructions to clone a github repository, which contains the scripts that run the
+analytical workflow.
+Lastly, it copies a gene set library from Reactome [206](Gillespie et al, 2022) with a clear version
+history to guarantee provenance of the annotation data.
 
-2. Analysis code packaged in a working Docker container.
+The `example_workflow.Rmd` is the Rmarkdown based script that undertakes the enrichment analysis.
+Rmarkdown allows the unification of extended descriptions, code and results into a single document,
+which is becoming ubiquitous in bioinformatics and data science more generally
+[207](Grolemund & Wickham 2017, p423).
+The ability to provide extended descriptions is helpful for providing contextualising information about
+the purpose of the work, detailed methods and interpretations derived from the analysis.
+The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
+data set from DEE2, inspecting data quality, differential expression and two types of enrichment
+analysis; over-representation analysis (ORA) and functional class scoring (FCS) (**Figure XX**).
+For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
+[22](Zhao & Rhee 2023).
 
-A Docker container is a minimal operating system that has the necessary software to complete a task.
-Docker is used extensively in web development, to rapidly set up many modular services into
-full stacks of functionality.
-For example using Docker containers, a website front end and back end can be set up once, and then
-deployed at huge scale (eg thousands of instances) with minimal work involved.
+```{r,diagram,fig.cap = "Figure XX: Flow chart of example workflow."}
 
-In addition, Docker containers ensure consistent behaviour across different types of computing
-infrastructure, so it doesn't matter whether the container is run on a linux workstation,
-desktop windows PC or cloud server, the end results should be identical.
+grViz("digraph flowchart {
+      node [fontname = Helvetica, shape = rectangle]
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
+      tab3 -> tab5 -> tab6 }
+      [1]: 'Fetch RNA-seq from DEE2'
+      [2]: 'Quality control'
+      [3]: 'Differential expression'
+      [4]: 'Over-representation analysis'
+      [5]: 'Functional class scoring'
+      [6]: 'Euler diagram'
+")
+
+```
+
+To demonstrate the utility of this workflow, we present the results of an example data analysis;
+it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
+in triplicate [208] (Lund et al, 2014).
+
+We use this analysis to try to understand better the effect of DNA methylatransferase inhibitors on
+genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
+the relationship between RNA expression and DNA methylation, but did not show results of a
+transcriptome-only enrichment analysis.
+
+Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
+FCS, and give guidance for interpreting results of such tests.
+
+After fetching the data from DEE2, the number of reads per sample was confirmed to be in the range of
+11-16M.
+Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
+genes.
+Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
+1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells.
+
+These differentially expressed gene sets were subjected to ORA analysis separately, using the list of
+13,168 genes as the background, and gene sets from REACTOME.
+
+With ORA, we observed 149 upregulated and 48 downregulated gene sets.
+
+We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
+which tends to highlight smaller and more specific gene sets undergoing more drastic differential
+expression, rather than large gene sets that undergo very small expression changes.
+This led to the identification of upregulated pathways related to nuclear RNA export,
+glucokinase regulation and G1/S DNA replication pathways; while downregulated pathways
+were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
+(host defence).
+
+After FCS, we identified 241 upregulated genesets and 155 downregulated genesets.
+FCS upregulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
+and defective homologous recombination repair.
+Downregulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
+
+Regarding upregulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
+Jaccard index of 0.54, however these method disagreed in downregulated pathways, with a Jaccard
+index of only 0.27.
+
+201.
+Ziemann M, Kaspi A, El-Osta A.
+Digital expression explorer 2: a repository of uniformly processed RNA sequencing data.
+Gigascience [Internet].
+2019;8(4).
+Available from: http://dx.doi.org/10.1093/gigascience/giz022
+202.
+Love MI, Huber W, Anders S.
+Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2.
+Genome Biol [Internet].
+2014;15(12):550.
+Available from: http://dx.doi.org/10.1186/s13059-014-0550-8
+203.
+Yu G, Wang LG, Han Y, He QY.
+clusterProfiler: an R package for comparing biological themes among gene clusters.
+OMICS [Internet].
+2012;16(5):284–7.
+Available from: http://dx.doi.org/10.1089/omi.2011.0118
+204.
+Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A.
+Fast gene set enrichment analysis [Internet].
+bioRxiv.
+2016.
+p.
+060012.
+Available from: http://biorxiv.org/content/early/2021/02/01/060012.abstract
+205.
+Kaspi A, Ziemann M.
+Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data.
+BMC Genomics [Internet].
+2020;21(1):447.
+Available from: http://dx.doi.org/10.1186/s12864-020-06856-9
+206.
+Gillespie M, Jassal B, Stephan R, Milacic M, Rothfels K, Senff-Ribeiro A, et al.
+The reactome pathway knowledgebase 2022.
+Nucleic Acids Res [Internet].
+2022;50(D1):D687–92.
+Available from: http://dx.doi.org/10.1093/nar/gkab1028
+207.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
+208.
+Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
+DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
+Genome Biol [Internet].
+2014;15(8):406.
+Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
+
+Table XX: The software bundles in the example image.
+R package dependancies not shown.
+
+| Software | Version | Purpose |
+| --- | --- | --- |
+| Ubuntu | 22.04 | Operating System |
+| R | 4.2.2 | Statistical computing language |
+| nano | 6.2 | text editing |
+| git | 2.34.1 | source control |
+| R/kableExtra | 1.3.4 | render formatted tables |
+| R/vioplot | 0.4.0 | violin charts |
+| R/gplots | 3.1.3 | heatmaps |
+| R/eulerr | 7.0.0 | Venn diagrams |
+| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
+| BioC/DESeq2 | 1.38.3 | differential expression |
+| BioC/fgsea | 1.24.0 | functional class scoring |
+| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
+| Bioc/mitch | 1.10.0 | multi-contrast enrichment |
 
@@
 2023 [cited 2023 Mar 1];0(0).
-Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0168952523000185%3Fshowall%3Dtrue
+Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5
 23.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
-
-23.
-Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
-DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
-Genome Biol [Internet].
-2014;15(8):406.
-Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
-24.
-Ziemann M, Kaspi A, El-Osta A.
-Digital expression explorer 2: a repository of uniformly processed RNA sequencing data.
-Gigascience [Internet].
-2019;8(4).
-Available from: http://dx.doi.org/10.1093/gigascience/giz022

### 2023-03-24 example.Rmd
update chart header

@@
   cex=0.6,pch=19,col="darkgray",
-  main="strict filter",
+  main="5-azacitidine treatment in AML3 cells",
   xlab="log10(basemean)",ylab="log2(fold change)")

### 2023-03-24 example.Rmd
added heat

@@
 ---
-title: "An example of extremely reproducible enrichment analysis."
+title: "An example of extremely reproducible enrichment analysis"
 author: "Mark Ziemann & Anusuiya Bora"
@@
   library("eulerr")
+  library("gplots")
 })
@@
 
-## Scatterplot
+## Drill in to a specific pathway
 
+In this case the top upregulated and downregulated pathways
+
+```{r,eplot}
+
+upname <- head(fgsea_up,1)$pathway
+plotEnrichment(upname, stat, gseaParam = 1, ticksSize = 0.2)
+
+dnname <- head(fgsea_dn,1)$pathway
+plotEnrichment(dnname, stat, gseaParam = 1, ticksSize = 0.2)
+
+```
+
+Now a heatmap of these gene sets.
+
+```{r,heat}
+
+# reads per million normalisation
+rpm <- apply(mxf,2, function(x) { x/sum(x) * 1000000} )
+gnames_up <- gsets[[which(names(gsets) == upname)]]
+gnames_dn <- gsets[[which(names(gsets) == dnname)]]
+gene_ids <- rownames(mxf)
+gene_names <- sapply(strsplit(gene_ids," "),"[[",2)
+rpm_up <- rpm[which(gene_names %in% gnames_up),]
+rpm_dn <- rpm[which(gene_names %in% gnames_dn),]
+
+colsidecols <- c("blue","blue","blue","red","red","red")
+heatmap.2(rpm_up,scale="row",trace="none",margins=c(10,10),main=upname,ColSideColors=colsidecols)
+heatmap.2(rpm_dn,scale="row",trace="none",margins=c(10,10),main=dnname,ColSideColors=colsidecols)
+
+```
 ## Session information

### 2023-03-24 example.Rmd
added enrichment plot

@@
 upname <- head(fgsea_up,1)$pathway
-plotEnrichment(upname, stat, gseaParam = 1, ticksSize = 0.2)
+plotEnrichment(gsets[[upname]], stat, gseaParam = 1, ticksSize = 0.2)
 
 dnname <- head(fgsea_dn,1)$pathway
-plotEnrichment(dnname, stat, gseaParam = 1, ticksSize = 0.2)
+plotEnrichment(gsets[[dnname]], stat, gseaParam = 1, ticksSize = 0.2)
 
@@
 rpm <- apply(mxf,2, function(x) { x/sum(x) * 1000000} )
+colnames(rpm) <- c("UNTR1","UNTR2","UNTR3","TRT1","TRT2","TRT3")
 gnames_up <- gsets[[which(names(gsets) == upname)]]
@@
 rpm_dn <- rpm[which(gene_names %in% gnames_dn),]
-
 colsidecols <- c("blue","blue","blue","red","red","red")
@@
 ```
+
 ## Session information

### 2023-03-24 example.Rmd
fig size

@@
     code_folding: hide
-    fig_width: 7
-    fig_height: 7
+    fig_width: 5
+    fig_height: 5
 theme: cosmo
@@
 
-```{r,venn1,fig.width=6,fig.height=6}
+```{r,venn1}
 
@@
 colsidecols <- c("blue","blue","blue","red","red","red")
-heatmap.2(rpm_up,scale="row",trace="none",margins=c(10,10),main=upname,ColSideColors=colsidecols)
-heatmap.2(rpm_dn,scale="row",trace="none",margins=c(10,10),main=dnname,ColSideColors=colsidecols)
+heatmap.2(rpm_up,scale="row",trace="none",margins=c(10,15),main=upname,ColSideColors=colsidecols)
+heatmap.2(rpm_dn,scale="row",trace="none",margins=c(10,15),main=dnname,ColSideColors=colsidecols)
 

### 2023-03-24 example.Rmd
fix heatmap gene names

@@
 rpm_up <- rpm[which(gene_names %in% gnames_up),]
+rownames(rpm_up) <- sapply(strsplit(rownames(rpm_up)," "),"[[",2)
 rpm_dn <- rpm[which(gene_names %in% gnames_dn),]
+rownames(rpm_dn) <- sapply(strsplit(rownames(rpm_dn)," "),"[[",2)
 colsidecols <- c("blue","blue","blue","red","red","red")

### 2023-03-24 example.Rmd
figsize
barplot
fixed number of pathways

@@
 
-topup2 <- rev(head(ora_up$es,20))
-names(topup2) <- rev(head(ora_up$ID,20))
+topup2 <- rev(head(ora_up$es,10))
+names(topup2) <- rev(head(ora_up$ID,10))
 
@@
 ```{r,oradn}
+n_pw=10
 
@@
 
-head(ora_dn) %>%
+head(ora_dn,n_pw) %>%
   kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
@@
 
-topdn2 <- head(ora_dn$es,20)
-names(topdn2) <- head(ora_dn$ID,20)
+topdn2 <- head(ora_dn$es,n_pw)
+names(topdn2) <- head(ora_dn$ID,n_pw)
 
@@
 
-```{r,topbarplot2}
+```{r,topbarplot2,fig.height=6}
 
-par(mar=c(5,25,5,1))
+par(mar=c(5,20,5,1))
 
-cols <- c(rep("blue",20),rep("red",20))
+cols <- c(rep("blue",n_pw),rep("red",n_pw))
 
@@
 
-head(fgsea_up,10) %>%
+head(fgsea_up,n_pw) %>%
   kbl(row.names = FALSE, caption="FGSEA:Top upregulated pathways in Aza treatment") %>%
@@
 
-head(fgsea_dn,10) %>%
+head(fgsea_dn,n_pw) %>%
   kbl(row.names = FALSE, caption="FGSEA:Top downregulated pathways in Aza treatment") %>%
@@
 
-fgsea_up <- head(fgsea_up,20)
+fgsea_up <- head(fgsea_up,n_pw)
 fgsea_up_vec <- fgsea_up$ES
@@
 
-fgsea_dn <- head(fgsea_dn,20)
+fgsea_dn <- head(fgsea_dn,n_pw)
 fgsea_dn_vec <- fgsea_dn$ES * -1
@@
 
-```{r,topbarplot3}
+```{r,topbarplot3,fig.height=6}
 
-par(mar=c(5,25,5,1))
+par(mar=c(5,20,5,1))
 
-cols <- c(rep("blue",20),rep("red",20))
+cols <- c(rep("blue",n_pw),rep("red",n_pw))
 

### 2023-03-24 manuscript.Rmd
updates

@@
     toc_float: true
+    code_folding: hide
     fig_width: 7
     fig_height: 5
-theme: cosmo
 ---
 
-<b>Alt titles:<b>
+<b>Alt titles:</b>
 
@@
 
-```{r,libs}
+```{r,libs,echo=FALSE}
 
@@
 reproducible virtual machines or containers remains only a small fraction.
+
+![Figure XX: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
 
@@
 
-```{r,diagram,fig.cap = "Figure XX: Flow chart of example workflow."}
+```{r,diagram,echo=FALSE,fig.cap = "Figure XX: Flow chart of example workflow."}
 
@@
 transcriptome-only enrichment analysis.
-
 Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
@@
 Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
-1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells.
-
+1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
+(Figure XXA).
 These differentially expressed gene sets were subjected to ORA analysis separately, using the list of
@@
 With ORA, we observed 149 upregulated and 48 downregulated gene sets.
-
 We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
@@
 were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
-(host defence).
+(host defence) (Figure XXB).
 
@@
 FCS upregulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
-and defective homologous recombination repair.
+and defective homologous recombination repair (Figure XXC).
 Downregulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
@@
 Jaccard index of 0.54, however these method disagreed in downregulated pathways, with a Jaccard
-index of only 0.27.
+index of only 0.27 (Figure XXD).
 
-201.
-Ziemann M, Kaspi A, El-Osta A.
-Digital expression explorer 2: a repository of uniformly processed RNA sequencing data.
-Gigascience [Internet].
-2019;8(4).
-Available from: http://dx.doi.org/10.1093/gigascience/giz022
-202.
-Love MI, Huber W, Anders S.
-Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2.
-Genome Biol [Internet].
-2014;15(12):550.
-Available from: http://dx.doi.org/10.1186/s13059-014-0550-8
-203.
-Yu G, Wang LG, Han Y, He QY.
-clusterProfiler: an R package for comparing biological themes among gene clusters.
-OMICS [Internet].
-2012;16(5):284–7.
-Available from: http://dx.doi.org/10.1089/omi.2011.0118
-204.
-Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A.
-Fast gene set enrichment analysis [Internet].
-bioRxiv.
-2016.
-p.
-060012.
-Available from: http://biorxiv.org/content/early/2021/02/01/060012.abstract
-205.
-Kaspi A, Ziemann M.
-Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data.
-BMC Genomics [Internet].
-2020;21(1):447.
-Available from: http://dx.doi.org/10.1186/s12864-020-06856-9
-206.
-Gillespie M, Jassal B, Stephan R, Milacic M, Rothfels K, Senff-Ribeiro A, et al.
-The reactome pathway knowledgebase 2022.
-Nucleic Acids Res [Internet].
-2022;50(D1):D687–92.
-Available from: http://dx.doi.org/10.1093/nar/gkab1028
-207.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
-208.
-Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
-DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
-Genome Biol [Internet].
-2014;15(8):406.
-Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
-
-Table XX: The software bundles in the example image.
-R package dependancies not shown.
-
-| Software | Version | Purpose |
-| --- | --- | --- |
-| Ubuntu | 22.04 | Operating System |
-| R | 4.2.2 | Statistical computing language |
-| nano | 6.2 | text editing |
-| git | 2.34.1 | source control |
-| R/kableExtra | 1.3.4 | render formatted tables |
-| R/vioplot | 0.4.0 | violin charts |
-| R/gplots | 3.1.3 | heatmaps |
-| R/eulerr | 7.0.0 | Venn diagrams |
-| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
-| BioC/DESeq2 | 1.38.3 | differential expression |
-| BioC/fgsea | 1.24.0 | functional class scoring |
-| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
-| Bioc/mitch | 1.10.0 | multi-contrast enrichment |
+![Figure XX: Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
 
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
+
+201.
+Ziemann M, Kaspi A, El-Osta A.
+Digital expression explorer 2: a repository of uniformly processed RNA sequencing data.
+Gigascience [Internet].
+2019;8(4).
+Available from: http://dx.doi.org/10.1093/gigascience/giz022
+202.
+Love MI, Huber W, Anders S.
+Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2.
+Genome Biol [Internet].
+2014;15(12):550.
+Available from: http://dx.doi.org/10.1186/s13059-014-0550-8
+203.
+Yu G, Wang LG, Han Y, He QY.
+clusterProfiler: an R package for comparing biological themes among gene clusters.
+OMICS [Internet].
+2012;16(5):284–7.
+Available from: http://dx.doi.org/10.1089/omi.2011.0118
+204.
+Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A.
+Fast gene set enrichment analysis [Internet].
+bioRxiv.
+2016.
+p.
+060012.
+Available from: http://biorxiv.org/content/early/2021/02/01/060012.abstract
+205.
+Kaspi A, Ziemann M.
+Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data.
+BMC Genomics [Internet].
+2020;21(1):447.
+Available from: http://dx.doi.org/10.1186/s12864-020-06856-9
+206.
+Gillespie M, Jassal B, Stephan R, Milacic M, Rothfels K, Senff-Ribeiro A, et al.
+The reactome pathway knowledgebase 2022.
+Nucleic Acids Res [Internet].
+2022;50(D1):D687–92.
+Available from: http://dx.doi.org/10.1093/nar/gkab1028
+207.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
+208.
+Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
+DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
+Genome Biol [Internet].
+2014;15(8):406.
+Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
+
+## Supplementary information
+
+Table XX: The software bundles in the example image.
+R package dependancies not shown.
+
+| Software | Version | Purpose |
+| --- | --- | --- |
+| Ubuntu | 22.04 | Operating System |
+| R | 4.2.2 | Statistical computing language |
+| nano | 6.2 | text editing |
+| git | 2.34.1 | source control |
+| R/kableExtra | 1.3.4 | render formatted tables |
+| R/vioplot | 0.4.0 | violin charts |
+| R/gplots | 3.1.3 | heatmaps |
+| R/eulerr | 7.0.0 | Venn diagrams |
+| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
+| BioC/DESeq2 | 1.38.3 | differential expression |
+| BioC/fgsea | 1.24.0 | functional class scoring |
+| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
+| Bioc/mitch | 1.10.0 | multi-contrast enrichment |

### 2023-03-30 manuscript.Rmd
Update manuscript.Rmd

@@
 <references to start at [101](Author Year)>
+<Wanted to check how this looks in the file>
+
+The selection of 20 articles from 2019 is based on the following factors:
+  * “Homo sapiens” is the organism of the datasets
+  * The tool used for enrichment analysis is “DAVID” (Database for Annotation, Visualization, and Integrated Discovery)
+  * Ensuring that the complete gene list is available to perform replication analysis successfully.
+
+To keep the work available and accessible at all times, google cloud storage is used to store several types of files. A google spreadsheet is maintained to track the
+progress. The sheet contains details like PMCID, type of organism, availability of gene set, usage of DAVID, and details about the utilization of the background gene
+list by the authors. The Google spreadsheet is attached as supplementary file 1 named as “Anusuiya_PMC_2019”.
+
+The articles are searched for the dataset of dysregulated (upregulated and downregulated, either combined or separately) genes.
+It is important to understand how the list of specific gene lists was obtained by the researchers.
+This is usually mentioned in the materials and methods section of the paper.
+The list of genes found in the articles is stored in the spreadsheet named “Anusuiya_gene_list”.
+The gene list is maintained for our analysis under their respective PMCIDs.
+Special attention is given to detecting gene name errors that are usually in converted date formats when procured from excel sheets from the original paper.
+On request, the gene list file can be shared.
 

### 2023-03-30 manuscript.Rmd
Update manuscript.Rmd

@@
 
-To keep the work available and accessible at all times, google cloud storage is used to store several types of files. A google spreadsheet is maintained to track the
-progress. The sheet contains details like PMCID, type of organism, availability of gene set, usage of DAVID, and details about the utilization of the background gene
-list by the authors. The Google spreadsheet is attached as supplementary file 1 named as “Anusuiya_PMC_2019”.
+To keep the work available and accessible at all times, google cloud storage is used to store several types of files. A google spreadsheet is maintained to 
+track the progress.
+The sheet contains details like PMCID, type of organism, availability of gene set, usage of DAVID, and details about the utilization of the background gene list by the authors.
+The Google spreadsheet is attached as supplementary file 1 named as “Anusuiya_PMC_2019”.
 
-The articles are searched for the dataset of dysregulated (upregulated and downregulated, either combined or separately) genes.
-It is important to understand how the list of specific gene lists was obtained by the researchers.
-This is usually mentioned in the materials and methods section of the paper.
-The list of genes found in the articles is stored in the spreadsheet named “Anusuiya_gene_list”.
-The gene list is maintained for our analysis under their respective PMCIDs.
-Special attention is given to detecting gene name errors that are usually in converted date formats when procured from excel sheets from the original paper.
-On request, the gene list file can be shared.
+The articles are searched for the dataset of dysregulated (upregulated and downregulated, either combined or separately) genes. It is important to understand 
+how the list of specific gene lists was obtained by the researchers. This is usually mentioned in the materials and methods section of the paper. The list of 
+genes found in the articles is stored in the spreadsheet named “Anusuiya_gene_list”. The gene list is maintained for our analysis under their respective PMCIDs.
+Special attention is given to detecting gene name errors that are usually in converted date formats when procured from excel sheets from the original paper. On
+request, the gene list file can be shared.
+
+The reproducibility analysis is performed using the respective methodologies stated by the 20 articles using DAVID. It is important to note that the DAVID 6.8 
+server has been replaced by DAVID 2021 server, starting from June 1, 2022. For our analysis, DAVID 6.8 version was used for replication procedures. It was 
+ensured that our analyses got completed by 31 May 2022.
+
+The following steps are performed using the DAVID web server:
+1. On google, the term “DAVID functional annotation” is searched. It can also be accessed through this link https://david.ncifcrf.gov/home.jsp.
+2. Click on “Start Analysis” present on top of the homepage.
+3. Click on the “Upload” tab.
+4. Select and copy the list of genes (specific to the paper being analyzed). Paste it under the “Enter Gene list/Paste a list” option.
+5. Select the identifier as “OFFICIAL_GENE_SYMBOL”
+6. Under the “Select Species” option, type “Homo sapiens” and click on the selection. 
+7. Select the list type as “Gene List”. Click on “Submit List”.
+8. Step 2 page comes up. Click on the “Functional Annotation Tool” option.
+9. The “Annotation Summary Results” page appears. 
+10. It can be observed that the current gene list and the current background show up. 
+11. Click on the tick next to “Check Defaults”, on top of the page, to unselect all the options that are selected by default by the server. 
+12. Under “Gene Ontology”, select the checkboxes of “GOTERM_BP_DIRECT”, GOTERM_CC_DIRECT” and “GOTERM_MF_DIRECT”.
+13. Click on “Functional Annotation Chart”. A new window opens up with a table of results containing the gene ontology terms. 
+14. Select the terms up to the prescribed P Value criteria from the paper. Other options like FDR values, Fischer Test values can be obtained by expanding 
+“options” above the results table and clicking on the required options. Click on Rerun using options.
+15. The table loads up with additional options. Select as per the P-Value criteria and keep the results in another document for future use.
+16. Go back to the “Annotation Summary Results” page again to obtain KEGG Pathway results.
+17. Expand the “Pathways” option. Click on “Chart” next to the KEGG_Pathway heading or the “Functional Chart”. 
+18. Step 14 can be performed accordingly for KEGG Pathway results as well.
+
+A live google document is used to store all the results that consisted of Gene ontology terms (Biological Process, Cellular Component, and Molecular Function) 
+and KEGG Pathway results. Terms and pathway results are included till the mentioned P values or corrected P Values from the paper (generally, P<0.005).
+
+Each article is re-read, keeping note of sections like Abstract, Results, Discussion, and Conclusion. It is checked whether the DAVID results obtained by the
+researchers are used in making the conclusions mentioned in the respective papers. This gives an idea of whether DAVID results were included to form conclusions 
+by the authors.
+
+Critical evaluation for each article is performed, and they are rated from 1 to 3. The following are the reasons for rating papers from 1 to 3:
+* Paper is rated “1” when it shows poor reproducibility. Terms for GO and/or KEGG did not match at all upon replication analysis.
+* Paper is rated “2” when at least 50% of the terms are observed after replication. 50% or more should be observed for both GO and KEGG analysis since 
+conclusions were drawn using both terms in the articles.
+* Paper is rated “3” when it shows maximum reproducibility. The terms upon replication are observed again according to the ones mentioned in the paper. The 
+conclusion of the paper is comprehensible and accurate.
+
+A bar graph is created denoting the number of articles that come under categories 1, 2, and 3. This is created using the “Chart” tool in google spreadsheet.
+
+A summary table is created that includes information about the parts of the paper that are reproduced using DAVID. It also states if replication analysis is 
+possible with subsequent reasoning.
 

### 2023-03-31 manuscript.Rmd
Update manuscript.Rmd
refining manuscript

--- 2023-03-24 manuscript.Rmd
@@
 To address this, many articles have been written to endorse the use of "virtual machines" or
-"Docker containers" to encapsulate the operating system and dependancies along with the research code
-to facilitate a high degree of reproducibility [4-16](Perkel 2020; Stein 2010; Dudley & Butte 2010;
-Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014; Lewis et al, 2016;
-da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018; Hung et al, 2019;
-Brito et al, 2020; Piccolo et al, 2021).
-Despite the availability of such tools for over a decade, bioinformatics analyses using fully
-reproducible virtual machines or containers remains only a small fraction.
+"containerisation" to encapsulate the "environment" including the operating system and dependancies
+along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
+Stein 2010; Dudley & Butte 2010; Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014;
+Lewis et al, 2016; da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018;
+Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
+Containers also allow for an easy installation process, as all working dependencies are shipped together
+and can be installed with a single command.
 
-![Figure XX: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
+In addition to shared code, data and containerisation, there are a few ways researchers can make their
+work more reproducible.
+Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
+to novice bioinformaticians.
+Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
+free text explanations together with the code and outputs such as charts and tables.
+Secondly, code and data can be linked. This means that the code knows the location of the the data,
+avoiding any "file not found" errors.
+Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
+for each figure sequentially, making the entire manuscript reproducible by executing just one command.
+Taken together, computational reproducibility is a spectrum (**Figure 1**).
+Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
+over a decade.
 
-In this article, we will focus on enrichment analysis, which is one of the most used methods in
+![Figure 1: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
+
+In this article, we focus on enrichment analysis, which is one of the most used methods in
 computational biology, and involves the summarisation of omics data to reflect biological changes,
@@
 of work [21](Wijesooriya et al, 2022).
-Web-based tools are a potential concern, firstly because sometimes methods are not completely described,
-and secondly, algorithms and functional annotation sets undergo regular updates.
+Web-based tools are a potential concern, firstly because sometimes methods, options and parameters
+are not completely described, and secondly, algorithms and functional annotation sets undergo
+regular updates.
 These historical versions are typically not archived, which hampers later reproducibility.
@@
 But is this extra work really necessary?
-Although our previous work suggests that enrichment analyses on the whole lack reproducibility
-[5](Wijesooriya et al, 2022), we lack the empirical data confirming this directly.
+Although our previous work suggests that enrichment analyses generally lack sufficient methodological
+detail [5](Wijesooriya et al, 2022), we don't have empirical evidence confirming irreproducibility.
 Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
-to assess directly whether current standards are appropriate, by trying to reproduce the enrichment
+to assess directly whether current practices are appropriate, by trying to reproduce the enrichment
 analyses of previously published studies, and examining whether the conclusions of those articles
@@
 
-## Methods
+## A systematic assessment of reproducibility in enrichment analysis
 
-### A systematic assessment of reproducibility in enrichment analysis
+### Methods
 
@@
 <references to start at [101](Author Year)>
+<Wanted to check how this looks in the file>
 
-### A container based approach to addressing reproducibility of enrichment analysis
+From a list of 186 PubMed central articles using enrichment analysis that we examined previously
+[5](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
+organism under study is human; (2) the tool used was DAVID (Database for Annotation, Visualization,
+and Integrated Discovery)[101](Huang et al, 2007); (3) the omics type was gene expression (array or
+RNA-seq); and (4) that the gene list was provided in the supplementary data.
+
+The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
+special attention to preventing gene name errors from occurring[102](Zeeberg et al, 2004).
+The reproducibility analysis was performed using the respective methodologies stated in the methods
+sections of the 20 articles with DAVID version 6.8 (March-May 2022).
+
+Briefly, we used the DAVID "Analysis Wizard" and pasted in the provided gene list into the text box,
+selecting this as "gene list", together with the appropriate gene identifier type.
+Typically this was the official gene symbol.
+The species was specified as *Homo sapiens*.
+Next we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
+page.
+We followed the steps indicated in the respective papers to select appropriate gene sets investigated
+in the articles such as gene ontology (biological process, molecular function), KEGG pathways, etc.
+Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
+enrichment results, which were saved for future reference.
+
+Firstly, the data presented in the paper was compared with the results we obtained, recording any
+discrepancies.
+Secondly, the statements made in the abstract, results, and discussion sections were carefully examined
+and cross-checked with our results to understand whether those assertions survive reproduction or not.
+
+Each article was critically evaluated and given a score of one (1) to three (3).
+Articles rated 1 showed overall poor reproducibility, for example most GO and/or KEGG terms did not
+match upon reproduction, and the conclusions of the study were overall not supported.
+Aticles were rated 2 when at least 50% of the terms are observed as significant after replication,
+according to the original study's significance threshold, and most conclusions were supported.
+Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
+shown in the article observed again in reproduction, and all conclusions were supported in reproduction.
+Results are summarised in table and bar chart form.
+
+### Results and discussion
+
+* bar graph
+
+* table
+
+*
+
+It is important to note that the DAVID 6.8 server has been replaced by DAVID 2022 server, starting
+from June 1, 2022.
+For our analysis, DAVID version 6.8 was used for replication procedures.
+It was ensured that our analyses got completed by 31 May 2022.
+It is no longer possible to reproduce analyses conducted in DAVID 6.8
+
+## A container based approach to addressing reproducibility of enrichment analysis
+
+### Implementation
 
@@
 ```
+
+### Findings
 
@@
 
+101.
+Huang DW, Sherman BT, Tan Q, Collins JR, Alvord WG, Roayaei J, et al.
+The DAVID Gene Functional Classification Tool: a novel biological module-centric algorithm to functionally analyze large gene lists.
+Genome Biol [Internet].
+2007;8(9):R183.
+Available from: http://dx.doi.org/10.1186/gb-2007-8-9-r183
+102.
+Zeeberg BR, Riss J, Kane DW, Bussey KJ, Uchio E, Linehan WM, et al.
+Mistaken identifiers: gene name errors can be introduced inadvertently when using Excel in bioinformatics.
+BMC Bioinformatics [Internet].
+2004;5:80.
+Available from: http://dx.doi.org/10.1186/1471-2105-5-80
+
 201.

### 2023-04-03 example.Rmd
updates
#3 solved - added an mds

@@
 barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")
+
+```
+
+Now make a MDS plot.
+
+```{r,mds1}
+
+mds <- cmdscale(dist(t(mx)))
+
+# expand plot area
+XMIN=min(mds[,1])*1.3
+XMAX=max(mds[,1])*1.3
+YMIN=min(mds[,2])*1.3
+YMAX=max(mds[,2])*1.3
+
+cols <- as.character(grepl("Treated",ss$Experiment_title))
+cols <- gsub("FALSE","lightblue",cols)
+cols <- gsub("TRUE","pink",cols)
+plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
+  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
+  pch=19,cex=2,col=cols, main="MDS plot")
+text(cmdscale(dist(t(mx))), labels=colnames(mx) )
 
@@
 
-```{r,topbarplot2,fig.height=6}
+```{r,topbarplot2}
 
@@
 
-```{r,topbarplot3,fig.height=6}
+```{r,topbarplot3}
 

### 2023-04-04 example.Rmd
adjustments to workflow

@@
 All genes that meet the detection threshold will comprise the background list.
+The first 6 rows of the count matrix are shows.
 
@@
 dim(mxf)
+
+head(mxf,6) %>%
+  kbl(caption="Count matrix format") %>%
+  kable_paper("hover", full_width = F)
+
 dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
@@
   kable_paper("hover", full_width = F)
+
+```
+
+Make a smear plot.
+
+```{r,smearplot}
 
@@
 mtext(HEADER)
-
-defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
-defup <- unique(sapply(strsplit(defup," "),"[[",2))
-
-defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
-defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
-
-bg <- rownames(def)
-bg <- unique(sapply(strsplit(bg," "),"[[",2))
 
@@
 
+Get the gene sets loaded in R.
+
 ```{r,gs}
@@
 
-Firstly with the upregulated genes
+Now filter the gene names into three lists, up-regulated, down-regulated and
+background.
+Background is simply all genes that were detected.
+
+```{r,subset1}
+
+defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
+defup <- unique(sapply(strsplit(defup," "),"[[",2))
+
+defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
+defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
+
+bg <- rownames(def)
+bg <- unique(sapply(strsplit(bg," "),"[[",2))
+
+```
+
+Enrichment firstly with upregulated genes
 
@@
 ```{r,oradn}
+
 n_pw=10

### 2023-04-08 manuscript.Rmd
Update manuscript.Rmd

@@
 * bar graph
+  <img src = "https://user-images.githubusercontent.com/88286275/230719934-4fecbbb5-6cde-4167-895b-ee5baabb2459.png" width = "600">
 
 * table
+
+  The following tables gives a summary of the replication results of the articles:
+  S.No.
+|	Pubmed ID	| Replication of (Table/Figure) | Replication was possible?
+1.
+Were the terms the same as authors?
+2.
+Were there other significant terms found which were not mentioned in the article?
+|	Summary | Replication Classified as 1, 2, 3 |
+| --- | -------- | ---- | ---- | ---- | ---- |
+| 1.	| PMC6349697 | Figure 2C and 2D |	1.
+Yes, however, corrected p values are different after replication.
+2.
+Yes, only BP terms were taken into account so several CC and MF significant terms were not mentioned in the paper.
+|	The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed when there is a higher amount of significant GOs and pathways are observed after replication.
+It was impossible to determine if the article utilised a background gene list or not.
+The steps to perform DAVID analysis could have been elaborated in the paper.
+The results shown in the paper are observed after replication but only 6 BP GO significant terms are shown in the paper.
+|	3 [Considering top 6 BP Terms and KEGG pathways] |
 

### 2023-04-08 manuscript.Rmd
Update manuscript.Rmd

@@
 * bar graph
+
   <img src = "https://user-images.githubusercontent.com/88286275/230719934-4fecbbb5-6cde-4167-895b-ee5baabb2459.png" width = "600">
@@
   The following tables gives a summary of the replication results of the articles:
+
 S.No.

### 2023-04-08 manuscript.Rmd
Update manuscript.Rmd

@@
 * table
+  The following table gives a summary of the replication results of the articles:
 
-  The following tables gives a summary of the replication results of the articles:
-
-S.No.
-|	Pubmed ID	| Replication of (Table/Figure) | Replication was possible?
+| S.No.| Pubmed ID | Replication of (Table/Figure) | Replication was possible?
 1.
@@
 | --- | -------- | ---- | ---- | ---- | ---- |
-| 1.	| PMC6349697 | Figure 2C and 2D |	1.
+| 1.
+| PMC6349697 | Figure 2C and 2D |	1.
 Yes, however, corrected p values are different after replication.

### 2023-04-12 manuscript.Rmd
minor updates

@@
 Perampalam & Dick 2020).
-This helps reproducibility greatly, but is insufficient as it lacks information about how the input
-data was used and a record of any parameter selections.
+This helps reproducibility greatly, but is not a complete solution as it lacks information about how
+the input data was used and a record of any parameter selections.
 
@@
 Each article was critically evaluated and given a score of one (1) to three (3).
-Articles rated 1 showed overall poor reproducibility, for example most GO and/or KEGG terms did not
-match upon reproduction, and the conclusions of the study were overall not supported.
-Aticles were rated 2 when at least 50% of the terms are observed as significant after replication,
+Articles rated 1 showed overall poor reproducibility.
+For example most GO and/or KEGG terms mentioned in the results and discussion were not confirmed
+upon reproduction, and the conclusions of the study were overall not supported.
+Articles were rated 2 when at least 50% of the terms are observed as significant after replication,
 according to the original study's significance threshold, and most conclusions were supported.

### 2023-04-13 manuscript.Rmd
Update manuscript.Rmd

@@
 
-| S.No.| Pubmed ID | Replication of (Table/Figure) | Replication was possible?
-1.
-Were the terms the same as authors?
+| S.No.| Pubmed ID | Replication of (Table/Figure) | 1.
+Was Replication possible?
 2.
@@
 |	Summary | Replication Classified as 1, 2, 3 |
-| --- | -------- | ---- | ---- | ---- | ---- |
+| --- | --- | --- | --- | --- | --- |
 | 1.
@@
 |	3 [Considering top 6 BP Terms and KEGG pathways] |
+| 2.
+| PMC6368841 | Figure 3 A - D |	1.
+Only GO Terms were the same as the paper.
+KEGG terms were highly inconsistent.
+2.
+Replicated KEGG Terms (using DAVID) were not mentioned in the results of the paper.
+Shows inconsistency due to the usage of the server.
+Some terms using clusterprofiler in R were mentioned.
+| GO results from the original manuscript is somewhat similar to reproduced analysis.
+The FDR values however are inconsistent.
+KEGG results were highly different.
+The list of genes as input into DAVID analysis is unclear.
+They should have mentioned both up and down-regulated proteins were considered.
+| 2 |
+| 3.
+| PMC6381667 | Figure 1 and Figure 2 | 1.
+Results are not reproducible and are highly inconsistent 2.
+N.A.
+|	1.
+GO terms obtained after replication showed inconsistent results compared to the results in the original article.
+KEGG results were similar.
+2.
+Better naming convention for supplementary file for 117 genes for DAVID input is necessary.
+3.
+GO Results are not reproducible and are highly inconsistent.
+| 2 |
+| 4.
+| PMC6405693	| Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways”	| 1.
+Results do not match.
+2.
+N.A.
+|	1.
+GO analysis from upregulated and downregulated proteins are less similar to the results given in the paper.
+The FDR values for some pathways are highly inconsistent.
+2.KEGG Analysis of all DEGs do not match at all.
+|	1 |
+| 5.
+| PMC6425008 |	Most of the tables and figures of the paper (Figure 5, 6, 7), Table 2	| 1.
+No.
+2.
+N.A.
+|	1.
+This paper could have done proper labelling of supplementary files as it was difficult the gene list for DAVID analysis.
+2.
+Lacks information on when control genes are used and when citral-affected gene lists are used.
+3.
+Replication of GO and KEGG analysis is not consistent with results from the paper.
+4.
+How were Figures 5-7 derived?
+5.
+The overall protocol is not so coherent.
+| 1 |
+| 6.
+| PMC6444048	| The text under “Pathway analysis of SUVmax related genes” And Table 2 |	1.
+Terms are the same but gene counts have been different.
+The DAVID results are somewhat reproducible.
+2.
+No |	1.
+Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” has changed.
+2.
+Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate.
+The results claimed are somewhat similar.
+All the processes have different FDR values.
+| 3 |
+| 7.
+| PMC6463127 |	Table 4, Table 5, Table 6	| 1.
+Yes but the significance differed.
+2.
+Yes, several terms from replication were not mentioned in the paper for GO Analysis.
+| Does not consider corrected p values.
+Overall, significant terms from GO analysis of up and downregulated DEGs are missing upon replication.
+| 2 |
+| 8.
+| PMC6535219 |	Figure 2 | 1.
+Somewhat similar.
+2.
+There were many top significant terms that were not mentioned.	Narrowing/Filtering the given gene list is a required step that could have been mentioned.
+Many terms were not so significant anymore.
+There were new terms that are of top positions after replication.
+| 1 |
+| 9.
+| PMC6539328 |	Figure 1A |	1.
+Mostly not.
+2.
+Yes, many	The significant pathways stated in the paper are not showing up once replicated.
+There is no clarity of the up-regulated genes used for GO and KEGG analysis.
+Only 21 upregulated genes are listed whereas the paper states 4831.
+Also, why are only upregulated genes are being used?
+| 1 |
+| 10. | PMC6542760 | Figure 4 |	1. No. 2. Yes, many significant terms were observed. | Replication results of GO terms are not at all compliant with the results shown in the paper. | 1 |
+| 11.
+| PMC6557785 | Figure 5 A-B |	1.
+Yes, but results in paper are not significant upon replication.
+2.
+Yes | For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
+For KEGG Pathways, there are other pathways that are more significant and contain a higher number of genes enriched for example “Salmonella infection” that has a higher count of genes but is not mentioned in the paper.
+| 2 |
+| 12. | PMC6561911 | Figure 5 a and b	| 1. No. 2. N.A. | The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done | 1 |
+| 13.
+| PMC6580941 | Table 1 and Table 2 | 1.
+Yes.
+2.
+Yes!
+Significant terms like “nucleoplasm”, “RNA Binding”, “protein binding” are missing from the paper.
+|	For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
+The Gene list for input is not so clearly mentioned, it should have been labelled properly in the name of the supplemental file.
+P-values are completely different.
+| 2 |
+| 14.
+| PMC6587650 | 	Figure 2 |	1.
+Yes.
+2.
+Yes, som.
+| The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
+They could have mentioned why they chose the type of sorting shown in the paper to display their results.
+Overall, all results have been replicated.	| 3 [Best] |
+| 15.
+| PMC6591946 | Figure 5 c and Figure 6 a |	1.
+Yes for GO analysis only.
+2.
+Yes for KEGG analysis |	The GO term results of replication are compliant with Figure 5 c of the paper.
+KEGG pathway results are semi-consistent.
+| 2 |
+| 16.
+| PMC6663624 |	Table 3 and Table 4	| 1.
+Yes.
+2.
+Yes | The top 2 significant GO terms in the paper are not significant after replication analysis is done.
+Other GO terms are the same as replicated results.
+For KEGG Analysis, some terms are more significant after replication.
+Eg: PD-L1 expression and PD-1 checkpoint pathway in cancer, Th1 and Th2 cell differentiation, Th17 cell differentiation were more significant after “natural killer (NK) cell-mediated cytotoxicity” KEGG Pathway.
+The paper is somewhat replicable.
+| 1 |
+| 17.
+| PMC6582306 |	Figure 5 |	1.
+Partly.
+2.
+Yes, many.
+|	The paper uses 44 sets of DEGs grouped in 4 major types.
+For replication, OFC Set 9 was used, DAVID GO analysis was done twice where a discrepancy of results was observed.
+Compared to the paper, the most significant term was missing after replication analysis was conducted.
+For OFC set 9, several pathways are seen after replication but only one pathway is shown in paper.
+This should be addressed.
+|	2 |
+| 18.
+| PMC6333352 |	Figure 7 |	1.
+Yes.
+2.
+Yes.
+There was a difference in gene counts as well.
+| DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
+From NERI - BAHN - KEGG Analysis: “Pathways of neurodegeneration - multiple diseases”, “Pathways in cancer” - pathways with the highest gene count (30 and 28 respectively) are not mentioned in the paper.
+Other pathways are present and the paper is somewhat reproducible.
+| 2 |
+| 19.
+| PMC6526186 |	Table 1 |	1.
+Yes.
+2.
+Yes.
+|	The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
+The top 2 KEGG Pathways after replication are “Human cytomegalovirus infection” followed by “Pathways in cancer”.
+The p-values have changed after the results were replicated, compared to the values given in the paper.
+The rest of the pathways are similar to the results given in the paper.
+|	2 |
+| 20.
+| PMC6607402 | Figure 1 A-D |	1.
+Yes.
+2.
+Yes, many!
+|	For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
+Most of the terms are mentioned in the replicated results, but different levels of enrichment.
+| 2 |
 

### 2023-04-13 manuscript.Rmd
Update manuscript.Rmd

@@
 |	Summary | Replication Classified as 1, 2, 3 |
-| --- | --- | --- | --- | --- | --- |
+| --- | --- | ----- | ----- | ----- | --- |
 | 1.
@@
 2.
-There were many top significant terms that were not mentioned.	Narrowing/Filtering the given gene list is a required step that could have been mentioned.
+There were many top significant terms that were not mentioned.
+|	Narrowing/Filtering the given gene list is a required step that could have been mentioned.
 Many terms were not so significant anymore.
@@
 2.
-Yes, many	The significant pathways stated in the paper are not showing up once replicated.
+Yes, many | The significant pathways stated in the paper are not showing up once replicated.
 There is no clarity of the up-regulated genes used for GO and KEGG analysis.

### 2023-04-13 manuscript.Rmd
Update manuscript.Rmd

@@
 
-* bar graph
+* Bar graph
 
   <img src = "https://user-images.githubusercontent.com/88286275/230719934-4fecbbb5-6cde-4167-895b-ee5baabb2459.png" width = "600">
+
+Bar graph denoting the level of reproducibility of the 20 journal articles.
+
+While going through the papers a lot of discrepancies were observed. Out of the 20 journal articles were analysed, 
+thirty five percent (35%) of the papers showed low levels of reproducibility, fifty percent (50%) of the papers showed 
+medium level of reproducibility and only fifteen percent (15 %) of the papers showed a high degree of reproducibility of 
+the DAVID analysis results.
 

### 2023-04-13 manuscript.Rmd
Update manuscript.Rmd

--- 2023-04-13 manuscript.Rmd
@@
 
-* bar graph
+* Bar graph
 
@@
 
-* table
+Bar graph denoting the level of reproducibility of the 20 journal articles.
+
+While going through the papers a lot of discrepancies were observed. Out of the 20 journal articles that were analysed, 
+thirty five percent (35%) of the papers showed low levels of reproducibility, fifty percent (50%) of the papers showed 
+medium level of reproducibility and only fifteen percent (15 %) of the papers showed a high degree of reproducibility of 
+the DAVID analysis results.
+
+* Summary Table
+
   The following table gives a summary of the replication results of the articles:

### 2023-04-19 manuscript.Rmd
manuscript updates

--- 2023-04-13 manuscript.Rmd
@@
     code_folding: hide
-    fig_width: 7
+    fig_width: 5
     fig_height: 5
@@
 
-![Figure 1: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
+![Figure 1. The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
 
@@
 [5](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
-organism under study is human; (2) the tool used was DAVID (Database for Annotation, Visualization,
-and Integrated Discovery)[101](Huang et al, 2007); (3) the omics type was gene expression (array or
-RNA-seq); and (4) that the gene list was provided in the supplementary data.
+organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
+Visualization, and Integrated Discovery)[101](Huang et al, 2007); (3) the omics type was gene
+expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
+The basis for these selection criteria is that this combination represents the most common application
+of enrichment analysis in our previous study, which could be reproduced without the need to
+reanalyse the dataset from raw data.
 
@@
 
-* bar graph
+After conducting this reproduction study according to the methods described in the respective articles,
+the 20 articles were classified based on the degree of reproducibility (Figure 2).
+Six studies were classified as low reproducibility, while 10 were classified as medium and just three
+were classified as high.
 
-  <img src = "https://user-images.githubusercontent.com/88286275/230719934-4fecbbb5-6cde-4167-895b-ee5baabb2459.png" width = "600">
+```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 2. Examining reproducibility in a sample of 20 journal articles."}
 
-* table
-  The following table gives a summary of the replication results of the articles:
+score1 <- c("PMC6405693","PMC6425008","PMC6535219","PMC6539328","PMC6542760","PMC6561911","PMC6663624")
+score2 <- c("PMC6368841","PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
+  "PMC6582306","PMC6333352","PMC6526186","PMC6607402")
+score3 <- c("PMC6349697","PMC6444048","PMC6587650")
+l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)
 
-| S.No.| Pubmed ID | Replication of (Table/Figure) | 1.
+barplot(unlist(lapply(l,length)),col="lightgray",
+  xlab="reproducibility score",ylab="no. articles",
+  ylim=c(0,11))
+
+grid()
+text(0.7,1:7-0.5,labels=score1)
+text(1.9,1:10-0.5,labels=score2)
+text(3.1,1:3-0.5,labels=score3)
+
+pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
+pc2 <- paste(pc,"%",sep="")
+text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)
+
+```
+
+| S.No.
+Pubmed ID: Replication of (Table/Figure) | A.
 Was Replication possible?
-2.
+B.
 Were there other significant terms found which were not mentioned in the article?
-|	Summary | Replication Classified as 1, 2, 3 |
-| --- | --- | ----- | ----- | ----- | --- |
-| 1.
-| PMC6349697 | Figure 2C and 2D |	1.
+| Replication score and justification |
+| --- | --- | --- |
+| 1.PMC6349697: Figure 2C and 2D | A.
 Yes, however, corrected p values are different after replication.
-2.
+B.
 Yes, only BP terms were taken into account so several CC and MF significant terms were not mentioned in the paper.
-|	The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed when there is a higher amount of significant GOs and pathways are observed after replication.
+|  **3.** Considering top 6 BP Terms and KEGG pathways.
+The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed when there is a higher amount of significant GOs and pathways are observed after replication.
 It was impossible to determine if the article utilised a background gene list or not.
@@
 The results shown in the paper are observed after replication but only 6 BP GO significant terms are shown in the paper.
-|	3 [Considering top 6 BP Terms and KEGG pathways] |
-| 2.
-| PMC6368841 | Figure 3 A - D |	1.
+|
+| 2.PMC6368841: Figure 3 A - D | A.
 Only GO Terms were the same as the paper.
 KEGG terms were highly inconsistent.
-2.
+B.
 Replicated KEGG Terms (using DAVID) were not mentioned in the results of the paper.
@@
 Some terms using clusterprofiler in R were mentioned.
-| GO results from the original manuscript is somewhat similar to reproduced analysis.
+| **2.** GO results from the original manuscript is somewhat similar to reproduced analysis.
 The FDR values however are inconsistent.
@@
 They should have mentioned both up and down-regulated proteins were considered.
-| 2 |
-| 3.
-| PMC6381667 | Figure 1 and Figure 2 | 1.
-Results are not reproducible and are highly inconsistent 2.
+|
+| 3.PMC6381667: Figure 1 and Figure 2 | A.
+Results are not reproducible and are highly inconsistent B.
 N.A.
-|	1.
-GO terms obtained after replication showed inconsistent results compared to the results in the original article.
+| **2.** GO terms obtained after replication showed inconsistent results compared to the results in the original article.
 KEGG results were similar.
-2.
 Better naming convention for supplementary file for 117 genes for DAVID input is necessary.
-3.
 GO Results are not reproducible and are highly inconsistent.
-| 2 |
-| 4.
-| PMC6405693	| Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways”	| 1.
+|
+| 4.PMC6405693: Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways” | A.
 Results do not match.
-2.
+B.
 N.A.
-|	1.
-GO analysis from upregulated and downregulated proteins are less similar to the results given in the paper.
+| **1.** GO analysis from upregulated and downregulated proteins are less similar to the results given in the paper.
 The FDR values for some pathways are highly inconsistent.
-2.KEGG Analysis of all DEGs do not match at all.
-|	1 |
-| 5.
-| PMC6425008 |	Most of the tables and figures of the paper (Figure 5, 6, 7), Table 2	| 1.
+KEGG Analysis of all DEGs do not match at all.
+|
+| 5.PMC6425008: Most of the tables and figures of the paper (Figure 5, 6, 7), Table 2 | A.
 No.
-2.
+B.
 N.A.
-|	1.
-This paper could have done proper labelling of supplementary files as it was difficult the gene list for DAVID analysis.
-2.
+| **1.** This paper could have done proper labelling of supplementary files as it was difficult the gene list for DAVID analysis.
 Lacks information on when control genes are used and when citral-affected gene lists are used.
-3.
 Replication of GO and KEGG analysis is not consistent with results from the paper.
-4.
 How were Figures 5-7 derived?
-5.
 The overall protocol is not so coherent.
-| 1 |
-| 6.
-| PMC6444048	| The text under “Pathway analysis of SUVmax related genes” And Table 2 |	1.
+|
+| 6.PMC6444048: The text under “Pathway analysis of SUVmax related genes” And Table 2 | A.
 Terms are the same but gene counts have been different.
 The DAVID results are somewhat reproducible.
-2.
-No |	1.
-Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” has changed.
-2.
+B.
+No | **3.** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” has changed.
 Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate.
@@
 All the processes have different FDR values.
-| 3 |
-| 7.
-| PMC6463127 |	Table 4, Table 5, Table 6	| 1.
+|
+| 7.PMC6463127: Table 4, Table 5, Table 6 | A.
 Yes but the significance differed.
-2.
+B.
 Yes, several terms from replication were not mentioned in the paper for GO Analysis.
-| Does not consider corrected p values.
+| **2.** Does not consider corrected p values.
 Overall, significant terms from GO analysis of up and downregulated DEGs are missing upon replication.
-| 2 |
-| 8.
-| PMC6535219 |	Figure 2 | 1.
+|
+| 8.PMC6535219: Figure 2 | A.
 Somewhat similar.
-2.
+B.
 There were many top significant terms that were not mentioned.
-|	Narrowing/Filtering the given gene list is a required step that could have been mentioned.
+| **1.** Narrowing/Filtering the given gene list is a required step that could have been mentioned.
 Many terms were not so significant anymore.
 There were new terms that are of top positions after replication.
-| 1 |
-| 9.
-| PMC6539328 |	Figure 1A |	1.
+|
+| 9.PMC6539328: Figure 1A | A.
 Mostly not.
-2.
-Yes, many | The significant pathways stated in the paper are not showing up once replicated.
+B.
+Yes, many | **1.** The significant pathways stated in the paper are not showing up once replicated.
 There is no clarity of the up-regulated genes used for GO and KEGG analysis.
@@
 Also, why are only upregulated genes are being used?
-| 1 |
-| 10. | PMC6542760 | Figure 4 |	1. No. 2. Yes, many significant terms were observed. | Replication results of GO terms are not at all compliant with the results shown in the paper. | 1 |
-| 11.
-| PMC6557785 | Figure 5 A-B |	1.
+|
+| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1.** Replication results of GO terms are not at all compliant with the results shown in the paper. |
+| 11.PMC6557785: Figure 5 A-B | A.
 Yes, but results in paper are not significant upon replication.
-2.
-Yes | For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
+B.
+Yes | **2.** For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
 For KEGG Pathways, there are other pathways that are more significant and contain a higher number of genes enriched for example “Salmonella infection” that has a higher count of genes but is not mentioned in the paper.
-| 2 |
-| 12. | PMC6561911 | Figure 5 a and b	| 1. No. 2. N.A. | The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done | 1 |
-| 13.
-| PMC6580941 | Table 1 and Table 2 | 1.
+|
+| 12.PMC6561911: Figure 5 a and b | A. No. B. N.A. | **1.** The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done |
+| 13.PMC6580941: Table 1 and Table 2 | A.
 Yes.
-2.
+B.
 Yes!
 Significant terms like “nucleoplasm”, “RNA Binding”, “protein binding” are missing from the paper.
-|	For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
+| **2.** For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
 The Gene list for input is not so clearly mentioned, it should have been labelled properly in the name of the supplemental file.
 P-values are completely different.
-| 2 |
-| 14.
-| PMC6587650 | 	Figure 2 |	1.
+|
+| 14.PMC6587650: Figure 2 | A.
 Yes.
-2.
+B.
 Yes, som.
-| The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
+| **3.** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
 They could have mentioned why they chose the type of sorting shown in the paper to display their results.
-Overall, all results have been replicated.	| 3 [Best] |
-| 15.
-| PMC6591946 | Figure 5 c and Figure 6 a |	1.
+Overall, all results have been replicated.
+|
+| 15.PMC6591946: Figure 5 c and Figure 6 a | A.
 Yes for GO analysis only.
-2.
-Yes for KEGG analysis |	The GO term results of replication are compliant with Figure 5 c of the paper.
+B.
+Yes for KEGG analysis | **2.** The GO term results of replication are compliant with Figure 5 c of the paper.
 KEGG pathway results are semi-consistent.
-| 2 |
-| 16.
-| PMC6663624 |	Table 3 and Table 4	| 1.
+|
+| 16.PMC6663624: Table 3 and Table 4 | A.
 Yes.
-2.
-Yes | The top 2 significant GO terms in the paper are not significant after replication analysis is done.
+B.
+Yes | **1.** The top 2 significant GO terms in the paper are not significant after replication analysis is done.
 Other GO terms are the same as replicated results.
@@
 The paper is somewhat replicable.
-| 1 |
-| 17.
-| PMC6582306 |	Figure 5 |	1.
+|
+| 17.PMC6582306: Figure 5 | A.
 Partly.
-2.
+B.
 Yes, many.
-|	The paper uses 44 sets of DEGs grouped in 4 major types.
+| **2.** The paper uses 44 sets of DEGs grouped in 4 major types.
 For replication, OFC Set 9 was used, DAVID GO analysis was done twice where a discrepancy of results was observed.
@@
 This should be addressed.
-|	2 |
-| 18.
-| PMC6333352 |	Figure 7 |	1.
+|
+| 18.PMC6333352: Figure 7 | A.
 Yes.
-2.
+B.
 Yes.
 There was a difference in gene counts as well.
-| DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
+| **2.** DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
 From NERI - BAHN - KEGG Analysis: “Pathways of neurodegeneration - multiple diseases”, “Pathways in cancer” - pathways with the highest gene count (30 and 28 respectively) are not mentioned in the paper.
 Other pathways are present and the paper is somewhat reproducible.
-| 2 |
-| 19.
-| PMC6526186 |	Table 1 |	1.
+|
+| 19.PMC6526186: Table 1 | A.
 Yes.
-2.
+B.
 Yes.
-|	The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
+| **2.** The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
 The top 2 KEGG Pathways after replication are “Human cytomegalovirus infection” followed by “Pathways in cancer”.
@@
 The rest of the pathways are similar to the results given in the paper.
-|	2 |
-| 20.
-| PMC6607402 | Figure 1 A-D |	1.
+|
+| 20.PMC6607402: Figure 1 A-D | A.
 Yes.
-2.
+B.
 Yes, many!
-|	For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
+| **2.** For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
 Most of the terms are mentioned in the replicated results, but different levels of enrichment.
-| 2 |
+|
 
-*
+: Table 1. Classification and justification for each article.
 
-It is important to note that the DAVID 6.8 server has been replaced by DAVID 2022 server, starting
-from June 1, 2022.
-For our analysis, DAVID version 6.8 was used for replication procedures.
-It was ensured that our analyses got completed by 31 May 2022.
-It is no longer possible to reproduce analyses conducted in DAVID 6.8
+Taken together, the ability to replicate DAVID enrichment analysis using the supplementary gene
+lists was highly variable.
+We note that DAVID 6.8 server has been replaced by DAVID 2022 server as of June 1, 2022, meaning
+that is no longer possible to reproduce DAVID 6.8 analyses.
 
@@
 * documentation
+
+* sharing/archiving
 
@@
 
+It is important to note that the DAVID 6.8 server has been replaced by DAVID 2022 server, starting
+from June 1, 2022.
+For our analysis, DAVID version 6.8 was used for replication procedures.
+It was ensured that our analyses got completed by 31 May 2022.
+It is no longer possible to reproduce analyses conducted in DAVID 6.8
+
 While there have been many general guides to computatonal reproducibility in the literature,
@@
 research resources and greater trust/support from the general community.
+
+* web tool developers could provide archived versions of the tool as docker images to facilitate
+reproducibility while keeping the online version up to date.
 

### 2023-04-26 manuscript.Rmd
manuscript update

@@
 To address this, many articles have been written to endorse the use of "virtual machines" or
-"containerisation" to encapsulate the "environment" including the operating system and dependancies
+"containerisation" to control the "environment" including the operating system and dependancies
 along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
@@
 to novice bioinformaticians.
+Version control keeps track of code and documentation changes over time and manages the edits made by
+multiple contributors, taking over some of the functions of a laboratory notebook for computer based
+projects.
 Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
@@
 Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
-for each figure sequentially, making the entire manuscript reproducible by executing just one command.
-Taken together, computational reproducibility is a spectrum (**Figure 1**).
+for each figure sequentially, making the entire manuscript reproducible by executing just one command
+any without any manual steps.
+Taken together, computational reproducibility is a spectrum (**Figure 1**), and rests upon five key
+pillars mentioned above (**Figure 2**).
 Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
@@
 ![Figure 1. The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
+
+![Figure 2. The five pillars of computational reproducibility.](img/pillarsV2.png "The five pillars of computational reproducibility. .")
 
@@
 Although our previous work suggests that enrichment analyses generally lack sufficient methodological
-detail [5](Wijesooriya et al, 2022), we don't have empirical evidence confirming irreproducibility.
-Therefore before recommending or prescribing Docker-based bioinformatics workflows, it is necessary
+detail [5](Wijesooriya et al, 2022), we don't have empirical proof confirming irreproducibility.
+Therefore before recommending or prescribing new workflows, it is necessary
 to assess directly whether current practices are appropriate, by trying to reproduce the enrichment
@@
 
-<Anusuiya to populate>
-<references to start at [101](Author Year)>
-<Wanted to check how this looks in the file>
-
 From a list of 186 PubMed central articles using enrichment analysis that we examined previously
@@
 of enrichment analysis in our previous study, which could be reproduced without the need to
-reanalyse the dataset from raw data.
+reanalyse the raw dataset from scratch.
 
@@
 The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles with DAVID version 6.8 (March-May 2022).
+sections of the 20 articles using DAVID version 6.8 (March-May 2022).
 
@@
 After conducting this reproduction study according to the methods described in the respective articles,
-the 20 articles were classified based on the degree of reproducibility (Figure 2).
+the 20 articles were classified based on the degree of reproducibility (Figure 3).
 Six studies were classified as low reproducibility, while 10 were classified as medium and just three
 were classified as high.
+Justifications for these classifications are provided in Table 1.
 
-```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 2. Examining reproducibility in a sample of 20 journal articles."}
+```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 3. Examining reproducibility in a sample of 20 journal articles."}
 
@@
 
-## A container based approach to addressing reproducibility of enrichment analysis
+## A container-based approach to addressing reproducibility of enrichment analysis
 
@@
 
-Infographic:
+We have provided to the community a step-by-step protocol for how to conduct an extremely reproducible
+enrichment analysis, which is attached as a supplementary file (Suppl Info 1), and published to
+protocols.io with a permissive license.
+This protocol has the features of reproducibility mentioned in **Figure 1** and **Figure 2** including
+version control with git and GitHub, compute environment control using Docker, literate programming
+with Rmarkdown, persistent code and data sharing with Zenodo and documentation of all of these to
+empower others to reproduce the workflow.
 
-* source control
-
-* containerisation
-
-* literate programming
-
-* documentation
-
-* sharing/archiving
-
-The templates provided include the Dockerfile and an Rmarkdown-based workflow script.
-These templates are intended to be altered by the end user, but here is an overview of the contents of
-each.
+Central to this protocol is the provision of template Dockerfile and an Rmarkdown script, which
+are intended to be remixed and altered by the end user to suit the needs of their project.
 
@@
 
-The `example_workflow.Rmd` is the Rmarkdown based script that undertakes the enrichment analysis.
+The Rmarkdown script called `example_workflow.Rmd` undertakes the transcriptome and enrichment analyses.
 Rmarkdown allows the unification of extended descriptions, code and results into a single document,
@@
 data set from DEE2, inspecting data quality, differential expression and two types of enrichment
-analysis; over-representation analysis (ORA) and functional class scoring (FCS) (**Figure XX**).
+analysis; over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
 For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
@@
 
-```{r,diagram,echo=FALSE,fig.cap = "Figure XX: Flow chart of example workflow."}
+```{r,diagram,echo=FALSE,fig.cap = "Figure 4: Flow chart of example workflow."}
 
@@
 
-### Findings
+To reproduce this work is relatively quick and simple.
+Installing Docker requires one command and takes about 1 min.
+Pulling the image takes about 1 minute.
+Launching a container takes a few seconds, and executing the Rmarkdown script takes 80 seconds.
+
+### Demonstration
 
@@
 in triplicate [208] (Lund et al, 2014).
-
 We use this analysis to try to understand better the effect of DNA methylatransferase inhibitors on
@@
 
-After fetching the data from DEE2, the number of reads per sample was confirmed to be in the range of
+For quality control the number of reads per sample was calculated, and found to be in the range of
 11-16M.
@@
 1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
-(Figure XXA).
+(Figure 5A).
 These differentially expressed gene sets were subjected to ORA analysis separately, using the list of
@@
 which tends to highlight smaller and more specific gene sets undergoing more drastic differential
-expression, rather than large gene sets that undergo very small expression changes.
+expression, rather than large gene sets that undergo small expression changes.
 This led to the identification of upregulated pathways related to nuclear RNA export,
@@
 were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
-(host defence) (Figure XXB).
+(host defence) (Figure 5B).
 
@@
 FCS upregulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
-and defective homologous recombination repair (Figure XXC).
+and defective homologous recombination repair (Figure 5C).
 Downregulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
@@
 Jaccard index of 0.54, however these method disagreed in downregulated pathways, with a Jaccard
-index of only 0.27 (Figure XXD).
+index of only 0.27 (Figure 5D).
 
-![Figure XX: Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
+![Figure 5: Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
 
 ## Concluding remarks
+
+Although our reproducibility study was relatively small, there are a few key observations.
+
+We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
+
+A literature search identified very few studies conducted to systematically examine the reproducibility
+of bioinformatics research.
+
+There were several case studies, but the overall lack of quantitative data appears to be out of
+proportion to the number of review and commentary articles recommending computational best-practices.
+
+Ionnidis et al (2009) examined a set of 10 gene expression studies with available data published in
+2005-2006, and found only two could be reproduced to a satisfactory level
+(https://www.nature.com/articles/ng.295).
+
+In the realm of social sciences,
+
+Only 40-50% of social psychology studies could be reproduced, depending on how reproduction is defined
+(https://plato.stanford.edu/entries/scientific-reproducibility/)
+
+Reproducibility in the Social Sciences
+(https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-090221-035954)
+
+we can conclude that just 15% of studies could be
 
@@
 reproducibility while keeping the online version up to date.
+
+As scholars, we should be following best practice in what we do and publish.
+
+Rather than doing the minimum amount of work to get our results published,
+
+Docker install in 32 seconds
+Docker pull in 47 seconds
 

### 2023-04-26 example.Rmd
refinements

@@
   main="top DE Reactomes",
-  xlab="ES")
+  xlab="ES",
+  cex.main=0.9)
 
@@
 
-```{r,venn1}
+```{r,venn1,fig.width=6}
 
@@
 par(mfrow=c(2,1))
-plot(euler(v1),quantities = TRUE)
-plot(euler(v2),quantities = TRUE)
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))
 

### 2023-04-26 manuscript.Rmd
refinements

@@
 Version control keeps track of code and documentation changes over time and manages the edits made by
-multiple contributors, taking over some of the functions of a laboratory notebook for computer based
-projects.
+multiple contributors.
 Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
@@
 of work [21](Wijesooriya et al, 2022).
-Web-based tools are a potential concern, firstly because sometimes methods, options and parameters
-are not completely described, and secondly, algorithms and functional annotation sets undergo
-regular updates.
-These historical versions are typically not archived, which hampers later reproducibility.
+Although web-based tools are convenient, they are a potential reproducibility concern.
+Firstly, incomplete reporting of methods, options and parameters is common.
+Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
+typically not made available hampering later reproducibility.
 In some cases the entire web resource can become unavailable, and hence irreproducible due to
@@
 
-Therefore the most complete reproducibility solution is to use containerisation to provide a scripted
+A solution to this is to use containerisation to provide a scripted
 workflow with prescribed links to research data, external databases and complete record of parameter
 selection.
-Designed correctly, these containers could encapsulate bioinformatics workflows and ensure their
+When designed correctly, these containers encapsulate bioinformatics workflows and could ensure their
 computational reproducibility for decades to come with just one or a few simple commands.
@@
 
-But is this extra work really necessary?
-Although our previous work suggests that enrichment analyses generally lack sufficient methodological
-detail [5](Wijesooriya et al, 2022), we don't have empirical proof confirming irreproducibility.
-Therefore before recommending or prescribing new workflows, it is necessary
+However, before recommending or prescribing such workflows, it is necessary
 to assess directly whether current practices are appropriate, by trying to reproduce the enrichment
@@
 
-Firstly, the data presented in the paper was compared with the results we obtained, recording any
+The data presented in the paper was compared with the results we obtained, recording any
 discrepancies.
-Secondly, the statements made in the abstract, results, and discussion sections were carefully examined
-and cross-checked with our results to understand whether those assertions survive reproduction or not.
+Additionally, the statements made in the abstract, results, and discussion sections were carefully
+examined and cross-checked with our results to understand whether those assertions survive reproduction
+or not.
 
@@
 shown in the article observed again in reproduction, and all conclusions were supported in reproduction.
-Results are summarised in table and bar chart form.
 
@@
 
-We have provided to the community a step-by-step protocol for how to conduct an extremely reproducible
+We have developed a step-by-step protocol for how to conduct an extremely reproducible
 enrichment analysis, which is attached as a supplementary file (Suppl Info 1), and published to
@@
 
-The Rmarkdown script called `example_workflow.Rmd` undertakes the transcriptome and enrichment analyses.
+The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
 Rmarkdown allows the unification of extended descriptions, code and results into a single document,
@@
 
-To reproduce this work is relatively quick and simple.
-Installing Docker requires one command and takes about 1 min.
-Pulling the image takes about 1 minute.
-Launching a container takes a few seconds, and executing the Rmarkdown script takes 80 seconds.
+To reproduce this work is relatively quick and simple, requiring six commands and approximately four
+minutes to install Docker, pull the image, run the container, run the Rmarkdown script exit the container
+and copy the newly created report from the container to the current working directory (Box 1).
+
+```bash
+sudo apt update && sudo apt install docker-ce -y  # 60 sec
+docker pull mziemann/enrichment_recipe  # 60 sec
+docker run -it --entrypoint /bin/bash mziemann/enrichment_recipe #few seconds
+Rscript -e 'rmarkdown::render("example.Rmd")' # 90 sec
+exit
+docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
+```
+<p style="text-align: center;">
+*Box 1. Reproduce a transcriptome and enrichment analysis with Docker.*
+</p>
 
@@
 
-To demonstrate the utility of this workflow, we present the results of an example data analysis;
+To demonstrate the utility of this workflow, we present the results of this example data analysis;
 it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
-in triplicate [208] (Lund et al, 2014).
-We use this analysis to try to understand better the effect of DNA methylatransferase inhibitors on
+in triplicate [208](Lund et al, 2014).
+We use this analysis to better understand the effect of DNA methylatransferase inhibitors on
 genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
@@
 (Figure 5A).
-These differentially expressed gene sets were subjected to ORA analysis separately, using the list of
+Up and down-regulated gene sets were subjected to ORA separately, using the list of
 13,168 genes as the background, and gene sets from REACTOME.
@@
 (https://www.nature.com/articles/ng.295).
+
+http://www.bioinformatics.deib.polimi.it/geco/workshop/reproducibility.pdf
 

### 2023-04-27 manuscript.Rmd
manuscript updates

@@
 
-<b>Alt titles:</b>
-
-* "Enrichment analysis with gold standard reproducibility"
-
-* "A guide to completely reproducible enrichment analysis"
-
 Mark Ziemann<sup>1</sup>*, Anusuiya Bora<sup>2</sup>
@@
 
-asddsadasd
+Irreproducible research is a significant problem that wastes resources undermines the public
+perception of science.
+Previous work has highlighted that published enrichment analyses suffers from statistical and reporting
+flaws.
+We sought to determine whether this translates into irreproducibility by examining whether the
+findings of 20 open-access articles describing enrichment analysis were reproducible.
+We find that only three articles exhibited a high degree of concordance, while seven exhibited poor
+reproducibility.
+The findings indicate that current enrichment analysis practices are inadequate.
+To address this, we provide a set of boilerplate scripts and a step-by-step protocol to adapt them to
+a new project's needs.
+This recipe is built according to a framework we call "five pillars of computational reproducibility"
+, which when implemented together facilitate extreme reproducibility.
 
@@
 [1](National Academies of Sciences 2019, p1).
-According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was "significant
+According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was a "significant
 crisis" of reproducibility [2](Baker 2016).
@@
 To address this, many articles have been written to endorse the use of "virtual machines" or
-"containerisation" to control the "environment" including the operating system and dependancies
+"containerisation" to control the computing environment including the operating system and dependancies
 along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
@@
 for each figure sequentially, making the entire manuscript reproducible by executing just one command
-any without any manual steps.
-Taken together, computational reproducibility is a spectrum (**Figure 1**), and rests upon five key
+and without any manual steps.
+Taken together, computational reproducibility is a spectrum (**Figure 1**) [17](Akalin 2018), and rests upon five key
 pillars mentioned above (**Figure 2**).
@@
 
-![Figure 1. The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")
+![Figure 1.
+The computational research reproducibility spectrum.
+Creator: Altuna Akalin.
+Republished with permission [17](Akalin 2018).](img/repro.webp "The computational research reproducibility spectrum.")
 
@@
 such as the detection of pathway activation in development or disease
-[17,18](Slonim 2002; Khatri et al, 2012).
+[18,19](Slonim 2002; Khatri et al, 2012).
 While enrichment analysis is undoubtedly a key method for the interpretation of large
@@
 (ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
-[19-22](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
+[20-23](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
 Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
@@
 point-and-click tools appear to be far more popular than computer script based methods for this type
-of work [21](Wijesooriya et al, 2022).
+of work [22](Wijesooriya et al, 2022).
 Although web-based tools are convenient, they are a potential reproducibility concern.
@@
 Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
-typically not made available hampering later reproducibility.
+typically not made available, hampering later reproducibility.
 In some cases the entire web resource can become unavailable, and hence irreproducible due to
-a phenomenon known as link decay [23](Hennessey & Ge 2013).
+a phenomenon known as link decay [24](Hennessey & Ge 2013).
 
 A small number of enrichment analysis applications have been developed to have beneficial properties
-of being "containerised" and with archived historical versions [24,25](Ge et al, 2020;
+of being "containerised" and with archived historical versions [25,26](Ge et al, 2020;
 Perampalam & Dick 2020).
@@
 
-We address this firstly by providing working templates for docker image and Rmarkdown scripts for
+We address this firstly by providing working templates for Docker image and Rmarkdown scripts for
 common transcriptome and enrichment analysis routines.
 Secondly, we provide step-by-step written and video guides to help users customise these templates for
-their own work, deploy the analysis, verify the results, and archive the Docker image.
+their own work, deploy the analysis, verify the results, and share the Docker image.
 
 However, before recommending or prescribing such workflows, it is necessary
-to assess directly whether current practices are appropriate, by trying to reproduce the enrichment
+to assess directly whether current practices are adequate, by trying to reproduce the enrichment
 analyses of previously published studies, and examining whether the conclusions of those articles
@@
 In the following section we investigate the computational reproducibility of enrichment analyses
-shown in a sample of 20 journal articles published in 2019.
+shown in a sample of 20 open-access journal articles published in 2019.
 This should provide a sound basis for recommendations.
@@
 From a list of 186 PubMed central articles using enrichment analysis that we examined previously
-[5](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
+[22](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
 organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
-Visualization, and Integrated Discovery)[101](Huang et al, 2007); (3) the omics type was gene
+Visualization, and Integrated Discovery)[27](Huang et al, 2007); (3) the omics type was gene
 expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
@@
 of enrichment analysis in our previous study, which could be reproduced without the need to
-reanalyse the raw dataset from scratch.
+reanalyse the raw dataset from scratch. These articles are listed in the bibliography [28-47]().
 
 The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
-special attention to preventing gene name errors from occurring[102](Zeeberg et al, 2004).
+special attention to preventing gene name errors from occurring [48](Zeeberg et al, 2004).
 The reproducibility analysis was performed using the respective methodologies stated in the methods
@@
 
-### Results and discussion
+### Results
 
@@
 lists was highly variable.
-We note that DAVID 6.8 server has been replaced by DAVID 2022 server as of June 1, 2022, meaning
-that is no longer possible to reproduce DAVID 6.8 analyses.
 
@@
 It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
-analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` [201](Ziemann et al, 2019),
-differential expression with `DESeq2` [202](Love & Huber, 2014),
+analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` [49](Ziemann et al, 2019),
+differential expression with `DESeq2` [50](Love & Huber, 2014),
 enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
-[203-205](Yu et al, 2012; Korotkevich et al, 2016; Kaspi & Ziemann 2020).
+[51-53](Yu et al, 2012; Korotkevich et al, 2016; Kaspi & Ziemann 2020).
 It also contains instructions to clone a github repository, which contains the scripts that run the
 analytical workflow.
-Lastly, it copies a gene set library from Reactome [206](Gillespie et al, 2022) with a clear version
+Lastly, it copies a gene set library from Reactome [54](Gillespie et al, 2022) with a clear version
 history to guarantee provenance of the annotation data.
@@
 which is becoming ubiquitous in bioinformatics and data science more generally
-[207](Grolemund & Wickham 2017, p423).
+[55](Grolemund & Wickham 2017, p423).
 The ability to provide extended descriptions is helpful for providing contextualising information about
@@
 For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
-[22](Zhao & Rhee 2023).
+[23](Zhao & Rhee 2023).
 
-```{r,diagram,echo=FALSE,fig.cap = "Figure 4: Flow chart of example workflow."}
+```{r,diagram,echo=FALSE,fig.cap = "Figure 4. Flow chart of example workflow."}
 
@@
 
+The software versions used in this example analysis are provided in Table 2.
+
+| Software | Version | Purpose |
+| --- | --- | --- |
+| Ubuntu | 22.04 | Operating System |
+| R | 4.2.2 | Statistical computing language |
+| nano | 6.2 | text editing |
+| git | 2.34.1 | source control |
+| R/kableExtra | 1.3.4 | render formatted tables |
+| R/vioplot | 0.4.0 | violin charts |
+| R/gplots | 3.1.3 | heatmaps |
+| R/eulerr | 7.0.0 | Venn diagrams |
+| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
+| BioC/DESeq2 | 1.38.3 | differential expression |
+| BioC/fgsea | 1.24.0 | functional class scoring |
+| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
+| Bioc/mitch | 1.10.0 | multi-contrast enrichment |
+
+: Table 2. The software bundled in the example Docker image.
+
 To reproduce this work is relatively quick and simple, requiring six commands and approximately four
@@
 it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
-in triplicate [208](Lund et al, 2014).
+in triplicate [56](Lund et al, 2014).
 We use this analysis to better understand the effect of DNA methylatransferase inhibitors on
@@
 
-![Figure 5: Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
+![Figure 5. Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
 
@@
 Although our reproducibility study was relatively small, there are a few key observations.
+We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
+In a way, this is surprising, given the simplicity of using DAVID, only
+requiring the user to provide a list of genes in a text box and select some options.
+These irregularities may have arisen inadvertently from several sources:
 
-We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
+* The methods stated in the respective articles are inconsistent with what was actually done.
 
-A literature search identified very few studies conducted to systematically examine the reproducibility
-of bioinformatics research.
+* The list of genes in the supplement is not the same as what was used for enrichment analysis.
 
-There were several case studies, but the overall lack of quantitative data appears to be out of
-proportion to the number of review and commentary articles recommending computational best-practices.
+* Ambiguity around whether up- and down-regulate genes were considered
+separately or together
 
-Ionnidis et al (2009) examined a set of 10 gene expression studies with available data published in
-2005-2006, and found only two could be reproduced to a satisfactory level
-(https://www.nature.com/articles/ng.295).
+* Ambiguity around whether all GO classes were considered or just one, such as "biological process".
 
-http://www.bioinformatics.deib.polimi.it/geco/workshop/reproducibility.pdf
+* Ambiguity around whether a custom background gene set was used.
 
-In the realm of social sciences,
+The observed lack of reproducibility might be expected given that substantial methodological flaws
+were present in 85% of enrichment analyses [22](Wijesooriya et al, 2022).
+A non-exhaustive literature search identified a small number of case studies, but a dearth of systematic
+examinations of bioinformatics research reproducibility.
+Ionnidis et al (2009)[57] examined a set of 10 gene expression studies with available data published in
+2005-2006, and found only two could be reproduced to a satisfactory level.
 
-Only 40-50% of social psychology studies could be reproduced, depending on how reproduction is defined
-(https://plato.stanford.edu/entries/scientific-reproducibility/)
+This problem is not limited to bioinformatics.
+In the field of drug development, according to a 2011 study by Bayer, only 25% of projects could be
+replicated to a high degree [58](Prinz et al, 2011).
+When the US Biotech firm Amgen set out to reproduce 53 landmark cancer studies, they were able to
+reproduce only 6 (11%) [59](Begley & Ellis 2012).
 
-Reproducibility in the Social Sciences
-(https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-090221-035954)
+It is clear that as a field, the life sciences need to raise the rate of reproducibility or risk the
+public perceiving it as unreliable and wasteful.
 
-we can conclude that just 15% of studies could be
+We propose the five pillars (**Figure 2**), as a framework to guarantee computational reproducibility.
+However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
+protocol to guide them on how to customise boilerplate Dockerfile and Rmarkdown scripts to suit their
+project's needs.
 
-It is important to note that the DAVID 6.8 server has been replaced by DAVID 2022 server, starting
-from June 1, 2022.
-For our analysis, DAVID version 6.8 was used for replication procedures.
-It was ensured that our analyses got completed by 31 May 2022.
-It is no longer possible to reproduce analyses conducted in DAVID 6.8
+Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
+we estimate at 15-20 hours.
+This is significantly longer than using a web-based tool that provides results in just a few
+minutes, but we argue the time investment is well spent.
+It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
+simple and fast (Box 1), which should aid the early identification of errors.
 
-While there have been many general guides to computatonal reproducibility in the literature,
-there is a need for boilerplate code templates that researchers can quickly adapt/tweak for their own
-projects.
+We also want to emphasise that there is still a definite need for graphical and web-based tools.
+These are good for data exploration, to quickly identify trends that can be followed up with more
+rigourous methods.
+Moreover, a lot of work is being done to make such point-and-click tools more reproducible.
+The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
+workflows that enable reproducibility [60-62](Kuehn et al; 2008; Goecks et al, 2010; Teichman et al, 2023).
+The Degust differential expression suite is a graphical interface to popular
+expression tools like edgeR and limma, and allows users to download the script to facilitate future
+non-interactive reproducibility [63](Powell 2019).
+Such functionality could be written into existing enrichment analysis tools like DAVID, and together with
+retention of historical gene annotation sets could provide a "good enough" solution to reproducibility.
 
-The demand for this is expected to be huge, as there are massive numbers of published articles featuring
-such analyses.
-Just in 2022, there were 9,917 PubMed articles with the keywords pathway/enrichment/ontology analysis
-in the title or abstract alone.
+Although we hope that extremely reproducible research will become more widespread, we do not
+anticipate any dramatic changes in researcher behaviour unless the motivating factors are augmented.
+Reward structures in academia encourage individuals to publish many articles, and reproducibility
+is not a significant factor unless the issues are extreme enough to provoke a retraction by the journal,
+which could tarnish the reputation of an author or their institution.
+Solutions to this could include mandates by funding agencies or institutions [64](Garijo et al, 2013),
+and/or badging of articles that comply with best practices [65](Blohowiak et al, 2023).
 
-The more rigourous, well documented and reproducible these studies are, we will find less wasted
-research resources and greater trust/support from the general community.
+## Availability of materials
 
-* web tool developers could provide archived versions of the tool as docker images to facilitate
-reproducibility while keeping the online version up to date.
+* Protocol: protocols.io
 
-As scholars, we should be following best practice in what we do and publish.
+* Boilerplate code: https://github.com/markziemann/enrichment_recipe
 
-Rather than doing the minimum amount of work to get our results published,
-
-Docker install in 32 seconds
-Docker pull in 47 seconds
+* Example Docker image: https://hub.docker.com/r/mziemann/enrichment_recipe
 
@@
 17.
+Akalin A.
+Scientific data analysis pipelines and reproducibility.
+Towards Data Science.
+2018.
+Available from: https://towardsdatascience.com/scientific-data-analysis-pipelines-and-reproducibility-75ff9df5b4c5
+18.
 Slonim DK.
@@
 Available from: https://www.nature.com/articles/ng1033z
-18.
+19.
 Khatri P, Sirota M, Butte AJ.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
-19.
+20.
 Timmons JA, Szkop KJ, Gallagher IJ.
@@
 Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
-20.
+21.
 Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al.
@@
 Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
-21.
+22.
 Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M.
@@
 Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
-22.
+23.
 Zhao K, Rhee SY.
@@
 Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5
-23.
+24.
 Hennessey J, Ge S.
@@
 Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
-24.
+25.
 Ge SX, Jung D, Yao R.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
-25.
+26.
 Perampalam P, Dick FA.
@@
 Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
-
-101.
+27.
 Huang DW, Sherman BT, Tan Q, Collins JR, Alvord WG, Roayaei J, et al.
@@
 Available from: http://dx.doi.org/10.1186/gb-2007-8-9-r183
-102.
+28.
+Liu H, Wu N, Zhang Z, Zhong X, Zhang H, Guo H, Nie Y, Liu Y.
+Long Non-coding RNA LINC00941 as a Potential Biomarker Promotes the Proliferation and Metastasis of Gastric Cancer.
+Front Genet.
+2019 Jan 22;10:5.
+doi: 10.3389/fgene.2019.00005.
+29.
+Zhang S, Cao R, Li Q, Yao M, Chen Y, Zhou H.
+Comprehensive analysis of lncRNA-associated competing endogenous RNA network in tongue squamous cell carcinoma.
+PeerJ.
+2019 Feb 6;7:e6397.
+doi: 10.7717/peerj.6397.
+30.
+Zhang L, Luo M, Yang H, Zhu S, Cheng X, Qing C.
+Next-generation sequencing-based genomic profiling analysis reveals novel mutations for clinical diagnosis in Chinese primary epithelial ovarian cancer patients.
+J Ovarian Res.
+2019 Feb 20;12(1):19.
+doi: 10.1186/s13048-019-0494-4.
+31.
+Wang K, Li L, Fu L, Yuan Y, Dai H, Zhu T, Zhou Y, Yuan F.
+Integrated Bioinformatics Analysis the Function of RNA Binding Proteins (RBPs) and Their Prognostic Value in Breast Cancer.
+Front Pharmacol.
+2019 Mar 1;10:140.
+doi: 10.3389/fphar.2019.00140.
+32.
+Balusamy SR, Ramani S, Natarajan S, Kim YJ, Perumalsamy H.
+Integrated transcriptome and in vitro analysis revealed anti-proliferative effect of citral in human stomach cancer through apoptosis.
+Sci Rep.
+2019 Mar 19;9(1):4883.
+doi: 10.1038/s41598-019-41406-8.
+33.
+Ahn KS, Kang KJ, Kim YH, Kim TS, Song BI, Kim HW, O'Brien D, Roberts LR, Lee JW, Won KS.
+Genetic features associated with 18F-FDG uptake in intrahepatic cholangiocarcinoma.
+Ann Surg Treat Res.
+2019 Apr;96(4):153-161.
+doi: 10.4174/astr.2019.96.4.153.
+34.
+Chai YJ, Chae H, Kim K, Lee H, Choi S, Lee KE, Kim SW.
+Comparative Gene Expression Profiles in Parathyroid Adenoma and Normal Parathyroid Tissue.
+J Clin Med.
+2019 Mar 2;8(3):297.
+doi: 10.3390/jcm8030297.
+35.
+Guo C, Xu LF, Li HM, Wang W, Guo JH, Jia MQ, Jia R, Jia J.
+Transcriptomic study of the mechanism of anoikis resistance in head and neck squamous carcinoma.
+PeerJ.
+2019 May 23;7:e6978.
+doi: 10.7717/peerj.6978.
+36.
+Alebrahim D, Nayak M, Ward A, Ursomanno P, Shams R, Corsica A, Sleiman R, Fils KH, Silvestro M, Boytard L, Hadi T, Gelb B, Ramkhelawon B.
+Mapping Semaphorins and Netrins in the Pathogenesis of Human Thoracic Aortic Aneurysms.
+Int J Mol Sci.
+2019 Apr 28;20(9):2100.
+doi: 10.3390/ijms20092100.
+37.
+Hong H, Liu T, Wu H, Zhang J, Shi X, Le X, Chen AM, Mo H, Huang Q, Zhou H, Rao X.
+MicroRNA-550a is associated with muscle system conferring poorer survival for esophageal cancer.
+Biosci Rep.
+2019 May 31;39(5):BSR20181173.
+doi: 10.1042/BSR20181173.
+38.
+Han MZ, Wang S, Zhao WB, Ni SL, Yang N, Kong Y, Huang B, Chen AJ, Li XG, Wang J, Wang DH.
+Immune checkpoint molecule herpes virus entry mediator is overexpressed and associated with poor prognosis in human glioblastoma.
+EBioMedicine.
+2019 May;43:159-170.
+doi: 10.1016/j.ebiom.2019.04.002.
+39.
+McSweeney KM, Bozza WP, Alterovitz WL, Zhang B.
+Transcriptomic profiling reveals p53 as a key regulator of doxorubicin-induced cardiotoxicity.
+Cell Death Discov.
+2019 Jun 12;5:102.
+doi: 10.1038/s41420-019-0182-6.
+40.
+Wang C, Li D, Zhang L, Jiang S, Liang J, Narita Y, Hou I, Zhong Q, Zheng Z, Xiao H, Gewurz BE, Teng M, Zhao B.
+RNA Sequencing Analyses of Gene Expression during Epstein-Barr Virus Infection of Primary B Lymphocytes.
+J Virol.
+2019 Jun 14;93(13):e00226-19.
+doi: 10.1128/JVI.00226-19.
+41.
+Pan Q, Wang L, Zhang H, Liang C, Li B.
+Identification of a 5-Gene Signature Predicting Progression and Prognosis of Clear Cell Renal Cell Carcinoma.
+Med Sci Monit.
+2019 Jun 13;25:4401-4413.
+doi: 10.12659/MSM.917399.
+42.  Wang Y, Zhao W, Liu X, Guan G, Zhuang M. ARL3 is downregulated and acts as a prognostic biomarker in glioma. J Transl Med. 2019 Jun 24;17(1):210. doi: 10.1186/s12967-019-1914-3.
+43.
+Weigt SS, Wang X, Palchevskiy V, Li X, Patel N, Ross DJ, Reynolds J, Shah PD, Danziger-Isakov LA, Sweet SC, Singer LG, Budev M, Palmer S, Belperio JA.
+Usefulness of gene expression profiling of bronchoalveolar lavage cells in acute lung allograft rejection.
+J Heart Lung Transplant.
+2019 Aug;38(8):845-855.
+doi: 10.1016/j.healun.2019.05.001.
+44.
+Najafi H, Naseri M, Zahiri J, Totonchi M, Sadeghizadeh M.
+Identification of the Molecular Events Involved in the Development of Prefrontal Cortex Through the Analysis of RNA-Seq Data From BrainSpan.
+ASN Neuro.
+2019 Jan-Dec;11:1759091419854627.
+doi: 10.1177/1759091419854627.
+45.
+Feltrin AS, Tahira AC, Simões SN, Brentani H, Martins DC Jr.
+Assessment of complementarity of WGCNA and NERI results for identification of modules associated to schizophrenia spectrum disorders.
+PLoS One.
+2019 Jan 15;14(1):e0210431.
+doi: 10.1371/journal.pone.0210431.
+46.
+Wang SS, Chen G, Li SH, Pang JS, Cai KT, Yan HB, Huang ZG, He RQ.
+Identification and validation of an individualized autophagy-clinical prognostic index in bladder cancer patients.
+Onco Targets Ther.
+2019 May 14;12:3695-3712.
+doi: 10.2147/OTT.S197676.
+47.
+Sun R, Meng X, Wang W, Liu B, Lv X, Yuan J, Zeng L, Chen Y, Yuan B, Yang S.
+Five genes may predict metastasis in non-small cell lung cancer using bioinformatics analysis.
+Oncol Lett.
+2019 Aug;18(2):1723-1732.
+doi: 10.3892/ol.2019.10498.
+48.
 Zeeberg BR, Riss J, Kane DW, Bussey KJ, Uchio E, Linehan WM, et al.
@@
 Available from: http://dx.doi.org/10.1186/1471-2105-5-80
-
-201.
+49.
 Ziemann M, Kaspi A, El-Osta A.
@@
 Available from: http://dx.doi.org/10.1093/gigascience/giz022
-202.
+50.
 Love MI, Huber W, Anders S.
@@
 Available from: http://dx.doi.org/10.1186/s13059-014-0550-8
-203.
+51.
 Yu G, Wang LG, Han Y, He QY.
@@
 Available from: http://dx.doi.org/10.1089/omi.2011.0118
-204.
+52.
 Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A.
@@
 Available from: http://biorxiv.org/content/early/2021/02/01/060012.abstract
-205.
+53.
 Kaspi A, Ziemann M.
@@
 Available from: http://dx.doi.org/10.1186/s12864-020-06856-9
-206.
+54.
 Gillespie M, Jassal B, Stephan R, Milacic M, Rothfels K, Senff-Ribeiro A, et al.
@@
 Available from: http://dx.doi.org/10.1093/nar/gkab1028
-207.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
-208.
+55.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
+56.
 Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
@@
 Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
-
-## Supplementary information
-
-Table XX: The software bundles in the example image.
-R package dependancies not shown.
-
-| Software | Version | Purpose |
-| --- | --- | --- |
-| Ubuntu | 22.04 | Operating System |
-| R | 4.2.2 | Statistical computing language |
-| nano | 6.2 | text editing |
-| git | 2.34.1 | source control |
-| R/kableExtra | 1.3.4 | render formatted tables |
-| R/vioplot | 0.4.0 | violin charts |
-| R/gplots | 3.1.3 | heatmaps |
-| R/eulerr | 7.0.0 | Venn diagrams |
-| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
-| BioC/DESeq2 | 1.38.3 | differential expression |
-| BioC/fgsea | 1.24.0 | functional class scoring |
-| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
-| Bioc/mitch | 1.10.0 | multi-contrast enrichment |
+57.  Ioannidis JPA, Allison DB, Ball CA, Coulibaly I, Cui X, Culhane AC, et al. Repeatability of published microarray gene expression analyses. Nat Genet. 2009;41(2):149–55.
+58.
+Prinz F, Schlange T, Asadullah K.
+Believe it or not: how much can we rely on published data on potential drug targets?
+Nat Rev Drug Discov [Internet].
+2011 [cited 2023 Apr 27];10(9):712.
+Available from: https://www.nature.com/articles/nrd3439-c1
+59.
+Begley CG, Ellis LM.
+Drug development: Raise standards for preclinical cancer research: Drug development.
+Nature [Internet].
+2012 [cited 2023 Apr 27];483(7391):531–3.
+Available from: https://www.nature.com/articles/483531a
+60.
+Kuehn H, Liberzon A, Reich M, Mesirov JP.
+Using GenePattern for gene expression analysis.
+Curr Protoc Bioinformatics [Internet].
+2008;Chapter 7(1):7.12.1-7.12.39.
+Available from: http://dx.doi.org/10.1002/0471250953.bi0712s22
+61.
+Goecks J, Nekrutenko A, Taylor J, Galaxy Team.
+Galaxy: a comprehensive approach for supporting accessible, reproducible, and transparent computational research in the life sciences.
+Genome Biol [Internet].
+2010;11(8):R86.
+Available from: http://dx.doi.org/10.1186/gb-2010-11-8-r86
+62.
+Teichman G, Cohen D, Ganon O, Dunsky N, Shani S, Gingold H, et al.
+RNAlysis: analyze your RNA sequencing data without writing a single line of code.
+BMC Biol [Internet].
+2023;21(1):74.
+Available from: http://dx.doi.org/10.1186/s12915-023-01574-6
+63.  Powell D. drpowell/degust 4.1.1. Zenodo; 2019.
+64.  Garijo D, Kinnings S, Xie L, Xie L, Zhang Y, Bourne PE, et al. Quantifying reproducibility in computational biology: The case of the tuberculosis drugome. PLoS One. 2013;8(11):e80278.
+65.  Blohowiak BB, Cohoon J, de-Wit L, Eich E, Farach FJ, Hasselman F, et al. Badges to Acknowledge Open Practices. OSF; 2023. Available from: osf.io/tvyxz

### 2023-04-27 manuscript.Rmd
referencing

@@
 ---
-title: "A recipe for extremely reproducible enrichment analysis"
+title: "A recipe for extremely reproducible enrichment analysis (and why we need it)"
 author: "Mark Ziemann and Anusuiya Bora"
@@
     fig_height: 5
+bibliography: references.bib
+csl: plos-computational-biology.csl
 ---
@@
 
-Irreproducible research is a significant problem that wastes resources undermines the public
+Irreproducible research is a significant problem that wastes resources and undermines the public
 perception of science.
-Previous work has highlighted that published enrichment analyses suffers from statistical and reporting
-flaws.
+Previous work has highlighted that published enrichment analyses frequently suffers from statistical
+and reporting flaws.
 We sought to determine whether this translates into irreproducibility by examining whether the
@@
 The findings indicate that current enrichment analysis practices are inadequate.
-To address this, we provide a set of boilerplate scripts and a step-by-step protocol to adapt them to
-a new project's needs.
-This recipe is built according to a framework we call "five pillars of computational reproducibility"
-, which when implemented together facilitate extreme reproducibility.
+To address this, we provide a set of template scripts and a step-by-step protocol to adapt them to
+a new project's enrichment analysis needs.
+This recipe is built according to a framework we call "five pillars", which when implemented together
+facilitate extreme reproducibility.
 
@@
 Reproducibility is essential to the scientific enterprise by confirming the validity of new discoveries
-[1](National Academies of Sciences 2019, p1).
+[@Committee_on_Reproducibility_and_Replicability_in_Science2019-vb p1].
 According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was a "significant
-crisis" of reproducibility [2](Baker 2016).
+crisis" of reproducibility @Baker2016-hm.
 While bioinformatics and other branches of computational research are theoretically amenable to
 complete reproducibility, in reality this is rarely achieved due to lack of detail provided in methods
-sections, lack of shared data and lack of code [3](Peng 2011).
+sections, lack of shared data and lack of code @Peng2011-ia.
 Indeed even if data and code are shared, reproducibility still cannot be guaranteed due to the
-differences in computing environments over time, and the availability of dependancies [4](Perkel 2020).
+differences in computing environments over time, and the availability of dependancies @Perkel2020-io.
 
@@
 "containerisation" to control the computing environment including the operating system and dependancies
-along with the research code to facilitate a high degree of reproducibility [4-16](Perkel 2020;
-Stein 2010; Dudley & Butte 2010; Sandve et al, 2013; Piccolo & Frampton 2016; Grüning et al, 2014;
-Lewis et al, 2016; da Veiga Leprevost et al, 2017; Kulkarni et al, 2018; Pasquier et al, 2018;
-Hung et al, 2019; Brito et al, 2020; Piccolo et al, 2021).
+along with the research code to facilitate a high degree of reproducibility [@Perkel2020-io;
+@Stein2010-of; @Dudley2010-ua; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
+@Da_Veiga_Leprevost2017-dw; @Pasquier2018-rt; @Kulkarni2018-km; @Hung2019-et; @Brito2020-ld;
+@Piccolo2021-ud]
 Containers also allow for an easy installation process, as all working dependencies are shipped together
@@
 and without any manual steps.
-Taken together, computational reproducibility is a spectrum (**Figure 1**) [17](Akalin 2018), and rests upon five key
+Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five key
 pillars mentioned above (**Figure 2**).
@@
 Creator: Altuna Akalin.
-Republished with permission [17](Akalin 2018).](img/repro.webp "The computational research reproducibility spectrum.")
+Republished with permission @Akalin2018-eg.](img/repro.webp "The computational research reproducibility spectrum.")
 
@@
 such as the detection of pathway activation in development or disease
-[18,19](Slonim 2002; Khatri et al, 2012).
+[@Slonim2002-sf; @Khatri2012-do].
 While enrichment analysis is undoubtedly a key method for the interpretation of large
@@
 (ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
-[20-23](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).
+[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
 Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
@@
 point-and-click tools appear to be far more popular than computer script based methods for this type
-of work [22](Wijesooriya et al, 2022).
+of work @Wijesooriya2022-li.
 Although web-based tools are convenient, they are a potential reproducibility concern.
@@
 In some cases the entire web resource can become unavailable, and hence irreproducible due to
-a phenomenon known as link decay [24](Hennessey & Ge 2013).
+a phenomenon known as link decay @Hennessey2013-tj.
 
 A small number of enrichment analysis applications have been developed to have beneficial properties
-of being "containerised" and with archived historical versions [25,26](Ge et al, 2020;
-Perampalam & Dick 2020).
+of being "containerised" and with archived historical versions [@Ge2020-qe; @Perampalam2020-qm].
 This helps reproducibility greatly, but is not a complete solution as it lacks information about how
@@
 From a list of 186 PubMed central articles using enrichment analysis that we examined previously
-[22](Wijesooriya et al, 2022), we selected 20 articles that fulfilled the following criteria; (1) the
+@Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
 organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
-Visualization, and Integrated Discovery)[27](Huang et al, 2007); (3) the omics type was gene
+Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
 expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
@@
 of enrichment analysis in our previous study, which could be reproduced without the need to
-reanalyse the raw dataset from scratch. These articles are listed in the bibliography [28-47]().
+reanalyse the raw dataset from scratch. These articles are listed in the bibliography
+[@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
+@Guo2019-fx; @Alebrahim2019-zc; @Hong2019-ym; @Han2019-ix; @McSweeney2019-ij; @Wang2019-yw; @Pan2019-iv;
+@Wang2019-jy; @Weigt2019-mk; @Najafi2019-ce; @Feltrin2019-to; @Wang2019-xx; @Sun2019-xr].
 
 The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
-special attention to preventing gene name errors from occurring [48](Zeeberg et al, 2004).
+special attention to preventing gene name errors from occurring @Zeeberg2004-na.
 The reproducibility analysis was performed using the respective methodologies stated in the methods
@@
 It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
-analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` [49](Ziemann et al, 2019),
-differential expression with `DESeq2` [50](Love & Huber, 2014),
+analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` @Ziemann2019-zo,
+differential expression with `DESeq2` @Love2014-hd,
 enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
-[51-53](Yu et al, 2012; Korotkevich et al, 2016; Kaspi & Ziemann 2020).
+[@Yu2012-ts; @Korotkevich2016-ls; @Kaspi2020-dq].
 It also contains instructions to clone a github repository, which contains the scripts that run the
 analytical workflow.
-Lastly, it copies a gene set library from Reactome [54](Gillespie et al, 2022) with a clear version
+Lastly, it copies a gene set library from Reactome @Gillespie2022-gz with a clear version
 history to guarantee provenance of the annotation data.
@@
 which is becoming ubiquitous in bioinformatics and data science more generally
-[55](Grolemund & Wickham 2017, p423).
+@Grolemund2017-zl p423.
 The ability to provide extended descriptions is helpful for providing contextualising information about
@@
 For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
-[23](Zhao & Rhee 2023).
+@Zhao2023-ok.
 
@@
 it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
-in triplicate [56](Lund et al, 2014).
+in triplicate @Lund2014-et.
 We use this analysis to better understand the effect of DNA methylatransferase inhibitors on
@@
 The observed lack of reproducibility might be expected given that substantial methodological flaws
-were present in 85% of enrichment analyses [22](Wijesooriya et al, 2022).
+were present in 85% of enrichment analyses @Wijesooriya2022-li.
 A non-exhaustive literature search identified a small number of case studies, but a dearth of systematic
 examinations of bioinformatics research reproducibility.
-Ionnidis et al (2009)[57] examined a set of 10 gene expression studies with available data published in
-2005-2006, and found only two could be reproduced to a satisfactory level.
+Ionnidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
+published in 2005-2006, and found only two could be reproduced to a satisfactory level.
 
@@
 In the field of drug development, according to a 2011 study by Bayer, only 25% of projects could be
-replicated to a high degree [58](Prinz et al, 2011).
+replicated to a high degree @Prinz2011-ht.
 When the US Biotech firm Amgen set out to reproduce 53 landmark cancer studies, they were able to
-reproduce only 6 (11%) [59](Begley & Ellis 2012).
+reproduce only 6 (11%) @Begley2012-di.
 
@@
 The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
-workflows that enable reproducibility [60-62](Kuehn et al; 2008; Goecks et al, 2010; Teichman et al, 2023).
+workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
 The Degust differential expression suite is a graphical interface to popular
 expression tools like edgeR and limma, and allows users to download the script to facilitate future
-non-interactive reproducibility [63](Powell 2019).
+non-interactive reproducibility @Powell2019-lv.
 Such functionality could be written into existing enrichment analysis tools like DAVID, and together with
@@
 which could tarnish the reputation of an author or their institution.
-Solutions to this could include mandates by funding agencies or institutions [64](Garijo et al, 2013),
-and/or badging of articles that comply with best practices [65](Blohowiak et al, 2023).
+Solutions to this could include mandates by funding agencies or institutions @Garijo2013-zb,
+and/or badging of articles that comply with best practices @Kidwell2016-pm.
+
+## Acknowledgements
 
@@
 ## Bibliography
-
-1.
-National Academies of Sciences, Engineering, Medicine, Policy, Global Affairs, Committee on Science, Engineering, Medicine, Public Policy, et al.
-Reproducibility and Replicability in Science.
-Washington, D.C., DC: National Academies Press; 2019.
-2.  Baker M. 1,500 scientists lift the lid on reproducibility. Nature [Internet]. 2016 [cited 2023 Mar 20];533(7604):452–4. Available from: https://www.nature.com/articles/533452a
-3.  Peng RD. Reproducible research in computational science. Science [Internet]. 2011;334(6060):1226–7. Available from: http://dx.doi.org/10.1126/science.1213847
-4.
-Perkel JM.
-Challenge to scientists: does your ten-year-old code still run?
-Nature [Internet].
-2020 [cited 2023 Mar 20];584(7822):656–8.
-Available from: https://www.nature.com/articles/d41586-020-02462-7
-5.  Stein LD. The case for cloud computing in genome informatics. Genome Biol [Internet]. 2010;11(5):207. Available from: http://dx.doi.org/10.1186/gb-2010-11-5-207
-6.  Dudley JT, Butte AJ. In silico research in the era of cloud computing. Nat Biotechnol [Internet]. 2010;28(11):1181–5. Available from: http://dx.doi.org/10.1038/nbt1110-1181
-7.
-Sandve GK, Nekrutenko A, Taylor J, Hovig E.
-Ten simple rules for reproducible computational research.
-PLoS Comput Biol [Internet].
-2013;9(10):e1003285.
-Available from: http://dx.doi.org/10.1371/journal.pcbi.1003285
-8.  Piccolo SR, Frampton MB. Tools and techniques for computational reproducibility. Gigascience [Internet]. 2016;5(1). Available from: http://dx.doi.org/10.1186/s13742-016-0135-4
-9.
-Grüning B, Chilton J, Köster J, Dale R, Soranzo N, van den Beek M, et al.
-Practical computational reproducibility in the life sciences.
-Cell Syst [Internet].
-2018;6(6):631–5.
-Available from: http://dx.doi.org/10.1016/j.cels.2018.03.014
-10.
-Lewis J, Breeze CE, Charlesworth J, Maclaren OJ, Cooper J.
-Where next for the reproducibility agenda in computational biology?
-BMC Syst Biol [Internet].
-2016;10(1).
-Available from: http://dx.doi.org/10.1186/s12918-016-0288-x
-11.
-da Veiga Leprevost F, Grüning BA, Alves Aflitos S, Röst HL, Uszkoreit J, Barsnes H, et al.
-BioContainers: an open-source and community-driven framework for software standardization.
-Bioinformatics [Internet].
-2017;33(16):2580–2.
-Available from: http://dx.doi.org/10.1093/bioinformatics/btx192
-12.
-Pasquier T, Lau MK, Han X, Fong E, Lerner BS, Boose ER, et al.
-Sharing and Preserving Computational Analyses for Posterity with encapsulator.
-Comput Sci Eng [Internet].
-2018;20(4):111–24.
-Available from: https://ieeexplore.ieee.org/abstract/document/8409369
-13.
-Kulkarni N, Alessandrì L, Panero R, Arigoni M, Olivero M, Ferrero G, et al.
-Reproducible bioinformatics project: a community for reproducible bioinformatics analysis pipelines.
-BMC Bioinformatics [Internet].
-2018;19(Suppl 10):349.
-Available from: http://dx.doi.org/10.1186/s12859-018-2296-x
-14.
-Hung LH, Hu J, Meiss T, Ingersoll A, Lloyd W, Kristiyanto D, et al.
-Building containerized workflows using the BioDepot-workflow-builder.
-Cell Syst [Internet].
-2019;9(5):508-514.e3.
-Available from: http://dx.doi.org/10.1016/j.cels.2019.08.007
-15.
-Brito JJ, Li J, Moore JH, Greene CS, Nogoy NA, Garmire LX, et al.
-Recommendations to enhance rigor and reproducibility in biomedical research.
-Gigascience [Internet].
-2020 [cited 2023 Mar 9];9(6).
-Available from: https://pubmed.ncbi.nlm.nih.gov/32479592/
-16.
-Piccolo SR, Ence ZE, Anderson EC, Chang JT, Bild AH.
-Simplifying the development of portable, scalable, and reproducible workflows.
-Elife [Internet].
-2021;10.
-Available from: http://dx.doi.org/10.7554/eLife.71069
-17.
-Akalin A.
-Scientific data analysis pipelines and reproducibility.
-Towards Data Science.
-2018.
-Available from: https://towardsdatascience.com/scientific-data-analysis-pipelines-and-reproducibility-75ff9df5b4c5
-18.
-Slonim DK.
-From patterns to pathways: gene expression data analysis comes of age.
-Nat Genet [Internet].
-2002 [cited 2023 Mar 3];32 Suppl(S4):502–8.
-Available from: https://www.nature.com/articles/ng1033z
-19.
-Khatri P, Sirota M, Butte AJ.
-Ten years of pathway analysis: current approaches and outstanding challenges.
-PLoS Comput Biol [Internet].
-2012 [cited 2023 Mar 3];8(2):e1002375.
-Available from: https://pubmed.ncbi.nlm.nih.gov/22383865/
-20.
-Timmons JA, Szkop KJ, Gallagher IJ.
-Multiple sources of bias confound functional enrichment analysis of global -omics data.
-Genome Biol [Internet].
-2015;16(1):186.
-Available from: http://dx.doi.org/10.1186/s13059-015-0761-7
-21.
-Reimand J, Isserlin R, Voisin V, Kucera M, Tannus-Lopes C, Rostamianfar A, et al.
-Pathway enrichment analysis and visualization of omics data using g:Profiler, GSEA, Cytoscape and EnrichmentMap.
-Nat Protoc [Internet].
-2019;14(2):482–517.
-Available from: http://dx.doi.org/10.1038/s41596-018-0103-9
-22.
-Wijesooriya K, Jadaan SA, Perera KL, Kaur T, Ziemann M.
-Urgent need for consistent standards in functional enrichment analysis.
-PLoS Comput Biol [Internet].
-2022;18(3):e1009935.
-Available from: http://dx.doi.org/10.1371/journal.pcbi.1009935
-23.
-Zhao K, Rhee SY.
-Interpreting omics data with pathway enrichment analysis.
-Trends Genet [Internet].
-2023 [cited 2023 Mar 1];0(0).
-Available from: https://www.cell.com/trends/genetics/fulltext/S0168-9525(23)00018-5
-24.
-Hennessey J, Ge S.
-A cross disciplinary study of link decay and the effectiveness of mitigation techniques.
-BMC Bioinformatics [Internet].
-2013;14 Suppl 14(S14):S5.
-Available from: http://dx.doi.org/10.1186/1471-2105-14-S14-S5
-25.
-Ge SX, Jung D, Yao R.
-ShinyGO: a graphical gene-set enrichment tool for animals and plants.
-Bioinformatics [Internet].
-2020 [cited 2023 Mar 9];36(8):2628–9.
-Available from: https://pubmed.ncbi.nlm.nih.gov/31882993/
-26.
-Perampalam P, Dick FA.
-BEAVR: a browser-based tool for the exploration and visualization of RNA-seq data.
-BMC Bioinformatics [Internet].
-2020 [cited 2023 Mar 9];21(1):221.
-Available from: https://pubmed.ncbi.nlm.nih.gov/32471392/
-27.
-Huang DW, Sherman BT, Tan Q, Collins JR, Alvord WG, Roayaei J, et al.
-The DAVID Gene Functional Classification Tool: a novel biological module-centric algorithm to functionally analyze large gene lists.
-Genome Biol [Internet].
-2007;8(9):R183.
-Available from: http://dx.doi.org/10.1186/gb-2007-8-9-r183
-28.
-Liu H, Wu N, Zhang Z, Zhong X, Zhang H, Guo H, Nie Y, Liu Y.
-Long Non-coding RNA LINC00941 as a Potential Biomarker Promotes the Proliferation and Metastasis of Gastric Cancer.
-Front Genet.
-2019 Jan 22;10:5.
-doi: 10.3389/fgene.2019.00005.
-29.
-Zhang S, Cao R, Li Q, Yao M, Chen Y, Zhou H.
-Comprehensive analysis of lncRNA-associated competing endogenous RNA network in tongue squamous cell carcinoma.
-PeerJ.
-2019 Feb 6;7:e6397.
-doi: 10.7717/peerj.6397.
-30.
-Zhang L, Luo M, Yang H, Zhu S, Cheng X, Qing C.
-Next-generation sequencing-based genomic profiling analysis reveals novel mutations for clinical diagnosis in Chinese primary epithelial ovarian cancer patients.
-J Ovarian Res.
-2019 Feb 20;12(1):19.
-doi: 10.1186/s13048-019-0494-4.
-31.
-Wang K, Li L, Fu L, Yuan Y, Dai H, Zhu T, Zhou Y, Yuan F.
-Integrated Bioinformatics Analysis the Function of RNA Binding Proteins (RBPs) and Their Prognostic Value in Breast Cancer.
-Front Pharmacol.
-2019 Mar 1;10:140.
-doi: 10.3389/fphar.2019.00140.
-32.
-Balusamy SR, Ramani S, Natarajan S, Kim YJ, Perumalsamy H.
-Integrated transcriptome and in vitro analysis revealed anti-proliferative effect of citral in human stomach cancer through apoptosis.
-Sci Rep.
-2019 Mar 19;9(1):4883.
-doi: 10.1038/s41598-019-41406-8.
-33.
-Ahn KS, Kang KJ, Kim YH, Kim TS, Song BI, Kim HW, O'Brien D, Roberts LR, Lee JW, Won KS.
-Genetic features associated with 18F-FDG uptake in intrahepatic cholangiocarcinoma.
-Ann Surg Treat Res.
-2019 Apr;96(4):153-161.
-doi: 10.4174/astr.2019.96.4.153.
-34.
-Chai YJ, Chae H, Kim K, Lee H, Choi S, Lee KE, Kim SW.
-Comparative Gene Expression Profiles in Parathyroid Adenoma and Normal Parathyroid Tissue.
-J Clin Med.
-2019 Mar 2;8(3):297.
-doi: 10.3390/jcm8030297.
-35.
-Guo C, Xu LF, Li HM, Wang W, Guo JH, Jia MQ, Jia R, Jia J.
-Transcriptomic study of the mechanism of anoikis resistance in head and neck squamous carcinoma.
-PeerJ.
-2019 May 23;7:e6978.
-doi: 10.7717/peerj.6978.
-36.
-Alebrahim D, Nayak M, Ward A, Ursomanno P, Shams R, Corsica A, Sleiman R, Fils KH, Silvestro M, Boytard L, Hadi T, Gelb B, Ramkhelawon B.
-Mapping Semaphorins and Netrins in the Pathogenesis of Human Thoracic Aortic Aneurysms.
-Int J Mol Sci.
-2019 Apr 28;20(9):2100.
-doi: 10.3390/ijms20092100.
-37.
-Hong H, Liu T, Wu H, Zhang J, Shi X, Le X, Chen AM, Mo H, Huang Q, Zhou H, Rao X.
-MicroRNA-550a is associated with muscle system conferring poorer survival for esophageal cancer.
-Biosci Rep.
-2019 May 31;39(5):BSR20181173.
-doi: 10.1042/BSR20181173.
-38.
-Han MZ, Wang S, Zhao WB, Ni SL, Yang N, Kong Y, Huang B, Chen AJ, Li XG, Wang J, Wang DH.
-Immune checkpoint molecule herpes virus entry mediator is overexpressed and associated with poor prognosis in human glioblastoma.
-EBioMedicine.
-2019 May;43:159-170.
-doi: 10.1016/j.ebiom.2019.04.002.
-39.
-McSweeney KM, Bozza WP, Alterovitz WL, Zhang B.
-Transcriptomic profiling reveals p53 as a key regulator of doxorubicin-induced cardiotoxicity.
-Cell Death Discov.
-2019 Jun 12;5:102.
-doi: 10.1038/s41420-019-0182-6.
-40.
-Wang C, Li D, Zhang L, Jiang S, Liang J, Narita Y, Hou I, Zhong Q, Zheng Z, Xiao H, Gewurz BE, Teng M, Zhao B.
-RNA Sequencing Analyses of Gene Expression during Epstein-Barr Virus Infection of Primary B Lymphocytes.
-J Virol.
-2019 Jun 14;93(13):e00226-19.
-doi: 10.1128/JVI.00226-19.
-41.
-Pan Q, Wang L, Zhang H, Liang C, Li B.
-Identification of a 5-Gene Signature Predicting Progression and Prognosis of Clear Cell Renal Cell Carcinoma.
-Med Sci Monit.
-2019 Jun 13;25:4401-4413.
-doi: 10.12659/MSM.917399.
-42.  Wang Y, Zhao W, Liu X, Guan G, Zhuang M. ARL3 is downregulated and acts as a prognostic biomarker in glioma. J Transl Med. 2019 Jun 24;17(1):210. doi: 10.1186/s12967-019-1914-3.
-43.
-Weigt SS, Wang X, Palchevskiy V, Li X, Patel N, Ross DJ, Reynolds J, Shah PD, Danziger-Isakov LA, Sweet SC, Singer LG, Budev M, Palmer S, Belperio JA.
-Usefulness of gene expression profiling of bronchoalveolar lavage cells in acute lung allograft rejection.
-J Heart Lung Transplant.
-2019 Aug;38(8):845-855.
-doi: 10.1016/j.healun.2019.05.001.
-44.
-Najafi H, Naseri M, Zahiri J, Totonchi M, Sadeghizadeh M.
-Identification of the Molecular Events Involved in the Development of Prefrontal Cortex Through the Analysis of RNA-Seq Data From BrainSpan.
-ASN Neuro.
-2019 Jan-Dec;11:1759091419854627.
-doi: 10.1177/1759091419854627.
-45.
-Feltrin AS, Tahira AC, Simões SN, Brentani H, Martins DC Jr.
-Assessment of complementarity of WGCNA and NERI results for identification of modules associated to schizophrenia spectrum disorders.
-PLoS One.
-2019 Jan 15;14(1):e0210431.
-doi: 10.1371/journal.pone.0210431.
-46.
-Wang SS, Chen G, Li SH, Pang JS, Cai KT, Yan HB, Huang ZG, He RQ.
-Identification and validation of an individualized autophagy-clinical prognostic index in bladder cancer patients.
-Onco Targets Ther.
-2019 May 14;12:3695-3712.
-doi: 10.2147/OTT.S197676.
-47.
-Sun R, Meng X, Wang W, Liu B, Lv X, Yuan J, Zeng L, Chen Y, Yuan B, Yang S.
-Five genes may predict metastasis in non-small cell lung cancer using bioinformatics analysis.
-Oncol Lett.
-2019 Aug;18(2):1723-1732.
-doi: 10.3892/ol.2019.10498.
-48.
-Zeeberg BR, Riss J, Kane DW, Bussey KJ, Uchio E, Linehan WM, et al.
-Mistaken identifiers: gene name errors can be introduced inadvertently when using Excel in bioinformatics.
-BMC Bioinformatics [Internet].
-2004;5:80.
-Available from: http://dx.doi.org/10.1186/1471-2105-5-80
-49.
-Ziemann M, Kaspi A, El-Osta A.
-Digital expression explorer 2: a repository of uniformly processed RNA sequencing data.
-Gigascience [Internet].
-2019;8(4).
-Available from: http://dx.doi.org/10.1093/gigascience/giz022
-50.
-Love MI, Huber W, Anders S.
-Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2.
-Genome Biol [Internet].
-2014;15(12):550.
-Available from: http://dx.doi.org/10.1186/s13059-014-0550-8
-51.
-Yu G, Wang LG, Han Y, He QY.
-clusterProfiler: an R package for comparing biological themes among gene clusters.
-OMICS [Internet].
-2012;16(5):284–7.
-Available from: http://dx.doi.org/10.1089/omi.2011.0118
-52.
-Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A.
-Fast gene set enrichment analysis [Internet].
-bioRxiv.
-2016.
-p.
-060012.
-Available from: http://biorxiv.org/content/early/2021/02/01/060012.abstract
-53.
-Kaspi A, Ziemann M.
-Mitch: Multi-contrast pathway enrichment for multi-omics and single-cell profiling data.
-BMC Genomics [Internet].
-2020;21(1):447.
-Available from: http://dx.doi.org/10.1186/s12864-020-06856-9
-54.
-Gillespie M, Jassal B, Stephan R, Milacic M, Rothfels K, Senff-Ribeiro A, et al.
-The reactome pathway knowledgebase 2022.
-Nucleic Acids Res [Internet].
-2022;50(D1):D687–92.
-Available from: http://dx.doi.org/10.1093/nar/gkab1028
-55.  Grolemund G, Wickham H. R for Data Science. Sebastopol, CA: O’Reilly Media; 2017.
-56.
-Lund K, Cole JJ, VanderKraats ND, McBryan T, Pchelintsev NA, Clark W, et al.
-DNMT inhibitors reverse a specific signature of aberrant promoter DNA methylation and associated gene silencing in AML.
-Genome Biol [Internet].
-2014;15(8):406.
-Available from: http://dx.doi.org/10.1186/s13059-014-0406-2
-57.  Ioannidis JPA, Allison DB, Ball CA, Coulibaly I, Cui X, Culhane AC, et al. Repeatability of published microarray gene expression analyses. Nat Genet. 2009;41(2):149–55.
-58.
-Prinz F, Schlange T, Asadullah K.
-Believe it or not: how much can we rely on published data on potential drug targets?
-Nat Rev Drug Discov [Internet].
-2011 [cited 2023 Apr 27];10(9):712.
-Available from: https://www.nature.com/articles/nrd3439-c1
-59.
-Begley CG, Ellis LM.
-Drug development: Raise standards for preclinical cancer research: Drug development.
-Nature [Internet].
-2012 [cited 2023 Apr 27];483(7391):531–3.
-Available from: https://www.nature.com/articles/483531a
-60.
-Kuehn H, Liberzon A, Reich M, Mesirov JP.
-Using GenePattern for gene expression analysis.
-Curr Protoc Bioinformatics [Internet].
-2008;Chapter 7(1):7.12.1-7.12.39.
-Available from: http://dx.doi.org/10.1002/0471250953.bi0712s22
-61.
-Goecks J, Nekrutenko A, Taylor J, Galaxy Team.
-Galaxy: a comprehensive approach for supporting accessible, reproducible, and transparent computational research in the life sciences.
-Genome Biol [Internet].
-2010;11(8):R86.
-Available from: http://dx.doi.org/10.1186/gb-2010-11-8-r86
-62.
-Teichman G, Cohen D, Ganon O, Dunsky N, Shani S, Gingold H, et al.
-RNAlysis: analyze your RNA sequencing data without writing a single line of code.
-BMC Biol [Internet].
-2023;21(1):74.
-Available from: http://dx.doi.org/10.1186/s12915-023-01574-6
-63.  Powell D. drpowell/degust 4.1.1. Zenodo; 2019.
-64.  Garijo D, Kinnings S, Xie L, Xie L, Zhang Y, Bourne PE, et al. Quantifying reproducibility in computational biology: The case of the tuberculosis drugome. PLoS One. 2013;8(11):e80278.
-65.  Blohowiak BB, Cohoon J, de-Wit L, Eich E, Farach FJ, Hasselman F, et al. Badges to Acknowledge Open Practices. OSF; 2023. Available from: osf.io/tvyxz

### 2023-04-27 manuscript.Rmd
minor edits

@@
 reproduce only 6 (11%) @Begley2012-di.
-
 It is clear that as a field, the life sciences need to raise the rate of reproducibility or risk the
@@
 However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
-protocol to guide them on how to customise boilerplate Dockerfile and Rmarkdown scripts to suit their
+protocol to guide them on how to customise template Dockerfile and Rmarkdown scripts to suit their
 project's needs.
-
 Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
@@
 
-Although we hope that extremely reproducible research will become more widespread, we do not
+Although we hope that extremely reproducible enrichment analysis will become more widespread, we do not
 anticipate any dramatic changes in researcher behaviour unless the motivating factors are augmented.
-Reward structures in academia encourage individuals to publish many articles, and reproducibility
-is not a significant factor unless the issues are extreme enough to provoke a retraction by the journal,
-which could tarnish the reputation of an author or their institution.
+Reward structures in academia favour career progression for individuals who publish many articles
+as opposed to individuals who publish fewer but more rigourously conducted articles.
 Solutions to this could include mandates by funding agencies or institutions @Garijo2013-zb,

### 2023-05-02 manuscript.Rmd
Nectar acknowledgement
fix typos and minor edits

@@
 
-Irreproducible research is a significant problem that wastes resources and undermines the public
-perception of science.
-Previous work has highlighted that published enrichment analyses frequently suffers from statistical
+Unreliable and irreproducible research is a significant problem that wastes resources and risks
+undermining the public perception of science.
+Previous work has highlighted that published enrichment analyses frequently suffer from statistical
 and reporting flaws.
@@
 This recipe is built according to a framework we call "five pillars", which when implemented together
-facilitate extreme reproducibility.
+facilitates extreme reproducibility.
 
@@
 sections, lack of shared data and lack of code @Peng2011-ia.
-Indeed even if data and code are shared, reproducibility still cannot be guaranteed due to the
-differences in computing environments over time, and the availability of dependancies @Perkel2020-io.
+Indeed, even if data and code are shared, reproducibility still cannot be guaranteed due to the
+differences in computing environments over time, and the availability of dependencies @Perkel2020-io.
 
 To address this, many articles have been written to endorse the use of "virtual machines" or
-"containerisation" to control the computing environment including the operating system and dependancies
+"containerisation" to control the computing environment including the operating system and dependencies
 along with the research code to facilitate a high degree of reproducibility [@Perkel2020-io;
-@Stein2010-of; @Dudley2010-ua; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
-@Da_Veiga_Leprevost2017-dw; @Pasquier2018-rt; @Kulkarni2018-km; @Hung2019-et; @Brito2020-ld;
-@Piccolo2021-ud]
+@Stein2010-of; @Dudley2010-ua; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg;
+@Lewis2016-qh; @Da_Veiga_Leprevost2017-dw; @Pasquier2018-rt; @Kulkarni2018-km; @Hung2019-et;
+@Brito2020-ld; @Piccolo2021-ud]
 Containers also allow for an easy installation process, as all working dependencies are shipped together
@@
 free text explanations together with the code and outputs such as charts and tables.
-Secondly, code and data can be linked. This means that the code knows the location of the the data,
-avoiding any "file not found" errors.
+Secondly, code and data can be linked. This means that the code knows the location of the data,
+avoiding any "file not found" errors Peng2011-ia.
 Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
 for each figure sequentially, making the entire manuscript reproducible by executing just one command
-and without any manual steps.
+and without any manual steps @Lasser2020-mm.
 Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five key
@@
 While enrichment analysis is undoubtedly a key method for the interpretation of large
-data sets, there are concerns that inadvertant misuse of these methods leads to
+data sets, there are concerns that inadvertent misuse of these methods leads to
 unreliable results.
@@
 We are concerned about the reliability of published enrichment analyses, as web-based and
-point-and-click tools appear to be far more popular than computer script based methods for this type
+point-and-click tools appear to be far more popular than computer script-based methods for this type
 of work @Wijesooriya2022-li.
@@
 typically not made available, hampering later reproducibility.
-In some cases the entire web resource can become unavailable, and hence irreproducible due to
+In some cases, the entire web resource can become unavailable, and hence irreproducible due to
 a phenomenon known as link decay @Hennessey2013-tj.
@@
 of being "containerised" and with archived historical versions [@Ge2020-qe; @Perampalam2020-qm].
-This helps reproducibility greatly, but is not a complete solution as it lacks information about how
+This helps reproducibility greatly but is not a complete solution as it lacks information about how
 the input data was used and a record of any parameter selections.
@@
 Some terms using clusterprofiler in R were mentioned.
-| **2.** GO results from the original manuscript is somewhat similar to reproduced analysis.
+| **2.** GO results from the original manuscript are somewhat similar to reproduced analysis.
 The FDR values however are inconsistent.
@@
 N.A.
-| **1.** GO analysis from upregulated and downregulated proteins are less similar to the results given in the paper.
+| **1.** GO analysis from up-regulated and down-regulated proteins are less similar to the results given in the paper.
 The FDR values for some pathways are highly inconsistent.
@@
 N.A.
-| **1.** This paper could have done proper labelling of supplementary files as it was difficult the gene list for DAVID analysis.
+| **1.** This paper could have done proper labelling of supplementary files as it was difficult to find the gene list for DAVID analysis.
 Lacks information on when control genes are used and when citral-affected gene lists are used.
@@
 B.
-No | **3.** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” has changed.
+No | **3.** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” have changed.
 Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate.
@@
 | **2.** Does not consider corrected p values.
-Overall, significant terms from GO analysis of up and downregulated DEGs are missing upon replication.
+Overall, significant terms from GO analysis of up and down-regulated DEGs are missing upon replication.
 |
@@
 There is no clarity of the up-regulated genes used for GO and KEGG analysis.
-Only 21 upregulated genes are listed whereas the paper states 4831.
-Also, why are only upregulated genes are being used?
+Only 21 up-regulated genes are listed whereas the paper states 4831.
+Also, why are only up-regulated genes used?
 |
@@
 B.
-Yes, som.
+Yes, some.
 | **3.** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
@@
 We have developed a step-by-step protocol for how to conduct an extremely reproducible
-enrichment analysis, which is attached as a supplementary file (Suppl Info 1), and published to
+enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
 protocols.io with a permissive license.
@@
 stable release version (currently 3.16), running on R v4.2.2 and Ubuntu 22.04.
-The image contains instruction for installing a few useful utilities including `nano` and `git` for
-modifying scripts, and `magic-wormhole` for transfering data between computers.
+The image contains instructions for installing a few useful utilities including `nano` and `git` for
+modifying scripts, and `magic-wormhole` for transferring data between computers.
 It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
@@
 data set from DEE2, inspecting data quality, differential expression and two types of enrichment
-analysis; over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
+analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
 For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
@@
 To reproduce this work is relatively quick and simple, requiring six commands and approximately four
-minutes to install Docker, pull the image, run the container, run the Rmarkdown script exit the container
+minutes to install Docker, pull and run the container, run the Rmarkdown script, exit the container
 and copy the newly created report from the container to the current working directory (Box 1).
@@
 sudo apt update && sudo apt install docker-ce -y  # 60 sec
-docker pull mziemann/enrichment_recipe  # 60 sec
-docker run -it --entrypoint /bin/bash mziemann/enrichment_recipe #few seconds
+sudo docker run -it --entrypoint /bin/bash mziemann/enrichment_recipe #60 sec
 Rscript -e 'rmarkdown::render("example.Rmd")' # 90 sec
@@
 in triplicate @Lund2014-et.
-We use this analysis to better understand the effect of DNA methylatransferase inhibitors on
+We use this analysis to better understand the effect of DNA methyltransferase inhibitors on
 genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
@@
 
-With ORA, we observed 149 upregulated and 48 downregulated gene sets.
+With ORA, we observed 149 up-regulated and 48 down-regulated gene sets.
 We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
@@
 expression, rather than large gene sets that undergo small expression changes.
-This led to the identification of upregulated pathways related to nuclear RNA export,
-glucokinase regulation and G1/S DNA replication pathways; while downregulated pathways
+This led to the identification of up-regulated pathways related to nuclear RNA export,
+glucokinase regulation and G1/S DNA replication pathways; while down-regulated pathways
 were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
@@
 
-After FCS, we identified 241 upregulated genesets and 155 downregulated genesets.
-FCS upregulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
+After FCS, we identified 241 up-regulated gene sets and 155 down-regulated gene sets.
+FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
 and defective homologous recombination repair (Figure 5C).
-Downregulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
+Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
 
-Regarding upregulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
-Jaccard index of 0.54, however these method disagreed in downregulated pathways, with a Jaccard
+Regarding up-regulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
+Jaccard index of 0.54, however these methods disagreed in down-regulated pathways, with a Jaccard
 index of only 0.27 (Figure 5D).
@@
 
-* Ambiguity around whether up- and down-regulate genes were considered
-separately or together
+* Ambiguity around whether up- and down-regulated genes were considered
+separately or together.
 
@@
 were present in 85% of enrichment analyses @Wijesooriya2022-li.
-A non-exhaustive literature search identified a small number of case studies, but a dearth of systematic
-examinations of bioinformatics research reproducibility.
+
+While researching the literature on computational reproducibility in bioinformatics, we identified
+some case studies but only one other systematic examination of multiple journal articles.
 Ionnidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
 published in 2005-2006, and found only two could be reproduced to a satisfactory level.
+More work in this direction is urgently needed to understand the causes of irreproducible bioinformatics
+research so it can be remedied, as reproducibility is considered only the first step towards
+reliability.
 
@@
 These are good for data exploration, to quickly identify trends that can be followed up with more
-rigourous methods.
+rigorous methods.
 Moreover, a lot of work is being done to make such point-and-click tools more reproducible.
@@
 Although we hope that extremely reproducible enrichment analysis will become more widespread, we do not
-anticipate any dramatic changes in researcher behaviour unless the motivating factors are augmented.
-Reward structures in academia favour career progression for individuals who publish many articles
-as opposed to individuals who publish fewer but more rigourously conducted articles.
+anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
+Currently, reward structures in academia favour career progression for individuals who publish many
+articles as opposed to individuals who publish fewer but more rigorously conducted works.
 Solutions to this could include mandates by funding agencies or institutions @Garijo2013-zb,
-and/or badging of articles that comply with best practices @Kidwell2016-pm.
+badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
+in track record evaluations.
 
 ## Acknowledgements
+
+This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
+platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
 

### 2023-05-02 manuscript.Rmd
minor edits

@@
 
-* Protocol: protocols.io
+* Protocol: https://github.com/markziemann/enrichment_recipe/blob/main/docs/guide_for_linux_users_with_docker.md
+A final version will be deposited to protocols.io.
 
-* Boilerplate code: https://github.com/markziemann/enrichment_recipe
+* Code repository including template Dockerfile and Rmarkdown script: https://github.com/markziemann/enrichment_recipe
 

### 2023-05-02 docs/5pillars.Rmd
5 pillars explainer started

--- 2023-05-02 manuscript.Rmd
@@
 ---
-title: "A recipe for extremely reproducible enrichment analysis (and why we need it)"
-author: "Mark Ziemann and Anusuiya Bora"
+title: "The five pillars of computational reproducibility"
+author: "Mark Ziemann"
 date: "`r Sys.Date()`"
@@
     fig_height: 5
-bibliography: references.bib
+bibliography: 5pillars.bib
 csl: plos-computational-biology.csl
@@
 
-Mark Ziemann<sup>1</sup>*, Anusuiya Bora<sup>2</sup>
+Mark Ziemann
 
-**Affiliations**
+Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
 
-1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
+Correspondance: m.ziemann@deakin.edu.au
 
-2. Vellore Institute of Technology, Vellore, India.
+## Summary
 
-(*) Corresponding author: m.ziemann@deakin.edu.au
-
-## Abstract
-
-Unreliable and irreproducible research is a significant problem that wastes resources and risks
-undermining the public perception of science.
-Previous work has highlighted that published enrichment analyses frequently suffer from statistical
-and reporting flaws.
-We sought to determine whether this translates into irreproducibility by examining whether the
-findings of 20 open-access articles describing enrichment analysis were reproducible.
-We find that only three articles exhibited a high degree of concordance, while seven exhibited poor
-reproducibility.
-The findings indicate that current enrichment analysis practices are inadequate.
-To address this, we provide a set of template scripts and a step-by-step protocol to adapt them to
-a new project's enrichment analysis needs.
-This recipe is built according to a framework we call "five pillars", which when implemented together
-facilitates extreme reproducibility.
+Computational reproducibility is in theory a simple premise, however in practice it is difficult to
+achieve.
+Here, we present a framework to improve reproducibility, called the five pillars.
+These are a combination of best practices which ensure that the work can be reproduced quickly and
+easily, and will endure relatively long time-frames.
 
@@
 
-Reproducibility is essential to the scientific enterprise by confirming the validity of new discoveries
+Computational reproducibility is the ability to use the data and code provided by a research study to
+regenerate the outputs including figures and tables to confirm the study's findings.
+Reproducibility is only the first step towards overall reliability.
+Replicability is the ability to repeat the entire study and come up with similar overall results.
+Robustness refers to the ability for the findings to be generalisable to other situations
+@Nosek2022-zc.
+
+If a study is irreproducible or unreplicable, then it is of little value.
+
+essential to the scientific enterprise by confirming the validity of new discoveries
 [@Committee_on_Reproducibility_and_Replicability_in_Science2019-vb p1].
@@
 Secondly, code and data can be linked. This means that the code knows the location of the data,
-avoiding any "file not found" errors Peng2011-ia.
+avoiding any "file not found" errors @Peng2011-ia.
 Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
@@
 docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
+firefox example.html
 ```
@@
 The observed lack of reproducibility might be expected given that substantial methodological flaws
-were present in 85% of enrichment analyses @Wijesooriya2022-li.
+were present in 85% of enrichment analyses as part of a wider survey of articles of a similar age
+@Wijesooriya2022-li.
 

### 2023-05-03 docs/5pillars.Rmd
working on 5 pillars

@@
 Here, we present a framework to improve reproducibility, called the five pillars.
-These are a combination of best practices which ensure that the work can be reproduced quickly and
-easily, and will endure relatively long time-frames.
+These are a combination of best practices which ensure that the computational research work can be
+reproduced quickly and easily, and will endure relatively long time-frames.
 
@@
 
-Computational reproducibility is the ability to use the data and code provided by a research study to
-regenerate the outputs including figures and tables to confirm the study's findings.
-Reproducibility is only the first step towards overall reliability.
-Replicability is the ability to repeat the entire study and come up with similar overall results.
+In research, Computational reproducibility is the ability to use the data and code provided by a
+study to regenerate the outputs including figures and tables to confirm the study's findings.
+Reproducibility is only the first step towards overall reliability, for example a study may be
+reproducible, but suffer from analytical problems that invalidate the results.
+If a study is reproducible, then at least these analytical issues can be more easily identified and
+rectified.
+So reproducibility is a necessary stepping-stone that in theory should enhance the overall reliability
+of computational research including replicability and robustness.
+Replicability being defined as the ability to repeat the entire study and come up with similar overall
+results.
 Robustness refers to the ability for the findings to be generalisable to other situations
@@
 
-If a study is irreproducible or unreplicable, then it is of little value.
+Unreliable research is a waste of resources and time, and can mislead readers to pursue doomed
+research directions.
+Much has been written on how irreproducibility has reached crisis levels in psychology and life
+sciences [@Baker2015-bs; @Begley2012-di].
+Indeed in bioinformatics, the situation is grim with only a very fraction of studies
+being able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of those
+studies.
+Irreproducible research has the potential to place patient safety at risk, as demonstrated by an
+earlier case study @Baggerly2010-nr.
+
+As computational researchers, we are theoretically able to achieve 100% analytical reproducibility,
+however in practice this is rarely achieved.
+Data and code availability, ability to install software easily, and other factors mean that most
+computational research works are unable to be reproduced.
+
+A number of guides recommending enhanced computational reproducibility practices have emerged
+[@Peng2011-ia; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
+@Brito2020-ld; @Akalin2018-eg ]
+and this article attempts to unify these concepts and assemble a group of best practices where
+computational reproducibility can be considered as complete.
+
+```{r,5pillars}
+
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 10))
+polygon(c(0,1.8,1.8,0),c(1,1,6,6))
+polygon(c(2,3.8,3.8,2),c(1,1,6,6))
+polygon(c(4,5.8,5.8,4),c(1,1,6,6))
+polygon(c(6,7.8,7.8,6),c(1,1,6,6))
+polygon(c(8,9.8,9.8,8),c(1,1,6,6))
+polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8))
+polygon(c(-0.2,5,10),c(6.5,8.5,6.5))
+
+```
 

### 2023-05-04 manuscript.Rmd
Fig2 now reproducible

@@
 
-![Figure 2. The five pillars of computational reproducibility.](img/pillarsV2.png "The five pillars of computational reproducibility. .")
+```{r,5pillars,fig.cap="Figure 2. The five pillars of computational reproducibility.",echo=FALSE,fig.height=6}
+
+library("RColorBrewer")
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
+polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
+polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
+polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
+polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
+polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
+polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
+polygon(c(-0.2,5,10),c(6.5,8.5,6.5),col="lightgray", border = NA)
+points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
+text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)
+
+labs <- c("Version\ncontrol &\npersistent\ncode sharing",
+  "Compute \nenvironment \ncontrol",
+  "Literate\nprogramming",
+  "Documentation",
+  "FAIR &\npersistent\ndata sharing")
+
+text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
+text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
+
+```
 
@@
 
+library("DiagrammeR")
 grViz("digraph flowchart {

### 2023-05-04 docs/5pillars.Rmd
progress on 5pillar idea

@@
 
-In research, Computational reproducibility is the ability to use the data and code provided by a
-study to regenerate the outputs including figures and tables to confirm the study's findings.
+In research, computational reproducibility is the ability to use the materials from a past study
+(such as data, code and documentation) to regenerate the outputs including figures and tables to
+confirm the study's findings.
 Reproducibility is only the first step towards overall reliability, for example a study may be
@@
 studies.
-Irreproducible research has the potential to place patient safety at risk, as demonstrated by an
+Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
 earlier case study @Baggerly2010-nr.
@@
 and this article attempts to unify these concepts and assemble a group of best practices where
-computational reproducibility can be considered as complete.
+computational reproducibility can be considered as complete, using a framework called the
+five pillars of reproducible computational research (**Figure 1**).
+These concepts are unpacked in the following sections.
 
-```{r,5pillars}
+```{r,5pillars,fig.cap="Figure 1. The five pillars of computational reproducibility.",echo=FALSE,fig.height=6}
 
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 10))
-polygon(c(0,1.8,1.8,0),c(1,1,6,6))
-polygon(c(2,3.8,3.8,2),c(1,1,6,6))
-polygon(c(4,5.8,5.8,4),c(1,1,6,6))
-polygon(c(6,7.8,7.8,6),c(1,1,6,6))
-polygon(c(8,9.8,9.8,8),c(1,1,6,6))
-polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8))
-polygon(c(-0.2,5,10),c(6.5,8.5,6.5))
+library("RColorBrewer")
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
+polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
+polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
+polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
+polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
+polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
+polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
+polygon(c(-0.2,5,10),c(6.5,8.5,6.5),col="lightgray", border = NA)
+points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
+text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)
+
+labs <- c("Code version\ncontrol &\npersistent\n sharing",
+  "Compute \nenvironment \ncontrol",
+  "Literate\nprogramming",
+  "Documentation",
+  "FAIR &\npersistent\ndata sharing")
+
+text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
+text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
 
@@
 
-essential to the scientific enterprise by confirming the validity of new discoveries
-[@Committee_on_Reproducibility_and_Replicability_in_Science2019-vb p1].
-According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was a "significant
-crisis" of reproducibility @Baker2016-hm.
-While bioinformatics and other branches of computational research are theoretically amenable to
-complete reproducibility, in reality this is rarely achieved due to lack of detail provided in methods
-sections, lack of shared data and lack of code @Peng2011-ia.
-Indeed, even if data and code are shared, reproducibility still cannot be guaranteed due to the
-differences in computing environments over time, and the availability of dependencies @Perkel2020-io.
+## 1. Code version control and persistent sharing
 
-To address this, many articles have been written to endorse the use of "virtual machines" or
-"containerisation" to control the computing environment including the operating system and dependencies
-along with the research code to facilitate a high degree of reproducibility [@Perkel2020-io;
-@Stein2010-of; @Dudley2010-ua; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg;
-@Lewis2016-qh; @Da_Veiga_Leprevost2017-dw; @Pasquier2018-rt; @Kulkarni2018-km; @Hung2019-et;
-@Brito2020-ld; @Piccolo2021-ud]
-Containers also allow for an easy installation process, as all working dependencies are shipped together
-and can be installed with a single command.
+A version control system is a type of program that "tracks changes" made to sets of files, typically
+other computer programs, scripts and documentation.
+A set of files under version control is called a "repository" and represents a project or sub-project.
+Version control is used extensively by software developers and is considered one of the key best
+practices in software engineering.
+Computational researchers can benefit significantly by adopting version control into their routine
+practices.
+Below I outline some of the benefits.
 
-In addition to shared code, data and containerisation, there are a few ways researchers can make their
-work more reproducible.
-Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
-to novice bioinformaticians.
-Version control keeps track of code and documentation changes over time and manages the edits made by
-multiple contributors.
-Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
-free text explanations together with the code and outputs such as charts and tables.
-Secondly, code and data can be linked. This means that the code knows the location of the data,
-avoiding any "file not found" errors @Peng2011-ia.
-Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
-for each figure sequentially, making the entire manuscript reproducible by executing just one command
-and without any manual steps @Lasser2020-mm.
-Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five key
-pillars mentioned above (**Figure 2**).
-Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
-over a decade.
+* All steps of the workflow are coded in a computer script that allows automation.
+There are no manual intervention steps.
+For example, spreadsheets are irreproducible and prone to data entry, manipulation and formula errors.
+Even gene names can be automatically converted to dates, and this is common in publlished supplementary
+files.
+Includes data processing, and all figures and tables that go into the the final manuscript.
+If there are manual intervention steps, such as data cleaning or curating metadata, then the cleaned
+data can be shared and incorporated into the workflow.
 
-![Figure 1.
-The computational research reproducibility spectrum.
-Creator: Altuna Akalin.
-Republished with permission @Akalin2018-eg.](img/repro.webp "The computational research reproducibility spectrum.")
+* When designed correctly, the code "knows" where to access the associated metadata and data sets are,
+which avoids "file not found" errors.
+Metadata are usually small, so they can also be included in the repository as flat text files.
+Small data files can also be included in the repository, although keep in mind that big data files
+(>10MB) such as raw array scans, fastq files and the like are unsuitable for retention in a version
+control repository as the large files will slow down code updates signifiantly.
+Larger data files belong in a dedicated data repository, discussed below.
+If the metadata and small data require some cleaning, then include the original and cleaned data in the
+repository.
+The code should refer only to the cleaned data and the original files are there only for reference.
 
-![Figure 2. The five pillars of computational reproducibility.](img/pillarsV2.png "The five pillars of computational reproducibility. .")
+* For a research project, code gets really complicated very quickly.
+Workflows are build in a step wise way and require step-by-step validation that they are actually doing
+what we want them to do.
+There may be separate scripts for data processing, modeling and visualsation.
+There may be many false starts, research ideas that were pursued, but didn't give desired results and
+were removed from the final manuscript.
+A project that is under development will have code that undergoes many changes each week over the course
+of months to years.
+There may be many contributors to a study and merging these code changes can get difficult without
+version control.
 
-In this article, we focus on enrichment analysis, which is one of the most used methods in
-computational biology, and involves the summarisation of omics data to reflect biological changes,
-such as the detection of pathway activation in development or disease
-[@Slonim2002-sf; @Khatri2012-do].
-While enrichment analysis is undoubtedly a key method for the interpretation of large
-data sets, there are concerns that inadvertent misuse of these methods leads to
-unreliable results.
-These potential problems include (i) the lack of correct background for over-representation tests,
-(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
-[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
-Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
-the validity of the methods and results.
-We are concerned about the reliability of published enrichment analyses, as web-based and
-point-and-click tools appear to be far more popular than computer script-based methods for this type
-of work @Wijesooriya2022-li.
-Although web-based tools are convenient, they are a potential reproducibility concern.
-Firstly, incomplete reporting of methods, options and parameters is common.
-Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
-typically not made available, hampering later reproducibility.
-In some cases, the entire web resource can become unavailable, and hence irreproducible due to
-a phenomenon known as link decay @Hennessey2013-tj.
+* Version control keeps a track of all these changes over time.
+We always know which version of the script is the most up-to-date.
+We can inspect the code base at a particular point in the past.
+Commit messages are retained, which gives a record of the motivations and purpose behind each
+modification.
 
-A small number of enrichment analysis applications have been developed to have beneficial properties
-of being "containerised" and with archived historical versions [@Ge2020-qe; @Perampalam2020-qm].
-This helps reproducibility greatly but is not a complete solution as it lacks information about how
-the input data was used and a record of any parameter selections.
+* Version control also works extremely well for documentation.
+Books, educational resources and other data types are well served using version control.
 
-A solution to this is to use containerisation to provide a scripted
-workflow with prescribed links to research data, external databases and complete record of parameter
-selection.
-When designed correctly, these containers encapsulate bioinformatics workflows and could ensure their
-computational reproducibility for decades to come with just one or a few simple commands.
-There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
-existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
-a lack of step-by-step guides and templates aimed at beginner users.
+* Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
+from team members who may be working ascynchrously in different timezones and on different servers.
+It also makes the user's work searchable, which is convenient when code snippets from a project
+need to be reused for a new project.
+These centralised platforms also enable discussions of code "issues" such as bugs and feature requests.
+In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
+and allows team members to raise potential issues with a colleagues code (or one's own).
 
-We address this firstly by providing working templates for Docker image and Rmarkdown scripts for
-common transcriptome and enrichment analysis routines.
-Secondly, we provide step-by-step written and video guides to help users customise these templates for
-their own work, deploy the analysis, verify the results, and share the Docker image.
+* Version control makes works easily sharable.
+Work is easier to reproduce because the source folder contains a complete workflow with linked code
+that doesn't need any modifications before being executed to reproduce some works.
 
-However, before recommending or prescribing such workflows, it is necessary
-to assess directly whether current practices are adequate, by trying to reproduce the enrichment
-analyses of previously published studies, and examining whether the conclusions of those articles
-are supported or not.
-In the following section we investigate the computational reproducibility of enrichment analyses
-shown in a sample of 20 open-access journal articles published in 2019.
-This should provide a sound basis for recommendations.
+* `git` is the defacto standard for version control currently although there are alternatives like
+`svn` (Subversion) and `hg` (Mercurial).
+Using `git` is advisable because there is a greater depth of support available on web forums, and more
+alternatives for centralised code hosting providers and graphical interfaces.
 
-## A systematic assessment of reproducibility in enrichment analysis
+## 2. Compute environment control
 
-### Methods
+* Conda
 
-From a list of 186 PubMed central articles using enrichment analysis that we examined previously
-@Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
-organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
-Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
-expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
-The basis for these selection criteria is that this combination represents the most common application
-of enrichment analysis in our previous study, which could be reproduced without the need to
-reanalyse the raw dataset from scratch. These articles are listed in the bibliography
-[@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
-@Guo2019-fx; @Alebrahim2019-zc; @Hong2019-ym; @Han2019-ix; @McSweeney2019-ij; @Wang2019-yw; @Pan2019-iv;
-@Wang2019-jy; @Weigt2019-mk; @Najafi2019-ce; @Feltrin2019-to; @Wang2019-xx; @Sun2019-xr].
+* Containers: Docker/Guix/others
 
-The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
-special attention to preventing gene name errors from occurring @Zeeberg2004-na.
-The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles using DAVID version 6.8 (March-May 2022).
+* Virtual machines
 
-Briefly, we used the DAVID "Analysis Wizard" and pasted in the provided gene list into the text box,
-selecting this as "gene list", together with the appropriate gene identifier type.
-Typically this was the official gene symbol.
-The species was specified as *Homo sapiens*.
-Next we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
-page.
-We followed the steps indicated in the respective papers to select appropriate gene sets investigated
-in the articles such as gene ontology (biological process, molecular function), KEGG pathways, etc.
-Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
-enrichment results, which were saved for future reference.
+* Wasm
 
-The data presented in the paper was compared with the results we obtained, recording any
-discrepancies.
-Additionally, the statements made in the abstract, results, and discussion sections were carefully
-examined and cross-checked with our results to understand whether those assertions survive reproduction
-or not.
+## 3. Literate programming
 
-Each article was critically evaluated and given a score of one (1) to three (3).
-Articles rated 1 showed overall poor reproducibility.
-For example most GO and/or KEGG terms mentioned in the results and discussion were not confirmed
-upon reproduction, and the conclusions of the study were overall not supported.
-Articles were rated 2 when at least 50% of the terms are observed as significant after replication,
-according to the original study's significance threshold, and most conclusions were supported.
-Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
-shown in the article observed again in reproduction, and all conclusions were supported in reproduction.
+The report is truth.
 
-### Results
+* Sweave, Rmarkdown, Quarto
 
-After conducting this reproduction study according to the methods described in the respective articles,
-the 20 articles were classified based on the degree of reproducibility (Figure 3).
-Six studies were classified as low reproducibility, while 10 were classified as medium and just three
-were classified as high.
-Justifications for these classifications are provided in Table 1.
+* Jupyter
 
-```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 3. Examining reproducibility in a sample of 20 journal articles."}
+* Compilation
 
-score1 <- c("PMC6405693","PMC6425008","PMC6535219","PMC6539328","PMC6542760","PMC6561911","PMC6663624")
-score2 <- c("PMC6368841","PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
-  "PMC6582306","PMC6333352","PMC6526186","PMC6607402")
-score3 <- c("PMC6349697","PMC6444048","PMC6587650")
-l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)
+* Posting reports to lab website and webarchive
 
-barplot(unlist(lapply(l,length)),col="lightgray",
-  xlab="reproducibility score",ylab="no. articles",
-  ylim=c(0,11))
+* Make the manuscript reproducible as well.
 
-grid()
-text(0.7,1:7-0.5,labels=score1)
-text(1.9,1:10-0.5,labels=score2)
-text(3.1,1:3-0.5,labels=score3)
+## 4. Documentation
 
-pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
-pc2 <- paste(pc,"%",sep="")
-text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)
+* Instructions needed
 
-```
+* Requirements
 
-| S.No.
-Pubmed ID: Replication of (Table/Figure) | A.
-Was Replication possible?
-B.
-Were there other significant terms found which were not mentioned in the article?
-| Replication score and justification |
-| --- | --- | --- |
-| 1.PMC6349697: Figure 2C and 2D | A.
-Yes, however, corrected p values are different after replication.
-B.
-Yes, only BP terms were taken into account so several CC and MF significant terms were not mentioned in the paper.
-|  **3.** Considering top 6 BP Terms and KEGG pathways.
-The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed when there is a higher amount of significant GOs and pathways are observed after replication.
-It was impossible to determine if the article utilised a background gene list or not.
-The steps to perform DAVID analysis could have been elaborated in the paper.
-The results shown in the paper are observed after replication but only 6 BP GO significant terms are shown in the paper.
-|
-| 2.PMC6368841: Figure 3 A - D | A.
-Only GO Terms were the same as the paper.
-KEGG terms were highly inconsistent.
-B.
-Replicated KEGG Terms (using DAVID) were not mentioned in the results of the paper.
-Shows inconsistency due to the usage of the server.
-Some terms using clusterprofiler in R were mentioned.
-| **2.** GO results from the original manuscript are somewhat similar to reproduced analysis.
-The FDR values however are inconsistent.
-KEGG results were highly different.
-The list of genes as input into DAVID analysis is unclear.
-They should have mentioned both up and down-regulated proteins were considered.
-|
-| 3.PMC6381667: Figure 1 and Figure 2 | A.
-Results are not reproducible and are highly inconsistent B.
-N.A.
-| **2.** GO terms obtained after replication showed inconsistent results compared to the results in the original article.
-KEGG results were similar.
-Better naming convention for supplementary file for 117 genes for DAVID input is necessary.
-GO Results are not reproducible and are highly inconsistent.
-|
-| 4.PMC6405693: Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways” | A.
-Results do not match.
-B.
-N.A.
-| **1.** GO analysis from up-regulated and down-regulated proteins are less similar to the results given in the paper.
-The FDR values for some pathways are highly inconsistent.
-KEGG Analysis of all DEGs do not match at all.
-|
-| 5.PMC6425008: Most of the tables and figures of the paper (Figure 5, 6, 7), Table 2 | A.
-No.
-B.
-N.A.
-| **1.** This paper could have done proper labelling of supplementary files as it was difficult to find the gene list for DAVID analysis.
-Lacks information on when control genes are used and when citral-affected gene lists are used.
-Replication of GO and KEGG analysis is not consistent with results from the paper.
-How were Figures 5-7 derived?
-The overall protocol is not so coherent.
-|
-| 6.PMC6444048: The text under “Pathway analysis of SUVmax related genes” And Table 2 | A.
-Terms are the same but gene counts have been different.
-The DAVID results are somewhat reproducible.
-B.
-No | **3.** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” have changed.
-Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate.
-The results claimed are somewhat similar.
-All the processes have different FDR values.
-|
-| 7.PMC6463127: Table 4, Table 5, Table 6 | A.
-Yes but the significance differed.
-B.
-Yes, several terms from replication were not mentioned in the paper for GO Analysis.
-| **2.** Does not consider corrected p values.
-Overall, significant terms from GO analysis of up and down-regulated DEGs are missing upon replication.
-|
-| 8.PMC6535219: Figure 2 | A.
-Somewhat similar.
-B.
-There were many top significant terms that were not mentioned.
-| **1.** Narrowing/Filtering the given gene list is a required step that could have been mentioned.
-Many terms were not so significant anymore.
-There were new terms that are of top positions after replication.
-|
-| 9.PMC6539328: Figure 1A | A.
-Mostly not.
-B.
-Yes, many | **1.** The significant pathways stated in the paper are not showing up once replicated.
-There is no clarity of the up-regulated genes used for GO and KEGG analysis.
-Only 21 up-regulated genes are listed whereas the paper states 4831.
-Also, why are only up-regulated genes used?
-|
-| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1.** Replication results of GO terms are not at all compliant with the results shown in the paper. |
-| 11.PMC6557785: Figure 5 A-B | A.
-Yes, but results in paper are not significant upon replication.
-B.
-Yes | **2.** For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
-For KEGG Pathways, there are other pathways that are more significant and contain a higher number of genes enriched for example “Salmonella infection” that has a higher count of genes but is not mentioned in the paper.
-|
-| 12.PMC6561911: Figure 5 a and b | A. No. B. N.A. | **1.** The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done |
-| 13.PMC6580941: Table 1 and Table 2 | A.
-Yes.
-B.
-Yes!
-Significant terms like “nucleoplasm”, “RNA Binding”, “protein binding” are missing from the paper.
-| **2.** For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
-The Gene list for input is not so clearly mentioned, it should have been labelled properly in the name of the supplemental file.
-P-values are completely different.
-|
-| 14.PMC6587650: Figure 2 | A.
-Yes.
-B.
-Yes, some.
-| **3.** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
-They could have mentioned why they chose the type of sorting shown in the paper to display their results.
-Overall, all results have been replicated.
-|
-| 15.PMC6591946: Figure 5 c and Figure 6 a | A.
-Yes for GO analysis only.
-B.
-Yes for KEGG analysis | **2.** The GO term results of replication are compliant with Figure 5 c of the paper.
-KEGG pathway results are semi-consistent.
-|
-| 16.PMC6663624: Table 3 and Table 4 | A.
-Yes.
-B.
-Yes | **1.** The top 2 significant GO terms in the paper are not significant after replication analysis is done.
-Other GO terms are the same as replicated results.
-For KEGG Analysis, some terms are more significant after replication.
-Eg: PD-L1 expression and PD-1 checkpoint pathway in cancer, Th1 and Th2 cell differentiation, Th17 cell differentiation were more significant after “natural killer (NK) cell-mediated cytotoxicity” KEGG Pathway.
-The paper is somewhat replicable.
-|
-| 17.PMC6582306: Figure 5 | A.
-Partly.
-B.
-Yes, many.
-| **2.** The paper uses 44 sets of DEGs grouped in 4 major types.
-For replication, OFC Set 9 was used, DAVID GO analysis was done twice where a discrepancy of results was observed.
-Compared to the paper, the most significant term was missing after replication analysis was conducted.
-For OFC set 9, several pathways are seen after replication but only one pathway is shown in paper.
-This should be addressed.
-|
-| 18.PMC6333352: Figure 7 | A.
-Yes.
-B.
-Yes.
-There was a difference in gene counts as well.
-| **2.** DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
-From NERI - BAHN - KEGG Analysis: “Pathways of neurodegeneration - multiple diseases”, “Pathways in cancer” - pathways with the highest gene count (30 and 28 respectively) are not mentioned in the paper.
-Other pathways are present and the paper is somewhat reproducible.
-|
-| 19.PMC6526186: Table 1 | A.
-Yes.
-B.
-Yes.
-| **2.** The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
-The top 2 KEGG Pathways after replication are “Human cytomegalovirus infection” followed by “Pathways in cancer”.
-The p-values have changed after the results were replicated, compared to the values given in the paper.
-The rest of the pathways are similar to the results given in the paper.
-|
-| 20.PMC6607402: Figure 1 A-D | A.
-Yes.
-B.
-Yes, many!
-| **2.** For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
-Most of the terms are mentioned in the replicated results, but different levels of enrichment.
-|
+* Credits
 
-: Table 1. Classification and justification for each article.
+## 5. FAIR and persistent data sharing
 
-Taken together, the ability to replicate DAVID enrichment analysis using the supplementary gene
-lists was highly variable.
+* NCBI
 
-## A container-based approach to addressing reproducibility of enrichment analysis
+* Zenodo
 
-### Implementation
+* Figshare
 
-We have developed a step-by-step protocol for how to conduct an extremely reproducible
-enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
-protocols.io with a permissive license.
-This protocol has the features of reproducibility mentioned in **Figure 1** and **Figure 2** including
-version control with git and GitHub, compute environment control using Docker, literate programming
-with Rmarkdown, persistent code and data sharing with Zenodo and documentation of all of these to
-empower others to reproduce the workflow.
+* Arweave/IPFS
 
-Central to this protocol is the provision of template Dockerfile and an Rmarkdown script, which
-are intended to be remixed and altered by the end user to suit the needs of their project.
+* Portico
 
-The Dockerfile uses a base image from Bioconductor which should correspond to the latest available
-stable release version (currently 3.16), running on R v4.2.2 and Ubuntu 22.04.
-The image contains instructions for installing a few useful utilities including `nano` and `git` for
-modifying scripts, and `magic-wormhole` for transferring data between computers.
-It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
-analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` @Ziemann2019-zo,
-differential expression with `DESeq2` @Love2014-hd,
-enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
-[@Yu2012-ts; @Korotkevich2016-ls; @Kaspi2020-dq].
-It also contains instructions to clone a github repository, which contains the scripts that run the
-analytical workflow.
-Lastly, it copies a gene set library from Reactome @Gillespie2022-gz with a clear version
-history to guarantee provenance of the annotation data.
+## Bringing it all together
 
-The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
-Rmarkdown allows the unification of extended descriptions, code and results into a single document,
-which is becoming ubiquitous in bioinformatics and data science more generally
-@Grolemund2017-zl p423.
-The ability to provide extended descriptions is helpful for providing contextualising information about
-the purpose of the work, detailed methods and interpretations derived from the analysis.
-The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
-data set from DEE2, inspecting data quality, differential expression and two types of enrichment
-analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
-For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
-@Zhao2023-ok.
+Extreme reproducibility should be the new standard.
 
-```{r,diagram,echo=FALSE,fig.cap = "Figure 4. Flow chart of example workflow."}
+It generally doesn't matter what type of computer you're using.
 
-grViz("digraph flowchart {
-      node [fontname = Helvetica, shape = rectangle]
-      tab1 [label = '@@1']
-      tab2 [label = '@@2']
-      tab3 [label = '@@3']
-      tab4 [label = '@@4']
-      tab5 [label = '@@5']
-      tab6 [label = '@@6']
-      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
-      tab3 -> tab5 -> tab6 }
-      [1]: 'Fetch RNA-seq from DEE2'
-      [2]: 'Quality control'
-      [3]: 'Differential expression'
-      [4]: 'Over-representation analysis'
-      [5]: 'Functional class scoring'
-      [6]: 'Euler diagram'
-")
+The benefit is that even if one of the pillars breaks, the overall study could be reproduced
+thanks to the remaining pillars.
 
-```
-
-The software versions used in this example analysis are provided in Table 2.
-
-| Software | Version | Purpose |
-| --- | --- | --- |
-| Ubuntu | 22.04 | Operating System |
-| R | 4.2.2 | Statistical computing language |
-| nano | 6.2 | text editing |
-| git | 2.34.1 | source control |
-| R/kableExtra | 1.3.4 | render formatted tables |
-| R/vioplot | 0.4.0 | violin charts |
-| R/gplots | 3.1.3 | heatmaps |
-| R/eulerr | 7.0.0 | Venn diagrams |
-| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
-| BioC/DESeq2 | 1.38.3 | differential expression |
-| BioC/fgsea | 1.24.0 | functional class scoring |
-| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
-| Bioc/mitch | 1.10.0 | multi-contrast enrichment |
-
-: Table 2. The software bundled in the example Docker image.
-
-To reproduce this work is relatively quick and simple, requiring six commands and approximately four
-minutes to install Docker, pull and run the container, run the Rmarkdown script, exit the container
-and copy the newly created report from the container to the current working directory (Box 1).
-
-```bash
-sudo apt update && sudo apt install docker-ce -y  # 60 sec
-sudo docker run -it --entrypoint /bin/bash mziemann/enrichment_recipe #60 sec
-Rscript -e 'rmarkdown::render("example.Rmd")' # 90 sec
-exit
-docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
-firefox example.html
-```
-<p style="text-align: center;">
-*Box 1. Reproduce a transcriptome and enrichment analysis with Docker.*
-</p>
-
-### Demonstration
-
-To demonstrate the utility of this workflow, we present the results of this example data analysis;
-it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
-in triplicate @Lund2014-et.
-We use this analysis to better understand the effect of DNA methyltransferase inhibitors on
-genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
-the relationship between RNA expression and DNA methylation, but did not show results of a
-transcriptome-only enrichment analysis.
-Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
-FCS, and give guidance for interpreting results of such tests.
-
-For quality control the number of reads per sample was calculated, and found to be in the range of
-11-16M.
-Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
-genes.
-Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
-1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
-(Figure 5A).
-Up and down-regulated gene sets were subjected to ORA separately, using the list of
-13,168 genes as the background, and gene sets from REACTOME.
-
-With ORA, we observed 149 up-regulated and 48 down-regulated gene sets.
-We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
-which tends to highlight smaller and more specific gene sets undergoing more drastic differential
-expression, rather than large gene sets that undergo small expression changes.
-This led to the identification of up-regulated pathways related to nuclear RNA export,
-glucokinase regulation and G1/S DNA replication pathways; while down-regulated pathways
-were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
-(host defence) (Figure 5B).
-
-After FCS, we identified 241 up-regulated gene sets and 155 down-regulated gene sets.
-FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
-and defective homologous recombination repair (Figure 5C).
-Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
-
-Regarding up-regulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
-Jaccard index of 0.54, however these methods disagreed in down-regulated pathways, with a Jaccard
-index of only 0.27 (Figure 5D).
-
-![Figure 5. Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
-
-## Concluding remarks
-
-Although our reproducibility study was relatively small, there are a few key observations.
-We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
-In a way, this is surprising, given the simplicity of using DAVID, only
-requiring the user to provide a list of genes in a text box and select some options.
-These irregularities may have arisen inadvertently from several sources:
-
-* The methods stated in the respective articles are inconsistent with what was actually done.
-
-* The list of genes in the supplement is not the same as what was used for enrichment analysis.
-
-* Ambiguity around whether up- and down-regulated genes were considered
-separately or together.
-
-* Ambiguity around whether all GO classes were considered or just one, such as "biological process".
-
-* Ambiguity around whether a custom background gene set was used.
-
-The observed lack of reproducibility might be expected given that substantial methodological flaws
-were present in 85% of enrichment analyses as part of a wider survey of articles of a similar age
-@Wijesooriya2022-li.
-
-While researching the literature on computational reproducibility in bioinformatics, we identified
-some case studies but only one other systematic examination of multiple journal articles.
-Ionnidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
-published in 2005-2006, and found only two could be reproduced to a satisfactory level.
-More work in this direction is urgently needed to understand the causes of irreproducible bioinformatics
-research so it can be remedied, as reproducibility is considered only the first step towards
-reliability.
-
-This problem is not limited to bioinformatics.
-In the field of drug development, according to a 2011 study by Bayer, only 25% of projects could be
-replicated to a high degree @Prinz2011-ht.
-When the US Biotech firm Amgen set out to reproduce 53 landmark cancer studies, they were able to
-reproduce only 6 (11%) @Begley2012-di.
-It is clear that as a field, the life sciences need to raise the rate of reproducibility or risk the
-public perceiving it as unreliable and wasteful.
-
-We propose the five pillars (**Figure 2**), as a framework to guarantee computational reproducibility.
-However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
-protocol to guide them on how to customise template Dockerfile and Rmarkdown scripts to suit their
-project's needs.
-Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
-we estimate at 15-20 hours.
-This is significantly longer than using a web-based tool that provides results in just a few
-minutes, but we argue the time investment is well spent.
-It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
-simple and fast (Box 1), which should aid the early identification of errors.
-
-We also want to emphasise that there is still a definite need for graphical and web-based tools.
-These are good for data exploration, to quickly identify trends that can be followed up with more
-rigorous methods.
-Moreover, a lot of work is being done to make such point-and-click tools more reproducible.
-The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
-workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
-The Degust differential expression suite is a graphical interface to popular
-expression tools like edgeR and limma, and allows users to download the script to facilitate future
-non-interactive reproducibility @Powell2019-lv.
-Such functionality could be written into existing enrichment analysis tools like DAVID, and together with
-retention of historical gene annotation sets could provide a "good enough" solution to reproducibility.
-
-Although we hope that extremely reproducible enrichment analysis will become more widespread, we do not
-anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
-Currently, reward structures in academia favour career progression for individuals who publish many
-articles as opposed to individuals who publish fewer but more rigorously conducted works.
-Solutions to this could include mandates by funding agencies or institutions @Garijo2013-zb,
-badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
-in track record evaluations.
+More systematic studies to assess reproducible practices are required.
 
@@
 
-This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
-platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+I am grateful to the individuals who provided feedback on an [earlier draft of the five pillars
+concept](https://twitter.com/mdziemann/status/1650371743113175040) including:
 
-## Availability of materials
+* Sriganesh Srihari
 
-* Protocol: https://github.com/markziemann/enrichment_recipe/blob/main/docs/guide_for_linux_users_with_docker.md
-A final version will be deposited to protocols.io.
+* Barbara Zdrazil
 
-* Code repository including template Dockerfile and Rmarkdown script: https://github.com/markziemann/enrichment_recipe
+* Aziz Khan
 
-* Example Docker image: https://hub.docker.com/r/mziemann/enrichment_recipe
+* Pierre Poulain
+
+* Irene López Santiago
 

### 2023-05-04 manuscript.Rmd
Update manuscript.Rmd

@@
 |  **3.** Considering top 6 BP Terms and KEGG pathways.
-The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed when there is a higher amount of significant GOs and pathways are observed after replication.
+The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs, to be addressed when there is a higher amount of significant GOs and pathways that are observed after replication.
 It was impossible to determine if the article utilised a background gene list or not.
@@
 B.
-Replicated KEGG Terms (using DAVID) were not mentioned in the results of the paper.
+When replicated, KEGG Terms (using DAVID) were not mentioned in the results of the paper.
 Shows inconsistency due to the usage of the server.
@@
 The list of genes as input into DAVID analysis is unclear.
-They should have mentioned both up and down-regulated proteins were considered.
+They should have mentioned that both up and down-regulated proteins were considered.
 |

### 2023-05-04 manuscript.Rmd
Update manuscript.Rmd

@@
 |  **3.** Considering top 6 BP Terms and KEGG pathways.
-The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs, to be addressed when there is a higher amount of significant GOs and pathways that are observed after replication.
+The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed, when there is a higher amount of significant GOs and pathways that are observed after replication.
 It was impossible to determine if the article utilised a background gene list or not.

### 2023-05-08 manuscript.Rmd
using docker.io

@@
 ```bash
-sudo apt update && sudo apt install docker-ce -y  # 60 sec
+sudo apt update && sudo apt install docker.io -y  # 30 sec
 sudo docker run -it --entrypoint /bin/bash mziemann/enrichment_recipe #60 sec

### 2023-05-08 manuscript.Rmd
slight mods

@@
 @Lewis2016-qh; @Da_Veiga_Leprevost2017-dw; @Pasquier2018-rt; @Kulkarni2018-km; @Hung2019-et;
-@Brito2020-ld; @Piccolo2021-ud]
+@Brito2020-ld; @Piccolo2021-ud].
 Containers also allow for an easy installation process, as all working dependencies are shipped together
@@
 Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
-for each figure sequentially, making the entire manuscript reproducible by executing just one command
+for each part sequentially, making the entire manuscript reproducible by executing just one command
 and without any manual steps @Lasser2020-mm.
-Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five key
-pillars mentioned above (**Figure 2**).
+Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five
+pillars (**Figure 2**).
 Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
@@
 unreliable results.
-These potential problems include (i) the lack of correct background for over-representation tests,
+These potential problems include (i) the lack of a correct background for over-representation tests,
 (ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
@@
 Secondly, we provide step-by-step written and video guides to help users customise these templates for
-their own work, deploy the analysis, verify the results, and share the Docker image.
+their own work, deploy the analysis, verify the results, and share the materials.
 

### 2023-05-08 docs/5pillars.Rmd
5pil update

@@
 sciences [@Baker2015-bs; @Begley2012-di].
-Indeed in bioinformatics, the situation is grim with only a very fraction of studies
-being able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of those
-studies.
+Indeed in bioinformatics, the situation is grim with a 2009 report showing only a small fraction
+of studies able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of
+those studies.
 Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
@@
 ```
+
+## Literate programming
+
+A common recommendation for reproducibility is to complete all works with computer scripts
+and remove the need for any manual steps.
+Computer programs (scripts) are preferred over graphical interface tools as the script contains
+a complete instruction set for how a piece of work was conducted.
+This allows auditing and reproduction which would be difficult for graphical tools like spreadsheets.
+Scripts also allow the work to be completely automated and reduce the chance for human error during
+any manual steps.
+On the other hand, data cleaning and curation is sometimes required, and so manual steps are unavoidable
+in some situations.
+
+Literate programming combines "chunks" of analytical code together with human readable text @Knuth1984-di.
+After compilation, the resulting output contains the code with computational results
+(eg: figures and tables) along with contextualising explanations and narratives.
+
+The Sweave project, concieved in 2001, was designed to give LaTeX documents embedded R code chunks and
+was envisaged as a way to generate R documentation materials and generate statistical analysis reports
+@Leisch2002-yc.
+Around 2015, R Markdown emerged as a more popular alternative, as formatting is simpler as compared to Sweave,
+saving time.
+Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
+R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
+In 2022, the company behind RStudio (Posit) released Quarto, the conceptual successor of R Markdown, but with
+enhanced support for other computer languages like Python, Julia, and Observable JavaScript
+(and more attractive reports as well) @Bauer2023-eh.
+
+Whether Rmarkdown, Jupyter or Quarto, literate programming offers some substantial benefits over alternative
+approaches:
+
+* The provenance of any result can be demonstrated.
+The resulting document produced contains the code executed, together with the results (eg: a chart).
+This is in contrast to a data analysis report assembled in a word processor, which due to the many
+copy-paste operations required might contain errors.
+
+* It saves time.
+For the case where a report needs to be run routinely, this would involve significant copy-paste to
+be assembled "manually" with a word processor.
+Using a literate script would mean only minimal changes are required between interations.
+
+* Literate programs accommodate extensive documentation.
+This allows the analyst to include rich descriptions of scientific works.
+For example, a report may contain background, methods, results, discussion and references.
+Not only text, but various objects can be included like links to other resources,
+tables, images, videos, etc.
+More information on documentation below.
+
+* Outputs are arranged.
+When using a regular script, an analyst might make dozens of charts which are written to the
+disk with names like "chart1.png" or "model8.svg".
+When the number of outputs is large, it becomes unwieldy and hard to place which plot corresponds
+to which part of the script.
+By arranging the outputs (eg: charts and tables) in a document in the sequence that they were generated,
+it helps the reader understand the logical steps taken in any analysis.
+
+* Literate scripts are amenable to version control.
+Version control is a useful best practice in software development (discussed below).
+
+* The resulting document is only rendered when the entire script has compiled without errors.
+In contrast, a regular script might generate some outputs and then encounter an error,
+so we are not sure whether the whole script is free of errors.
+
+* Flexible output formats.
+These include PDF, DOC/DOCX and HTML, with HTML having some notable benefits to other formats
+including better usability for mobile devices like phones and tablets, and the ability to
+support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
 

### 2023-05-09 docs/5pillars.Rmd
5 pillars work

@@
 
-## Literate programming
+## The five pillars: a code-first approach
 
@@
 and remove the need for any manual steps.
-Computer programs (scripts) are preferred over graphical interface tools as the script contains
+Computer scripts are preferred over graphical interface tools as the script contains
 a complete instruction set for how a piece of work was conducted.
 This allows auditing and reproduction which would be difficult for graphical tools like spreadsheets.
+Indeed spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
+somewhat overused in the life sciences.
 Scripts also allow the work to be completely automated and reduce the chance for human error during
 any manual steps.
-On the other hand, data cleaning and curation is sometimes required, and so manual steps are unavoidable
-in some situations.
+One thing to keep in mind is that, data cleaning and curation is often required, and in many cases
+can only be conducted using manual intervention.
+In those cases, retaining raw and cleaned data and a cleaning protocol is recommended.
 
-Literate programming combines "chunks" of analytical code together with human readable text @Knuth1984-di.
-After compilation, the resulting output contains the code with computational results
-(eg: figures and tables) along with contextualising explanations and narratives.
+Another principle of the five pillars approach is that the publishing researchers should put in the
+extra effort over and above the minimum standards of a classical journal to make the reproduction
+process easier and faster for those who wish to subsequently reproduce the analysis.
+
+## Literate programming
+
+Literate programming combines "chunks" of analytical code with human readable text @Knuth1984-di.
+After compilation, the resulting output document contains the code together with computational results
+such as figures and tables along with contextualising explanations and narratives.
 
@@
 @Leisch2002-yc.
-Around 2015, R Markdown emerged as a more popular alternative, as formatting is simpler as compared to Sweave,
-saving time.
+Around 2015, R Markdown emerged as a more popular alternative, as formatting is simpler as compared to
+Sweave, saving time.
 Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
 R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
-In 2022, the company behind RStudio (Posit) released Quarto, the conceptual successor of R Markdown, but with
-enhanced support for other computer languages like Python, Julia, and Observable JavaScript
-(and more attractive reports as well) @Bauer2023-eh.
+In 2022, the company behind RStudio (Posit) released Quarto, the conceptual successor of R Markdown,
+but with enhanced support for other computer languages like Python, Julia, and Observable JavaScript
+@Bauer2023-eh.
 
-Whether Rmarkdown, Jupyter or Quarto, literate programming offers some substantial benefits over alternative
-approaches:
+Whether Rmarkdown, Jupyter or Quarto, literate programming offers some substantial benefits over
+alternative approaches:
 
@@
 This is in contrast to a data analysis report assembled in a word processor, which due to the many
-copy-paste operations required might contain errors.
+copy-paste operations required might contain errors, or version mismatches.
 
@@
 to which part of the script.
-By arranging the outputs (eg: charts and tables) in a document in the sequence that they were generated,
+By embedding the outputs (eg: charts and tables) in a document in the sequence that they were generated,
 it helps the reader understand the logical steps taken in any analysis.
@@
 
-* The resulting document is only rendered when the entire script has compiled without errors.
+* Literate programming output reports are free from code errors.
+The resulting document is only rendered when the entire script has compiled without errors.
 In contrast, a regular script might generate some outputs and then encounter an error,
-so we are not sure whether the whole script is free of errors.
+so we are not sure whether the whole script is free of errors and completed successfully.
 
@@
 support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
+Outputs include HTML documents, but also slideshows and even e-books.
+
+These features make literate programming a quantum leap for science communication in a range of
+situations.
+Whether this is sharing a data analysis report, giving a presentation at a meeting,
+writing a research article or self-publishing an e-book, literate programming provides the
+ability to construct transparent data narratives with clear provenance in a convieniently sharable
+form.
+
+Literate programming also works neatly with the concept of the "executable paper", the idea that
+the data analysis underlying an entire study can be reproduced with one or a few commands @Lasser2020-mm.
+A typical genomics/bioinformatics study could involve one large literate script, or be broken down into
+smaller scripts, where each one contains the code for generating a part of the article.
+A multi-script approach would benefit from a master script that executes each component.
+This further makes the job of reproducibility easier.
 
@@
 
-A version control system is a type of program that "tracks changes" made to sets of files, typically
-other computer programs, scripts and documentation.
+A version control system (sometimes called "source control") is a type of program that "tracks changes"
+made to sets of files, typically other computer program source code, scripts and documentation.
 A set of files under version control is called a "repository" and represents a project or sub-project.
@@
 practices in software engineering.
+Distributed version control systems involve a central web-accessible server to hosting a repository,
+and each team member possesses a mirror copy on their local system.
+Having a central, publicly available node assists in disseminating changes within teams and releasing
+code to consumers.
+There are many such distributed version control systems but `git` has emerged as the most popular
+solution due to it's many powerful features, speed and large community and ecosystem.
+
 Computational researchers can benefit significantly by adopting version control into their routine
@@
 
-* All steps of the workflow are coded in a computer script that allows automation.
-There are no manual intervention steps.
-For example, spreadsheets are irreproducible and prone to data entry, manipulation and formula errors.
-Even gene names can be automatically converted to dates, and this is common in publlished supplementary
-files.
-Includes data processing, and all figures and tables that go into the the final manuscript.
-If there are manual intervention steps, such as data cleaning or curating metadata, then the cleaned
-data can be shared and incorporated into the workflow.
+* Version control keeps a track of all these changes over time.
+We always know which version of the script is the most up-to-date.
+We can inspect and execute the code and reproduce the analysis at any particular point in the past.
+Commit messages are retained, which gives a record of the motivations and purpose behind each
+modification.
+
+* Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
+from team members who may be working ascynchrously in different timezones and on different servers.
+It also makes the user's work searchable, which is convenient when code snippets from a project
+need to be reused for a new project.
+These centralised platforms also enable discussions of code "issues" such as bugs and feature requests.
+In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
+and allows team members to raise potential issues with a colleagues code (or one's own).
+
+* Version control also works extremely well for documentation.
+Books, educational resources and other data types are well served using version control (see below).
+
+* Version control makes works easily sharable.
+Work is easier to reproduce because the source folder contains a complete workflow with linked code
+that doesn't need any modifications before being executed to reproduce some works.
+
+* Distributed version control systems protect against code loss.
+Computers sometimes fail, and if there are
+
+## 2. Compute environment control
+
+* Conda
+
+* Containers: Docker/Guix/others
+
+* Virtual machines
+
+* Wasm
+
+## 4. Documentation
+
+* Instructions needed
+
+* Requirements
+
+* Credits
+
+## 5. FAIR and persistent data sharing
 
@@
 The code should refer only to the cleaned data and the original files are there only for reference.
-
-* For a research project, code gets really complicated very quickly.
-Workflows are build in a step wise way and require step-by-step validation that they are actually doing
-what we want them to do.
-There may be separate scripts for data processing, modeling and visualsation.
-There may be many false starts, research ideas that were pursued, but didn't give desired results and
-were removed from the final manuscript.
-A project that is under development will have code that undergoes many changes each week over the course
-of months to years.
-There may be many contributors to a study and merging these code changes can get difficult without
-version control.
-
-* Version control keeps a track of all these changes over time.
-We always know which version of the script is the most up-to-date.
-We can inspect the code base at a particular point in the past.
-Commit messages are retained, which gives a record of the motivations and purpose behind each
-modification.
-
-* Version control also works extremely well for documentation.
-Books, educational resources and other data types are well served using version control.
-
-* Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
-from team members who may be working ascynchrously in different timezones and on different servers.
-It also makes the user's work searchable, which is convenient when code snippets from a project
-need to be reused for a new project.
-These centralised platforms also enable discussions of code "issues" such as bugs and feature requests.
-In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
-and allows team members to raise potential issues with a colleagues code (or one's own).
-
-* Version control makes works easily sharable.
-Work is easier to reproduce because the source folder contains a complete workflow with linked code
-that doesn't need any modifications before being executed to reproduce some works.
-
-* `git` is the defacto standard for version control currently although there are alternatives like
-`svn` (Subversion) and `hg` (Mercurial).
-Using `git` is advisable because there is a greater depth of support available on web forums, and more
-alternatives for centralised code hosting providers and graphical interfaces.
-
-## 2. Compute environment control
-
-* Conda
-
-* Containers: Docker/Guix/others
-
-* Virtual machines
-
-* Wasm
-
-## 3. Literate programming
-
-The report is truth.
-
-* Sweave, Rmarkdown, Quarto
-
-* Jupyter
-
-* Compilation
-
-* Posting reports to lab website and webarchive
-
-* Make the manuscript reproducible as well.
-
-## 4. Documentation
-
-* Instructions needed
-
-* Requirements
-
-* Credits
-
-## 5. FAIR and persistent data sharing
 

### 2023-05-10 docs/5pillars.Rmd
e2e code coverage

@@
 polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
-polygon(c(-0.2,5,10),c(6.5,8.5,6.5),col="lightgray", border = NA)
+polygon(c(-0.2,5,10),c(6.3,8.5,6.3),col="lightgray", border = NA)
 points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
@@
 text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
+text( 5 , 0.3 , lab="End-to-end code coverage")
 text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
@@
 
-## The five pillars: a code-first approach
+## End-to-end code coverage
 
-A common recommendation for reproducibility is to complete all works with computer scripts
-and remove the need for any manual steps.
-Computer scripts are preferred over graphical interface tools as the script contains
-a complete instruction set for how a piece of work was conducted.
-This allows auditing and reproduction which would be difficult for graphical tools like spreadsheets.
+End-to-end code coverage is a foundation for reproducibility as the codebase should contain all
+instructions for completing a data analysis task.
+This removes the need for any manual steps, which are not only time-consuming but also prone to errors.
+Scripts enable better auditing and easier reproduction which would be difficult for graphical tools
+like spreadsheets.
 Indeed spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
 somewhat overused in the life sciences.
-Scripts also allow the work to be completely automated and reduce the chance for human error during
-any manual steps.
+Transfering data between compute platforms is also discouraged.
+For example, while there are extensive bioinformatic tools on Galaxy, GenePattern and other web
+sources, the data needs to be transfered between platforms which is inefficient and could lead to
+errors.
 One thing to keep in mind is that, data cleaning and curation is often required, and in many cases
@@
 
-## 1. Code version control and persistent sharing
+## Code version control and persistent sharing
 
@@
 
-* Version control keeps a track of all these changes over time.
+* Keeps a complete history of all code changes over time.
 We always know which version of the script is the most up-to-date.
@@
 
-* Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
+* Helps collaboration and project management.
+Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
 from team members who may be working ascynchrously in different timezones and on different servers.
@@
 * Distributed version control systems protect against code loss.
-Computers sometimes fail, and if there are
+Computers sometimes fail and we sometimes delete or overwrite important files.
+If the code is unrecoverable it can be a significant loss of work.
+Although version control systems are not a back-up solution, they do add a layer of protection to the
+code.
+By preserving a copy of the repository on each team members computer in addition to the central
+repo, means the code can be easily recovered if any one of those computers fail.
+
+Need to mention long term (persistent) sharing.
 

### 2023-05-10 manuscript.Rmd
e2e code coverage

@@
 
-labs <- c("Version\ncontrol &\npersistent\ncode sharing",
+labs <- c("Code version\ncontrol &\npersistent\n sharing",
   "Compute \nenvironment \ncontrol",

### 2023-05-10 docs/5pillars.Rmd
apping content

@@
 [@Peng2011-ia; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
-@Brito2020-ld; @Akalin2018-eg ]
+@Brito2020-ld; @Akalin2018-eg ; @Rodrigues2023-fw ]
 and this article attempts to unify these concepts and assemble a group of best practices where
@@
 text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
-text( 5 , 0.3 , lab="End-to-end code coverage")
+text( 5 , 0.2 , lab="End-to-end code coverage")
 text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
@@
 
-End-to-end code coverage is a foundation for reproducibility as the codebase should contain all
+End-to-end code coverage is the foundation for reproducibility as the codebase should contain all
 instructions for completing a data analysis task.
 This removes the need for any manual steps, which are not only time-consuming but also prone to errors.
-Scripts enable better auditing and easier reproduction which would be difficult for graphical tools
-like spreadsheets.
+Without end-to-end code coverage, most of the other best practices described here are not possible.
+Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
+which would be difficult for graphical tools like spreadsheets.
 Indeed spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
 somewhat overused in the life sciences.
+
 Transfering data between compute platforms is also discouraged.
-For example, while there are extensive bioinformatic tools on Galaxy, GenePattern and other web
-sources, the data needs to be transfered between platforms which is inefficient and could lead to
-errors.
-One thing to keep in mind is that, data cleaning and curation is often required, and in many cases
-can only be conducted using manual intervention.
-In those cases, retaining raw and cleaned data and a cleaning protocol is recommended.
+For example, having workflows that involve combinations of web-based and scripted tools require
+data transfer steps which are inefficient and error-prone.
 
-Another principle of the five pillars approach is that the publishing researchers should put in the
-extra effort over and above the minimum standards of a classical journal to make the reproduction
-process easier and faster for those who wish to subsequently reproduce the analysis.
+Related to this, the code and data need to be "linked", which means the code is aware of the location
+of the data and how to get it.
+The code should be able to fetch the data from a publicly available location and conduct the downstream
+processing and analysis without having to aquire the data some other way.
+
+A caveat for end-to-end code coverage is that sometimes manual data cleaning is unavoidable.
+In those cases, retaining raw and cleaned data and providing a cleaning protocol is recommended.
+
+A guiding principle of the five pillars approach is that the publishing researchers should strive
+to make the reproduction process easier and faster for those who wish to.
 
@@
 
-* Literate programs accommodate extensive documentation.
+* It accommodates extensive documentation.
 This allows the analyst to include rich descriptions of scientific works.
@@
 
-* Literate scripts are amenable to version control.
+* Amenable to version control.
 Version control is a useful best practice in software development (discussed below).
 
-* Literate programming output reports are free from code errors.
+* Output reports are free from code errors.
 The resulting document is only rendered when the entire script has compiled without errors.
@@
 practices in software engineering.
-Distributed version control systems involve a central web-accessible server to hosting a repository,
-and each team member possesses a mirror copy on their local system.
+Distributed version control systems (DVCS) involve a central web-accessible server to hosting a
+repository, and each team member possesses a mirror copy on their local system.
 Having a central, publicly available node assists in disseminating changes within teams and releasing
 code to consumers.
-There are many such distributed version control systems but `git` has emerged as the most popular
-solution due to it's many powerful features, speed and large community and ecosystem.
+There are many such DVCSs but `git` has emerged as the most popular
+solution due to it's many powerful features, speed/efficiency and large community and ecosystem.
 
-Computational researchers can benefit significantly by adopting version control into their routine
+Computational researchers can benefit significantly by adopting DVCS into their routine
 practices.
@@
 We can inspect and execute the code and reproduce the analysis at any particular point in the past.
-Commit messages are retained, which gives a record of the motivations and purpose behind each
+Commit messages are retained, which gives a record of the motivations, purpose and person behind each
 modification.
@@
 
-* Version control also works extremely well for documentation.
-Books, educational resources and other data types are well served using version control (see below).
+* Supports documentation best practices.
+Books, educational resources and other data types are well served using DVCS (see below).
 
-* Version control makes works easily sharable.
+* Makes works easily sharable.
 Work is easier to reproduce because the source folder contains a complete workflow with linked code
@@
 
-* Distributed version control systems protect against code loss.
+* Protect against code loss.
 Computers sometimes fail and we sometimes delete or overwrite important files.
-If the code is unrecoverable it can be a significant loss of work.
-Although version control systems are not a back-up solution, they do add a layer of protection to the
+If the code is unrecoverable it can be a significant loss of work that is costly to replace.
+Although DVCSs are not a back-up solution, they do add a layer of protection to the
 code.
@@
 
-Need to mention long term (persistent) sharing.
+Although DVCSs assist with code sharing, they are not considered a long term solution and might not
+survive the "ten year challenge" of code reproducibility @Perkel2020-io.
+To achieve that, the code needs to be deposited to a long term
+repository.
+Zenodo and Figshare repositories are both good options for long term archiving of software, and the
+link to these should be provided in the respective journal article/pre-print.
+
+* Provide some links to practical guides.
 
@@
 
+We provide some practical resources for those interested in adopting various
+
+| Topic | Resource |
+| --- | --- |
+| General coding for science | eg |
+| Version control | eg |
+| Literate programming | eg |
+| Environment control | eg |
+| Documentation | eg |
+| Data sharing | eg |
+
 Extreme reproducibility should be the new standard.
@@
 
+More step-by-step guides required for frequently used routines.
+
+Adoption of these practices would require a change in incentives. https://www.nature.com/articles/s41562-016-0021
+
 ## Acknowledgements
@@
 
-* Sriganesh Srihari
+* Sriganesh Srihari @srigsri23
 
-* Barbara Zdrazil
+* Barbara Zdrazil @BZdrazil
 
-* Aziz Khan
+* Aziz Khan @khanaziz84
 
-* Pierre Poulain
+* Pierre Poulain @pierrepo
 
-* Irene López Santiago
+* Irene López Santiago @IreneLopSci
 

### 2023-05-11 docs/5pillars.Rmd
updates

@@
 achieve.
-Here, we present a framework to improve reproducibility, called the five pillars.
+Building upon past efforts and proposals to maximise reproducibility and rigour, we
+present a framework called the five pillars of reproducible computational research.
 These are a combination of best practices which ensure that the computational research work can be
@@
 [@Peng2011-ia; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
-@Brito2020-ld; @Akalin2018-eg ; @Rodrigues2023-fw ]
+@Brito2020-ld; @Rodrigues2023-fw; @Wilson2017-uq; @Akalin2018-eg ]
 and this article attempts to unify these concepts and assemble a group of best practices where
@@
 
-End-to-end code coverage is the foundation for reproducibility as the codebase should contain all
+End-to-end code coverage is the foundation for reproducibility as the codebase should ideally contain all
 instructions for completing a data analysis task.
@@
 of the data and how to get it.
-The code should be able to fetch the data from a publicly available location and conduct the downstream
+The code should be able to fetch the data from a publicly accessible location and conduct the downstream
 processing and analysis without having to aquire the data some other way.
@@
 
+Practical guides for jupyter and R Markdown - put these on webarchive and provide persistent links.
+
+* https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074
+
+* https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/
+
+* https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html
+
+* https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html
+
+* https://quarto.org/docs/computations/r.html
+
+* https://quarto.org/docs/computations/python.html
+
+* https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/
+
+* https://www.dataquest.io/blog/jupyter-notebook-tutorial/
+
+* https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af
+
 ## Code version control and persistent sharing
@@
 
-* Provide some links to practical guides.
+Practical guides for `git` and GitHub - put these on webarchive and provide persistent links.
 
-## 2. Compute environment control
+* https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/
 
-* Conda
+* https://www.freecodecamp.org/news/introduction-to-git-and-github/
+
+* https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/
+
+* https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/
+
+* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
+
+* https://blog.reviewnb.com/github-jupyter-notebook/
+
+* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
+
+## Compute environment control
+
+Most popular software undergoes regular updates to patch bugs and add new features.
+Bioinformatics software is no different, and it is well known that such changes have the
+potential to cause changes to results @Seyednasrollah2015-cc.
+This is why it is best practice to report the exact version of all programs used in an analysis
+(and packages therein) and even make archival copies for future reference @Sandve2013-lk.
+
+In R, such reporting is possible using the `sessionInfo()` command;
+while for python this is possible using the `session_info` package.
+Using literate programming ensures that the output document contains a record of this important
+information.
+
+Although regular software updates are overall a good thing, it poses a problem for future
+reproducibility.
+A researcher trying to reproduce a 10 year old study in R v3.0 could have a headache, as they would need
+to roll back their R version, and possibly their operating system as well, as R and other languages
+requires certain system libraries like BLAS and LAPACK for low-level routines.
+
+To avoid this, a virtual machine (VM) could be used to run a system-in-a-system.
+This means that the "host" machine can run another "guest" operating system with the right R version,
+without needing to change the host R version.
+While this provides good reproducibility, setting up an environment with a 10 year old operating system,
+R and packages would take a few hours to accomplish.
+Researchers could take a snapshot of their VM system and share it to help reproducibility and
+auditability, however the size of these images is relatively large due to the fact it contains the OS,
+software stack and project data.
+Moreover, the performance of computation in guest system is typically slower than when run direct on
+the host.
+
+Another potential solution to this is to use the Conda package environment management system that
+allows researchers to easily create, manage, and deploy software packages, dependencies, and
+environments across different computing platforms @Perkel2023-cv.
+It was initially developed to solve the problem of package dependency management in Python, but it now
+supports many other programming languages, including R, C/C++, Java, and more.
+Conda allows researchers to create isolated environments with specific versions of
+packages and dependencies, so users can have different versions of R or python in different environments
+on the same host system.
+That said, some have reported difficulties with Conda solving some environments.
 
@@
 
-## 4. Documentation
+## Documentation
 
@@
 
-## 5. FAIR and persistent data sharing
+## FAIR and persistent data sharing
 
@@
 The code should refer only to the cleaned data and the original files are there only for reference.
+
+Archive versions of critical reference datasets.
 

### 2023-05-15 docs/5pillars.Rmd
containers

@@
 five pillars of reproducible computational research (**Figure 1**).
-These concepts are unpacked in the following sections.
+
+The intended audience is bioinformatics data analysts (not tool developers), however the
+principles described here should translate to other domains of study.
 
@@
 A caveat for end-to-end code coverage is that sometimes manual data cleaning is unavoidable.
-In those cases, retaining raw and cleaned data and providing a cleaning protocol is recommended.
+In those cases, retaining raw and cleaned data and providing a cleaning protocol is recommended,
+then the computational workflow can begin with the cleaned dataset.
 
@@
 to make the reproduction process easier and faster for those who wish to.
+
+Getting started with R and python:
+
+* Intro to R and Rstudio
+
+* Intro to python and ipython IDE
+
+* Intro to vscode
 
@@
 
-Another potential solution to this is to use the Conda package environment management system that
-allows researchers to easily create, manage, and deploy software packages, dependencies, and
-environments across different computing platforms @Perkel2023-cv.
-It was initially developed to solve the problem of package dependency management in Python, but it now
-supports many other programming languages, including R, C/C++, Java, and more.
+Another potential solution to this is to use a package environment management system
+(eg: Conda or Guix).
+These allow users to easily create, manage, and deploy software packages,
+dependencies, and environments across different computing platforms @Perkel2023-cv.
+Conda was initially developed to solve the problem of package dependency management in Python,
+but it now supports many other programming languages, including R, C/C++, Java, and more.
 Conda allows researchers to create isolated environments with specific versions of
-packages and dependencies, so users can have different versions of R or python in different environments
+packages, so users can have different versions of R or python in different environments
 on the same host system.
-That said, some have reported difficulties with Conda solving some environments.
+Conda however does not typically disseminate the research code, as it is up to the reproducer to
+set that up inside the environment.
+Moreover Conda environments have limitations around portability and consistency, as Conda environments
+are tied to specific operating systems and hardware architecture.
+Guix on the other hand, gives system-level reproducibility, but has a smaller community and ecosystem
+of supported packages.
 
-* Containers: Docker/Guix/others
+Containers are an attempt to resolve some of the downsides of virtual machines and package managers.
+Container images can be thought of as similar to virtual machines, but are more lightweight as they
+share parts of the host operating system.
+This means that execution in a container entails only a small overhead as compared to virtual machines
+where the overhead is much larger, meaning that performance in containers is similar to the host system.
+Containers provide a consistent runtime environment, ensuring that applications runs the same way
+across different types of systems and environments.
+This is because containers include everything needed to run the application, including the system tools
+and libraries, ensuring that the environment is consistent across different systems.
+For example, the most popular containerisation system, Docker, has cross platform support, so it is
+possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
+promise of reproducibility.
+While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
+containerisation system for web development and research.
+Due to its early adoption, Docker now has a large community of users, extensive documentation and
+a vast collection of pre-built container images called DockerHub.
+Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
+and typically within a few minutes.
+This accellerates the installation procedure dramatically which is a known bottleneck for bioinformatic
+reproducibility.
 
-* Virtual machines
-
-* Wasm
+We are also watching with interest the early development of WebAssembly (Wasm) based approaches for
+compute environment control.
+The ability to reproduce operating systems, programming languages and workflows in the browser
+opens up intriguing possibilities for more widespread reproduction and auditing without the need to
+install any software at all.
+For example, it is now possible embed a Wasm-enabled R terminal inside a Quarto HTML document, which
+enables in-browser reproducibility of R workflows @Balamuta2023-ma.
 
@@
 ## Bringing it all together
+
+Computational reproducibility of
+Jupyter notebooks from biomedical
+publications
+https://arxiv.org/pdf/2209.04308.pdf
+
+AlNoamany, Yasmin, and John A. Borghi. 2018. “Towards Computational Reproducibility: Researcher Perspectives on the Use and Sharing of
+Software.” Article. PEERJ COMPUTER SCIENCE, September. https://doi.org/ghkb9j (https://doi.org/ghkb9j).
+
+Archmiller, Althea A., Andrew D. Johnson, Jane Nolan, Margaret Edwards, Lisa H. Elliott, Jake M. Ferguson, Fabiola Iannarilli, et al. 2020.
+“Computational Reproducibility in the Wildlife Society’s Flagship Journals.” Article. JOURNAL OF WILDLIFE MANAGEMENT 84 (5): 1012–17.
+https://doi.org/gg66q7 (https://doi.org/gg66q7).
+
+Beaulieu-Jones, Brett K, and Casey S Greene. 2017. “Reproducibility of Computational Workflows Is Automated Using Continuous Analysis.”
+Nature Biotechnology, no. 4: 342.
+
+Catlin, Ann Christine, Chandima HewaNadungodage, and Andres Bejarano. 2019. “Lifecycle Support for Scientific Investigations: Integrating
+Data, Computing, and Workflows.” Article. COMPUTING IN SCIENCE & ENGINEERING 21 (4): 49–61. https://doi.org/gjt725 (https://doi.org
+/gjt725).
+
+Culina, Antica et al. 2018. “Navigating the Unfolding Open Data Landscape in Ecology and Evolution.” Nature Ecology & Evolution, 1–7.
+de Bayser, Maximilien, Leonardo G. Azevedo, and Renato Cerqueira. 2015. “ResearchOps: The Case for DevOps in Scientific Applications.” In
+2015 IFIP/IEEE International Symposium on Integrated Network Management (IM), 1398–1404. Ottawa, ON, Canada: IEEE. https://doi.org
+/ghkb9s (https://doi.org/ghkb9s).
+
+Díaz, Jessica, Jorge Pérez-Martínez, Agustin Yague, Andrea Villegas, and Antonio Antona. 2019. DevOps in Practice – A Preliminary Analysis of
+Two Multinational Companies.
+
+Hatton, L., and M. Van Genuchten. 2019. “Computational Reproducibility: The Elephant in the Room.” IEEE Software 36 (2): 137–44.
+https://doi.org/ggkvtr (https://doi.org/ggkvtr).
+
+Hocquet, Alexandre, and Frederic Wieber. 2021a. “Epistemic Issues in Computational Reproducibility: Software as the Elephant in the Room.”
+Article. EUROPEAN JOURNAL FOR PHILOSOPHY OF SCIENCE 11 (2). https://doi.org/gkm94m (https://doi.org/gkm94m).
+
+Hocquet, Alexandre, and Frédéric Wieber. 2021b. “Epistemic Issues in Computational Reproducibility: Software as the Elephant in the Room.”
+European Journal for Philosophy of Science 11 (2). https://doi.org/gkm94m (https://doi.org/gkm94m).
+
+Koehler Leman, Julia, Brian D. Weitzner, P. Douglas Renfrew, Steven M. Lewis, Rocco Moretti, Andrew M. Watkins, Vikram Khipple Mulligan, et al.
+2020. “Better Together: Elements of Successful Scientific Software Development in a Distributed Collaborative Community.” PLoS
+Computational Biology 16 (5): e1007507. https://doi.org/ggt62r (https://doi.org/ggt62r).
+
+Leipzig, Jeremy, Daniel Nüst, Charles Tapley Hoyt, Stian Soiland-Reyes, Karthik Ram, and Jane Greenberg. 2020. “The Role of Metadata in
+Reproducible Computational Research.”
+
+Maer-Matei, Monica Mihaela, Tiberiu Marian Georgescu, Cristina Mocanu, and Ana-Maria Zamfir. 2019/01/01////. “Skill Needs for Early Career
+Researchers.” Sustainability 11 (10). https://doi.org/ghkb9m (https://doi.org/ghkb9m).
+
+Murray, NJ et al. 2021. “Data Freshness in Ecology and Conservation.” Trends Ecol Evol 36 (6): 485–87.
+
+Obels, Pepijn, Daniël Lakens, Nicholas Coles, Jaroslav Gottfried, and Seth Green. 2020. “Analysis of Open Data and Computational
+
+Reproducibility in Registered Reports in Psychology.” Advances in Methods and Practices in Psychological Science 3 (May):
+251524592091887. https://doi.org/gg4vw4 (https://doi.org/gg4vw4).
+
+Peikert, Aaron, and Andreas Markus Brandmaier. 2019. “A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker.”
+
+PsyArXiv. https://doi.org/10.31234/osf.io/8xzqy (https://doi.org/10.31234/osf.io/8xzqy).
+@aaron_willcox ! " # || @Elliot_Gould_ ! " #
 

### 2023-05-15 docs/5pillars.Rmd
goodman

@@
 (such as data, code and documentation) to regenerate the outputs including figures and tables to
-confirm the study's findings.
+confirm the study's findings Goodman2016-wi.
 Reproducibility is only the first step towards overall reliability, for example a study may be
@@
 
+RESEARCHOPS: A. PRINCIPLED FRAMEWORK. AND GUIDE TO. COMPUTATIONAL. REPRODUCIBILITY.
+https://osf.io/ubj76/download
+
 AlNoamany, Yasmin, and John A. Borghi. 2018. “Towards Computational Reproducibility: Researcher Perspectives on the Use and Sharing of

### 2023-05-15 docs/5pillars.Rmd
progress

@@
 five pillars of reproducible computational research (**Figure 1**).
-
 The intended audience is bioinformatics data analysts (not tool developers), however the
@@
 to make the reproduction process easier and faster for those who wish to.
-
-Getting started with R and python:
-
-* Intro to R and Rstudio
-
-* Intro to python and ipython IDE
-
-* Intro to vscode
 
@@
 
-Practical guides for jupyter and R Markdown - put these on webarchive and provide persistent links.
-
-* https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074
-
-* https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/
-
-* https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html
-
-* https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html
-
-* https://quarto.org/docs/computations/r.html
-
-* https://quarto.org/docs/computations/python.html
-
-* https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/
-
-* https://www.dataquest.io/blog/jupyter-notebook-tutorial/
-
-* https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af
-
 ## Code version control and persistent sharing
@@
 link to these should be provided in the respective journal article/pre-print.
-
-Practical guides for `git` and GitHub - put these on webarchive and provide persistent links.
-
-* https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/
-
-* https://www.freecodecamp.org/news/introduction-to-git-and-github/
-
-* https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/
-
-* https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/
-
-* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
-
-* https://blog.reviewnb.com/github-jupyter-notebook/
-
-* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
 
@@
 
+Use webarchive to archive important resources.
+
 * NCBI
@@
 
+These innovations, when brought together facilitate extreme reproducibility, ie: makes reproducing a previous study relatively
+easy in terms of complexity, number of commands, and in time.
+For example, we made an example transcriptome analysis workflow based on the five pillars approach and it could could be installed
+and generate differential expression and pathway enrichment results in a matter of minutes.
+
+Extreme reproducibility should be the new standard.
+
+It generally doesn't matter what type of computer you're using.
+
+The benefit is that even if one of the pillars breaks, the overall study could be reproduced
+thanks to the remaining pillars.
+
+More systematic studies to assess reproducible practices are required.
+
+More step-by-step guides required for frequently used routines.
+
+Adoption of these practices would require a change in incentives. https://www.nature.com/articles/s41562-016-0021
+
+With the above innovations, it's possible to reproduce a computational study with just a few
+
+## References to read
+
 Computational reproducibility of
@@
 
-RESEARCHOPS: A. PRINCIPLED FRAMEWORK. AND GUIDE TO. COMPUTATIONAL. REPRODUCIBILITY.
+RESEARCHOPS: A PRINCIPLED FRAMEWORK AND GUIDE TO COMPUTATIONAL REPRODUCIBILITY.
 https://osf.io/ubj76/download
@@
 
+## Practical guides
+
 We provide some practical resources for those interested in adopting various
+
+Put these on webarchive and provide persistent links
 
@@
 
-Extreme reproducibility should be the new standard.
+### Getting started with data analysis in R and python
 
-It generally doesn't matter what type of computer you're using.
+* Intro to R and Rstudio
 
-The benefit is that even if one of the pillars breaks, the overall study could be reproduced
-thanks to the remaining pillars.
+* Intro to python and ipython IDE
 
-More systematic studies to assess reproducible practices are required.
+* Intro to vscode
 
-More step-by-step guides required for frequently used routines.
+### Literate programming with jupyter, R Markdown and Quarto
 
-Adoption of these practices would require a change in incentives. https://www.nature.com/articles/s41562-016-0021
+* https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074
+
+* https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/
+
+* https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html
+
+* https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html
+
+* https://quarto.org/docs/computations/r.html
+
+* https://quarto.org/docs/computations/python.html
+
+* https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/
+
+* https://www.dataquest.io/blog/jupyter-notebook-tutorial/
+
+* https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af
+
+### Practical guides for `git` and GitHub
+
+* https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/
+
+* https://www.freecodecamp.org/news/introduction-to-git-and-github/
+
+* https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/
+
+* https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/
+
+* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
+
+* https://blog.reviewnb.com/github-jupyter-notebook/
+
+* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
+
+### Practical guides for Conda, Guix and Docker
+
+### Practical guides for documenting computational research
+
+### Practical guides to data sharing
 

### 2023-05-16 docs/5pillars.Rmd
progress

@@
 Correspondance: m.ziemann@deakin.edu.au
+
+Source: https://github.com/markziemann/enrichment_recipe/blob/main/docs/5pillars.Rmd
 
@@
 Data and code availability, ability to install software easily, and other factors mean that most
-computational research works are unable to be reproduced.
+computational research works are unable to be reproduced @Samuel2022-ha.
 
@@
 text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)
+#mtext("Five pillars of reproducible computational research", cex=1.5)
 
@@
 text( 5 , 0.2 , lab="End-to-end code coverage")
+#text(5,7,lab="Code testing")
 text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
@@
 data transfer steps which are inefficient and error-prone.
-
 Related to this, the code and data need to be "linked", which means the code is aware of the location
@@
 Sweave, saving time.
+An example R Markdown script and output report is given in Figure 2.
+
+![Figure 2. An example R Markdown script (left) and output document (right).](../img/markdownchunk.png)
+
 Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
@@
 Literate programming also works neatly with the concept of the "executable paper", the idea that
-the data analysis underlying an entire study can be reproduced with one or a few commands @Lasser2020-mm.
+the data analysis underlying an entire study can be reproduced with one or a few commands
+[@Strijkers2011-py ; @Editorial2017-yt; @Lasser2020-mm].
 A typical genomics/bioinformatics study could involve one large literate script, or be broken down into
@@
 need to be reused for a new project.
-These centralised platforms also enable discussions of code "issues" such as bugs and feature requests.
+
+* Helps resolve issues.
+These centralised platforms also enable discussions of code issues such as bugs and feature requests.
 In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
 and allows team members to raise potential issues with a colleagues code (or one's own).
+This also allows the scientific community to raise an issue with the software maintainers.
+For example, this feature is used extensively during the peer review process for the Journal of
+Open-Source Software @Editorial2018-kt.
 
@@
 
-Although DVCSs assist with code sharing, they are not considered a long term solution and might not
-survive the "ten year challenge" of code reproducibility @Perkel2020-io.
-To achieve that, the code needs to be deposited to a long term
-repository.
+Although DVCSs assist with code sharing, they are not considered a long term archiving solution and
+might not survive the "ten year challenge" of code reproducibility @Perkel2020-io.
+To achieve that, the code needs to be deposited to a long term repository.
 Zenodo and Figshare repositories are both good options for long term archiving of software, and the
@@
 Conda was initially developed to solve the problem of package dependency management in Python,
-but it now supports many other programming languages, including R, C/C++, Java, and more.
+but it now supports many other programming languages, including R, C/C++, Java, and others.
 Conda allows researchers to create isolated environments with specific versions of
@@
 on the same host system.
-Conda however does not typically disseminate the research code, as it is up to the reproducer to
-set that up inside the environment.
-Moreover Conda environments have limitations around portability and consistency, as Conda environments
+Conda environments have limitations around portability and consistency, as such environments
 are tied to specific operating systems and hardware architecture.
@@
 share parts of the host operating system.
-This means that execution in a container entails only a small overhead as compared to virtual machines
-where the overhead is much larger, meaning that performance in containers is similar to the host system.
-Containers provide a consistent runtime environment, ensuring that applications runs the same way
-across different types of systems and environments.
-This is because containers include everything needed to run the application, including the system tools
-and libraries, ensuring that the environment is consistent across different systems.
-For example, the most popular containerisation system, Docker, has cross platform support, so it is
+In contrast to VMs, running containers incurs only a small reduction in performance as compared
+to running directly on the host system @Di_Tommaso2015.
+Container images are highly portable because containers include everything needed to run the
+application, including the system tools and libraries, ensuring that the environment is consistent
+across different systems.
+For example, the most popular containerisation system, Docker, makes it
 possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
@@
 enables in-browser reproducibility of R workflows @Balamuta2023-ma.
+
+It is not sufficient to simply use these tools, they need to be shared as part of the publication
+process.
+Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
+make the environment.
+As they are a recipe, these are small files which are easily shared in the project code repository.
+Docker images are larger, and cannot be shared in the code repository, so these need to be archived
+in a long term repository, such as Zenodo or FigShare.
 
@@
 
-These innovations, when brought together facilitate extreme reproducibility, ie: makes reproducing a previous study relatively
-easy in terms of complexity, number of commands, and in time.
-For example, we made an example transcriptome analysis workflow based on the five pillars approach and it could could be installed
-and generate differential expression and pathway enrichment results in a matter of minutes.
+When these features are brought together they facilitate extreme reproducibility, ie: makes reproducing a
+previous study relatively easy in terms of complexity, number of commands, and in time.
+For example, we made an example transcriptome analysis workflow based on the five pillars approach and
+it could could be installed and generate differential expression and pathway enrichment results in a
+matter of minutes @Ziemann_undated-zx.
+
+When not integrated correctly, problems readily emerge.
+A recent systematic examination of Jupyter notebooks in GitHub repositories accompanying research papers
+found that environment problems (eg: "module not found", "import error") and lack of data links
+(eg: "file not found" errors) accounted for 80% of code errors @Samuel2022-ha.
+
+This is why regular code testing is another recommended best practice.
+By regularly executing the workflow after changes are made ensures the absence of code errors.
+Executing literate scripts is the simplest approach, as if there is an error incurred the output report
+won't be generated, and the source of the error can be investigated.
+It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
+scientific community would fare reproducing such a workflow.
+
+Some researchers have suggested to take it even further, by adopting the software development
+principles of unit tests and continuous integration/continuous delivery into the research process.
+By doing this, the entire project workflow is automatically executed and checked whenever a change is
+made to the code base.
+If the workflow executes successfully and tests are passed, the repository can be decorated with a
+"badge" to inform users of the lack of errors in the code.
+If those tests are designed cleverly such that certain conclusions drawn in the corresponding research
+article can be assessed computationally, then such tests could be used to demonstrate computational
+reproducibility without the need for individuals to execute the workflow themselves.
 
 Extreme reproducibility should be the new standard.
-
-It generally doesn't matter what type of computer you're using.
-
-The benefit is that even if one of the pillars breaks, the overall study could be reproduced
-thanks to the remaining pillars.
 

### 2023-05-16 docs/5pillars.Rmd
progress

@@
 (such as data, code and documentation) to regenerate the outputs including figures and tables to
-confirm the study's findings Goodman2016-wi.
+confirm the study's findings @Goodman2016-wi.
 Reproducibility is only the first step towards overall reliability, for example a study may be
@@
 
-As computational researchers, we are theoretically able to achieve 100% analytical reproducibility,
+Researchers are theoretically able to achieve 100% computational reproducibility,
 however in practice this is rarely achieved.
@@
 @Leisch2002-yc.
-Around 2015, R Markdown emerged as a more popular alternative, as formatting is simpler as compared to
-Sweave, saving time.
+Around 2015, R Markdown emerged as a more popular alternative, as formatting content is simpler with 
+Markdown as compared to LaTeX, saving time.
 An example R Markdown script and output report is given in Figure 2.
@@
 
-Whether Rmarkdown, Jupyter or Quarto, literate programming offers some substantial benefits over
+Whether R Markdown, Jupyter or Quarto, literate programming offers some substantial benefits over
 alternative approaches:
@@
 be assembled "manually" with a word processor.
-Using a literate script would mean only minimal changes are required between interations.
+Using a literate script would mean only minimal changes are required between interations @Perkel2022-af.
 
@@
 By embedding the outputs (eg: charts and tables) in a document in the sequence that they were generated,
-it helps the reader understand the logical steps taken in any analysis.
+it helps the reader understand the logical steps taken in an analysis.
 
@@
 * Output reports are free from code errors.
-The resulting document is only rendered when the entire script has compiled without errors.
+The output document is only rendered when the entire script has compiled without errors.
 In contrast, a regular script might generate some outputs and then encounter an error,
 so we are not sure whether the whole script is free of errors and completed successfully.
+Therefore it is good practice to routinely execute these scripts during the development process,
+and not rely too heavily on the interactive execution of individual chunks or lines.
 
@@
 practices in software engineering.
-Distributed version control systems (DVCS) involve a central web-accessible server to hosting a
+Distributed version control systems (DVCS) involve a central web-accessible server hosting a
 repository, and each team member possesses a mirror copy on their local system.
@@
 * Makes works easily sharable.
-Work is easier to reproduce because the source folder contains a complete workflow with linked code
-that doesn't need any modifications before being executed to reproduce some works.
+Work is easier to reproduce because the source folder contains a complete workflow with code linked
+to data and metadata, so it doesn't need any modifications before reproduction.
 
@@
 Computers sometimes fail and we sometimes delete or overwrite important files.
-If the code is unrecoverable it can be a significant loss of work that is costly to replace.
+If code is lost and unrecoverable it can be a significant cost to replace.
 Although DVCSs are not a back-up solution, they do add a layer of protection to the
 code.
-By preserving a copy of the repository on each team members computer in addition to the central
-repo, means the code can be easily recovered if any one of those computers fail.
+By preserving a copy of the repository on each team members' computer in addition to the central
+repository, means the code can be easily recovered if any one of those computers fail.
 
@@
 Bioinformatics software is no different, and it is well known that such changes have the
-potential to cause changes to results @Seyednasrollah2015-cc.
+potential to affect results @Seyednasrollah2015-cc.
 This is why it is best practice to report the exact version of all programs used in an analysis
@@
 In R, such reporting is possible using the `sessionInfo()` command;
-while for python this is possible using the `session_info` package.
+while for Python this is possible using the `session_info` package.
 Using literate programming ensures that the output document contains a record of this important
@@
 to roll back their R version, and possibly their operating system as well, as R and other languages
-requires certain system libraries like BLAS and LAPACK for low-level routines.
+requires certain system libraries for low-level routines.
 
@@
 
-Another potential solution to this is to use a package environment management system
+Another potential solution to this is to use a package/environment management system
 (eg: Conda or Guix).
-These allow users to easily create, manage, and deploy software packages,
+These allow users to create, manage, and deploy software packages,
 dependencies, and environments across different computing platforms @Perkel2023-cv.
@@
 Conda allows researchers to create isolated environments with specific versions of
-packages, so users can have different versions of R or python in different environments
+packages, so users can have different versions of R or Python in different environments
 on the same host system.
@@
 In contrast to VMs, running containers incurs only a small reduction in performance as compared
-to running directly on the host system @Di_Tommaso2015.
+to running directly on the host system @Di_Tommaso2015-pw.
 Container images are highly portable because containers include everything needed to run the
@@
 This accellerates the installation procedure dramatically which is a known bottleneck for bioinformatic
-reproducibility.
+reproducibility @Mangul2019-ei.
 
@@
 
-Some researchers have suggested to take it even further, by adopting the software development
-principles of unit tests and continuous integration/continuous delivery into the research process.
-By doing this, the entire project workflow is automatically executed and checked whenever a change is
-made to the code base.
+Some researchers have suggested to take it even further, by adapting software development
+principles of unit tests and continuous integration/continuous delivery.
+This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
+made to the project's code base @Beaulieu-Jones2017-vj.
+
 If the workflow executes successfully and tests are passed, the repository can be decorated with a
@@
 
-Extreme reproducibility should be the new standard.
+## Where to from here
 
-More systematic studies to assess reproducible practices are required.
+The key technological advances enabling the five pillars approach have been around for about a decade,
+yet progress on integrating these into computational research publications has been minimal.
+Part of the problem is that funding bodies and journals are focused on supporting novel research
+directions, instead of validating previous findings.
+Indeed the number of systematic replication studies in bioinformatics remains very small
+(eg: [@Ioannidis2009-hv; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the problem
+actually is.
 
-More step-by-step guides required for frequently used routines.
+A part of the problem lies in the lack of technical computer skills in many research groups.
+Specific training in these areas could help, but the most scalable solution would be to
+provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
+For example, my group has just released such a protocol that walks novice users through the process
+of implementing functional enrichment analysis of gene expression data with the five pillars principles
+@Ziemann2023-nm.
+More such protocols are needed to make extreme reproducibility mainstream.
 
 Adoption of these practices would require a change in incentives. https://www.nature.com/articles/s41562-016-0021
-
-With the above innovations, it's possible to reproduce a computational study with just a few
 
@@
 
-### Getting started with data analysis in R and python
+### Getting started with data analysis in R and Python
 
@@
 
-* Intro to python and ipython IDE
+* Intro to Python and iPython IDE
 

### 2023-05-16 docs/5pillars.Rmd
added stuff

@@
 Markdown as compared to LaTeX, saving time.
-An example R Markdown script and output report is given in Figure 2.
+An example R Markdown script and output report is given in **Figure 2**.
 
-![Figure 2. An example R Markdown script (left) and output document (right).](../img/markdownchunk.png)
+![Figure 2. An example R Markdown script (left) and output document (right) @RStudio2016-fw.](../img/markdownchunk.png)
 
@@
 Distributed version control systems (DVCS) involve a central web-accessible server hosting a
-repository, and each team member possesses a mirror copy on their local system.
+repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
 Having a central, publicly available node assists in disseminating changes within teams and releasing
@@
 solution due to it's many powerful features, speed/efficiency and large community and ecosystem.
+
+```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}
+
+cols2 <- adjustcolor( cols, alpha.f = 0.4)
+
+plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
+polygon(c(-0.6,3,3,-0.6),c(0.8,0.8,5.2,5.2), col=cols2[1], border = NA)
+polygon(c(-0.2,2.8,2.8,-0.2),c(1,1,2,2), col=cols[1], border = NA)
+polygon(c(-0.2,2.8,2.8,-0.2),c(4,4,5,5), col=cols[1], border = NA)
+
+polygon(c(3.1,6.7,6.7,3.1),c(0.8,0.8,5.2,5.2), col=cols2[2], border = NA)
+polygon(c(3.3,6.5,6.5,3.3),c(1,1,2,2), col=cols[2], border = NA)
+polygon(c(3.3,6.5,6.5,3.3),c(4,4,5,5), col=cols[2], border = NA)
+
+polygon(c(6.8,10.2,10.2,6.8),c(0.8,0.8,5.2,5.2), col=cols2[3], border = NA)
+polygon(c(7,10,10,7),c(1,1,2,2), col=cols[3], border = NA)
+polygon(c(7,10,10,7),c(4,4,5,5), col=cols[3], border = NA)
+
+polygon(c(2.5,7.3,7.3,2.5),c(7,7,8.5,8.5), col=cols[4], border = NA)
+
+labs <- c("Developer 1",
+  "Local Repo",
+  "Working Files",
+  "Developer 2",
+  "Local Repo",
+  "Working Files",
+  "Developer 3",
+  "Local Repository",
+  "Working Files")
+
+text( c(1.2,1.2,1.2,4.9,4.9,4.9,8.5,8.5,8.5), c(0,1.5,4.5,0,1.5,4.5,0,1.5,4.5), labels=labs )
+text(4.9,7.7,labels="Main Repository",cex=1.5)
+
+arrows( 1, 2, 1, 4 ,lwd=3)
+arrows( 1.4, 4, 1.4, 2 ,lwd=3)
+arrows( 4.7, 2, 4.7, 4 ,lwd=3)
+arrows( 5.1, 4, 5.1, 2 ,lwd=3)
+arrows( 8.3, 2, 8.3, 4 ,lwd=3)
+arrows( 8.7, 4, 8.7, 2 ,lwd=3)
+arrows( 1, 5, 2.5, 7 ,lwd=3)
+arrows( 2.9, 7 , 1.4, 5 ,lwd=3)
+arrows( 4.7, 5, 4.7, 7 ,lwd=3)
+arrows( 5.1, 7 , 5.1, 5 ,lwd=3)
+arrows( 7.3, 7, 8.7, 5 ,lwd=3)
+arrows( 8.3, 5, 6.9, 7 ,lwd=3)
+
+text(c(0.7,4.4,8) ,c(3,3,3), labels="commit",srt=90)
+text(c(1.7,5.4,9) ,c(3,3,3), labels="update",srt=-90)
+text(c(1.5) ,c(6.2), labels="push",srt=50)
+text(c(2.4) ,c(5.9), labels="pull",srt=50)
+text(c(4.4) ,c(6), labels="push",srt=90)
+text(c(5.4) ,c(6), labels="pull",srt=-90)
+text(c(7.3) ,c(5.9), labels="push",srt=-60)
+text(c(8.2) ,c(6.2), labels="pull",srt=-60)
+
+```
 
@@
 Container images can be thought of as similar to virtual machines, but are more lightweight as they
-share parts of the host operating system.
+share parts of the host operating system (**Figure 4**).
 In contrast to VMs, running containers incurs only a small reduction in performance as compared
@@
 This accellerates the installation procedure dramatically which is a known bottleneck for bioinformatic
-reproducibility @Mangul2019-ei.
+reproducibility [@Mangul2019-ei;@Mangul2019-xp].
+
+![Figure 4. Architecture of containers (left) compared to virtual machines (right)](../img/docker-containerized-and-vm-transparent-bg.webp)
 
@@
 reproducibility without the need for individuals to execute the workflow themselves.
+Therefore continuous analysis might be a future pediment (crowning feature) to the five-pillared
+"temple" of reproducibl research.
 
@@
 ## References to read
+
+https://www.minot.bio/
+https://us02web.zoom.us/rec/play/zZsJzFME-9Uu5S5BKwX-9DRXQfHq1wnZjH4Mrgy2v1l3e2iE9BkA_Bwj8ZDcoz19BADqmZ6KU3hLA-oP.RKy6CJCCZD_iRjV2?canPlayFromShare=true&componentName=rec-play&continueMode=true&from=share_recording_detail&originRequestUrl=https%3A%2F%2Fus02web.zoom.us%2Frec%2Fshare%2FRbeIXIFsAd_oFNttD7zBeC-W8vswX-rq3r4XoUmlp2V2RFJLTd_wut9qhfuzrESF.wM6AyVmmcQks_cdy
 

### 2023-05-16 docs/5pillars.Rmd
minor works

@@
 then the computational workflow can begin with the cleaned dataset.
+
+Another challenge is the fact that different parts of the workflow have different compute needs.
+For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
+reference genome and variant calling which are demanding in memory and compute.
+These computationally demanding steps are typically conducted on high end servers and workstations.
+After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
+so downstream analysis could be conducted on commodity PCs or laptops.
+In those situations,
+
+Later stage
 
@@
 This also allows the scientific community to raise an issue with the software maintainers.
-For example, this feature is used extensively during the peer review process for the Journal of
-Open-Source Software @Editorial2018-kt.
+For example, this feature is used extensively during the peer review process for the *Journal of
+Open-Source Software* @Editorial2018-kt.
 
@@
 Container images can be thought of as similar to virtual machines, but are more lightweight as they
-share parts of the host operating system (**Figure 4**).
-In contrast to VMs, running containers incurs only a small reduction in performance as compared
+share parts of the host operating system.
+In the example shown in **Figure 4**, the six containerised applications share the same operating
+system, while the three VM applications each involve their own operating system which incurs a
+significant performance overhead @noauthor_2021-lt.
+Therefore, running workflows in containers incurs only a small reduction in performance as compared
 to running directly on the host system @Di_Tommaso2015-pw.
@@
 
-![Figure 4. Architecture of containers (left) compared to virtual machines (right)](../img/docker-containerized-and-vm-transparent-bg.webp)
+![Figure 4. Architecture of containers (left) compared to virtual machines (right) @noauthor_2021-lt](../img/docker-containerized-and-vm-transparent-bg.webp)
 
@@
 As they are a recipe, these are small files which are easily shared in the project code repository.
-Docker images are larger, and cannot be shared in the code repository, so these need to be archived
-in a long term repository, such as Zenodo or FigShare.
+Docker images are larger, and cannot be shared in the code repository.
+They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
+commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
+A better option for bioinformatics workflows would be to use the BioContainers registry
+@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
+*Global Alliance for Genomics and Health* that will underpin its longevity.
+Still, these are not considered permanent archival services, so it is advised to deposit the image
+used in anarticle to a suitable persitent long-term archive (discussed below).
+The recipe used to build the Docker image, called the Dockerfile, is sufficiently small and can be added
+to the project repository.
 
@@
 
-When these features are brought together they facilitate extreme reproducibility, ie: makes reproducing a
-previous study relatively easy in terms of complexity, number of commands, and in time.
-For example, we made an example transcriptome analysis workflow based on the five pillars approach and
+This work is not designed to give strong recommendations on the exact combination of tools or resources
+to use in computational research, but to set out key principles and highlight some useful solutions
+that users can incorporate into their toolkit to improve the overall reproducibility of their work.
+
+When these five considerations are addressed, they facilitate extreme reproducibility, ie: makes
+reproduction relatively quick, and simple.
+For example, we made a transcriptome analysis workflow based on the five pillars approach and
 it could could be installed and generate differential expression and pathway enrichment results in a
-matter of minutes @Ziemann_undated-zx.
+matter of minutes with only a few commands @Ziemann_undated-zx.
 
@@
 This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
-made to the project's code base @Beaulieu-Jones2017-vj.
+made to a project's code base @Beaulieu-Jones2017-vj.
 
@@
 Therefore continuous analysis might be a future pediment (crowning feature) to the five-pillared
-"temple" of reproducibl research.
+"temple" of reproducible research.
 
@@
 
-Adoption of these practices would require a change in incentives. https://www.nature.com/articles/s41562-016-0021
+There are other challenges worth mentioning.
+Different parts of the workflow may have different compute needs.
+For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
+reference genome and variant calling which are demanding in memory and compute.
+These computationally demanding steps are typically conducted on high end servers and workstations.
+After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
+so downstream analysis could be conducted on commodity PCs or laptops.
+In those situations, a workflow manager such as Snakemake or Nextflow could be useful to
+coordinate those hybrid computate workflows.
+
+Adoption of these practices would require a change in incentives.
+https://www.nature.com/articles/s41562-016-0021
 

### 2023-05-18 docs/5pillars.Rmd
data sharing

@@
 then the computational workflow can begin with the cleaned dataset.
-
-Another challenge is the fact that different parts of the workflow have different compute needs.
-For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
-reference genome and variant calling which are demanding in memory and compute.
-These computationally demanding steps are typically conducted on high end servers and workstations.
-After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
-so downstream analysis could be conducted on commodity PCs or laptops.
-In those situations,
-
-Later stage
 
@@
 Still, these are not considered permanent archival services, so it is advised to deposit the image
-used in anarticle to a suitable persitent long-term archive (discussed below).
+used in an article to a suitable persitent long-term archive (discussed below).
 The recipe used to build the Docker image, called the Dockerfile, is sufficiently small and can be added
@@
 
-## Documentation
+## FAIR and persistent data sharing
 
-* Instructions needed
+FAIR data sharing is one of the key features of open science, which characterises itself as science
+that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].
 
-* Requirements
+The Universal Declaration of Human Rights states [@United_Nations1948-ua]:
 
-* Credits
+>Everyone has the right freely to participate in the cultural life of the
+>community, to enjoy the arts and to share in scientific advancement and
+>its benefits.
 
-## FAIR and persistent data sharing
+In this context, access to research data may be interpreted as a benefit of scientific advancement.
+In particular, when large data sets are generated with public funds, the public might expect that the
+benefts of that study, like the publications and underlying data are released.
+
+Data sharing has dual purposes, firstly this data may be useful in answering other research questions,
+and secondly to facilitate reproduction and auditing [@Wilkinson2016-qq].
+Data reuse prevents redundant research and facilitates new research ideas that were impossible without
+it.
+In terms of research rigour, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
+in addition to validation of scientific findings by independent computational reproduction [@Ioannidis2009-hv].
+In light of this, a research article without supporting data and code is much like a press release or
+advertisement and the claims made cannot be verified as true or not.
+
+Although it is common to see journal articles with "Data available upon reasonable request", this is less
+than ideal.
+I, and many fellow researchers have had the displeasure of attempting to contact corresponding
+authors of these studies and learn that they have moved on to another institute, retired, passed away,
+or the data has been lost or otherwise unavailable (if they receive a response at all).
+In the rare cases that the data has been retained, arranging access can be tedious as such datasets can
+run into hundreds of gigabytes which cannot be easily transferred across the internet.
+If many requests are made for a dataset, this can become time intensive for the sharers which is
+wasteful considering they could have shared the data from the start and reached a larger audience.
+
+FAIR principles: @Wilkinson2016-qq
+
+Long term data archiving (and sharing) not only faciliitates this,
+
+Using secondary data is
+
+Moreover there is
+
+Data available on request?
+
+NIH data sharing requirements.
 
@@
 * Portico
+
+## Documentation
+
+* Instructions needed
+
+* Requirements
+
+* Credits
 
@@
 https://www.minot.bio/
-https://us02web.zoom.us/rec/play/zZsJzFME-9Uu5S5BKwX-9DRXQfHq1wnZjH4Mrgy2v1l3e2iE9BkA_Bwj8ZDcoz19BADqmZ6KU3hLA-oP.RKy6CJCCZD_iRjV2?canPlayFromShare=true&componentName=rec-play&continueMode=true&from=share_recording_detail&originRequestUrl=https%3A%2F%2Fus02web.zoom.us%2Frec%2Fshare%2FRbeIXIFsAd_oFNttD7zBeC-W8vswX-rq3r4XoUmlp2V2RFJLTd_wut9qhfuzrESF.wM6AyVmmcQks_cdy
 

### 2023-05-19 docs/5pillars.Rmd
data sharing

--- 2023-05-16 docs/5pillars.Rmd
@@
 
-Another challenge is the fact that different parts of the workflow have different compute needs.
-For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
-reference genome and variant calling which are demanding in memory and compute.
-These computationally demanding steps are typically conducted on high end servers and workstations.
-After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
-so downstream analysis could be conducted on commodity PCs or laptops.
-In those situations,
-
-Later stage
-
 A guiding principle of the five pillars approach is that the publishing researchers should strive
@@
 
-```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}
+```{r,dvcs,fig.cap="Figure 3. Distributed version control allows. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}
 
@@
 Still, these are not considered permanent archival services, so it is advised to deposit the image
-used in anarticle to a suitable persitent long-term archive (discussed below).
-The recipe used to build the Docker image, called the Dockerfile, is sufficiently small and can be added
-to the project repository.
-
-## Documentation
-
-* Instructions needed
-
-* Requirements
-
-* Credits
+used in an article to a suitable persitent long-term archive (discussed below).
+The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
+added to the project source code repository.
 
 ## FAIR and persistent data sharing
+
+Without data sharing, computational research is not reproducible nor auditable.
+Data sharing is also one of the key features of "open science", which is characterised as science
+that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].
+
+The Universal Declaration of Human Rights states [@United_Nations1948-ua]:
+
+>Everyone has the right freely to participate in the cultural life of the
+>community, to enjoy the arts and to share in scientific advancement and
+>its benefits.
+
+In this context, access to research data may be interpreted as a benefit of scientific advancement.
+In particular, when large research projects are funded by the public, the public might expect that the
+benefts of that study, like the publications and underlying data are made publicly accessible.
+
+Data sharing has dual purposes, firstly to facilitate reproduction and auditing, and secondly to
+enable reuse in other contexts [@Wilkinson2016-qq].
+Data reuse increases efficiency, as it prevents redundant research and facilitates new research ideas
+that were impossible without it.
+In terms of research rigour, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
+in addition to validating scientific findings by independent computational reproduction [@Ioannidis2009-hv].
+In light of this, a research article without supporting data and code is much like a press release or
+advertisement where the claims made cannot be verified as true or not.
+
+Although it is common to see journal articles with "Data available upon reasonable request", this is less
+than ideal.
+A systematic study of such data availability statements found that the data was successfully provided
+in just 6.8% of requests [@Gabelica2022-ci].
+In the rare cases that data is shared in accordance with the data availability statement, it poses
+a burden in terms of labour to lodge and respond to such requests, and if the data is large, may incur
+an additional cost for hosting/transferring these large files.
+
+The field of genomics has has a long history of data sharing @Brazma2001-yk, which is supported by
+policy leadership by funding bodies
+[@National_Health_and_Medical_Research_Council2019-yx;
+@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
+(eg [@Packer2018-ru; @Hanson2011-ub]), although there are some challenges.
+
+Some researchers would rather keep data for their private use.
+Some may keep data private due to concerns over participant privacy.
+This is a well-founded concern for identifiable data, however once identifying features are removed,
+it is nearly impossible to reidentify genomic data @Venkatesaramani2021-tn unless the attackers have
+preexisting genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
+Therefore sharing deidentified genomic data and metadata is likely to carry little risk to participant
+privacy.
+Still, much human genomic data and metadata resides in restricted databases due to the scope of
+data sharing that was agreed to in the consent process.
+For example the consent form may only allow sharing of data to researchers investigating certain
+diseases, which means a vetting process must be undertaken before sharing data with external
+investigators.
+In the interests of reuse and reproducibility, future studies could consider a consent form that allows
+more broad sharing of deidentified data.
+
+In order to maximise the value of shared data for reuse and reproducibility, it needs to be Findable,
+Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
+High throughput biology has already built a number of data repositories for specific data types
+conforming to FAIR principles (eg: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
+However other domains lack well known specialised organised data repositories, such as evolutionary
+biology, ecology, cell biology, social sciences and computing.
+The re3data.org registry of data repositories may be useful to find a repository that accepts data
+from these fields [@Pampel2013-ep].
+Moreover general recommendations have been proposed to enhance data FAIRness in ecology, but which
+can also be applied to other disciplines [@Roche2015-aq].
+
+Key recommendations include:
+
+1. Depositing data to a specialised repository if possible, otherwise a general purpose repository.
+
+2. If possible, avoid large supplementary data files accompanying journal articles.
+
+2. Provide detailed metadata.
+
+3. Preferably archive and share raw data, and use existing standards for the discipline.
+
+Once the quality of the data set has been established researchers may consider depositing it
+to a data repository early, before publicisation, as it has some advantages.
+Most data repositories have an optional 12-month embargo period so researchers can share publicly
+at a later date once they're happy with the overall project.
+The data repository acts as an additional backup to the host institutions' own in case of data loss
+or calamity.
+
+By assuming the data and metadata can be sourced from a public location, the workflow developers ensure
+that the code they share won't suffer from "file not found errors" during reproduction.
+This ensures code and data are linked, which has been a  [@Peng2011-ia]
 
@@
 
+* Dryad
+
+* GigaDB
+
 * Figshare

### 2023-05-22 docs/5pillars.Rmd
development

@@
 ---
-title: "The five pillars of computational reproducibility"
+title: "The five pillars of computational reproducibility: A bioinformatics perspective"
 author: "Mark Ziemann"
@@
 [@Peng2011-ia; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
-@Brito2020-ld; @Rodrigues2023-fw; @Wilson2017-uq; @Akalin2018-eg ]
-and this article attempts to unify these concepts and assemble a group of best practices where
-computational reproducibility can be considered as complete, using a framework called the
+@Brito2020-ld; @Rodrigues2023-fw; @Wilson2017-uq; @Akalin2018-eg ; @Peikert2021-eq ].
+This article attempts to unify these concepts into five key principles called the
 five pillars of reproducible computational research (**Figure 1**).
@@
 
-```{r,dvcs,fig.cap="Figure 3. Distributed version control allows. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}
+```{r,dvcs,fig.cap="Figure 3. Distributed version control. Adapted from @Chacon2014-ys.",echo=FALSE,fig.height=6}
 
@@
 Without data sharing, computational research is not reproducible nor auditable.
+Lack of data sharing is one of the key reasons why research is irreproducible
+[@Ioannidis2009-hv, @Archmiller2020-er ].
 Data sharing is also one of the key features of "open science", which is characterised as science
 that is collaborative, transparent, accessible and inclusive [@noauthor_2023-uq].
-
-The Universal Declaration of Human Rights states [@United_Nations1948-ua]:
-
->Everyone has the right freely to participate in the cultural life of the
->community, to enjoy the arts and to share in scientific advancement and
->its benefits.
-
-In this context, access to research data may be interpreted as a benefit of scientific advancement.
-In particular, when large research projects are funded by the public, the public might expect that the
-benefts of that study, like the publications and underlying data are made publicly accessible.
-
-Data sharing has dual purposes, firstly to facilitate reproduction and auditing, and secondly to
-enable reuse in other contexts [@Wilkinson2016-qq].
-Data reuse increases efficiency, as it prevents redundant research and facilitates new research ideas
-that were impossible without it.
+Indeed, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
+In addition to facilitating reproduction and auditing, sharing enables reuse in other contexts.
+Data reuse increases efficiency, as it prevents redundant research expenditure and facilitates new
+research ideas that were previously impossible [@Wilkinson2016-qq].
 In terms of research rigour, inspecting raw data can uncover research integrity problems [@Miyakawa2020-yj],
-in addition to validating scientific findings by independent computational reproduction [@Ioannidis2009-hv].
 In light of this, a research article without supporting data and code is much like a press release or
@@
 Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
+
+![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb](../img/gr2)
+
 High throughput biology has already built a number of data repositories for specific data types
@@
 biology, ecology, cell biology, social sciences and computing.
-The re3data.org registry of data repositories may be useful to find a repository that accepts data
+The **re3data.org** registry of data repositories may be useful to find a repository that accepts data
 from these fields [@Pampel2013-ep].
-Moreover general recommendations have been proposed to enhance data FAIRness in ecology, but which
+If no specialised repository exists, then a general purpose repository such as
+[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
+should be considered.
+Moreover, general recommendations have been proposed to enhance data FAIRness in ecology, but which
 can also be applied to other disciplines [@Roche2015-aq].
@@
 
-2. If possible, avoid large supplementary data files accompanying journal articles.
+2. Avoid commodity cloud storage as these are impermanent.
 
-2. Provide detailed metadata.
+3. Avoid large supplementary data files accompanying journal articles as these are less findable and
+accessible.
 
-3. Preferably archive and share raw data, and use existing standards for the discipline.
+4. Provide detailed metadata.
+
+5. Preferably archive and share raw data, and use existing standards for the discipline.
 
@@
 or calamity.
-
 By assuming the data and metadata can be sourced from a public location, the workflow developers ensure
 that the code they share won't suffer from "file not found errors" during reproduction.
-This ensures code and data are linked, which has been a  [@Peng2011-ia]
+This ensures code and data are linked, which has been a priority for reproducibility [@Peng2011-ia],
+and a source of problems for shared code [@Samuel2022-ha].
 
-* When designed correctly, the code "knows" where to access the associated metadata and data sets are,
-which avoids "file not found" errors.
-Metadata are usually small, so they can also be included in the repository as flat text files.
-Small data files can also be included in the repository, although keep in mind that big data files
-(>10MB) such as raw array scans, fastq files and the like are unsuitable for retention in a version
-control repository as the large files will slow down code updates signifiantly.
-Larger data files belong in a dedicated data repository, discussed below.
-If the metadata and small data require some cleaning, then include the original and cleaned data in the
-repository.
-The code should refer only to the cleaned data and the original files are there only for reference.
-
-Archive versions of critical reference datasets.
-
-Use webarchive to archive important resources.
-
-* NCBI
-
-* Zenodo
-
-* Dryad
-
-* GigaDB
-
-* Figshare
-
-* Arweave/IPFS
-
-* Portico
+In addition to the experimental data, reference data that is critical to a project should be archived.
+For example, in the field of genomics, gene function annotation sets are relied upon for a common
+proceduce called "pathway analysis".
+These annotation sets are regularly updated, but versions are rarely reported [@Wijesooriya2022-ly]
+and finding historical data versions is sometimes difficult.
+So in order to guarantee future reproducibility, shapshots of those reference data should be archived &
+shared if the licence permits it.
 
@@
 
-This work is not designed to give strong recommendations on the exact combination of tools or resources
+Due to the rapidly devloping software landscape, this work is not designed to give strong
+recommendations on the exact combination of tools or resources
 to use in computational research, but to set out key principles and highlight some useful solutions
 that users can incorporate into their toolkit to improve the overall reproducibility of their work.
-
-When these five considerations are addressed, they facilitate extreme reproducibility, ie: makes
-reproduction relatively quick, and simple.
-For example, we made a transcriptome analysis workflow based on the five pillars approach and
-it could could be installed and generate differential expression and pathway enrichment results in a
-matter of minutes with only a few commands @Ziemann_undated-zx.
 
@@
 
-Computational reproducibility of
-Jupyter notebooks from biomedical
-publications
-https://arxiv.org/pdf/2209.04308.pdf
+https://www.researchgate.net/publication/349603799_Making_radiotherapy_more_efficient_with_FAIR_data/figures?lo=1
 
-RESEARCHOPS: A PRINCIPLED FRAMEWORK AND GUIDE TO COMPUTATIONAL REPRODUCIBILITY.
-https://osf.io/ubj76/download
-
-AlNoamany, Yasmin, and John A. Borghi. 2018. “Towards Computational Reproducibility: Researcher Perspectives on the Use and Sharing of
-Software.” Article. PEERJ COMPUTER SCIENCE, September. https://doi.org/ghkb9j (https://doi.org/ghkb9j).
-
-Archmiller, Althea A., Andrew D. Johnson, Jane Nolan, Margaret Edwards, Lisa H. Elliott, Jake M. Ferguson, Fabiola Iannarilli, et al. 2020.
-“Computational Reproducibility in the Wildlife Society’s Flagship Journals.” Article. JOURNAL OF WILDLIFE MANAGEMENT 84 (5): 1012–17.
-https://doi.org/gg66q7 (https://doi.org/gg66q7).
-
-Beaulieu-Jones, Brett K, and Casey S Greene. 2017. “Reproducibility of Computational Workflows Is Automated Using Continuous Analysis.”
-Nature Biotechnology, no. 4: 342.
-
-Catlin, Ann Christine, Chandima HewaNadungodage, and Andres Bejarano. 2019. “Lifecycle Support for Scientific Investigations: Integrating
-Data, Computing, and Workflows.” Article. COMPUTING IN SCIENCE & ENGINEERING 21 (4): 49–61. https://doi.org/gjt725 (https://doi.org
-/gjt725).
-
-Culina, Antica et al. 2018. “Navigating the Unfolding Open Data Landscape in Ecology and Evolution.” Nature Ecology & Evolution, 1–7.
-de Bayser, Maximilien, Leonardo G. Azevedo, and Renato Cerqueira. 2015. “ResearchOps: The Case for DevOps in Scientific Applications.” In
-2015 IFIP/IEEE International Symposium on Integrated Network Management (IM), 1398–1404. Ottawa, ON, Canada: IEEE. https://doi.org
-/ghkb9s (https://doi.org/ghkb9s).
-
-Díaz, Jessica, Jorge Pérez-Martínez, Agustin Yague, Andrea Villegas, and Antonio Antona. 2019. DevOps in Practice – A Preliminary Analysis of
-Two Multinational Companies.
-
-Hatton, L., and M. Van Genuchten. 2019. “Computational Reproducibility: The Elephant in the Room.” IEEE Software 36 (2): 137–44.
-https://doi.org/ggkvtr (https://doi.org/ggkvtr).
-
-Hocquet, Alexandre, and Frederic Wieber. 2021a. “Epistemic Issues in Computational Reproducibility: Software as the Elephant in the Room.”
-Article. EUROPEAN JOURNAL FOR PHILOSOPHY OF SCIENCE 11 (2). https://doi.org/gkm94m (https://doi.org/gkm94m).
-
-Hocquet, Alexandre, and Frédéric Wieber. 2021b. “Epistemic Issues in Computational Reproducibility: Software as the Elephant in the Room.”
-European Journal for Philosophy of Science 11 (2). https://doi.org/gkm94m (https://doi.org/gkm94m).
-
-Koehler Leman, Julia, Brian D. Weitzner, P. Douglas Renfrew, Steven M. Lewis, Rocco Moretti, Andrew M. Watkins, Vikram Khipple Mulligan, et al.
-2020. “Better Together: Elements of Successful Scientific Software Development in a Distributed Collaborative Community.” PLoS
-Computational Biology 16 (5): e1007507. https://doi.org/ggt62r (https://doi.org/ggt62r).
-
-Leipzig, Jeremy, Daniel Nüst, Charles Tapley Hoyt, Stian Soiland-Reyes, Karthik Ram, and Jane Greenberg. 2020. “The Role of Metadata in
-Reproducible Computational Research.”
-
-Maer-Matei, Monica Mihaela, Tiberiu Marian Georgescu, Cristina Mocanu, and Ana-Maria Zamfir. 2019/01/01////. “Skill Needs for Early Career
-Researchers.” Sustainability 11 (10). https://doi.org/ghkb9m (https://doi.org/ghkb9m).
+Maer-Matei, Monica Mihaela, Tiberiu Marian Georgescu, Cristina Mocanu, and Ana-Maria Zamfir. 2019/01/01////.
+“Skill Needs for Early Career Researchers.” Sustainability 11 (10). https://doi.org/ghkb9m (https://doi.org/ghkb9m).
 

### 2023-05-22 docs/5pillars.Rmd
fixing typos

@@
 ---
-title: "The five pillars of computational reproducibility: A bioinformatics perspective"
+title: "The five pillars of computational reproducibility: Bioinformatics and beyond"
 author: "Mark Ziemann"
@@
 
+ORCID: 0000-0002-7688-6974
+
 Correspondance: m.ziemann@deakin.edu.au
@@
 achieve.
-Building upon past efforts and proposals to maximise reproducibility and rigour, we
-present a framework called the five pillars of reproducible computational research.
-These are a combination of best practices which ensure that the computational research work can be
-reproduced quickly and easily, and will endure relatively long time-frames.
+Building upon past efforts and proposals to maximise reproducibility and rigour in
+bioinformatics, we present a framework called the five pillars of reproducible computational
+research.
+These include (1) code version control and sharing, (2) compute environment control,
+(3) literate programming, (4) documentation and (5) FAIR data sharing.
+These best practices will ensure that computational research work can be
+reproduced quickly and easily, long into the future.
+This guide is designed for bioinformatic data analyists, but should be relevant for other
+domains.
 
@@
 rectified.
-So reproducibility is a necessary stepping-stone that in theory should enhance the overall reliability
-of computational research including replicability and robustness.
+Therefore, reproducibility is a necessary stepping-stone that in theory should enhance the overall
+reliability of computational research including replicability and robustness.
 Replicability being defined as the ability to repeat the entire study and come up with similar overall
@@
 
-Unreliable research is a waste of resources and time, and can mislead readers to pursue doomed
+Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
 research directions.
@@
 sciences [@Baker2015-bs; @Begley2012-di].
-Indeed in bioinformatics, the situation is grim with a 2009 report showing only a small fraction
+Indeed, in bioinformatics, the situation is grim with a 2009 report showing only a small fraction
 of studies able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of
@@
 Researchers are theoretically able to achieve 100% computational reproducibility,
-however in practice this is rarely achieved.
+however, in practice this is rarely achieved.
 Data and code availability, ability to install software easily, and other factors mean that most
@@
 The intended audience is bioinformatics data analysts (not tool developers), however the
-principles described here should translate to other domains of study.
+principles described here could equally apply to other domains of study.
 
@@
 which would be difficult for graphical tools like spreadsheets.
-Indeed spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
+Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko, and are
 somewhat overused in the life sciences.
 
-Transfering data between compute platforms is also discouraged.
+Transferring data between compute platforms is also discouraged.
 For example, having workflows that involve combinations of web-based and scripted tools require
@@
 The code should be able to fetch the data from a publicly accessible location and conduct the downstream
-processing and analysis without having to aquire the data some other way.
+processing and analysis without having to acquire the data some other way.
 
@@
 
-The Sweave project, concieved in 2001, was designed to give LaTeX documents embedded R code chunks and
+The Sweave project, conceived in 2001, was designed to give LaTeX documents embedded R code chunks and
 was envisaged as a way to generate R documentation materials and generate statistical analysis reports
@@
 * The provenance of any result can be demonstrated.
-The resulting document produced contains the code executed, together with the results (eg: a chart).
+The resulting document produced contains the code executed, together with the results (e.g.: a chart).
 This is in contrast to a data analysis report assembled in a word processor, which due to the many
@@
 be assembled "manually" with a word processor.
-Using a literate script would mean only minimal changes are required between interations @Perkel2022-af.
+Using a literate script would mean only minimal changes are required between iterations @Perkel2022-af.
 
@@
 to which part of the script.
-By embedding the outputs (eg: charts and tables) in a document in the sequence that they were generated,
-it helps the reader understand the logical steps taken in an analysis.
+By embedding the outputs such as charts and tables in a document in the sequence that they were
+generated, it helps the reader understand the logical steps taken in an analysis.
 
@@
 so we are not sure whether the whole script is free of errors and completed successfully.
-Therefore it is good practice to routinely execute these scripts during the development process,
+Therefore, it is good practice to routinely execute these scripts during the development process,
 and not rely too heavily on the interactive execution of individual chunks or lines.
@@
 writing a research article or self-publishing an e-book, literate programming provides the
-ability to construct transparent data narratives with clear provenance in a convieniently sharable
+ability to construct transparent data narratives with clear provenance in a conveniently sharable
 form.
@@
 code to consumers.
-There are many such DVCSs but `git` has emerged as the most popular
-solution due to it's many powerful features, speed/efficiency and large community and ecosystem.
+There are many such DVCSs, but `git` has emerged as the most popular
+solution due to its many powerful features, speed/efficiency and large community and ecosystem.
 
@@
 * Keeps a complete history of all code changes over time.
-We always know which version of the script is the most up-to-date.
+We always know which version of the script is the most up to date.
 We can inspect and execute the code and reproduce the analysis at any particular point in the past.
@@
 Using a centralised code hosting platform such as "GitHub" or "BitBucket" can help manage contributions
-from team members who may be working ascynchrously in different timezones and on different servers.
+from team members who may be working asynchronously in different time zones and on different servers.
 It also makes the user's work searchable, which is convenient when code snippets from a project
@@
 These centralised platforms also enable discussions of code issues such as bugs and feature requests.
-In bioinformatics teams, issues can be used to track the progress of a project through it's milestones
-and allows team members to raise potential issues with a colleagues code (or one's own).
+In bioinformatics teams, issues can be used to track the progress of a project through its milestones
+and allow team members to raise potential issues with the code.
 This also allows the scientific community to raise an issue with the software maintainers.
@@
 code.
-By preserving a copy of the repository on each team members' computer in addition to the central
-repository, means the code can be easily recovered if any one of those computers fail.
+Preserving a copy of the repository on each team members' computer in addition to the central
+repository means the code can be easily recovered if any one of those computers fails.
 
-Although DVCSs assist with code sharing, they are not considered a long term archiving solution and
-might not survive the "ten year challenge" of code reproducibility @Perkel2020-io.
-To achieve that, the code needs to be deposited to a long term repository.
-Zenodo and Figshare repositories are both good options for long term archiving of software, and the
+Although DVCSs assist with code sharing, they are not considered a long-term archiving solution and
+might not survive the "ten-year challenge" of code reproducibility @Perkel2020-io.
+To achieve that, the code needs to be deposited to a long-term repository.
+Zenodo and Figshare repositories are both good options for long-term archiving of software, and the
 link to these should be provided in the respective journal article/pre-print.
@@
 
-In R, such reporting is possible using the `sessionInfo()` command;
+In R, such reporting is possible using the `sessionInfo()` command,
 while for Python this is possible using the `session_info` package.
-Using literate programming ensures that the output document contains a record of this important
-information.
+Using literate programming and sharing output documents ensures that a record of this important
+information is made available.
 
@@
 reproducibility.
-A researcher trying to reproduce a 10 year old study in R v3.0 could have a headache, as they would need
+A researcher trying to reproduce a ten-year-old study in R v3.0 could have a headache, as they would need
 to roll back their R version, and possibly their operating system as well, as R and other languages
@@
 without needing to change the host R version.
-While this provides good reproducibility, setting up an environment with a 10 year old operating system,
+While this provides good reproducibility, setting up an environment with a ten-year-old operating system,
 R and packages would take a few hours to accomplish.
@@
 Another potential solution to this is to use a package/environment management system
-(eg: Conda or Guix).
+(e.g.: [Conda](https://conda.org/) or [Guix](https://guix.gnu.org/)).
 These allow users to create, manage, and deploy software packages,
@@
 
-Containers are an attempt to resolve some of the downsides of virtual machines and package managers.
-Container images can be thought of as similar to virtual machines, but are more lightweight as they
+Containers are an attempt to resolve some of the downsides of VMs and package managers.
+Container images can be thought of as similar to VMs, but are more lightweight as they
 share parts of the host operating system.
@@
 
-![Figure 4. Architecture of containers (left) compared to virtual machines (right) @noauthor_2021-lt](../img/docker-containerized-and-vm-transparent-bg.webp)
+![Figure 4. Architecture of containers (left) compared to VMs (right) @noauthor_2021-lt](../img/docker-containerized-and-vm-transparent-bg.webp)
 
@@
 As they are a recipe, these are small files which are easily shared in the project code repository.
-Docker images are larger, and cannot be shared in the code repository.
+Docker images are larger and cannot be shared in the code repository.
 They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
@@
 Still, these are not considered permanent archival services, so it is advised to deposit the image
-used in an article to a suitable persitent long-term archive (discussed below).
+used in an article to a suitable persistent long-term archive (discussed below).
 The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
@@
 
-The field of genomics has has a long history of data sharing @Brazma2001-yk, which is supported by
+The field of genomics has a long history of data sharing @Brazma2001-yk, which is supported by
 policy leadership by funding bodies
@@
 preexisting genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
-Therefore sharing deidentified genomic data and metadata is likely to carry little risk to participant
+Therefore, sharing deidentified genomic data and metadata is likely to carry little risk to participant
 privacy.
@@
 data sharing that was agreed to in the consent process.
-For example the consent form may only allow sharing of data to researchers investigating certain
+For example, the consent form may only allow sharing of data to researchers investigating certain
 diseases, which means a vetting process must be undertaken before sharing data with external
@@
 Accessible, Interoperable and Reusable (FAIR) for people and for computers @Wilkinson2016-qq.
+These principles are summarised in **Figure 5**.
 
-![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb](../img/gr2)
+![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb](../img/gr2.jpg)
 
 High throughput biology has already built a number of data repositories for specific data types
-conforming to FAIR principles (eg: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
+conforming to FAIR principles (e.g.: GEO, SRA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
 However other domains lack well known specialised organised data repositories, such as evolutionary
@@
 from these fields [@Pampel2013-ep].
-If no specialised repository exists, then a general purpose repository such as
+If no specialised repository exists, then a general-purpose repository such as
 [Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
@@
 
-1. Depositing data to a specialised repository if possible, otherwise a general purpose repository.
+1. Depositing data to a specialised repository if possible, otherwise, a general-purpose repository.
 
@@
 
-5. Preferably archive and share raw data, and use existing standards for the discipline.
+5. Preferably archive and share raw data and use existing standards for the discipline.
 
@@
 For example, in the field of genomics, gene function annotation sets are relied upon for a common
-proceduce called "pathway analysis".
+procedure called "pathway analysis" @Khatri2012-ye.
 These annotation sets are regularly updated, but versions are rarely reported [@Wijesooriya2022-ly]
 and finding historical data versions is sometimes difficult.
-So in order to guarantee future reproducibility, shapshots of those reference data should be archived &
-shared if the licence permits it.
+In order to guarantee future reproducibility, snapshots of those reference data should be archived
+and shared if the licence permits it.
 
@@
 
-* Instructions needed
+Documentation is the glue that binds a data science project together.
+The manuscript/published article is the central artifact that outlines the research project and
+links to the supporting materials.
+From a reproducibility standpoint, the methods section is the most important section.
+It should be detailed enough so that other researchers can understand and replicate the
+experiments/analysis and yield similar results.
+The "Materials Design Analysis Reporting" (MDAR) checklist for authors has been developed to assist in
+comprehensive methodological reporting in the life sciences [@Mellor2022-up].
 
-* Requirements
+Researchers should consider depositing their thorough laboratory/*in silico* protocols as separate
+outputs to relevant repositories such as [protocols.io](https://www.protocols.io/),
+[RIO Journal](https://riojournal.com/) or [Zenodo](https://zenodo.org/), minting a Digital Object
+Identifier (DOI) that can be cited in the manuscript.
+This is particularly helpful when there are strict word limits on articles.
 
-* Credits
+The manuscript should have clear links to supporting materials including datasets, software code and
+other resources like computational environments (e.g.: Docker images).
+Each of these supporting materials also require documentation to ensure that their context can be
+understood and that they can be reused.
 
-## Bringing it all together
+The Code repository should have a detailed "README" file which should communicate:
 
-Due to the rapidly devloping software landscape, this work is not designed to give strong
-recommendations on the exact combination of tools or resources
+* The purpose of the code/software/overall project.
+
+* How to reproduce the analysis.
+
+* How to contribute/report issues.
+
+* Credits to the contributors.
+
+Dataset raw and intermediate files with metadata
+
+* Compute environment
+
+* Credits different authors.
+
+## Limitations and challenges
+
+Due to the constantly developing software landscape, this review is not designed to give strong
+recommendations on the exact combination of tools or resources.
 to use in computational research, but to set out key principles and highlight some useful solutions
@@
 A recent systematic examination of Jupyter notebooks in GitHub repositories accompanying research papers
-found that environment problems (eg: "module not found", "import error") and lack of data links
-(eg: "file not found" errors) accounted for 80% of code errors @Samuel2022-ha.
+found that environment problems (e.g.: "module not found", "import error") and lack of data links
+(e.g.: "file not found" errors) accounted for 80% of code errors @Samuel2022-ha.
 
@@
 
-Some researchers have suggested to take it even further, by adapting software development
+Some researchers have suggested adapting software development
 principles of unit tests and continuous integration/continuous delivery.
@@
 reproducibility without the need for individuals to execute the workflow themselves.
-Therefore continuous analysis might be a future pediment (crowning feature) to the five-pillared
+Continuous analysis might therefore be a future pediment (crowning feature) to the five-pillared
 "temple" of reproducible research.
-
-## Where to from here
 
@@
 Indeed the number of systematic replication studies in bioinformatics remains very small
-(eg: [@Ioannidis2009-hv; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the problem
-actually is.
+(e.g.: [@Ioannidis2009-hv; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
+problem actually is.
 
@@
 provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
-For example, my group has just released such a protocol that walks novice users through the process
+For example, our group has just released such a protocol that walks novice users through the process
 of implementing functional enrichment analysis of gene expression data with the five pillars principles
@@
 In those situations, a workflow manager such as Snakemake or Nextflow could be useful to
-coordinate those hybrid computate workflows.
+coordinate those hybrid computational workflows.
 
@@
 https://www.minot.bio/
-
-https://www.researchgate.net/publication/349603799_Making_radiotherapy_more_efficient_with_FAIR_data/figures?lo=1
-
-Maer-Matei, Monica Mihaela, Tiberiu Marian Georgescu, Cristina Mocanu, and Ana-Maria Zamfir. 2019/01/01////.
-“Skill Needs for Early Career Researchers.” Sustainability 11 (10). https://doi.org/ghkb9m (https://doi.org/ghkb9m).
-
-Murray, NJ et al. 2021. “Data Freshness in Ecology and Conservation.” Trends Ecol Evol 36 (6): 485–87.
-
-Obels, Pepijn, Daniël Lakens, Nicholas Coles, Jaroslav Gottfried, and Seth Green. 2020. “Analysis of Open Data and Computational
-
-Reproducibility in Registered Reports in Psychology.” Advances in Methods and Practices in Psychological Science 3 (May):
-251524592091887. https://doi.org/gg4vw4 (https://doi.org/gg4vw4).
-
-Peikert, Aaron, and Andreas Markus Brandmaier. 2019. “A Reproducible Data Analysis Workflow with R Markdown, Git, Make, and Docker.”
-
-PsyArXiv. https://doi.org/10.31234/osf.io/8xzqy (https://doi.org/10.31234/osf.io/8xzqy).
-@aaron_willcox ! " # || @Elliot_Gould_ ! " #
 

### 2023-05-23 docs/5pillars.Rmd
edit

@@
 reproduced quickly and easily, long into the future.
-This guide is designed for bioinformatic data analyists, but should be relevant for other
+This guide is designed for bioinformatic data analysts, but should be relevant for other
 domains.
@@
 to make the reproduction process easier and faster for those who wish to.
+A "master script" that coordinates the execution of individual parts of an analysis is an excellent
+way to reduce the number of commands required for reproduction.
+This can be coded in standard languages like R/R Markdown, Python/Jupyter notebook or the Unix Shell.
+Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
+`snakemake` [@Koster2012-sq; @Molder2021-mp] or `nextflow` [@Di_Tommaso2017-yo].
+The advantage of such tools is that if the analysis were to fail mid-way,
+for example at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't need to be repeated.
+After fixing the issue with code at step 8, re-running the analysis would pick up at step 8 again,
+which saves labour and compute time.
 
@@
 * Output reports are free from code errors.
-The output document is only rendered when the entire script has compiled without errors.
+The output document is only rendered when the entire script has been compiled without errors.
 In contrast, a regular script might generate some outputs and then encounter an error,
@@
 support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
-Outputs include HTML documents, but also slideshows and even e-books.
+Outputs include HTML documents, but also slideshows and even ebooks.
 
@@
 Whether this is sharing a data analysis report, giving a presentation at a meeting,
-writing a research article or self-publishing an e-book, literate programming provides the
-ability to construct transparent data narratives with clear provenance in a conveniently sharable
+writing a research article or self-publishing an ebook, literate programming provides the
+ability to construct transparent data narratives with clear provenance in a conveniently shareable
 form.
@@
 
-* Makes works easily sharable.
+* Makes works easily shareable.
 Work is easier to reproduce because the source folder contains a complete workflow with code linked
@@
 to roll back their R version, and possibly their operating system as well, as R and other languages
-requires certain system libraries for low-level routines.
+require certain system libraries for low-level routines.
 
@@
 software stack and project data.
-Moreover, the performance of computation in guest system is typically slower than when run direct on
-the host.
+Moreover, the performance of computation in the guest system is typically slower than when run directly
+on the host.
 
@@
 and typically within a few minutes.
-This accellerates the installation procedure dramatically which is a known bottleneck for bioinformatic
+This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
 reproducibility [@Mangul2019-ei;@Mangul2019-xp].
 
-![Figure 4. Architecture of containers (left) compared to VMs (right) @noauthor_2021-lt](../img/docker-containerized-and-vm-transparent-bg.webp)
+![Figure 4. Architecture of containers (left) compared to VMs (right) @noauthor_2021-lt.](../img/docker-containerized-and-vm-transparent-bg.webp)
 
@@
 install any software at all.
-For example, it is now possible embed a Wasm-enabled R terminal inside a Quarto HTML document, which
+For example, it is now possible to embed a Wasm-enabled R terminal inside a Quarto HTML document, which
 enables in-browser reproducibility of R workflows @Balamuta2023-ma.
@@
 
-![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb](../img/gr2.jpg)
+![Figure 5. Overview of the FAIR data principles @Kalendralis2021-jb.](../img/gr2.jpg)
 
@@
 biology, ecology, cell biology, social sciences and computing.
-The **re3data.org** registry of data repositories may be useful to find a repository that accepts data
-from these fields [@Pampel2013-ep].
+The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
+repository that accepts data from these fields [@Pampel2013-ep].
 If no specialised repository exists, then a general-purpose repository such as
@@
 5. Preferably archive and share raw data and use existing standards for the discipline.
+
+6. Use file formats that are compatible with many different types of software; e.g.: comma-separated
+values (CSV) files are preferred over Excel files.
 
@@
 Documentation is the glue that binds a data science project together.
-The manuscript/published article is the central artifact that outlines the research project and
+The manuscript/published article is the central artefact that outlines the research project and
 links to the supporting materials.
-From a reproducibility standpoint, the methods section is the most important section.
+From a reproducibility standpoint, the methods section is the most critical part.
 It should be detailed enough so that other researchers can understand and replicate the
@@
 The manuscript should have clear links to supporting materials including datasets, software code and
-other resources like computational environments (e.g.: Docker images).
+other resources like computational environments (Table 1).
 Each of these supporting materials also require documentation to ensure that their context can be
@@
 
-The Code repository should have a detailed "README" file which should communicate:
+| Item | Location |
+| --- | --- |
+| Data set and metadata | Persistent repository |
+| Code base and documentation | DVCS (eg: GitHub) and snapshot deposited to persistent repository |
+| Container image | Container registry and persistent repository |
 
-* The purpose of the code/software/overall project.
+: Table 1. Supporting information for a five-pillar research article.
 
-* How to reproduce the analysis.
+The code repository should have a detailed `README` file which plays a critical role in reproducibility.
+It should outline the purpose of the code/software/overall project and how it relates to the
+manuscript/article.
+For example, some articles may rely on more than one repository for various parts of the research
+project, so these need to be explained.
 
-* How to contribute/report issues.
+The `README` should outline exactly what is required in order to reproduce the
+analysis, including the requirements and the instructions to reproduce.
+Hardware requirements such as RAM, CPU architecture, GPU needs, etc need to be outlined.
+Software requirements, such as operating system, container engine, workflow manager,
+etc, also need to be described.
+The exact instructions for reproduction should be clearly outlined.
+These instructions should be tested whenever changes are made to the codebase.
 
-* Credits to the contributors.
+As outlined above, one of the goals is to reduce the complexity of reproduction, using a master script.
+For example, the pathway analysis workflow we previously developed could be reproduced and inspected
+with just six commands, starting from a new Ubuntu linux install [@Ziemann2023-nm].
+By minimising the complexity of reproduction and writing the instructions clearly in the `README`,
+reproducibility is made possible to a larger segment of the computational research community,
+especially those who are not experts in the languages used.
 
-Dataset raw and intermediate files with metadata
+In addition, the `README` should outline how users can contribute to the codebase and how to report
+issues, such as code errors and discrepancies.
+The `README` should also list the contributors to the codebase and acknowledge any financial support.
 
-* Compute environment
-
-* Credits different authors.
+The need for documentation extends to the other supporting information.
+The deposited data set needs thorough descriptive metadata to provide context.
+The container image and code snapshot should be deposited to a persistent repository with
+a thorough description of their purpose.
+Both the data and container image descriptions need to contain links to the DOI of the manuscript,
+so that machines or individuals browsing these resources can understand the context of the overall
+research project.
 
@@
 For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
-reference genome and variant calling which are demanding in memory and compute.
+reference genome and variant calling which are demanding in memory and computaion.
 These computationally demanding steps are typically conducted on high end servers and workstations.

### 2023-05-24 docs/5pillars.Rmd
1st draft

@@
 
-Computational reproducibility is in theory a simple premise, however in practice it is difficult to
+Computational reproducibility is a simple premise in theory, but in practice it is difficult to
 achieve.
@@
 ## Introduction
-
-```{r,libs,echo=FALSE}
-
-library("DiagrammeR")
-
-```
 
@@
 sciences [@Baker2015-bs; @Begley2012-di].
-Indeed, in bioinformatics, the situation is grim with a 2009 report showing only a small fraction
-of studies able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of
+Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
+2 of 18 articles could be reproduced (11%) @Ioannidis2009-hv, bringing into question the reliability of
 those studies.
+Workshops conducted by NIH intramural researchers in 2018/2019 sought to reproduce five bioinformatics
+studies and could not reproduce *any*, citing missing data, software and documentation
+@Zaringhalam2020-ms.
+A recent systamatic analysis of Jupyter notebooks in biomedical articles had similar observations, with
+only 245/4169 (5.9%) notebooks giving similar results compared to the original, with irreproducibility
+blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
 Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
 earlier case study @Baggerly2010-nr.
-
-Researchers are theoretically able to achieve 100% computational reproducibility,
-however, in practice this is rarely achieved.
-Data and code availability, ability to install software easily, and other factors mean that most
-computational research works are unable to be reproduced @Samuel2022-ha.
+Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
+in practice this is rarely achieved.
 
 A number of guides recommending enhanced computational reproducibility practices have emerged
-[@Peng2011-ia; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Lewis2016-qh;
-@Brito2020-ld; @Rodrigues2023-fw; @Wilson2017-uq; @Akalin2018-eg ; @Peikert2021-eq ].
-This article attempts to unify these concepts into five key principles called the
-five pillars of reproducible computational research (**Figure 1**).
+[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
+@Wilson2017-uq ; @Akalin2018-eg ; @Gruning2018-sg; @Brito2020-ld; @Peikert2021-eq ; @Rodrigues2023-fw]
+This article, inspired by Peikert and BrandMaier @Peikert2021-eq, attempts to unify these concepts into
+five key principles called the five pillars of reproducible computational research (**Figure 1**).
 The intended audience is bioinformatics data analysts (not tool developers), however the
@@
 
-End-to-end code coverage is the foundation for reproducibility as the codebase should ideally contain all
+End-to-end code coverage is the foundation for reproducibility as the codebase should contain all
 instructions for completing a data analysis task.
@@
 A caveat for end-to-end code coverage is that sometimes manual data cleaning is unavoidable.
-In those cases, retaining raw and cleaned data and providing a cleaning protocol is recommended,
-then the computational workflow can begin with the cleaned dataset.
+In those cases, retaining raw and cleaned data along with a cleaning protocol is recommended.
+Then the computational workflow can begin with the cleaned dataset.
 
@@
 Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
-`snakemake` [@Koster2012-sq; @Molder2021-mp] or `nextflow` [@Di_Tommaso2017-yo].
+`snakemake` [@Koster2012-sq; @Molder2021-mp], `drake` [@Michael_Landau2018-kk] or
+`nextflow` [@Di_Tommaso2017-yo].
 The advantage of such tools is that if the analysis were to fail mid-way,
@@
 R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
+Similarly, the [Streamlit](https://streamlit.io/) tool, which was originally invented for creating
+Python-based web dashboards but is equally capable of generating data science reports
+[e.g.: @Nantasenamat2023-wr].
+
 In 2022, the company behind RStudio (Posit) released Quarto, the conceptual successor of R Markdown,
@@
 tables, images, videos, etc.
-More information on documentation below.
+This means it is possible to author an entire journal articles using literate programming.
 
@@
 * Flexible output formats.
-These include PDF, DOC/DOCX and HTML, with HTML having some notable benefits to other formats
+These include PDF, DOC/DOCX and HTML, the latter having some notable benefits
 including better usability for mobile devices like phones and tablets, and the ability to
@@
 to data and metadata, so it doesn't need any modifications before reproduction.
+It is relatively easy to add a software license with a DVCS like `GitHub`, so consumers can understand
+the conditions of using it.
 
@@
 link to these should be provided in the respective journal article/pre-print.
+
+`git` is typically used at the command line, however it is also integrated into Rstudio, JupyterLab and
+VS Code, and there are also several `git` clients with graphical interfaces which better allow inspection
+of code changes.
 
@@
 across different systems.
-For example, the most popular containerisation system, Docker, makes it
+For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
 possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
@@
 The manuscript should have clear links to supporting materials including datasets, software code and
-other resources like computational environments (Table 1).
-Each of these supporting materials also require documentation to ensure that their context can be
-understood and that they can be reused.
-
-| Item | Location |
-| --- | --- |
-| Data set and metadata | Persistent repository |
-| Code base and documentation | DVCS (eg: GitHub) and snapshot deposited to persistent repository |
-| Container image | Container registry and persistent repository |
-
-: Table 1. Supporting information for a five-pillar research article.
-
+other resources like computational environments.
 The code repository should have a detailed `README` file which plays a critical role in reproducibility.
@@
 analysis, including the requirements and the instructions to reproduce.
+Typically it is written in the markdown format, which should be familiar
 Hardware requirements such as RAM, CPU architecture, GPU needs, etc need to be outlined.
@@
 The exact instructions for reproduction should be clearly outlined.
+This should include what the output objects of the analysis are, and instructions on how to access the
+results and what should be expected.
+Literate programming allows for thorough documentation of workflow methods and results which makes
+it more accessible for reproducers to comprehend the workflow details, which is not possible with
+standard code comments [@Rule2019-eh].
 These instructions should be tested whenever changes are made to the codebase.
 
-As outlined above, one of the goals is to reduce the complexity of reproduction, using a master script.
+As outlined above, one of the goals is to reduce the complexity of reproduction, and using a master
+script is a simple way to do this.
+It also makes the README documentation much simpler.
 For example, the pathway analysis workflow we previously developed could be reproduced and inspected
@@
 
-In addition, the `README` should outline how users can contribute to the codebase and how to report
-issues, such as code errors and discrepancies.
+In addition, the `README` should outline the contents of the repository, how users can contribute to
+the codebase and how to report issues, such as code errors and discrepancies.
 The `README` should also list the contributors to the codebase and acknowledge any financial support.
+It may also include a link to the research article, citation instructions and a mention of the software
+license.
 
@@
 Both the data and container image descriptions need to contain links to the DOI of the manuscript,
-so that machines or individuals browsing these resources can understand the context of the overall
-research project.
+so that machines or individuals browsing these resources can understand the context of these objects
+relative to the overall research project.
 
@@
 Due to the constantly developing software landscape, this review is not designed to give strong
-recommendations on the exact combination of tools or resources.
-to use in computational research, but to set out key principles and highlight some useful solutions
-that users can incorporate into their toolkit to improve the overall reproducibility of their work.
+recommendations on the exact combination of tools or resources to use in computational research,
+but to set out key principles and highlight some useful solutions that users can incorporate
+into their toolkit to improve the overall reproducibility of their work.
 
-When not integrated correctly, problems readily emerge.
-A recent systematic examination of Jupyter notebooks in GitHub repositories accompanying research papers
-found that environment problems (e.g.: "module not found", "import error") and lack of data links
-(e.g.: "file not found" errors) accounted for 80% of code errors @Samuel2022-ha.
-
-This is why regular code testing is another recommended best practice.
-By regularly executing the workflow after changes are made ensures the absence of code errors.
-Executing literate scripts is the simplest approach, as if there is an error incurred the output report
-won't be generated, and the source of the error can be investigated.
+When these principles are not integrated correctly and checked, problems readily emerge
+[@Barba2016-kk; @Samuel2022-ha; @Zaringhalam2020-ms].
+This is why regular code testing is another recommended best practice, especially after making updates
+to the code or data [@Rule2019-eh; @The_Turing_Way_Community2019-ig].
 It is worthwhile to conduct such testing on a separate computer to understand how someone else in the
@@
 
-Some researchers have suggested adapting software development
+Moreover, some have suggested adapting software development
 principles of unit tests and continuous integration/continuous delivery.
 This approach, called "continuous analysis" involves automatic execution and testing whenever a change is
-made to a project's code base @Beaulieu-Jones2017-vj.
-
+made to a project's code base [@Beaulieu-Jones2017-vj; @Blischak2019-dq].
 If the workflow executes successfully and tests are passed, the repository can be decorated with a
@@
 yet progress on integrating these into computational research publications has been minimal.
-Part of the problem is that funding bodies and journals are focused on supporting novel research
-directions, instead of validating previous findings.
-Indeed the number of systematic replication studies in bioinformatics remains very small
-(e.g.: [@Ioannidis2009-hv; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
+Part of the problem is the incentive structures in academia, with funding bodies and journals are
+focused on novel research directions rather than confirming previous findings
+[@Kaelin2017-id; @Munafo2017-ty].
+Indeed the number of systematic replication studies in bioinformatics remains small
+(e.g.: [@Ioannidis2009-hv; ; @Samuel2022-ha]), so we don't have much of a grasp as to how severe the
 problem actually is.
 
-A part of the problem lies in the lack of technical computer skills in many research groups.
-Specific training in these areas could help, but the most scalable solution would be to
+A lack of technical computer skills in many research groups is another culprit.
+Specific workshop-based training in these areas could help, but the most scalable solution would be to
 provide comprehensive beginner-friendly step-by-step guides for frequently used routines.
@@
 More such protocols are needed to make extreme reproducibility mainstream.
-
-There are other challenges worth mentioning.
-Different parts of the workflow may have different compute needs.
-For example, genomic sequencing analysis workflows typically involve the mapping of reads to the
-reference genome and variant calling which are demanding in memory and computaion.
-These computationally demanding steps are typically conducted on high end servers and workstations.
-After initial processing, the resulting variant files are ~100X smaller than the raw sequence files,
-so downstream analysis could be conducted on commodity PCs or laptops.
-In those situations, a workflow manager such as Snakemake or Nextflow could be useful to
-coordinate those hybrid computational workflows.
-
-Adoption of these practices would require a change in incentives.
-https://www.nature.com/articles/s41562-016-0021
-
-## References to read
-
-https://www.minot.bio/
-
-## Practical guides
-
-We provide some practical resources for those interested in adopting various
-
-Put these on webarchive and provide persistent links
-
-| Topic | Resource |
-| --- | --- |
-| General coding for science | eg |
-| Version control | eg |
-| Literate programming | eg |
-| Environment control | eg |
-| Documentation | eg |
-| Data sharing | eg |
-
-### Getting started with data analysis in R and Python
-
-* Intro to R and Rstudio
-
-* Intro to Python and iPython IDE
-
-* Intro to vscode
-
-### Literate programming with jupyter, R Markdown and Quarto
-
-* https://medium.com/@itsjzt/beginner-guide-to-markdown-229adce30074
-
-* https://www.dataquest.io/blog/r-markdown-guide-cheatsheet/
-
-* https://tysonbarrett.com/Rstats/chapter-9-reproducible-workflow-with-rmarkdown.html
-
-* https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html
-
-* https://quarto.org/docs/computations/r.html
-
-* https://quarto.org/docs/computations/python.html
-
-* https://jupyter-notebook-beginner-guide.readthedocs.io/en/latest/
-
-* https://www.dataquest.io/blog/jupyter-notebook-tutorial/
-
-* https://medium.com/velotio-perspectives/the-ultimate-beginners-guide-to-jupyter-notebooks-6b00846ed2af
-
-### Practical guides for `git` and GitHub
-
-* https://www.datacamp.com/tutorial/github-and-git-tutorial-for-beginners/
-
-* https://www.freecodecamp.org/news/introduction-to-git-and-github/
-
-* https://www.geo.uzh.ch/microsite/reproducible_research/post/rr-rstudio-git/
-
-* https://anderfernandez.com/en/blog/how-to-link-and-use-git-with-rstudio/
-
-* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
-
-* https://blog.reviewnb.com/github-jupyter-notebook/
-
-* https://andgarc.github.io/basic-git-jupyter-lesson/index.html
-
-### Practical guides for Conda, Guix and Docker
-
-### Practical guides for documenting computational research
-
-### Practical guides to data sharing
 

### 2023-06-22 example.Rmd
minor adjustment to euler diagram
bunch of changes

@@
 (Ziemann et al, 2019).
+
+The gene counts have also been deposited to the `/data` folder in the `example.Rdata` file in case
+the DEE2 resource becomes unavailable.
+To import it, use the command: `load("data/example.Rdata")`
 
@@
 Get the gene sets loaded in R.
+They are located in the `/ref` folder of the repo.
 
@@
 # from https://reactome.org/download/current/ReactomePathways.gmt.zip
-genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")
-gsets <- gmtPathways("ReactomePathways_2023-03-06.gmt")
+genesets2 <- read.gmt("ref/ReactomePathways_2023-03-06.gmt")
+gsets <- gmtPathways("ref/ReactomePathways_2023-03-06.gmt")
 
@@
 
+message("number of genes in each group")
+lapply(list("background"=bg,"up-regulated"=defup,"down-regulated"=defdn),length)
+
 ```
 
-Enrichment firstly with upregulated genes
+One of the quirky things I learned about clusterprofiler is that
+not all genes specified by the user as background are included in
+the enrichmentcalculation.
+It appears as though the genes not included in the annotation set are
+discarded.
+I believe this is a subtle but significant problem.
+To fix it is relatively easy from the user's point of view.
+Just create a new gene set that consists of all detected genes and
+
+Adding all detected genes to the background appears to improve results!
+
+```{r,gs2}
+
+bgdf <- data.frame("background",bg)
+colnames(bgdf) <- c("term","gene")
+genesets2 <- rbind(genesets2,bgdf)
+
+```
+
+Enrichment firstly with upregulated genes.
+Note that the BgRatio contains the correct number of genes in the background now.
 
@@
 
+# show 10 pathways only
+n_pw=10
+
 ora_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  universe = bg,  maxGSSize = 50000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 
-head(ora_up) %>%
+head(ora_up,n_pw) %>%
   kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
@@
 
-n_pw=10
-
 ora_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
+  universe = bg,  maxGSSize = 50000, TERM2GENE = genesets2,
   pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
@@
 
-```{r,venn1,fig.width=6}
+```{r,venn1,fig.width=8,fig.height=5}
 

### 2023-06-22 manuscript/manuscript.Rmd
update article

--- 2023-05-08 manuscript.Rmd
@@
 
-Mark Ziemann<sup>1</sup>*, Anusuiya Bora<sup>2</sup>
+[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
+[Anusuiya Bora<sup>2</sup>](https://orcid.org/0009-0006-2908-1352)
 
@@
 a new project's enrichment analysis needs.
-This recipe is built according to a framework we call "five pillars", which when implemented together
-facilitates extreme reproducibility.
+This recipe is built according to the "five pillars" framework, which facilitates extreme
+reproducibility.
 
@@
 and without any manual steps @Lasser2020-mm.
-Taken together, computational reproducibility is a spectrum (**Figure 1**) @Akalin2018-eg, and rests upon five
-pillars (**Figure 2**).
-Currently, bioinformatics work at the gold standard is rare, despite the availability of such tools for
+Taken together, computational reproducibility is a spectrum @Akalin2018-eg, and the *five pillars*
+framework has been described as a way of achieving extreme reproducibility @Ziemann2023-pt.
+Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
 over a decade.
-
-![Figure 1.
-The computational research reproducibility spectrum.
-Creator: Altuna Akalin.
-Republished with permission @Akalin2018-eg.](img/repro.webp "The computational research reproducibility spectrum.")
-
-```{r,5pillars,fig.cap="Figure 2. The five pillars of computational reproducibility.",echo=FALSE,fig.height=6}
-
-library("RColorBrewer")
-cols <- brewer.pal(5, "Set2")
-cols2 <- adjustcolor( cols, alpha.f = 0.7)
-plot(0,type='n',axes=FALSE,ann=FALSE, xlim=c(0, 10), ylim=c(0, 8.5))
-polygon(c(0,1.8,1.8,0),c(1,1,6,6), col=cols2[1], border = NA)
-polygon(c(2,3.8,3.8,2),c(1,1,6,6), col=cols2[2], border = NA)
-polygon(c(4,5.8,5.8,4),c(1,1,6,6), col=cols2[3], border = NA)
-polygon(c(6,7.8,7.8,6),c(1,1,6,6), col=cols2[4], border = NA)
-polygon(c(8,9.8,9.8,8),c(1,1,6,6), col=cols2[5], border = NA)
-polygon(c(-0.2,10,10,-0.2),c(-0.4,-0.4,0.8,0.8),col="lightgray", border = NA)
-polygon(c(-0.2,5,10),c(6.5,8.5,6.5),col="lightgray", border = NA)
-points( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5), pch=19,cex=18,col=cols )
-text(5,7,lab="Five pillars of reproducible computational research", cex=1.5)
-
-labs <- c("Version\ncontrol &\npersistent\ncode sharing",
-  "Compute \nenvironment \ncontrol",
-  "Literate\nprogramming",
-  "Documentation",
-  "FAIR &\npersistent\ndata sharing")
-
-text( c(0.9,2.9,4.9,6.9,8.9) , c(4.5,2.5,4.5,2.5,4.5) , lab=labs)
-text( 9, -0.2 , lab=expression(italic("created by Mark Ziemann")),cex=0.5)
-
-```
 
@@
 [@Slonim2002-sf; @Khatri2012-do].
-While enrichment analysis is undoubtedly a key method for the interpretation of large
+While enrichment analysis is undoubtedly a key method for the interpretation of large omics
 data sets, there are concerns that inadvertent misuse of these methods leads to
@@
 After conducting this reproduction study according to the methods described in the respective articles,
-the 20 articles were classified based on the degree of reproducibility (Figure 3).
+the 20 articles were classified based on the degree of reproducibility (Figure 1).
 Six studies were classified as low reproducibility, while 10 were classified as medium and just three
@@
 
-```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 3. Examining reproducibility in a sample of 20 journal articles."}
+```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility in a sample of 20 journal articles."}
 
@@
 protocols.io with a permissive license.
-This protocol has the features of reproducibility mentioned in **Figure 1** and **Figure 2** including
+This protocol is in accordance with the *five pillars* including
 version control with git and GitHub, compute environment control using Docker, literate programming
@@
 data set from DEE2, inspecting data quality, differential expression and two types of enrichment
-analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 4**).
+analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 2**).
 For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
@@
 
-```{r,diagram,echo=FALSE,fig.cap = "Figure 4. Flow chart of example workflow."}
+```{r,diagram,echo=FALSE,fig.cap = "Figure 2. Flow chart of example workflow."}
 
@@
 1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
-(Figure 5A).
+(Figure 3A).
 Up and down-regulated gene sets were subjected to ORA separately, using the list of
@@
 were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
-(host defence) (Figure 5B).
+(host defence) (Figure 3B).
 
@@
 FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
-and defective homologous recombination repair (Figure 5C).
+and defective homologous recombination repair (Figure 3C).
 Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
@@
 Jaccard index of 0.54, however these methods disagreed in down-regulated pathways, with a Jaccard
-index of only 0.27 (Figure 5D).
+index of only 0.27 (Figure 3D).
 
-![Figure 5. Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")
+![Figure 3. Differential expression and enrichment analysis with the example workflow.](../img/fig_example.png "Example enrichment analysis on")
 
@@
 
-We propose the five pillars (**Figure 2**), as a framework to guarantee computational reproducibility.
+This article is an attempt to foster increased reproducibility in genomics by demonstrating the
+implementation of the five pillars principles into a commonly used routine, namely differential
+expression and pathway enrichment analysis.
 However, these concepts may be difficult for novice analysts, and so we provide a step-by-step

### 2023-06-22 manuscript/manuscript.Rmd
updates to article

@@
 After conducting this reproduction study according to the methods described in the respective articles,
-the 20 articles were classified based on the degree of reproducibility (Figure 1).
+the 20 articles were classified based on the degree of reproducibility (**Figure 1**).
 Six studies were classified as low reproducibility, while 10 were classified as medium and just three
@@
 genes.
+A multi-dimensional scaling plot showed clustering of samples into their treatment
+groups, suggesting a robust effect of the chemical treatment (**Figure 3A**).
 Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
 1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
-(Figure 3A).
+(**Figure 3B**).
 Up and down-regulated gene sets were subjected to ORA separately, using the list of
@@
 
-With ORA, we observed 149 up-regulated and 48 down-regulated gene sets.
+With ORA, we observed 201 up-regulated and 99 down-regulated gene sets.
 We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
@@
 expression, rather than large gene sets that undergo small expression changes.
-This led to the identification of up-regulated pathways related to nuclear RNA export,
-glucokinase regulation and G1/S DNA replication pathways; while down-regulated pathways
-were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
-(host defence) (Figure 3B).
+This led to the identification of up-regulated pathways related to nuclear RNA import and
+glucokinase regulation; while down-regulated pathways
+were associated with interferon signaling, interleukin signaling and NADPH oxidase
+(host defence) (**Figure 3C**).
 
-After FCS, we identified 241 up-regulated gene sets and 155 down-regulated gene sets.
-FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
-and defective homologous recombination repair (Figure 3C).
-Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.
+After FCS, we identified 239 up-regulated gene sets and 158 down-regulated gene sets.
+FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA
+transport and defective homologous recombination repair (**Figure 3D**).
+Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF
+signaling.
 
 Regarding up-regulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
-Jaccard index of 0.54, however these methods disagreed in down-regulated pathways, with a Jaccard
-index of only 0.27 (Figure 3D).
+Jaccard index of 0.58, while this agreement was lower for down-regulated pathways, with a Jaccard
+index of 0.46 (**Figure 3E**).
 

### 2023-06-29 manuscript/manuscript.Rmd
editing

@@
 reproducibility.
-The findings indicate that current enrichment analysis practices are inadequate.
+The findings indicate that current enrichment analysis practices and reporting are inadequate.
 To address this, we provide a set of template scripts and a step-by-step protocol to adapt them to
@@
 Indeed, even if data and code are shared, reproducibility still cannot be guaranteed due to the
-differences in computing environments over time, and the availability of dependencies @Perkel2020-io.
+differences in computing environments over time, and the availability of dependencies and
+web-tools @Perkel2020-io.
 
@@
 
-In addition to shared code, data and containerisation, there are a few ways researchers can make their
-work more reproducible.
+In addition to shared code, data and containerisation, there are a few other ways researchers can make
+their work more reproducible.
 Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
@@
 multiple contributors.
-Literate programming using Rmarkdown or Jupyter notebooks provides optimal documentation by embedding
+Literate programming using R Markdown or Jupyter notebooks provides optimal documentation by embedding
 free text explanations together with the code and outputs such as charts and tables.
-Secondly, code and data can be linked. This means that the code knows the location of the data,
+Secondly, code and data can be linked. This means that the code "knows" the location of the data,
 avoiding any "file not found" errors @Peng2011-ia.
 Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
-for each part sequentially, making the entire manuscript reproducible by executing just one command
-and without any manual steps @Lasser2020-mm.
+for each part sequentially, making the entire manuscript reproducible by executing just one or a few
+commands, and without any manual steps @Lasser2020-mm.
 Taken together, computational reproducibility is a spectrum @Akalin2018-eg, and the *five pillars*
@@
 [@Slonim2002-sf; @Khatri2012-do].
+As of August 2020, this type of analysis was cited in 131,332 articles [@Xie2021-ch], illustrating
+how crucial these tools are in the interpretation of omics data.
+Moreover, analysis of related keywords in PubMed suggests the number of articles describing
+"enrichment analysis" is doubling every two years, as omics analysis becomes accessible to
+more researchers around the world.
+
 While enrichment analysis is undoubtedly a key method for the interpretation of large omics
@@
 
-A solution to this is to use containerisation to provide a scripted
-workflow with prescribed links to research data, external databases and complete record of parameter
-selection.
+A solution to this is to use containerisation to provide a complete scripted workflow with prescribed
+links to research data, external databases and complete record of parameter selection.
 When designed correctly, these containers encapsulate bioinformatics workflows and could ensure their
-computational reproducibility for decades to come with just one or a few simple commands.
+computational reproducibility for decades to come with a few simple commands.
 There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
 existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
-a lack of step-by-step guides and templates aimed at beginner users.
+a lack of step-by-step guides and templates aimed at non-expert bioinformaticians.
 
-We address this firstly by providing working templates for Docker image and Rmarkdown scripts for
+We address this firstly by providing working templates for Docker image and R Markdown scripts for
 common transcriptome and enrichment analysis routines.
@@
 However, before recommending or prescribing such workflows, it is necessary
-to assess directly whether current practices are adequate, by trying to reproduce the enrichment
+to assess *directly* whether current practices are adequate, by trying to reproduce the enrichment
 analyses of previously published studies, and examining whether the conclusions of those articles
@@
 upon reproduction, and the conclusions of the study were overall not supported.
-Articles were rated 2 when at least 50% of the terms are observed as significant after replication,
+Articles were rated 2 when at least 50% of the terms are observed as significant after replication
 according to the original study's significance threshold, and most conclusions were supported.
@@
 
-```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility in a sample of 20 journal articles."}
+```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility of enrichment analysis in a sample of 20 journal articles."}
 
@@
 Yes, only BP terms were taken into account so several CC and MF significant terms were not mentioned in the paper.
-|  **3.** Considering top 6 BP Terms and KEGG pathways.
+|  **3 (high).** Considering top 6 BP Terms and KEGG pathways.
 The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed, when there is a higher amount of significant GOs and pathways that are observed after replication.
@@
 Some terms using clusterprofiler in R were mentioned.
-| **2.** GO results from the original manuscript are somewhat similar to reproduced analysis.
+| **2 (medium).** GO results from the original manuscript are somewhat similar to reproduced analysis.
 The FDR values however are inconsistent.
@@
 N.A.
-| **2.** GO terms obtained after replication showed inconsistent results compared to the results in the original article.
+| **2 (medium).** GO terms obtained after replication showed inconsistent results compared to the results in the original article.
 KEGG results were similar.
@@
 N.A.
-| **1.** GO analysis from up-regulated and down-regulated proteins are less similar to the results given in the paper.
+| **1 (low).** GO analysis from up-regulated and down-regulated proteins are less similar to the results given in the paper.
 The FDR values for some pathways are highly inconsistent.
@@
 N.A.
-| **1.** This paper could have done proper labelling of supplementary files as it was difficult to find the gene list for DAVID analysis.
+| **1 (low).** This paper could have done proper labelling of supplementary files as it was difficult to find the gene list for DAVID analysis.
 Lacks information on when control genes are used and when citral-affected gene lists are used.
@@
 B.
-No | **3.** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” have changed.
+No | **3 (high).** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” have changed.
 Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate.
@@
 Yes, several terms from replication were not mentioned in the paper for GO Analysis.
-| **2.** Does not consider corrected p values.
+| **2 (medium).** Does not consider corrected p values.
 Overall, significant terms from GO analysis of up and down-regulated DEGs are missing upon replication.
@@
 There were many top significant terms that were not mentioned.
-| **1.** Narrowing/Filtering the given gene list is a required step that could have been mentioned.
+| **1 (low).** Narrowing/Filtering the given gene list is a required step that could have been mentioned.
 Many terms were not so significant anymore.
@@
 B.
-Yes, many | **1.** The significant pathways stated in the paper are not showing up once replicated.
+Yes, many | **1 (low).** The significant pathways stated in the paper are not showing up once replicated.
 There is no clarity of the up-regulated genes used for GO and KEGG analysis.
@@
 |
-| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1.** Replication results of GO terms are not at all compliant with the results shown in the paper. |
+| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1 (low).** Replication results of GO terms are not at all compliant with the results shown in the paper. |
 | 11.PMC6557785: Figure 5 A-B | A.
@@
 B.
-Yes | **2.** For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
+Yes | **2 (medium).** For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper.
 For KEGG Pathways, there are other pathways that are more significant and contain a higher number of genes enriched for example “Salmonella infection” that has a higher count of genes but is not mentioned in the paper.
 |
-| 12.PMC6561911: Figure 5 a and b | A. No. B. N.A. | **1.** The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done |
+| 12.PMC6561911: Figure 5 a and b | A. No. B. N.A. | **1 (low).** The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done |
 | 13.PMC6580941: Table 1 and Table 2 | A.
@@
 Significant terms like “nucleoplasm”, “RNA Binding”, “protein binding” are missing from the paper.
-| **2.** For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
+| **2 (medium).** For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper.
 The Gene list for input is not so clearly mentioned, it should have been labelled properly in the name of the supplemental file.
@@
 Yes, some.
-| **3.** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
+| **3 (high).** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results.
 They could have mentioned why they chose the type of sorting shown in the paper to display their results.
@@
 B.
-Yes for KEGG analysis | **2.** The GO term results of replication are compliant with Figure 5 c of the paper.
+Yes for KEGG analysis | **2 (medium).** The GO term results of replication are compliant with Figure 5 c of the paper.
 KEGG pathway results are semi-consistent.
@@
 B.
-Yes | **1.** The top 2 significant GO terms in the paper are not significant after replication analysis is done.
+Yes | **1 (low).** The top 2 significant GO terms in the paper are not significant after replication analysis is done.
 Other GO terms are the same as replicated results.
@@
 Yes, many.
-| **2.** The paper uses 44 sets of DEGs grouped in 4 major types.
+| **2 (medium).** The paper uses 44 sets of DEGs grouped in 4 major types.
 For replication, OFC Set 9 was used, DAVID GO analysis was done twice where a discrepancy of results was observed.
@@
 There was a difference in gene counts as well.
-| **2.** DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
+| **2 (medium).** DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper.
 From NERI - BAHN - KEGG Analysis: “Pathways of neurodegeneration - multiple diseases”, “Pathways in cancer” - pathways with the highest gene count (30 and 28 respectively) are not mentioned in the paper.
@@
 Yes.
-| **2.** The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
+| **2 (medium).** The top BP GO terms from the paper are no longer significantly higher in the replicated GO results.
 The top 2 KEGG Pathways after replication are “Human cytomegalovirus infection” followed by “Pathways in cancer”.
@@
 Yes, many!
-| **2.** For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
+| **2 (medium).** For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned.
 Most of the terms are mentioned in the replicated results, but different levels of enrichment.
@@
 
-: Table 1. Classification and justification for each article.
+: Table 1. Classification and justification of reproducibility scores for each article.
 
-Taken together, the ability to replicate DAVID enrichment analysis using the supplementary gene
+Taken together, the ability to reproduce DAVID enrichment analysis using the supplementary gene
 lists was highly variable.
+The reliability of enrichment analysis results would benefit by adopting best practices in computational
+research.
 
@@
 enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
-protocols.io with a permissive license.
-This protocol is in accordance with the *five pillars* including
+protocols.io with a permissive license [@Ziemann2023-kz].
+This protocol is in accordance with the *five pillars*, and includes
 version control with git and GitHub, compute environment control using Docker, literate programming
-with Rmarkdown, persistent code and data sharing with Zenodo and documentation of all of these to
+with R Markdown, persistent code and data sharing with Zenodo and documentation of all of these to
 empower others to reproduce the workflow.
 
-Central to this protocol is the provision of template Dockerfile and an Rmarkdown script, which
-are intended to be remixed and altered by the end user to suit the needs of their project.
+Central to this protocol is the provision of template Dockerfile and an R Markdown script, which
+are intended to be remixed and altered by end users to suit their project's needs.
 
 The Dockerfile uses a base image from Bioconductor which should correspond to the latest available
-stable release version (currently 3.16), running on R v4.2.2 and Ubuntu 22.04.
+stable release version, on the Ubuntu operating system.
 The image contains instructions for installing a few useful utilities including `nano` and `git` for
@@
 
-The Rmarkdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
-Rmarkdown allows the unification of extended descriptions, code and results into a single document,
+The R Markdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
+R Markdown allows the unification of extended descriptions, code and results into a single document,
 which is becoming ubiquitous in bioinformatics and data science more generally
-@Grolemund2017-zl p423.
+@Grolemund2017-zl.
 The ability to provide extended descriptions is helpful for providing contextualising information about
 the purpose of the work, detailed methods and interpretations derived from the analysis.
-The Rmarkdown script contains code chunks that perform specific tasks including fetching an example
+The R Markdown script contains code chunks that perform specific tasks including fetching an example
 data set from DEE2, inspecting data quality, differential expression and two types of enrichment
 analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 2**).
-For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
-@Zhao2023-ok.
 
@@
 
-The software versions used in this example analysis are provided in Table 2.
+For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
+[@Wijesooriya2021-lb; @Zhao2023-ok].
+An example checklist is shown in **Table 2**.
+
+| Reporting criteria | Method or resource used |
+| -- | -- |
+| Origin of gene sets | Reactome (2023-03-06) |
+| Tool used | ClusterProfiler v4.4.4 |
+| Statistical test used | hypergeometric test |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Selection Criteria | DESeq2 FDR<0.05 |
+| ORA directionality | Separate tests for up- and down-regulation |
+| Data availability | via dee2.io at accession SRP038101 |
+| Other parameters | Min gene set size of 10 |
+
+: Table 2. An example reporting checklist for enrichment analysis.
+
+The software versions used in this example analysis are provided in Table 3.
 
@@
 
-: Table 2. The software bundled in the example Docker image.
+: Table 3. The software bundled in the example Docker image.
 
-To reproduce this work is relatively quick and simple, requiring six commands and approximately four
-minutes to install Docker, pull and run the container, run the Rmarkdown script, exit the container
-and copy the newly created report from the container to the current working directory (Box 1).
+Reproducing this workflow is relatively quick and simple, requiring six commands and approximately four
+minutes starting from a base Linux install.
+The steps include installing Docker, pulling and running the container, running the R Markdown script,
+exiting the container and copy the newly created report from the container to the current working
+directory (Box 1).
 
@@
 <p style="text-align: center;">
-*Box 1. Reproduce a transcriptome and enrichment analysis with Docker.*
+*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.*
 </p>
@@
 We use this analysis to better understand the effect of DNA methyltransferase inhibitors on
-genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
+gene expression and elaborate upon previous findings [@Lund2014-et], which mainly focused on
 the relationship between RNA expression and DNA methylation, but did not show results of a
@@
 
-With ORA, we observed 201 up-regulated and 99 down-regulated gene sets.
+![Figure 3. Differential expression and enrichment analysis with the example workflow.
+(A) An MDS plot of gene expression data for the largest two principal components.
+Blue and pink points represent control and azacitidine-treated
+samples respectively.
+(B) An MA plot.
+The x-axis represents the log<sub>10</sub> base mean expression, and the y-axis represents the
+log<sub>2</sub> fold change.
+Points in red indicate genes with statistically significant differential expression (FDR<0.05), while
+those in grey indicate did not meet this threshold.
+(C) Bar plot of ORA-based enrichment test results.
+(D) Bar plot of FCS-based enrichment test results.
+(C,D) REACTOME gene sets with FDR>0.05 were discarded and the remaining gene sets were
+ranked by enrichment score (ES)
+The most extreme up- and down-regulated genes sets are shown (based on ES).
+Red bars indicate up-regulation and blue bars indicate down-regulation.
+(E) Euler diagrams of differentially regulated gene sets identified with ORA and FCS approaches.
+](../img/fig_example.png "Example enrichment analysis on")
+
+With ORA, we observed 201 up-regulated and 99 down-regulated gene sets (FDR<0.05).
 We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
@@
 
-After FCS, we identified 239 up-regulated gene sets and 158 down-regulated gene sets.
+After FCS, we identified 239 up-regulated gene sets and 158 down-regulated gene sets (FDR<0.05).
 FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA
 transport and defective homologous recombination repair (**Figure 3D**).
-Down-regulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF
-signaling.
+This finding is consistent with changes to the cell cycle observed in 5-azacytidine treated cells,
+where the proportion of cells in G2/M phase is reduced, and proxy measures of apoptosis appear
+slightly higher @Lund2014-et.
+
+Down-regulated pathways related to endosomal/vacuolar (antigen presentation) pathway, interferon
+signaling and TLR/TNF signaling, which was not described in the original publication @Lund2014-et.
+Interestingly, these pathways are reported to be *upregulated* in other cancer cell studies of DNA
+hypomethylating agents for similar durations [@Li2014-op; @Chiappinelli2015-tv; @Okoye-Okafor2019-iq].
 
@@
 
-![Figure 3. Differential expression and enrichment analysis with the example workflow.](../img/fig_example.png "Example enrichment analysis on")
-
 ## Concluding remarks
 
-Although our reproducibility study was relatively small, there are a few key observations.
+Although this reproducibility study is relatively small, there are a few key observations.
 We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
-In a way, this is surprising, given the simplicity of using DAVID, only
-requiring the user to provide a list of genes in a text box and select some options.
-These irregularities may have arisen inadvertently from several sources:
+In a way, this is unexpected, given the simplicity of using DAVID.
+Users provide a list of genes in a text box and select some options.
+There are several posible explanations for this inconsistency:
 
@@
 
-The observed lack of reproducibility might be expected given that substantial methodological flaws
-were present in 85% of enrichment analyses as part of a wider survey of articles of a similar age
-@Wijesooriya2022-li.
+This irreproducibility might be expected given that substantial methodological flaws and reporting
+deficiencies were present in 85% of enrichment analyses from which this set of 20 studies were
+drawn @Wijesooriya2022-li.
 
 While researching the literature on computational reproducibility in bioinformatics, we identified
-some case studies but only one other systematic examination of multiple journal articles.
+many case studies, but only two systematic examinations of multiple journal articles.
 Ionnidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
 published in 2005-2006, and found only two could be reproduced to a satisfactory level.
+A recent preprint reports that only 245 out of 9,625 Jupyter notebooks (2.5%) in biomedical journals
+successfully reproduced results identical to the respective journal article [@Samuel2022-dy].
 More work in this direction is urgently needed to understand the causes of irreproducible bioinformatics
 research so it can be remedied, as reproducibility is considered only the first step towards
-reliability.
-
-This problem is not limited to bioinformatics.
-In the field of drug development, according to a 2011 study by Bayer, only 25% of projects could be
-replicated to a high degree @Prinz2011-ht.
-When the US Biotech firm Amgen set out to reproduce 53 landmark cancer studies, they were able to
-reproduce only 6 (11%) @Begley2012-di.
-It is clear that as a field, the life sciences need to raise the rate of reproducibility or risk the
-public perceiving it as unreliable and wasteful.
+reliable research [@Nosek2022-hw].
 
@@
 However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
-protocol to guide them on how to customise template Dockerfile and Rmarkdown scripts to suit their
-project's needs.
+protocol to guide them on how to customise template Dockerfile and R Markdown scripts to suit their
+project's needs [@Ziemann2023-kz].
 Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
@@
 These are good for data exploration, to quickly identify trends that can be followed up with more
-rigorous methods.
-Moreover, a lot of work is being done to make such point-and-click tools more reproducible.
+rigorous and reproducible methods, such as this protocol.
+
+Significant progress is being made to make such point-and-click tools more reproducible.
 The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
@@
 The Degust differential expression suite is a graphical interface to popular
-expression tools like edgeR and limma, and allows users to download the script to facilitate future
-non-interactive reproducibility @Powell2019-lv.
-Such functionality could be written into existing enrichment analysis tools like DAVID, and together with
-retention of historical gene annotation sets could provide a "good enough" solution to reproducibility.
+expression tools like edgeR and limma, and allows users to download the computer script to facilitate
+future non-interactive reproducibility @Powell2019-lv.
+Such functionality could be written into existing enrichment analysis tools like DAVID, and
+could provide a "good enough" solution to reproducibility.
 
@@
 
-* Protocol: https://github.com/markziemann/enrichment_recipe/blob/main/docs/guide_for_linux_users_with_docker.md
-A final version will be deposited to protocols.io.
+* Protocol: can be found at protocols.io @Ziemann2023-kz.
 
-* Code repository including template Dockerfile and Rmarkdown script: https://github.com/markziemann/enrichment_recipe
+* Code repository including template Dockerfile and R Markdown script: https://github.com/markziemann/enrichment_recipe
 

### 2023-06-29 manuscript/manuscript.Rmd
fixing bibliography

@@
 [@Slonim2002-sf; @Khatri2012-do].
-As of August 2020, this type of analysis was cited in 131,332 articles [@Xie2021-ch], illustrating
-how crucial these tools are in the interpretation of omics data.
-Moreover, analysis of related keywords in PubMed suggests the number of articles describing
+As of August 2020, this type of analysis was cited 131,332 times [@Xie2021-ch].
+The most popular tool for enrichment analysis appears to be DAVID [@Sherman2022-hf].
+Its 11 related journal articles have received a total of 37,035 citations from 27,129
+separate PubMed articles (as at 29th June 2023).
+Examination of related keywords in PubMed suggests the number of articles describing
 "enrichment analysis" is doubling every two years, as omics analysis becomes accessible to
@@
 
-While enrichment analysis is undoubtedly a key method for the interpretation of large omics
-data sets, there are concerns that inadvertent misuse of these methods leads to
-unreliable results.
+These observations highlight that these tools are essential in the interpretation of omics data.
+However, there are concerns that inadvertent misuse of these methods leads to unreliable results.
 These potential problems include (i) the lack of a correct background for over-representation tests,
@@
 The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles using DAVID version 6.8 (March-May 2022).
+sections of the 20 articles using DAVID version 6.8 (March-May 2022) [@Sherman2022-hf].
 

### 2023-06-30 example2.Rmd
code edits

--- 2023-06-22 example.Rmd
@@
 ---
-title: "An example of extremely reproducible enrichment analysis"
+title: "A second example of extremely reproducible enrichment analysis"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 For this guide I will be using bulk RNA-seq data from a previous study, which is
-deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
-The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.
+deposited at NCBI GEO with [accession GSE93236](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE93236).
 
-The raw data have been processed by the DEE2 project, and the summary gene expression counts are
-available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
-(Ziemann et al, 2019).
+The experiment is designed to investigate the effect of knowckdown of the histone methyltransferase set7.
 
-The gene counts have also been deposited to the `/data` folder in the `example.Rdata` file in case
-the DEE2 resource becomes unavailable.
-To import it, use the command: `load("data/example.Rdata")`
-
-Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
-storage drive.
+We will be fetching the gene expression count matrix directly from NCBI GEO for downstream analysis.
+Link: https://ftp.ncbi.nlm.nih.gov/geo/series/GSE93nnn/GSE93236/suppl/GSE93236_Keating_Set7_KD_counts.tsv.gz
 
@@
 
-mdat <- getDEE2Metadata("hsapiens")
+download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE93nnn/GSE93236/suppl/GSE93236_Keating_Set7_KD_counts.tsv.gz",
+  destfile="GSE93236_Keating_Set7_KD_counts.tsv.gz")
 
-# get sample sheet
-ss <- subset(mdat,SRP_accession=="SRP038101")
+x <- read.table("GSE93236_Keating_Set7_KD_counts.tsv.gz")
 
-# fetch the whole set of RNA-seq data
-x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
-mx <- x$GeneCounts
-rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
+dim(x)
+
+# there is a row we need to remove
+mx <- x[rownames(x)!="NOTINBED",]
+
+head(mx)
+
 dim(mx)
 
-# aza no filtering
-ss$trt <- grepl("Treated",ss$Experiment_title)
+```
+
+The row names contains an underscore that we need to remove.
+
+```{r,geneid_fix}
+
+rownames(mx) <- sub("_"," ",rownames(mx))
+
+head(mx)
+
+```
+
+Curate sample sheet from the colnames.
+
+```{r,samplesheet1}
+
+ss <- as.data.frame(colnames(x))
+ss$trt <- as.numeric(grepl("Set7KD",colnames(x)))
 
 ss %>%
-  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
+  kbl(caption="sample sheet for Set7KD experiment") %>%
   kable_paper("hover", full_width = F)
@@
 
-cols <- as.character(grepl("Treated",ss$Experiment_title))
-cols <- gsub("FALSE","lightblue",cols)
-cols <- gsub("TRUE","pink",cols)
+cols <- as.character(ss$trt)
+cols <- gsub("0","lightblue",cols)
+cols <- gsub("1","pink",cols)
 plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
@@
   cex=0.6,pch=19,col="darkgray",
-  main="5-azacitidine treatment in AML3 cells",
+  main="Set7KD in endothelial cells",
   xlab="log10(basemean)",ylab="log2(fold change)")
@@
 | ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
+| Data availability | GEO under accession GSE93236 |
 | Other parameters | Min gene set size of 10 |
@@
 To fix it is relatively easy from the user's point of view.
-Just create a new gene set that consists of all detected genes and
+Just create a new gene set that consists of all detected genes and add it
+to the main gene set data frame.
 

### 2023-07-17 manuscript/manuscript.Rmd
ms edits

@@
 library("DiagrammeR")
+library("RColorBrewer")
 
@@
 their work more reproducible.
-Firstly, thorough documentation makes the reproducibility process a great deal easier and more accessible
-to novice bioinformaticians.
+Thorough documentation makes the reproducibility process a great deal easier and more
+accessible, especially to novice bioinformaticians.
 Version control keeps track of code and documentation changes over time and manages the edits made by
@@
 free text explanations together with the code and outputs such as charts and tables.
-Secondly, code and data can be linked. This means that the code "knows" the location of the data,
+Code and data can be linked. This means that the code "knows" the location of the data,
 avoiding any "file not found" errors @Peng2011-ia.
@@
 Although web-based tools are convenient, they are a potential reproducibility concern.
-Firstly, incomplete reporting of methods, options and parameters is common.
-Secondly, algorithms and functional annotation sets undergo regular updates, and previous versions are
+Incomplete reporting of methods, options and parameters is common.
+Algorithms and functional annotation sets undergo regular updates, and previous versions are
 typically not made available, hampering later reproducibility.
@@
 We followed the steps indicated in the respective papers to select appropriate gene sets investigated
-in the articles such as gene ontology (biological process, molecular function), KEGG pathways, etc.
+in the articles such as gene ontology (GO; biological process, molecular function), Kyoto Encyclopedia
+of Genes and Genomes (KEGG) pathways, etc.
 Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
@@
 
-The data presented in the paper was compared with the results we obtained, recording any
+The results presented in the paper were compared to those we obtained, recording any
 discrepancies.
 Additionally, the statements made in the abstract, results, and discussion sections were carefully
-examined and cross-checked with our results to understand whether those assertions survive reproduction
+examined and cross-checked with our results to understand whether those assertions survived reproduction
 or not.
@@
 
-### Implementation
+### Method overview
 
@@
 empower others to reproduce the workflow.
-
 Central to this protocol is the provision of template Dockerfile and an R Markdown script, which
 are intended to be remixed and altered by end users to suit their project's needs.
+The general steps involved are shown in **Figure 2**.
 
-The Dockerfile uses a base image from Bioconductor which should correspond to the latest available
+```{r,method_diagram,echo=FALSE,fig.height=6,fig.cap = "Figure 2. Flow chart of example workflow. An estimated amount of time to conduct each step is given."}
+
+library("RColorBrewer")
+library("shape")
+
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xaxs="i", yaxs="i", xlim=c(0, 11), ylim=c(-3, 8.5))
+
+polygon(c(0,6,6,0),c(8.5,8.5,7.5,7.5), col=cols2[1], border = NA)
+text( 3, 8.0, lab="1. Install Docker: 7 mins",cex=1.5)
+
+polygon(c(0.5,7.5,7.5,0.5),c(7.0,7.0,6.0,6.0), col=cols2[2], border = NA)
+text( 3.9, 6.5, lab="2. Validate example: 14 mins",cex=1.5)
+
+polygon(c(1,8.5,8.5,1),c(5.5,5.5,4.5,4.5), col=cols2[3], border = NA)
+text( 4.8, 5.0, lab="3. Create GitHub repo: 29 mins",cex=1.5)
+
+polygon(c(1.5,10.5,10.5,1.5),c(4,4,3,3), col=cols2[4], border = NA)
+text( 6.1, 3.5, lab="4. Build custom Docker image: 45 mins",cex=1.5)
+
+polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
+text( 6.3, 2, lab="5. Customize R Markdown: 3-12 hrs",cex=1.5)
+
+polygon(c(2.5,9.5,9.5,2.5),c(1,1,0,0),col="lightgray", border = NA)
+text( 6.1, 0.5, lab="6. Update repository: 70 mins",cex=1.5)
+
+polygon(c(3,11,11,3),c(-0.5,-0.5,-1.5,-1.5), col=cols2[1], border = NA)
+text( 7.0, -1, lab="7. Update Docker image: 40 mins",cex=1.5)
+
+polygon(c(3.5,12,12,3.5),c(-2,-2,-3,-3), col=cols2[2], border = NA)
+text( 7.6, -2.5, lab="8. Long Term Archiving: 180 mins",cex=1.5)
+
+#Arrows(2, 7.5, 2.5, 7.0, lwd=2, arr.type="triangle")
+#Arrows(2.5, 6, 3, 5.5, lwd=2, arr.type="triangle")
+#Arrows(3, 4.5, 3.5, 4, lwd=2, arr.type="triangle")
+#Arrows(3.5, 3, 4, 2.5, lwd=2, arr.type="triangle")
+#Arrows(4, 1.5, 4.5, 1, lwd=2, arr.type="triangle")
+
+```
+
+The Dockerfile uses a base image from Bioconductor that corresponds to the latest available
 stable release version, on the Ubuntu operating system.
@@
 differential expression with `DESeq2` @Love2014-hd,
-enrichment analysis with `clusterprofiler`, `fgsea` and `mitch`
+enrichment analysis with `clusterProfiler`, `fgsea` and `mitch`
 [@Yu2012-ts; @Korotkevich2016-ls; @Kaspi2020-dq].
@@
 data set from DEE2, inspecting data quality, differential expression and two types of enrichment
-analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 2**).
+analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 3**).
 
-```{r,diagram,echo=FALSE,fig.cap = "Figure 2. Flow chart of example workflow."}
+```{r,diagram2,echo=FALSE,fig.cap = "Figure 3. Flow chart of example workflow."}
 
@@
 | Origin of gene sets | Reactome (2023-03-06) |
-| Tool used | ClusterProfiler v4.4.4 |
+| Tool used | `clusterProfiler` v4.4.4 |
 | Statistical test used | hypergeometric test |
@@
 | BioC/clusterProfiler | 4.6.1 | over-representation analysis |
-| Bioc/mitch | 1.10.0 | multi-contrast enrichment |
+| BioC/mitch | 1.10.0 | multi-contrast enrichment |
 
@@
 
-Reproducing this workflow is relatively quick and simple, requiring six commands and approximately four
+Reproducing this workflow is relatively quick and simple, requiring six commands and approximately 4-7
 minutes starting from a base Linux install.
@@
 A multi-dimensional scaling plot showed clustering of samples into their treatment
-groups, suggesting a robust effect of the chemical treatment (**Figure 3A**).
+groups, suggesting a robust effect of the chemical exposure (**Figure 4A**).
 Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
 1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
-(**Figure 3B**).
+(**Figure 4B**).
 Up and down-regulated gene sets were subjected to ORA separately, using the list of
@@
 
-![Figure 3. Differential expression and enrichment analysis with the example workflow.
+![Figure 4. Differential expression and enrichment analysis with the example workflow.
 (A) An MDS plot of gene expression data for the largest two principal components.
@@
 were associated with interferon signaling, interleukin signaling and NADPH oxidase
-(host defence) (**Figure 3C**).
+(host defence) (**Figure 4C**).
 
@@
 FCS up-regulated pathways were associated with prometaphase chromosome condensation, nuclear RNA
-transport and defective homologous recombination repair (**Figure 3D**).
+transport and defective homologous recombination repair (**Figure 4D**).
 This finding is consistent with changes to the cell cycle observed in 5-azacytidine treated cells,
@@
 Jaccard index of 0.58, while this agreement was lower for down-regulated pathways, with a Jaccard
-index of 0.46 (**Figure 3E**).
+index of 0.46 (**Figure 4E**).
 

### 2023-07-20 manuscript/manuscript.Rmd
video playlist provided
minor edits

@@
 
+We thank Ms Megan Soria (Deakin University) for discussions on the concept and manuscript.
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research
@@
 
+* Instructional video guides are available on YouTube: https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db
+
 * Code repository including template Dockerfile and R Markdown script: https://github.com/markziemann/enrichment_recipe
@@
 
+* The code repository and Docker image have been uploaded to Zenodo for lon-term archiving and are available from: [TBA].
+
 ## Bibliography

### 2023-07-20 manuscript/manuscript.Rmd
tidy and edits for F1000 Res

@@
 title: "A recipe for extremely reproducible enrichment analysis (and why we need it)"
-author: "Mark Ziemann and Anusuiya Bora"
 date: "`r Sys.Date()`"
@@
 
-## Abstract
+# Abstract
 
@@
 findings of 20 open-access articles describing enrichment analysis were reproducible.
-We find that only three articles exhibited a high degree of concordance, while seven exhibited poor
-reproducibility.
+We find that only three articles exhibited a high degree of concordance, while seven exhibited major
+discrepancies.
 The findings indicate that current enrichment analysis practices and reporting are inadequate.
@@
 reproducibility.
+
+# Keywords
+
+* Bioinformatics
+
+* Reproducibility
+
+* Enrichment analysis
+
+* Pathway analysis
+
+* Gene set enrichment analysis
+
+* Docker
+
+# Main Body
 
@@
 
+# Data and Software Availability
+
+## Underlying data
+
+## Availability of materials
+
+* Protocol: can be found at protocols.io @Ziemann2023-kz.
+
+* Instructional video guides are available on YouTube:
+https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db
+
+## Software and Code
+
+* Code repository including template Dockerfile and R Markdown script:
+https://github.com/markziemann/enrichment_recipe
+
+* Example Docker image: https://hub.docker.com/r/mziemann/enrichment_recipe
+
+* The code repository and Docker image have been uploaded to Zenodo for lon-term archiving and are
+available from: [TBA].
+
 ## Acknowledgements
@@
 
-## Availability of materials
-
-* Protocol: can be found at protocols.io @Ziemann2023-kz.
-
-* Instructional video guides are available on YouTube: https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db
-
-* Code repository including template Dockerfile and R Markdown script: https://github.com/markziemann/enrichment_recipe
-
-* Example Docker image: https://hub.docker.com/r/mziemann/enrichment_recipe
-
-* The code repository and Docker image have been uploaded to Zenodo for lon-term archiving and are available from: [TBA].
-
 ## Bibliography

### 2023-07-21 manuscript/manuscript.Rmd
minor edits for journal style

@@
 | Ubuntu | 22.04 | Operating System |
-| R | 4.2.2 | Statistical computing language |
+| R | 4.3.0 | Statistical computing language |
 | nano | 6.2 | text editing |
@@
 | R/eulerr | 7.0.0 | Venn diagrams |
-| BioC/getDEE2 | 1.8.0 | fetch example RNA-seq data |
-| BioC/DESeq2 | 1.38.3 | differential expression |
-| BioC/fgsea | 1.24.0 | functional class scoring |
-| BioC/clusterProfiler | 4.6.1 | over-representation analysis |
+| BioC/getDEE2 | 1.6.0 | fetch example RNA-seq data |
+| BioC/DESeq2 | 1.36.0 | differential expression |
+| BioC/fgsea | 1.22.0 | functional class scoring |
+| BioC/clusterProfiler | 4.4.4 | over-representation analysis |
 | BioC/mitch | 1.10.0 | multi-contrast enrichment |
@@
 
-## Availability of materials
+Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io).
+Expression data from untreated and Aza treated AML3 cells.
+Accession Number SRP038101.
 
-* Protocol: can be found at protocols.io @Ziemann2023-kz.
-
-* Instructional video guides are available on YouTube:
-https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db
+Data are available under the terms of the GNU General Public License v3.0.
 
@@
 
-* Code repository including template Dockerfile and R Markdown script:
-https://github.com/markziemann/enrichment_recipe
+* The code repository including template Dockerfile and R Markdown script are available on GitHub
+(https://github.com/markziemann/enrichment_recipe).
 
-* Example Docker image: https://hub.docker.com/r/mziemann/enrichment_recipe
+* The example Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/enrichment_recipe).
 
-* The code repository and Docker image have been uploaded to Zenodo for lon-term archiving and are
+* The code repository and Docker image have been uploaded to Zenodo for long-term archiving and are
 available from: [TBA].
+
+## Availability of other materials
+
+* The full protocol can be found at protocols.io
+(https://www.protocols.io/view/a-recipe-for-extremely-reproducible-enrichment-ana-j8nlkwpdxl5r/v2) @Ziemann2023-kz.
+
+* Instructional video guides are available on YouTube
+(https://www.youtube.com/playlist?list=PLAAydBPtqFMXDpLa796q7f7W1HK4t_6Db).
+
+## Author Contributions
+
+Conceptualization: MZ, AB.
+Data Curation: MZ.
+Formal Analysis: MZ.
+Funding Acquisition: N/A.
+Investigation: MZ, AB.
+Methodology: MZ, AB.
+Project Administration: MZ.
+Resources: MZ.
+Software: MZ, AB.
+Supervision: MZ.
+Validation: AB.
+Visualization: MZ, AB.
+Writing – Original Draft Preparation: MZ, AB.
+Writing – Review & Editing: MZ, AB.
+
+## Competing Interests
+
+No competing interests were disclosed.
+
+## Grant Information
+
+The authors declared that no grants were involved in supporting this work.
 
@@
 
-## Bibliography
+# Bibliography

### 2023-11-30 manuscript/manuscript.Rmd
added zenodo link
progress

@@
 
-[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
-[Anusuiya Bora<sup>2</sup>](https://orcid.org/0009-0006-2908-1352)
+[Anusuiya Bora<sup>1,2,3</sup>](https://orcid.org/0009-0006-2908-1352),
+[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)
 
@@
 
-1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.
+1. Burnet Institute, Melbourne, Australia.
 
-2. Vellore Institute of Technology, Vellore, India.
+2. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.
+
+3. Vellore Institute of Technology, Vellore, India.
 
@@
 We sought to determine whether this translates into irreproducibility by examining whether the
-findings of 20 open-access articles describing enrichment analysis were reproducible.
+findings of 20 open-access articles published in 2019 describing enrichment analysis with the
+popular DAVID suite were reproducible.
 We find that only three articles exhibited a high degree of concordance, while seven exhibited major
-discrepancies.
-The findings indicate that current enrichment analysis practices and reporting are inadequate.
-To address this, we provide a set of template scripts and a step-by-step protocol to adapt them to
-a new project's enrichment analysis needs.
-This recipe is built according to the "five pillars" framework, which facilitates extreme
-reproducibility.
+discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
+As the tool version used is no longer available, all articles using this tool pre-2021 (~20,800 studies)
+including this sample of 20 articles are no longer able to be reproduced with the original tools.
+Based on this, we suggest that results from web-based tools without long-term preservation features should
+not be included in scientific publications due to the threat of link decay and short reproducibility horizon.
+We advocate for the adoption of extremely reproducible research workflows, and we provide a detailed protocol
+for how to achieve it using a combination of best practices including literate programming, version control,
+containerisation, documentation and persistent sharing of data and software.
 
@@
 commands, and without any manual steps @Lasser2020-mm.
-Taken together, computational reproducibility is a spectrum @Akalin2018-eg, and the *five pillars*
-framework has been described as a way of achieving extreme reproducibility @Ziemann2023-pt.
+Taken together, computational reproducibility is a spectrum, and the *five pillars*
+framework has been described as a way of achieving extreme reproducibility @Ziemann2023-dg.
 Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
@@
 [@Slonim2002-sf; @Khatri2012-do].
-As of August 2020, this type of analysis was cited 131,332 times [@Xie2021-ch].
+As of August 2020, the various types of enrichment analysis were collectively cited 131,332 times [@Xie2021-ch].
 The most popular tool for enrichment analysis appears to be DAVID [@Sherman2022-hf].
@@
 
-These observations highlight that these tools are essential in the interpretation of omics data.
+These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
 However, there are concerns that inadvertent misuse of these methods leads to unreliable results.
@@
 @Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
-organism under study is human; (2) the tool used was DAVID Version 6.8 (Database for Annotation,
+organism under study is human; (2) the tool used was DAVID (Database for Annotation,
 Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
@@
 The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles using DAVID version 6.8 (March-May 2022) [@Sherman2022-hf].
+sections of the 20 articles using the DAVID version specified in the articles (March-May 2022) [@Sherman2022-hf].
 
@@
 the 20 articles were classified based on the degree of reproducibility (**Figure 1**).
-Six studies were classified as low reproducibility, while 10 were classified as medium and just three
+Seven studies were classified as low reproducibility, while 10 were classified as medium and just three
 were classified as high.
-Justifications for these classifications are provided in Table 1.
+Justifications for these classifications are provided in Table S1.
 
@@
 
-: Table 1. Classification and justification of reproducibility scores for each article.
+: Table S1. Classification and justification of reproducibility scores for each article.
 
@@
 lists was highly variable.
-The reliability of enrichment analysis results would benefit by adopting best practices in computational
-research.
+As DAVID version 6.8 has been retired, these and approximately 20,000 other PubMed articles have enrichment analyses
+that are no longer reproducible with the original tools.
+These observations highlight the need for better methodological reporting to reduce ambiguity in methods sections,
+and suggest that web-based tools without long-term reproducibility features are no longer suitable for inclusion in
+scientific publications.
 
-## A container-based approach to addressing reproducibility of enrichment analysis
+## A five-pillars approach to addressing reproducibility of enrichment analysis
 
@@
 protocols.io with a permissive license [@Ziemann2023-kz].
-This protocol is in accordance with the *five pillars*, and includes
+This protocol is prepared in accordance with the *five pillars* principles, and includes
 version control with git and GitHub, compute environment control using Docker, literate programming
@@
 | Data availability | via dee2.io at accession SRP038101 |
-| Other parameters | Min gene set size of 10 |
+| Other parameters | Minimum gene set size of 10 |
 
@@
 
-Reproducing this workflow is relatively quick and simple, requiring six commands and approximately 4-7
+Reproducing the example workflow is relatively quick and simple, requiring six commands and approximately 4-7
 minutes starting from a base Linux install.
@@
 
-We also want to emphasise that there is still a definite need for graphical and web-based tools.
-These are good for data exploration, to quickly identify trends that can be followed up with more
-rigorous and reproducible methods, such as this protocol.
+With the retirement of DAVID 6.8 and earlier versions, which occurred in June 2022, all DAVID analyses
+prior to 2021 have become irreproducible, including the sample of 20 studies featured here.
+This is concerning, as these studies were only three years old at the time of the tool's retirement.
+Moreover, according to PubMed citations, the DAVID project accrued 27,572 citations before 2021 from 20,878
+PubMed articles, suggesting that ~20,000 omics studies can no longer be reproducible with the original tools.
 
-Significant progress is being made to make such point-and-click tools more reproducible.
+With this, we must reconsider whether web-based tools are valid for scientific publications if they are at risk
+of becoming unavailable due to link decay.
+Some progress is being made to make web-based point-and-click tools more reproducible.
 The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
@@
 future non-interactive reproducibility @Powell2019-lv.
-Such functionality could be written into existing enrichment analysis tools like DAVID, and
-could provide a "good enough" solution to reproducibility.
+The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
+downloads for all past versions[@Ge2020-qe], which aids future reproducibility.
+Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"
+approach detailed here.
 
@@
 articles as opposed to individuals who publish fewer but more rigorously conducted works.
-Solutions to this could include mandates by funding agencies or institutions @Garijo2013-zb,
+Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
 badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
@@
 
-* The code repository and Docker image have been uploaded to Zenodo for long-term archiving and are
-available from: [TBA].
+* The code repository and Docker image have been uploaded to Zenodo for long-term archiving
+(https://zenodo.org/record/8170984).
 

### 2023-12-01 manuscript/letter.Rmd
letter

---
title: "No publication without preservation: the argument against web-based tools in genomics"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

Alt title: "Time to abandon (most) web-based tools in genomics"

[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)

**Affiliations**

1. Burnet Institute, Melbourne, Australia.

2. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

(*) Corresponding author: mark.ziemann@burnet.edu.au

According to a 2016 survey, 52% of researchers believe we are in a reproducibility crisis [@Baker2016-hm].
In genomics we don't have a lot of data on study reproducibility, but a small survey from 2009 indicates that most
studies are irreproducible due to a lack of shared data, software and documentation [@Ioannidis2009-hv].
The types of tools we use for the analysis also play a role.
Web-based point-and-click tools have become very popular in genomics for the analysis of gene lists, to
interpret whether certain biochemical or signaling pathways are over-represented in a particular genomic study.
These tools are so popular because they do not require the installation of any software, and a gene list analysis
can be conducted in a matter of seconds.
They also require little training to use, which on one hand simplifies bioinformatics tasks, but on the other hand
these tools can easily be misused and results misreported, which can lead to misleading conclusions
[ @Timmons2015-ex ; @Wijesooriya2022-li].
Last year we sought to test whether a small group of articles published in 2019 involving enrichment analysis with the DAVID tool [@Sherman2022-hf]
were reproducible using the authors' own methods.
We found only 3/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies.
After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
This was disturbing, as according to analysis of PubMed citations, some 20,000 articles citing DAVID enrichment tools will no longer be reproducible with
the original tools, some articles just one or two years old.
This is a prime example of "link decay", a phenomenon where internet based resources are lost over time, which has been raised as a significant and ongoing problem for bioinformatic reproducibility [@Hennessey2013-tj].

Therefore, we caution against using data from web-based tools for publication, unless there is a system in place to enable future reproducibility.
Scripted workflows are perhaps the most suitable solution for fostering long-term reproduction, but require expertise in computer programming.
We have written a protocol for extremely reproducible enrichment analysis in R, designed for novice bioinformaticians to fill in this gap [@Ziemann2023-kz], but we understand this isn't for everyone.
ShinyGO is another potential remedy, as it provides most of the necessary features in a point-and-click web interface but also allows users to download all historical versions of the tool to be run as
a Docker based web page on the users own hardware [@Ge2020-qe].
In order to get genomics research to meet the ten-year reproducibility challenge [@Perkel2020-io], we need progress towards better tooling, a greater emphasis on investigator training, institutional support,
funder mandates, stronger publication criteria, and more incentives for rigourous and reliable practices [@Lewis2016-qh; @Diong2021-ma].

## Competing Interests

No competing interests were disclosed.

## Bibliography

### 2023-12-02 manuscript/letter.Rmd
update with data retention mandates

@@
 
-Therefore, we caution against using data from web-based tools for publication, unless there is a system in place to enable future reproducibility.
-Scripted workflows are perhaps the most suitable solution for fostering long-term reproduction, but require expertise in computer programming.
+Researchers should keep in mind that institutional and funding mandates for data retention also apply to software and algorithms [@NHMRC2019-wu ; @Harvard_2023-cl].
+Therefore, to remain compliant with these mandates and support reproducibility, we caution against using data from web-based tools for publication unless there is a system in place to enable future reproducibility.
+Scripted workflows are perhaps the most suitable solution for enabling reproduction in the long-term, but require expertise in computer programming.
 We have written a protocol for extremely reproducible enrichment analysis in R, designed for novice bioinformaticians to fill in this gap [@Ziemann2023-kz], but we understand this isn't for everyone.

### 2023-12-03 manuscript/letter.Rmd
update title

@@
 ---
-title: "No publication without preservation: the argument against web-based tools in genomics"
+title: "How vanishing web-based tools hamper reproducibility"
 date: "`r Sys.Date()`"
@@
 ---
+
+Alt title: "No publication without preservation: the argument against web-based tools in genomics"
 
@@
 a Docker based web page on the users own hardware [@Ge2020-qe].
-In order to get genomics research to meet the ten-year reproducibility challenge [@Perkel2020-io], we need progress towards better tooling, a greater emphasis on investigator training, institutional support,
-funder mandates, stronger publication criteria, and more incentives for rigourous and reliable practices [@Lewis2016-qh; @Diong2021-ma].
+In order to get genomics research to meet the ten-year reproducibility challenge [@Perkel2020-io], we need progress towards better tooling, a greater emphasis on investigator training, better institutional support,
+clearer funder mandates, stronger publication criteria, and more incentives for rigourous and reliable practices [@Lewis2016-qh; @Diong2021-ma].
 
@@
 
+## Acknowledgements
+
+We thank Ms Claudia Beyer for advice.
+
 ## Bibliography

### 2023-12-04 manuscript/manuscript.Rmd
ack
letter
mandates

@@
 
-(*) Corresponding author: m.ziemann@deakin.edu.au
+(*) Corresponding author: mark.ziemann@burnet.edu.au
 
@@
 not be included in scientific publications due to the threat of link decay and short reproducibility horizon.
+Relying exclusively on webtools for analysis may also be in breach of institutional and funder data management mandates.
 We advocate for the adoption of extremely reproducible research workflows, and we provide a detailed protocol
@@
 platform supported by the NCRIS-funded Australian Research Data Commons (ARDC).
+The authors gratefully acknowledge the contribution to this work of the
+Victorian Operational Infrastructure Support Program received by the Burnet Institute.
 

### 2023-12-04 manuscript/manuscript.Rmd
mandates2

@@
 We followed the steps indicated in the respective papers to select appropriate gene sets investigated
-in the articles such as gene ontology (GO; biological process, molecular function), Kyoto Encyclopedia
+in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
 of Genes and Genomes (KEGG) pathways, etc.
@@
 |
-| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1 (low).** Replication results of GO terms are not at all compliant with the results shown in the paper. |
+| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1 (low).** Replication results of GO terms are not at all similar with the results shown in the paper. |
 | 11.PMC6557785: Figure 5 A-B | A.
@@
 protocols.io with a permissive license [@Ziemann2023-kz].
-This protocol is prepared in accordance with the *five pillars* principles, and includes
+This protocol is prepared in accordance with the *five pillars* principles [@Ziemann2023-dg], and includes
 version control with git and GitHub, compute environment control using Docker, literate programming
@@
 [@Yu2012-ts; @Korotkevich2016-ls; @Kaspi2020-dq].
-It also contains instructions to clone a github repository, which contains the scripts that run the
+It also contains instructions to clone a GitHub repository, which contains the scripts that run the
 analytical workflow.
@@
 | Origin of gene sets | Reactome (2023-03-06) |
-| Tool used | `clusterProfiler` v4.4.4 |
+| Tool used | `clusterProfiler` |
 | Statistical test used | hypergeometric test |
@@
 Points in red indicate genes with statistically significant differential expression (FDR<0.05), while
-those in grey indicate did not meet this threshold.
+those in grey did not meet this threshold.
 (C) Bar plot of ORA-based enrichment test results.
@@
 In a way, this is unexpected, given the simplicity of using DAVID.
-Users provide a list of genes in a text box and select some options.
+Users provide a list of gene idetifiers in a text box and select some options.
 There are several posible explanations for this inconsistency:
@@
 many case studies, but only two systematic examinations of multiple journal articles.
-Ionnidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
+Ioannidis et al @Ioannidis2009-hv examined a set of 10 gene expression studies with available data
 published in 2005-2006, and found only two could be reproduced to a satisfactory level.
@@
 PubMed articles, suggesting that ~20,000 omics studies can no longer be reproducible with the original tools.
+
+The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly describes a decade ago [@Hennessey2013-tj].
+We are worried that a reliance on web tools can leave researchers in violation of their data management responsibilities, as preservation mandates include software as well as original research data [@NHMRC2019-wu ; @Harvard_2023-cl ].
 

### 2023-12-05 manuscript/manuscript.Rmd
edits
Update manuscript.Rmd
Changes to PMC6368841

@@
 | 2.PMC6368841: Figure 3 A - D | A.
-Only GO Terms were the same as the paper.
-KEGG terms were highly inconsistent.
+A.
+Only GO Terms were the similar as the paper.
+The FDR values are inconsistent.
+However, the paper should have clearly mentioned the different tools under figure captions to bring more clarity of the terms obtained by respective tools like DAVID and ClusterProfiler.
+The KEGG terms do not come up in DAVID at all.
 B.
-When replicated, KEGG Terms (using DAVID) were not mentioned in the results of the paper.
-Shows inconsistency due to the usage of the server.
-Some terms using clusterprofiler in R were mentioned.
-| **2 (medium).** GO results from the original manuscript are somewhat similar to reproduced analysis.
+No | **3 (high).** 3 (high).
+The article receives a score of 3 solely based on GO results.
+GO results from the original manuscript are somewhat similar to reproduced analysis in DAVID.
 The FDR values however are inconsistent.
-KEGG results were highly different.
 The list of genes as input into DAVID analysis is unclear.
-They should have mentioned that both up and down-regulated proteins were considered.
+The paper should have mentioned that both up and down-regulated proteins were considered for DAVID analysis.
+Why were two different tools used for GO and KEGG when both can be performed in DAVID or clusterprofiler?
 |
@@
 
-We thank Ms Megan Soria (Deakin University) for discussions on the concept and manuscript.
+We thank Ms Megan Soria and Ms Kaumadi Wijesooriya (Deakin University) for discussions on the concept and manuscript.
 This research was supported by use of the Nectar Research Cloud, a collaborative Australian research

### 2023-12-05 manuscript/manuscript.Rmd
Update manuscript.Rmd
Update manuscript.Rmd
PMC6381667 changes - DAVID and IPA

--- 2023-12-04 manuscript/manuscript.Rmd
@@
 Better naming convention for supplementary file for 117 genes for DAVID input is necessary.
-GO Results are not reproducible and are highly inconsistent.
-|
+The paper mentions pathway analysis was done using  Ingenuity Pathway Analysis (IPA).
+Even DAVID provides pathway enrichment results.
+It is unclear which figures and results were obtained from either of the tools.
+Figures should have mentioned the name of the tools to avoid confusion for source of results.|
 | 4.PMC6405693: Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways” | A.

### 2023-12-05 manuscript/manuscript.Rmd

--- 2023-12-04 manuscript/manuscript.Rmd
@@
 score1 <- c("PMC6405693","PMC6425008","PMC6535219","PMC6539328","PMC6542760","PMC6561911","PMC6663624")
-score2 <- c("PMC6368841","PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
+score2 <- c("PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
   "PMC6582306","PMC6333352","PMC6526186","PMC6607402")
-score3 <- c("PMC6349697","PMC6444048","PMC6587650")
+score3 <- c("PMC6349697","PMC6444048","PMC6587650", "PMC6368841")
 l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)

### 2023-12-05 manuscript/manuscript.Rmd

--- 2023-12-05 manuscript/manuscript.Rmd
@@
 | 2.PMC6368841: Figure 3 A - D | A.
-Only GO Terms were the same as the paper.
-KEGG terms were highly inconsistent.
+Only GO Terms were the similar as the paper.
+The FDR values are inconsistent.
+However, the paper should have clearly mentioned the different tools under figure captions to bring more clarity of the terms obtained by respective tools like DAVID and ClusterProfiler.
+The KEGG terms do not come up in DAVID at all.
 B.
-When replicated, KEGG Terms (using DAVID) were not mentioned in the results of the paper.
-Shows inconsistency due to the usage of the server.
-Some terms using clusterprofiler in R were mentioned.
-| **2 (medium).** GO results from the original manuscript are somewhat similar to reproduced analysis.
+No | **3 (high).** 3 (high).
+The article receives a score of 3 solely based on GO results.
+GO results from the original manuscript are somewhat similar to reproduced analysis in DAVID.
 The FDR values however are inconsistent.
-KEGG results were highly different.
 The list of genes as input into DAVID analysis is unclear.
-They should have mentioned that both up and down-regulated proteins were considered.
+The paper should have mentioned that both up and down-regulated proteins were considered for DAVID analysis.
+Another important question is why were two different tools used for GO and KEGG when both can be performed in DAVID or clusterprofiler?
 |
@@
 GO Results are not reproducible and are highly inconsistent.
+It was not made clear which results were obtained with DAVID or IPA.
 |

### 2023-12-05 manuscript/manuscript.Rmd
slight edits

@@
 text(0.7,1:7-0.5,labels=score1)
-text(1.9,1:10-0.5,labels=score2)
-text(3.1,1:3-0.5,labels=score3)
+text(1.9,1:9-0.5,labels=score2)
+text(3.1,1:4-0.5,labels=score3)
 
@@
 B.
-No | **3 (high).** 3 (high).
-The article receives a score of 3 solely based on GO results.
+No | **3 (high).** The article receives a score of 3 solely based on GO results.
 GO results from the original manuscript are somewhat similar to reproduced analysis in DAVID.

### 2023-12-06 manuscript/letter.Rmd
mods to letter

@@
 were reproducible using the authors' own methods.
-We found only 3/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies.
+We found only 4/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies.
 After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
@@
 This is a prime example of "link decay", a phenomenon where internet based resources are lost over time, which has been raised as a significant and ongoing problem for bioinformatic reproducibility [@Hennessey2013-tj].
+Of course DAVID isn't the only webserver without the option of preservation; our analysis of the latest NAR Webserver issue shows 28% (23/81) of webserver tools lacked source code or other features to allow preservation [@Seelow2023-xk].
 

### 2023-12-08 manuscript/manuscript.Rmd
data

@@
 popular DAVID suite were reproducible.
-We find that only three articles exhibited a high degree of concordance, while seven exhibited major
+We find that only four articles exhibited a high degree of concordance, while seven exhibited major
 discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
@@
 library("RColorBrewer")
+#library("kableExtra") there is a bug with kableextra - waiting for patch
+library("gt")
 
@@
 
+We screened reporting criteria of these articles, including whether the version of the application
+is given, whether p-values underwent correction for multiple testing, whether a background list was
+described and whether gene selection was described unambiguously.
+
 The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
@@
 
-Briefly, we used the DAVID "Analysis Wizard" and pasted in the provided gene list into the text box,
+Briefly, we used the DAVID v6.8 "Analysis Wizard" and pasted in the provided gene list into the text box,
 selecting this as "gene list", together with the appropriate gene identifier type.
@@
 
-After conducting this reproduction study according to the methods described in the respective articles,
-the 20 articles were classified based on the degree of reproducibility (**Figure 1**).
-Seven studies were classified as low reproducibility, while 10 were classified as medium and just three
+As reproducibility is associated with the quality of documentation we collected key methodological information
+about how the enrichment analysis was done and present in Table 1.
+From this, we can see that only seven articles reported the version of DAVID that was used.
+Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
+Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
+the remaining nine articles did not take into consideration the importance of p-value correction for
+minimising false positives.
+Of the 20 articles, none mentioned explicitly that a background list was used for enrichment analysis.
+We also examined whether the gene selection criteria was unambiguously clear, which was true in 16 articles.
+
+```{r,table1,echo=FALSE}
+
+x <- read.table("../data/checklist.tsv",header=TRUE,row.names=1)
+
+colnames(x) <- gsub("_"," ",colnames(x))
+x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
+#x %>% kbl(caption = "Table 1. Article reporting checklist results.") %>% kable_paper("hover", full_width = F)
+
+```
+
+We attempted reproduction according to the methods described in the respective articles, with the limitation
+that we had to use DAVID v6.8 as v6.7 was no longer available.
+The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
+Seven studies were classified as low reproducibility, while 9 were classified as medium and just four
 were classified as high.
@@
 Although this reproducibility study is relatively small, there are a few key observations.
-We observe that only 3/20(15%) of studies achieved a high degree of reproducibility.
+We observe that only 4/20(20%) of studies achieved a high degree of reproducibility.
 In a way, this is unexpected, given the simplicity of using DAVID.
@@
 * Ambiguity around whether a custom background gene set was used.
+
+* Differences between DAVID version 6.7 and 6.8.
 
@@
 This is significantly longer than using a web-based tool that provides results in just a few
-minutes, but we argue the time investment is well spent.
+minutes, but we consider the time investment is well spent.
 It ensures analytical reproducibility, assists with auditing and makes the process of reproduction

### 2023-12-08 manuscript/manuscript.Rmd
closer to submission

@@
 In a way, this is unexpected, given the simplicity of using DAVID.
-Users provide a list of gene idetifiers in a text box and select some options.
+Users provide a list of gene identifiers in a text box and select some options.
 There are several posible explanations for this inconsistency:
@@
 
-The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly describes a decade ago [@Hennessey2013-tj].
+The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a decade ago [@Hennessey2013-tj].
 We are worried that a reliance on web tools can leave researchers in violation of their data management responsibilities, as preservation mandates include software as well as original research data [@NHMRC2019-wu ; @Harvard_2023-cl ].
@@
 Expression data from untreated and Aza treated AML3 cells.
-Accession Number SRP038101.
-
-Data are available under the terms of the GNU General Public License v3.0.
+NCBI Sequence Read Archive accession Number SRP038101.
 

### 2023-12-11 manuscript/manuscript.Rmd
manuscript updates

@@
 multiple contributors.
-Literate programming using R Markdown or Jupyter notebooks provides optimal documentation by embedding
+Literate programming using R Markdown or Jupyter notebooks enables thorough documentation by embedding
 free text explanations together with the code and outputs such as charts and tables.
@@
 drawn @Wijesooriya2022-li.
+Authors can prevent these reporting decifiencies providing a checklist of methodological details similar
+to ours (**Table 2**).
+Checklists like these are a useful resource as they are simple for journals to enforce and reviewers to check.
 
@@
 
-The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a decade ago [@Hennessey2013-tj].
-We are worried that a reliance on web tools can leave researchers in violation of their data management responsibilities, as preservation mandates include software as well as original research data [@NHMRC2019-wu ; @Harvard_2023-cl ].
+The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a
+decade ago [@Hennessey2013-tj].
+We are worried that a reliance on web tools can leave researchers in violation of their data management
+responsibilities, as preservation mandates include software as well as original research data
+[@NHMRC2019-wu ; @Harvard_2023-cl ].
 
-With this, we must reconsider whether web-based tools are valid for scientific publications if they are at risk
-of becoming unavailable due to link decay.
+With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
+risk of becoming unavailable wthout notice due to link decay.
 Some progress is being made to make web-based point-and-click tools more reproducible.
@@
 The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
-downloads for all past versions[@Ge2020-qe], which aids future reproducibility.
+downloads for all past versions[@Ge2020-qe], facilitating preservation and aiding future reproducibility.
 Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"

### 2023-12-11 manuscript/manuscript.Rmd
ms edits

@@
 date: "`r Sys.Date()`"
-output:
-  rmdformats::robobook:
-  html_document:
-    toc: true
-    toc_float: true
-    code_folding: hide
+output: word_document
     fig_width: 5
@@
 While bioinformatics and other branches of computational research are theoretically amenable to
-complete reproducibility, in reality this is rarely achieved due to lack of detail provided in methods
+complete reproducibility, this is rarely achieved due to lack of detail provided in methods
 sections, lack of shared data and lack of code @Peng2011-ia.
@@
 In this article, we focus on enrichment analysis, which is one of the most used methods in
-computational biology, and involves the summarisation of omics data to reflect biological changes,
+computational biology and involves the summarisation of omics data to reflect biological changes,
 such as the detection of pathway activation in development or disease
@@
 Its 11 related journal articles have received a total of 37,035 citations from 27,129
-separate PubMed articles (as at 29th June 2023).
+separate PubMed articles (as of 29th June 2023).
 Examination of related keywords in PubMed suggests the number of articles describing
@@
 selecting this as "gene list", together with the appropriate gene identifier type.
-Typically this was the official gene symbol.
+Typically, this was the official gene symbol.
 The species was specified as *Homo sapiens*.
-Next we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
+Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
 page.
@@
 Articles rated 1 showed overall poor reproducibility.
-For example most GO and/or KEGG terms mentioned in the results and discussion were not confirmed
+For example, most GO and/or KEGG terms mentioned in the results and discussion were not confirmed
 upon reproduction, and the conclusions of the study were overall not supported.
@@
 
-As reproducibility is associated with the quality of documentation we collected key methodological information
+As reproducibility is associated with the quality of documentation, we collected key methodological information
 about how the enrichment analysis was done and present in Table 1.
@@
 Of the 20 articles, none mentioned explicitly that a background list was used for enrichment analysis.
-We also examined whether the gene selection criteria was unambiguously clear, which was true in 16 articles.
+We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.
 
@@
 that are no longer reproducible with the original tools.
-These observations highlight the need for better methodological reporting to reduce ambiguity in methods sections,
+These observations highlight the need for better methodological reporting to reduce ambiguity in methods sections
 and suggest that web-based tools without long-term reproducibility features are no longer suitable for inclusion in
@@
 gene expression and elaborate upon previous findings [@Lund2014-et], which mainly focused on
-the relationship between RNA expression and DNA methylation, but did not show results of a
+the relationship between RNA expression and DNA methylation but did not show results of a
 transcriptome-only enrichment analysis.
@@
 
-For quality control the number of reads per sample was calculated, and found to be in the range of
+For quality control the number of reads per sample was calculated and found to be in the range of
 11-16M.
@@
 Users provide a list of gene identifiers in a text box and select some options.
-There are several posible explanations for this inconsistency:
+There are several possible explanations for this inconsistency:
 
@@
 drawn @Wijesooriya2022-li.
-Authors can prevent these reporting decifiencies providing a checklist of methodological details similar
+Authors can prevent these reporting deficiencies providing a checklist of methodological details similar
 to ours (**Table 2**).
@@
 responsibilities, as preservation mandates include software as well as original research data
-[@NHMRC2019-wu ; @Harvard_2023-cl ].
+[@NHMRC2019-wu ; @Harvard_2023-cl].
 
 With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
-risk of becoming unavailable wthout notice due to link decay.
+risk of becoming unavailable without notice due to link decay.
 Some progress is being made to make web-based point-and-click tools more reproducible.
@@
 The Degust differential expression suite is a graphical interface to popular
-expression tools like edgeR and limma, and allows users to download the computer script to facilitate
+expression tools like edgeR and limma and allows users to download the computer script to facilitate
 future non-interactive reproducibility @Powell2019-lv.
 The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
-downloads for all past versions[@Ge2020-qe], facilitating preservation and aiding future reproducibility.
+downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
 Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"

### 2023-12-12 manuscript/letter.Rmd
abstract

@@
 (*) Corresponding author: mark.ziemann@burnet.edu.au
+
+## Abstract
+
+Web-based tools are an indispensable resource for conducting genomic and bioinformatics analysis, but if they disappear
+it can leave reproducibility gaps in research projects.
+Unless such tools allow for preservation of key materials for reproduction, researchers may unwittingly be in breach of
+data management mandates.
+
+## Main text
 

### 2023-12-13 manuscript/letter.Rmd
2013 data

@@
 bibliography: references.bib
-csl: plos-computational-biology.csl
+csl: genome-biology.csl
 ---
 
-Alt title: "No publication without preservation: the argument against web-based tools in genomics"
-
-Alt title: "Time to abandon (most) web-based tools in genomics"
-
 [Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
+[Nhi Nguyen<sup>2</sup>](https://orcid.org/XXXX-XXXX-XXXX-XXXX),
 [Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)
@@
 Unless such tools allow for preservation of key materials for reproduction, researchers may unwittingly be in breach of
-data management mandates.
+data retention mandates.
 
@@
 Of course DAVID isn't the only webserver without the option of preservation; our analysis of the latest NAR Webserver issue shows 28% (23/81) of webserver tools lacked source code or other features to allow preservation [@Seelow2023-xk].
+A look at NAR Webservers from 2013 shows 68% of tools did not provide any preservation options, and 35% (33/95) are no longer online.
 

### 2023-12-13 manuscript/letter.Rmd
add nhi

@@
 [Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
-[Nhi Nguyen<sup>2</sup>](https://orcid.org/XXXX-XXXX-XXXX-XXXX),
+[Nhi Ngoc Lan Nguyen<sup>2</sup>](https://orcid.org/0009-0007-5710-4117),
 [Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)
@@
 it can leave reproducibility gaps in research projects.
-Unless such tools allow for preservation of key materials for reproduction, researchers may unwittingly be in breach of
+Unless such tools allow for preservation of key materials for future reproduction, researchers may unwittingly be in breach of
 data retention mandates.
@@
 After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
-This was disturbing, as according to analysis of PubMed citations, some 20,000 articles citing DAVID enrichment tools will no longer be reproducible with
+This was disturbing, as according to an analysis of PubMed citations, some 20,000 articles citing DAVID enrichment tools will no longer be reproducible with
 the original tools, some articles just one or two years old.

### 2023-12-14 manuscript/manuscript.Rmd
figure updates

@@
 date: "`r Sys.Date()`"
-output: word_document
+output:
+  rmdformats::robobook:
+  html_document:
+    toc: true
+    toc_float: true
+    code_folding: hide
     fig_width: 5
@@
 
+pdf("fig1.pdf",width=6,height=5)
+barplot(unlist(lapply(l,length)),col="lightgray",
+  xlab="reproducibility score",ylab="no. articles",
+  ylim=c(0,11))
+grid()
+text(0.7,1:7-0.5,labels=score1)
+text(1.9,1:9-0.5,labels=score2)
+text(3.1,1:4-0.5,labels=score3)
+pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
+pc2 <- paste(pc,"%",sep="")
+text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)
+dev.off()
+
 ```
@@
 
+pdf("fig2.pdf",width=8,height=6)
+cols <- brewer.pal(5, "Set2")
+cols2 <- adjustcolor( cols, alpha.f = 0.7)
+plot(0,type='n',axes=FALSE,ann=FALSE, xaxs="i", yaxs="i", xlim=c(0, 11), ylim=c(-3, 8.5))
+
+polygon(c(0,6,6,0),c(8.5,8.5,7.5,7.5), col=cols2[1], border = NA)
+text( 3, 8.0, lab="1. Install Docker: 7 mins",cex=1.5)
+
+polygon(c(0.5,7.5,7.5,0.5),c(7.0,7.0,6.0,6.0), col=cols2[2], border = NA)
+text( 3.9, 6.5, lab="2. Validate example: 14 mins",cex=1.5)
+
+polygon(c(1,8.5,8.5,1),c(5.5,5.5,4.5,4.5), col=cols2[3], border = NA)
+text( 4.8, 5.0, lab="3. Create GitHub repo: 29 mins",cex=1.5)
+
+polygon(c(1.5,10.5,10.5,1.5),c(4,4,3,3), col=cols2[4], border = NA)
+text( 6.1, 3.5, lab="4. Build custom Docker image: 45 mins",cex=1.5)
+
+polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
+text( 6.3, 2, lab="5. Customize R Markdown: 3-12 hrs",cex=1.5)
+
+polygon(c(2.5,9.5,9.5,2.5),c(1,1,0,0),col="lightgray", border = NA)
+text( 6.1, 0.5, lab="6. Update repository: 70 mins",cex=1.5)
+
+polygon(c(3,11,11,3),c(-0.5,-0.5,-1.5,-1.5), col=cols2[1], border = NA)
+text( 7.0, -1, lab="7. Update Docker image: 40 mins",cex=1.5)
+
+polygon(c(3.5,12,12,3.5),c(-2,-2,-3,-3), col=cols2[2], border = NA)
+text( 7.6, -2.5, lab="8. Long Term Archiving: 180 mins",cex=1.5)
+dev.off()
+
 ```
@@
 ")
+
+grViz("digraph flowchart {
+      node [fontname = Helvetica, shape = rectangle]
+      tab1 [label = '@@1']
+      tab2 [label = '@@2']
+      tab3 [label = '@@3']
+      tab4 [label = '@@4']
+      tab5 [label = '@@5']
+      tab6 [label = '@@6']
+      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
+      tab3 -> tab5 -> tab6 }
+      [1]: 'Fetch RNA-seq from DEE2'
+      [2]: 'Quality control'
+      [3]: 'Differential expression'
+      [4]: 'Over-representation analysis'
+      [5]: 'Functional class scoring'
+      [6]: 'Euler diagram'
+") %>%
+  DiagrammeRsvg::export_svg() %>%
+  charToRaw() %>%
+  rsvg::rsvg_pdf("fig3.pdf")
 

### 2023-12-14 manuscript/example_pub.Rmd
adding example scipt to manuscript folder; this generates PDF figures for the journal article

--- 2023-06-22 example.Rmd
@@
 
+pdf("fig5a.pdf",width=5,height=5)
+plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
+  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
+  pch=19,cex=2,col=cols, main="MDS plot")
+text(cmdscale(dist(t(mx))), labels=colnames(mx) )
+dev.off()
+
 ```
@@
 mtext(HEADER)
+
+pdf("fig5b.pdf",width=5,height=5)
+plot(log10(def$baseMean) ,def$log2FoldChange,
+  cex=0.6,pch=19,col="darkgray",
+  main="5-azacitidine treatment in AML3 cells",
+  xlab="log10(basemean)",ylab="log2(fold change)")
+points(log10(sigf$baseMean) ,sigf$log2FoldChange,
+  cex=0.6,pch=19,col="red")
+mtext(HEADER)
+dev.off()
 
@@
 
+pdf("fig5c.pdf",width=7,height=5)
+barplot(c(topdn2,topup2),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES",
+  cex.main=0.9)
+mtext("ORA test")
+dev.off()
+
 ```
@@
 
+pdf("fig5d.pdf",width=5,height=5)
+barplot(c(fgsea_dn_vec,fgsea_up_vec),
+  horiz=TRUE,las=1,cex.names=0.65,col=cols,
+  main="top DE Reactomes",
+  xlab="ES",
+  cex.main=0.9)
+mtext("FCS test")
+dev.off()
+
 ```
@@
 par(mfrow=c(2,1))
+pdf("fig5e.pdf",width=6,height=5)
 plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
 plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))
+dev.off()
 

### 2023-12-15 manuscript/letter.Rmd
update

@@
 were reproducible using the authors' own methods.
-We found only 4/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies.
+We found only 4/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies [@Bora2023-ab].
 After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
@@
 Scripted workflows are perhaps the most suitable solution for enabling reproduction in the long-term, but require expertise in computer programming.
-We have written a protocol for extremely reproducible enrichment analysis in R, designed for novice bioinformaticians to fill in this gap [@Ziemann2023-kz], but we understand this isn't for everyone.
+We have written a protocol for extremely reproducible enrichment analysis in R, designed for novice bioinformaticians to fill in this gap [@Bora2023-ab], but we understand this isn't for everyone.
 ShinyGO is another potential remedy, as it provides most of the necessary features in a point-and-click web interface but also allows users to download all historical versions of the tool to be run as

### 2023-12-16 manuscript/letter.Rmd
important reference

@@
 the original tools, some articles just one or two years old.
-This is a prime example of "link decay", a phenomenon where internet based resources are lost over time, which has been raised as a significant and ongoing problem for bioinformatic reproducibility [@Hennessey2013-tj].
+This is a prime example of "link decay", a phenomenon where internet based resources are lost over time, which has been raised as a significant and ongoing problem for bioinformatic reproducibility [@Hennessey2013-tj; @Kern2020-iz].
 Of course DAVID isn't the only webserver without the option of preservation; our analysis of the latest NAR Webserver issue shows 28% (23/81) of webserver tools lacked source code or other features to allow preservation [@Seelow2023-xk].

### 2024-01-24 manuscript/manuscript.Rmd
shorten docker command / make mac compatible

@@
 sudo apt update && sudo apt install docker.io -y  # 30 sec
-sudo docker run -it --entrypoint /bin/bash mziemann/enrichment_recipe #60 sec
-Rscript -e 'rmarkdown::render("example.Rmd")' # 90 sec
-exit
+sudo docker run -it --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")'
 docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant

### 2024-01-24 manuscript/manuscript.Rmd
forgot timing

@@
 sudo apt update && sudo apt install docker.io -y  # 30 sec
-sudo docker run -it --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")'
+sudo docker run -it --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
 docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant

### 2024-01-29 manuscript/manuscript.Rmd

--- 2024-01-24 manuscript/manuscript.Rmd
@@
 sudo apt update && sudo apt install docker.io -y  # 30 sec
-sudo docker run -it --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")'
+sudo docker run --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
 docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant

### 2024-02-29 manuscript/manuscript.Rmd
addressing reviewer points

@@
 ---
-title: "A recipe for extremely reproducible enrichment analysis (and why we need it)"
+title: "A recipe for highly reproducible enrichment analysis (and why we need it)"
 date: "`r Sys.Date()`"
@@
 
-Unreliable and irreproducible research is a significant problem that wastes resources and risks
+Irreproducible research is a significant problem that wastes resources and risks
 undermining the public perception of science.
@@
 discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
-As the tool version used is no longer available, all articles using this tool pre-2021 (~20,800 studies)
-including this sample of 20 articles are no longer able to be reproduced with the original tools.
-Based on this, we suggest that results from web-based tools without long-term preservation features should
-not be included in scientific publications due to the threat of link decay and short reproducibility horizon.
-Relying exclusively on webtools for analysis may also be in breach of institutional and funder data management mandates.
-We advocate for the adoption of extremely reproducible research workflows, and we provide a detailed protocol
-for how to achieve it using a combination of best practices including literate programming, version control,
-containerisation, documentation and persistent sharing of data and software.
+To help combat such issues, we provide a detailed beginner-friendly enrichment analysis protocol
+exemplifying current best practices including literate programming, version control, containerisation,
+documentation and persistent sharing of data and code.
+Finally, we discuss how web-based tools without long-term preservation features might inadvertently put
+researchers in breach of their data retention mandates.
 
@@
 avoiding any "file not found" errors @Peng2011-ia.
-Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
+Lastly, reanalysis can be made as simple as possible by creating a "main" script that executes analysis
 for each part sequentially, making the entire manuscript reproducible by executing just one or a few
@@
 Taken together, computational reproducibility is a spectrum, and the *five pillars*
-framework has been described as a way of achieving extreme reproducibility @Ziemann2023-dg.
+framework has been described as a way of achieving a high degree of reproducibility @Ziemann2023-dg.
 Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
@@
 These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
-However, there are concerns that inadvertent misuse of these methods leads to unreliable results.
+However, there are concerns that inadvertent misuse of these methods leads to invalid results.
 These potential problems include (i) the lack of a correct background for over-representation tests,
@@
 the validity of the methods and results.
-We are concerned about the reliability of published enrichment analyses, as web-based and
+We are concerned about the validity of published enrichment analyses, as web-based and
 point-and-click tools appear to be far more popular than computer script-based methods for this type
@@
 
-A solution to this is to use containerisation to provide a complete scripted workflow with prescribed
+One potential solution to this is to use containerisation to provide a complete scripted workflow with prescribed
 links to research data, external databases and complete record of parameter selection.
@@
 
-We address this firstly by providing working templates for Docker image and R Markdown scripts for
+We address this firstly by providing working example templates for Docker image and R Markdown scripts for
 common transcriptome and enrichment analysis routines.
@@
 barplot(unlist(lapply(l,length)),col="lightgray",
-  xlab="reproducibility score",ylab="no. articles",
+  xlab="Reproducibility score",ylab="no. articles",
   ylim=c(0,11))
 
-grid()
+#grid()
 text(0.7,1:7-0.5,labels=score1)
@@
 
-We have developed a step-by-step protocol for how to conduct an extremely reproducible
+We have developed a step-by-step protocol for how to conduct a highly reproducible
 enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
@@
 <p style="text-align: center;">
-*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.*
+*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.
+Ubuntu 22.04 was used.*
 </p>
@@
 
-## Concluding remarks
+## Discussion
 
@@
 decade ago [@Hennessey2013-tj].
-We are worried that a reliance on web tools can leave researchers in violation of their data management
-responsibilities, as preservation mandates include software as well as original research data
+We are worried that a reliance on web tools that lack open source code or containers can leave
+researchers in violation of their data management responsibilities, as preservation mandates
+include software as well as original research data
 [@NHMRC2019-wu ; @Harvard_2023-cl].
@@
 downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
-Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"
+Still, such tools leave documentation and reproducibility gaps which are resolved with the highly reproducible
 approach detailed here.
 
-Although we hope that extremely reproducible enrichment analysis will become more widespread, we do not
+Although the recipe described here is designed for gene expression analysis, it is modifiable to accommodate
+other omics, for example genomic, epigenomic and proteomic data, given the names of analytes measured are
+consistent with the annotation databases.
+
+Although we hope that highly reproducible enrichment analysis will become more widespread, we do not
 anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.

### 2024-03-01 manuscript/manuscript.Rmd
ms work

@@
 The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles using the DAVID version specified in the articles (March-May 2022) [@Sherman2022-hf].
+sections of the 20 articles using DAVID version 6.8 (conducted March-May 2022) [@Sherman2022-hf].
 
@@
 
-Although this reproducibility study is relatively small, there are a few key observations.
+There are some notable limitations of this reproduction work.
+The sample size of 20 is relatively small.
+The articles were selected randomly from a list of articles that cited DAVID, and they are mostly from
+low and mid-tier journals (mean SJR 2022: 1.2, SD=0.68) and the results may not be representative of
+higher-tier journal articles.
+This is not a complete reproduction of all work in those papers; we only reproduced the DAVID functional enrichment analysis,
+which typically is just one component of an article that can have several other parts, including
+experimental validation of key findings.
+Therefore, if we identified that the enrichment analysis was not reproducible, it does not automatically mean that
+all results and conclusions are flawed.
+
+Despite these limitations, there are a few key observations.
 We observe that only 4/20(20%) of studies achieved a high degree of reproducibility.

### 2024-03-05 manuscript/manuscript.Rmd
intro work

@@
 ---
-title: "A recipe for highly reproducible enrichment analysis (and why we need it)"
+title: "Tackling irreproducible enrichment analysis"
 date: "`r Sys.Date()`"
@@
 We find that only four articles exhibited a high degree of concordance, while seven exhibited major
-discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
-To help combat such issues, we provide a detailed beginner-friendly enrichment analysis protocol
+discrepancies, which we mainly ascribe to deficiencies in methodological reporting and lack of
+preservation of past DAVID versions.
+Better reporting and a long-term view of reproducibility are needed, which are satisfied with
+scripted workflows and shared code.
+We provide a step-by-step beginner-friendly enrichment analysis protocol
 exemplifying current best practices including literate programming, version control, containerisation,
 documentation and persistent sharing of data and code.
-Finally, we discuss how web-based tools without long-term preservation features might inadvertently put
-researchers in breach of their data retention mandates.
 
@@
 
-Reproducibility is essential to the scientific enterprise by confirming the validity of new discoveries
-[@Committee_on_Reproducibility_and_Replicability_in_Science2019-vb p1].
-According to a 2016 survey of *Nature* readers, 52% of respondents agreed that there was a "significant
-crisis" of reproducibility @Baker2016-hm.
-While bioinformatics and other branches of computational research are theoretically amenable to
-complete reproducibility, this is rarely achieved due to lack of detail provided in methods
-sections, lack of shared data and lack of code @Peng2011-ia.
-Indeed, even if data and code are shared, reproducibility still cannot be guaranteed due to the
-differences in computing environments over time, and the availability of dependencies and
-web-tools @Perkel2020-io.
-
-To address this, many articles have been written to endorse the use of "virtual machines" or
-"containerisation" to control the computing environment including the operating system and dependencies
-along with the research code to facilitate a high degree of reproducibility [@Perkel2020-io;
-@Stein2010-of; @Dudley2010-ua; @Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg;
-@Lewis2016-qh; @Da_Veiga_Leprevost2017-dw; @Pasquier2018-rt; @Kulkarni2018-km; @Hung2019-et;
-@Brito2020-ld; @Piccolo2021-ud].
-Containers also allow for an easy installation process, as all working dependencies are shipped together
-and can be installed with a single command.
-
-In addition to shared code, data and containerisation, there are a few other ways researchers can make
-their work more reproducible.
-Thorough documentation makes the reproducibility process a great deal easier and more
-accessible, especially to novice bioinformaticians.
-Version control keeps track of code and documentation changes over time and manages the edits made by
-multiple contributors.
-Literate programming using R Markdown or Jupyter notebooks enables thorough documentation by embedding
-free text explanations together with the code and outputs such as charts and tables.
-Code and data can be linked. This means that the code "knows" the location of the data,
-avoiding any "file not found" errors @Peng2011-ia.
-Lastly, reanalysis can be made as simple as possible by creating a "main" script that executes analysis
-for each part sequentially, making the entire manuscript reproducible by executing just one or a few
-commands, and without any manual steps @Lasser2020-mm.
-Taken together, computational reproducibility is a spectrum, and the *five pillars*
-framework has been described as a way of achieving a high degree of reproducibility @Ziemann2023-dg.
-Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
-over a decade.
-
-In this article, we focus on enrichment analysis, which is one of the most used methods in
+Functional enrichment analysis is one of the most frequently used methods in
 computational biology and involves the summarisation of omics data to reflect biological changes,
@@
 These potential problems include (i) the lack of a correct background for over-representation tests,
-(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
+(ii) lack of p-value correction for multiple testing and (iii) general lack of reporting detail
 [@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
@@
 the validity of the methods and results.
-We are concerned about the validity of published enrichment analyses, as web-based and
-point-and-click tools appear to be far more popular than computer script-based methods for this type
-of work @Wijesooriya2022-li.
-Although web-based tools are convenient, they are a potential reproducibility concern.
-Incomplete reporting of methods, options and parameters is common.
-Algorithms and functional annotation sets undergo regular updates, and previous versions are
-typically not made available, hampering later reproducibility.
-In some cases, the entire web resource can become unavailable, and hence irreproducible due to
-a phenomenon known as link decay @Hennessey2013-tj.
 
-A small number of enrichment analysis applications have been developed to have beneficial properties
-of being "containerised" and with archived historical versions [@Ge2020-qe; @Perampalam2020-qm].
-This helps reproducibility greatly but is not a complete solution as it lacks information about how
-the input data was used and a record of any parameter selections.
-
-One potential solution to this is to use containerisation to provide a complete scripted workflow with prescribed
-links to research data, external databases and complete record of parameter selection.
-When designed correctly, these containers encapsulate bioinformatics workflows and could ensure their
-computational reproducibility for decades to come with a few simple commands.
-There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
-existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
-a lack of step-by-step guides and templates aimed at non-expert bioinformaticians.
-
-We address this firstly by providing working example templates for Docker image and R Markdown scripts for
-common transcriptome and enrichment analysis routines.
-Secondly, we provide step-by-step written and video guides to help users customise these templates for
-their own work, deploy the analysis, verify the results, and share the materials.
-
-However, before recommending or prescribing such workflows, it is necessary
-to assess *directly* whether current practices are adequate, by trying to reproduce the enrichment
-analyses of previously published studies, and examining whether the conclusions of those articles
-are supported or not.
-In the following section we investigate the computational reproducibility of enrichment analyses
-shown in a sample of 20 open-access journal articles published in 2019.
-This should provide a sound basis for recommendations.
+To determine whether web based enrichment analysis is reproducible, we systematically examine the
+reproducibilty of DAVID enrichment analysis in 20 published articles.
+The results show that poor reporting and lack of preservation of web-tools is to blame for poor
+reproducibility.
+There is a need to bring enrichment analysis in line with bioinformatics best practices, so
+we present a beginner-friendly protocol that guides researchers through the process of
+literate scripting, version control, containerisation, documentation and code/data sharing.
 

### 2024-03-06 manuscript/manuscript.Rmd
intro

@@
 Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
-the validity of the methods and results.
+the validity of the methods and results [@Sandve2013-lk].
 
-To determine whether web based enrichment analysis is reproducible, we systematically examine the
-reproducibilty of DAVID enrichment analysis in 20 published articles.
-The results show that poor reporting and lack of preservation of web-tools is to blame for poor
-reproducibility.
-There is a need to bring enrichment analysis in line with bioinformatics best practices, so
-we present a beginner-friendly protocol that guides researchers through the process of
-literate scripting, version control, containerisation, documentation and code/data sharing.
+To determine whether web based enrichment analysis is reproducible, here we systematically examine the
+reproducibilty of DAVID enrichment analysis in 20 published articles, investigate root causes of
+reproduction failure and describe some potential solutions.
 
@@
 The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles using DAVID version 6.8 (conducted March-May 2022) [@Sherman2022-hf].
+sections of the 20 articles using DAVID v6.8 as at this time v6.7 was not available
+(conducted March-May 2022) [@Sherman2022-hf].
 
@@
 The species was specified as *Homo sapiens*.
-Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results”
+Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results"
 page.
@@
 
-```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility of enrichment analysis in a sample of 20 journal articles."}
+```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1.
+Examining reproducibility of enrichment analysis in a sample of 20 journal articles.
+Beside the PubMed Central identifier, the reported version of DAVID used in the original publication (blank if none given)."}
 
-score1 <- c("PMC6405693","PMC6425008","PMC6535219","PMC6539328","PMC6542760","PMC6561911","PMC6663624")
+score1 <- c("PMC6405693 (v6.7)","PMC6425008 (v6.8)","PMC6535219","PMC6539328 (v6.8)",
+  "PMC6542760","PMC6561911 (v6.8)","PMC6663624")
 score2 <- c("PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
-  "PMC6582306","PMC6333352","PMC6526186","PMC6607402")
-score3 <- c("PMC6349697","PMC6444048","PMC6587650", "PMC6368841")
+  "PMC6582306","PMC6333352 (v6.8)","PMC6526186","PMC6607402 (v6.7)")
+score3 <- c("PMC6349697 (v6.8)","PMC6444048","PMC6587650", "PMC6368841")
 l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)
@@
 #grid()
-text(0.7,1:7-0.5,labels=score1)
-text(1.9,1:9-0.5,labels=score2)
-text(3.1,1:4-0.5,labels=score3)
+text(0.7,1:7-0.5,labels=score1,cex=0.85)
+text(1.9,1:9-0.5,labels=score2,cex=0.85)
+text(3.1,1:4-0.5,labels=score3,cex=0.85)
 
@@
 ## A five-pillars approach to addressing reproducibility of enrichment analysis
+
+The results show that poor reporting and lack of preservation of web-tools is to blame for poor
+reproducibility.
+There is a need to bring enrichment analysis in line with bioinformatics best practices, so
+we present a beginner-friendly protocol that guides researchers through the process of
+literate scripting, version control, containerisation, documentation and code/data sharing.
 

### 2024-03-12 manuscript/manuscript.Rmd
ms work

@@
 ---
+
+Alt: *Tackling irreproducible pathway enrichment analysis*
 
@@
 We find that only four articles exhibited a high degree of concordance, while seven exhibited major
-discrepancies, which we mainly ascribe to deficiencies in methodological reporting and lack of
+discrepancies.
+These inconsistencies can mostly be ascribed to deficiencies in methodological reporting and lack of
 preservation of past DAVID versions.
-Better reporting and a long-term view of reproducibility are needed, which are satisfied with
-scripted workflows and shared code.
+Better reporting and a long-term view of reproducibility are needed.
+Scripted workflows and shared code are a potential solution to this problem, although there is a
+steep learning curve for novice analysts.
 We provide a step-by-step beginner-friendly enrichment analysis protocol
@@
 The basis for these selection criteria is that this combination represents the most common application
-of enrichment analysis in our previous study, which could be reproduced without the need to
-reanalyse the raw dataset from scratch. These articles are listed in the bibliography
+of enrichment analysis according to our previous study, which could be reproduced without the need to
+reanalyse raw data. These articles are listed in the bibliography
 [@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
@@
 The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
-special attention to preventing gene name errors from occurring @Zeeberg2004-na.
+special attention to preventing gene name errors from occurring [@Zeeberg2004-na,Ziemann2016-pv, Abeysooriya2021-uz].
 The reproducibility analysis was performed using the respective methodologies stated in the methods
-sections of the 20 articles using DAVID v6.8 as at this time v6.7 was not available
+sections of the 20 articles using DAVID v6.8, as at this time v6.7 was not available
 (conducted March-May 2022) [@Sherman2022-hf].
@@
 Articles rated 1 showed overall poor reproducibility.
-For example, most GO and/or KEGG terms mentioned in the results and discussion were not confirmed
+For example, less than 50% of GO and/or KEGG terms mentioned in the results and discussion were confirmed
 upon reproduction, and the conclusions of the study were overall not supported.
@@
 
-As reproducibility is associated with the quality of documentation, we collected key methodological information
-about how the enrichment analysis was done and present in Table 1.
+As reproducibility is closely associated with the quality of documentation, we collected key methodological
+information about how the enrichment analysis was done and present in Table 1.
 From this, we can see that only seven articles reported the version of DAVID that was used.
@@
 minimising false positives.
-Of the 20 articles, none mentioned explicitly that a background list was used for enrichment analysis.
+Of the 20 articles, none mentioned explicitly that a custom background list was used for enrichment analysis.
 We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.
@@
 The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
-Seven studies were classified as low reproducibility, while 9 were classified as medium and just four
+Seven studies were classified as low reproducibility, while nine were classified as medium and just four
 were classified as high.
@@
 
-Taken together, the ability to reproduce DAVID enrichment analysis using the supplementary gene
+The ability to reproduce DAVID enrichment analysis using the supplementary gene
 lists was highly variable.
-As DAVID version 6.8 has been retired, these and approximately 20,000 other PubMed articles have enrichment analyses
-that are no longer reproducible with the original tools.
-These observations highlight the need for better methodological reporting to reduce ambiguity in methods sections
-and suggest that web-based tools without long-term reproducibility features are no longer suitable for inclusion in
-scientific publications.
+The version of DAVID used is likely to be a big contributor to irreproducibility as we know that there are major
+differences in the underlying annotation set (knowledgebase) used by version 6.7 and 6.8.
+This is supported by the fact that none of the v6.7 studies received a high reproduicibility score.
+On the other hand these versions did not receive regular updates, so the knowledgebases were static for relatively
+long times; v6.7 (Jan. 2010 - Oct 2020) and v6.8 (Oct 2016 - Jun 2022).
+We can then deduce that knowledgebase version is likely not the cause of irreproducibility for PMC6561911,
+PMC6539328 and PMC6425008; studies that reportedly used v6.8 but received low reproducibility scores.
+
+We note that DAVID version 6.8 was retired in June 2022, so these 20 studies are currently no longer reproducible
+with the original tools.
+The implications of this are outlined in the Discussion section below.
 
@@
 
-The results show that poor reporting and lack of preservation of web-tools is to blame for poor
-reproducibility.
-There is a need to bring enrichment analysis in line with bioinformatics best practices, so
-we present a beginner-friendly protocol that guides researchers through the process of
-literate scripting, version control, containerisation, documentation and code/data sharing.
+Our results emphasise the need for better methodological reporting to reduce ambiguity in methods sections
+and highlight that webtools can suddenly become unavailable.
+Although there are various solutions to these issues (refer to the Discussion section), here we will outline
+a potential solution whose design is guided by current best practices
+[@Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Ziemann2023-dg].
+Our aim is to provide a beginner-friendly guide to enrichment analysis that will be
+reproducible for many years to come.
 
@@
 sudo apt update && sudo apt install docker.io -y  # 30 sec
-sudo docker run --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
+sudo docker run --platform linux/amd64 mziemann/enrichment_recipe
+Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
 docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
@@
 With ORA, we observed 201 up-regulated and 99 down-regulated gene sets (FDR<0.05).
-We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
+We prioritised enrichment results by enrichment score, a surrogate measure for effect size,
 which tends to highlight smaller and more specific gene sets undergoing more drastic differential
@@
 
-The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a
-decade ago [@Hennessey2013-tj].
+The problem of bioinformatic resources including web services becoming unavailable is not a new one, being
+raised over a decade ago [@Hennessey2013-tj] and revisited in 2020 [@Kern2020-iz].
 We are worried that a reliance on web tools that lack open source code or containers can leave
@@
 expression tools like edgeR and limma and allows users to download the computer script to facilitate
-future non-interactive reproducibility @Powell2019-lv.
+future non-interactive reproducibility [@Powell2019-lv].
 The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
@@
 Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
-badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
+badging of articles that comply with best practices [@Kidwell2016-pm], and recognition of research rigour
 in track record evaluations.

### 2024-03-14 manuscript/manuscript.Rmd
ms workl

@@
 ---
-title: "Tackling irreproducible enrichment analysis"
+title: "Tackling irreproducible pathway enrichment analysis"
 date: "`r Sys.Date()`"
@@
 ---
-
-Alt: *Tackling irreproducible pathway enrichment analysis*
 
@@
 undermining the public perception of science.
-Previous work has highlighted that published enrichment analyses frequently suffer from statistical
-and reporting flaws.
+Previous work has highlighted that published pathway enrichment analyses frequently suffer from
+statistical and reporting flaws.
 We sought to determine whether this translates into irreproducibility by examining whether the
@@
 polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
-text( 6.3, 2, lab="5. Customize R Markdown: 3-12 hrs",cex=1.5)
+text( 6.3, 2, lab="5. Customise R Markdown: 3-12 hrs",cex=1.5)
 
@@
 
+Furthermore, our reproducibility assessment is based on "head-counting" pathways that meet an arbitrary significance threshold.
+This binary classification of results is less powerful as opposed to comparison of standardised effect sizes.
+
 Despite these limitations, there are a few key observations.

### 2024-04-15 manuscript/manuscript.Rmd
updates

@@
 These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
-However, there are concerns that inadvertent misuse of these methods leads to invalid results.
-These potential problems include (i) the lack of a correct background for over-representation tests,
-(ii) lack of p-value correction for multiple testing and (iii) general lack of reporting detail
+However, there are concerns that inadvertent misuse of these methods leads to invalid results
 [@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
-Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
-the validity of the methods and results [@Sandve2013-lk].
+According to a recent survey of 186 articles, incorrect methods are very common [@Wijesooriya2022-li].
+For example, a suitable background list for over-representation tests is reported in only 4.1% of studies.
+Correction of p-values to account for multiple testing was reported in only 54% of studies.
+This survey also revealed a general lack of reporting detail around other aspects of enrichment analysis,
+such as the tool version used.
 
-To determine whether web based enrichment analysis is reproducible, here we systematically examine the
-reproducibilty of DAVID enrichment analysis in 20 published articles, investigate root causes of
-reproduction failure and describe some potential solutions.
+We hypothesise that the quality of enrichment analysis studies could be dramatically improved by adopting
+established simple rules for reproducible computing [@Sandve2013-lk], and specific guidelines for
+reporting enrichment analysis [@Zhao2023-ok].
+
+Our aim is to determine how reproducible published enrichment analyses are.
+To achieve this, we systematically examined the reproducibility of 20 published articles that used a web-based
+enrichment tool.
+We investigate root causes of reproduction failure and describe some potential solutions.
 
@@
 
-The results presented in the paper were compared to those we obtained, recording any
-discrepancies.
-Additionally, the statements made in the abstract, results, and discussion sections were carefully
-examined and cross-checked with our results to understand whether those assertions survived reproduction
-or not.
+The pathway enrichment results presented in the paper were compared to those we obtained.
+It would be ideal to compare standardised effect sizes (fold enrichment scores) between original and
+reproduction results, however these were reported in the main text of only 3/20 articles.
+Therefore we adopted a binary classification system, based on head-counting pathways that were presented as
+statistically significant in the original article.
+We defined a pathway as confirmed when it reached a similar level of statistical significance
+upon reproduction.
+We define "similar significance level" to be a value within 50% of the value presented in the original article.
+For example if an article shows a pathway with FDR=0.04, then an acceptable FDR in reproduction could be in
+the range of 0.02 to 0.06.
+Additionally, statements made in the abstract, results, and discussion sections based on specific pathway
+enrichment results were carefully examined and cross-checked with our results to understand whether those
+assertions survived reproduction or not.
 
-Each article was critically evaluated and given a score of one (1) to three (3).
-Articles rated 1 showed overall poor reproducibility.
-For example, less than 50% of GO and/or KEGG terms mentioned in the results and discussion were confirmed
-upon reproduction, and the conclusions of the study were overall not supported.
-Articles were rated 2 when at least 50% of the terms are observed as significant after replication
-according to the original study's significance threshold, and most conclusions were supported.
-Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
-shown in the article observed again in reproduction, and all conclusions were supported in reproduction.
+Each article was critically evaluated and given a score of one (1; low) to three (3; high).
+Articles were rated 1 when <50% of pathways shown in the results were confirmed
+upon reproduction, and most conclusions derived from enrichment analysis were not supported by reproduction.
+Articles were rated 2 when ≥50% of pathways shown in the results were confirmed
+upon reproduction, and most conclusions derived from the pathway enrichment analysis were supported.
+Articles were rated 3 when ≥80% of terms shown in the original article were observed again in
+reproduction with a similar significance value, and all conclusions derived from the pathway analysis were
+supported in reproduction.
 
@@
 As reproducibility is closely associated with the quality of documentation, we collected key methodological
-information about how the enrichment analysis was done and present in Table 1.
-From this, we can see that only seven articles reported the version of DAVID that was used.
+information about how the enrichment analysis was done and reported (Table 1).
+Only seven articles reported the version of DAVID that was used.
 Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
@@
 colnames(x) <- gsub("_"," ",colnames(x))
+x[,3]=NULL
 x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
@@
 @Grolemund2017-zl.
+R Markdown supports languages other than R, including Python, Shell, Julia and others.
 The ability to provide extended descriptions is helpful for providing contextualising information about
@@
 The articles were selected randomly from a list of articles that cited DAVID, and they are mostly from
-low and mid-tier journals (mean SJR 2022: 1.2, SD=0.68) and the results may not be representative of
+low and mid-tier journals (mean Scimago Journal Rank for 2022: 1.2, SD=0.68) and the results may not be representative of
 higher-tier journal articles.
@@
 
-Furthermore, our reproducibility assessment is based on "head-counting" pathways that meet an arbitrary significance threshold.
-This binary classification of results is less powerful as opposed to comparison of standardised effect sizes.
+Furthermore, our reproducibility assessment is based on "head-counting" pathways that meet an arbitrary significance range.
+This binary classification of reproduction results is less powerful as opposed to comparison of standardised effect sizes [@Lakens2013-fx; @Goodman2016-ie].
+For DAVID analysis, the fold enrichment score could be reported as a type of effect size.
+Unfortunately, fold enrichment scores are infrequently reported, so it was not a feasible approach for a systematic and representative
+reproduction study like this.
+In future, we recommend researchers use both significance values and enrichment scores to prioritise and report
+enrichment results.
 
@@
 * Differences between DAVID version 6.7 and 6.8.
+
+* Ambiguity around which tool was used, as some articles described using other tools in addition to DAVID
+but did not label figures and tables with the tool used.
 
@@
 simple and fast (Box 1), which should aid the early identification of errors.
+Using a literate scripting approach such as R Markdown, Jupyter or Quarto notebook, enables very high
+levels of workflow transparency including inspection of intermediate data objects, and evidence that
+the published methods were followed.
 
@@
 approach detailed here.
+
+Although container-based workflows are frequently described as being a complete solution to computational
+reproducibility, they have some limitations.
+For example, building a Docker image relies on a network of servers providing software packages which may work now
+but not in a decade's time.
+Therefore building such images has a short reproducibility horizon.
+The Guix package and container manager, together with the Software Heritage code repository enables reproducible
+image builds, which enhances the transparency of compute environments, and lengthening the reproducibility horizon
+[@Vallet2022-gw].
+It is also not guaranteed that Docker images published now will be compatible with future Docker versions,
+whereas Guix can build an older version of itself.
 

### 2024-04-15 manuscript/letter.Rmd
enrichment analysis -> gene list analysis

@@
 [ @Timmons2015-ex ; @Wijesooriya2022-li].
-Last year we sought to test whether a small group of articles published in 2019 involving enrichment analysis with the DAVID tool [@Sherman2022-hf]
+Last year we sought to test whether a small group of articles published in 2019 involving gene list analysis with the DAVID tool [@Sherman2022-hf]
 were reproducible using the authors' own methods.
-We found only 4/20 enrichment studies had a high degree of reproduction, while 7/20 had severe discrepancies [@Bora2023-ab].
+We found only 4/20 gene list studies had a high degree of reproduction, while 7/20 had severe discrepancies [@Bora2023-ab].
 After our pilot study completed we were surprised to hear that the version of DAVID used for all of these studies (v6.8) would no longer be available from June 2022 onwards.
-This was disturbing, as according to an analysis of PubMed citations, some 20,000 articles citing DAVID enrichment tools will no longer be reproducible with
+This was disturbing, as according to an analysis of PubMed citations, some 20,000 articles citing DAVID tools will no longer be reproducible with
 the original tools, some articles just one or two years old.
@@
 Scripted workflows are perhaps the most suitable solution for enabling reproduction in the long-term, but require expertise in computer programming.
-We have written a protocol for extremely reproducible enrichment analysis in R, designed for novice bioinformaticians to fill in this gap [@Bora2023-ab], but we understand this isn't for everyone.
+We have written a protocol for extremely reproducible gene list analysis in R, designed for novice bioinformaticians to fill in this gap [@Bora2023-ab], but we understand this isn't for everyone.
 ShinyGO is another potential remedy, as it provides most of the necessary features in a point-and-click web interface but also allows users to download all historical versions of the tool to be run as

### 2024-05-23 misc/pval_dist.R
pvals

@@ -1,0 +1,99 @@
+---
+title: "Examination of test statistics and p-values"
+author: "Mark Ziemann & Anusuiya Bora"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    toc: true
+    toc_float: true
+    code_folding: hide
+    fig_width: 5
+    fig_height: 5
+theme: cosmo
+---
+
+Source: https://github.com/markziemann/enrichment_recipe
+
+## Introduction
+
+The issue is that the 50% difference which we used as a classifier for
+reproduction of pathway p-values is not great when p-values are very small.
+
+One possible solution is to use the test statistics instead.
+We can use the chi-squared test for this case, with repeated tests with
+different hypothetical tests, to ascertain the relationship between p-values
+and the test statistics.
+
+## Running the chi-squared test
+
+In the next chunk, we are running a set of chi-squred tests, starting off with
+balanced data and making it progressively more skewed, which will degrease
+p-values.
+
+```{r,chi1}
+
+l <- lapply(0:40, function(x) {
+  tulip <- c(100+x, 100-x)
+  res <- chisq.test(tulip, p = c(1/2, 1/2))
+  c("stat"=res$stat,"p"=res$p.value)
+})
+
+res <- as.data.frame(do.call(rbind,l))
+res$logp <- -log10(res$p)
+
+```
+
+Now we can make a plot of the relationship of the -log10(p-values) and the
+test statistics.
+
+```{r,plot1}
+
+plot(res$logp,res$`stat.X-squared`)
+
+```
+
+## Testing the bounds
+
+Now we'll test plus and minus 50% the of the test statistic value.
+
+```{r,bound1}
+
+bound=0.5
+
+res$lower <- res$`stat.X-squared` - ( res$`stat.X-squared` * bound )
+res$upper <- res$`stat.X-squared` + ( res$`stat.X-squared` * bound )
+
+res
+
+```
+
+That looks good but the bounds can be quite wide, approximately two orders of
+magnitude in either direction.
+
+```{r,bound2}
+
+bound=0.25
+
+res$lower <- res$`stat.X-squared` - ( res$`stat.X-squared` * bound )
+res$upper <- res$`stat.X-squared` + ( res$`stat.X-squared` * bound )
+
+res
+
+```
+
+Now it is 1.3 orders of magnitude.
+
+```{r,bound2}
+
+bound=0.15
+
[…]

### 2024-05-23 misc/pval_dist.Rmd
rename, tidy up

---
title: "Examination of test statistics and p-values"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/enrichment_recipe

## Introduction

The issue is that the 50% difference which we used as a classifier for
reproduction of pathway p-values is not great when p-values are very small.

One possible solution is to use the test statistics instead.
We can use the chi-squared test for this case, with repeated tests with
different hypothetical tests, to ascertain the relationship between p-values
and the test statistics.

```{r,libs}

library("kableExtra")

```

## Running the chi-squared test

In the next chunk, we are running a set of chi-squred tests, starting off with
balanced data and making it progressively more skewed, which will decrease
p-values.

```{r,chi1}

l <- lapply(0:80, function(x) {
  tulip <- c(200+x, 200-x)
  res <- chisq.test(tulip, p = c(1/2, 1/2))
  c("stat"=res$stat,"p"=res$p.value)
})

res <- as.data.frame(do.call(rbind,l))
res$logp <- -log10(res$p)

```

Now we can make a plot of the relationship of the chi-squared test stats againse the -log10(p-values).

```{r,plot1}

plot(res$`stat.X-squared`,  res$logp ,
  xlab="chi-squared test stat", ylab="-log10(p-value)")

```

## Testing the bounds

Now we'll test plus and minus 50% the of the test statistic value.

```{r,bound1}

bound=0.5

res$lower <- res$`stat.X-squared` - ( res$`stat.X-squared` * bound )
res$upper <- res$`stat.X-squared` + ( res$`stat.X-squared` * bound )

plot(res$`stat.X-squared`, res$logp,
  xlab="test stat", ylab="-log10(p)",
  xlim=c(0,max(res$upper)))

points(res$lower, res$logp, col="blue")
points(res$upper, res$logp, col="blue")

res$logp_lower <- approx(x = res$`stat.X-squared`, y =res$logp, xout = res$lower)[["y"]]
res$logp_upper <- approx(x = res$`stat.X-squared`, y =res$logp, xout = res$upper)[["y"]]

res$p_lower <- 10^-res$logp_lower
res$p_upper <- 10^-res$logp_upper

format(res,digits=3) %>%
  kbl(caption="p-value ranges considered reproduced with a bound of 50% around the test statistic") %>%
  kable_paper("hover", full_width = F)

```

That looks good but the bounds can be quite wide, approximately two orders of
magnitude in either direction.

```{r,bound2}

bound=0.2

res$lower <- res$`stat.X-squared` - ( res$`stat.X-squared` * bound )
res$upper <- res$`stat.X-squared` + ( res$`stat.X-squared` * bound )

plot(res$`stat.X-squared`, res$logp,
  xlab="chisq test stat", ylab="-log10(p)",
  xlim=c(0,max(res$upper)))

points(res$lower, res$logp, col="blue")
points(res$upper, res$logp, col="blue")

res$logp_lower <- approx(x = res$`stat.X-squared`, y =res$logp, xout = res$lower)[["y"]]
res$logp_upper <- approx(x = res$`stat.X-squared`, y =res$logp, xout = res$upper)[["y"]]

res$p_lower <- 10^-res$logp_lower
res$p_upper <- 10^-res$logp_upper

format(res,digits=3) %>%
  kbl(caption="p-value ranges considered reproduced with a bound of 20% around the test statistic") %>%
  kable_paper("hover", full_width = F)

```

## Session information

```{r,session}

sessionInfo()

```

### 2024-05-23 misc/pval_dist.R

@@ -1,99 +1,0 @@
----
-title: "Examination of test statistics and p-values"
-author: "Mark Ziemann & Anusuiya Bora"
-date: "`r Sys.Date()`"
-output:
-  html_document:
-    toc: true
-    toc_float: true
-    code_folding: hide
-    fig_width: 5
-    fig_height: 5
-theme: cosmo
----
-
-Source: https://github.com/markziemann/enrichment_recipe
-
-## Introduction
-
-The issue is that the 50% difference which we used as a classifier for
-reproduction of pathway p-values is not great when p-values are very small.
-
-One possible solution is to use the test statistics instead.
-We can use the chi-squared test for this case, with repeated tests with
-different hypothetical tests, to ascertain the relationship between p-values
-and the test statistics.
-
-## Running the chi-squared test
-
-In the next chunk, we are running a set of chi-squred tests, starting off with
-balanced data and making it progressively more skewed, which will degrease
-p-values.
-
-```{r,chi1}
-
-l <- lapply(0:40, function(x) {
-  tulip <- c(100+x, 100-x)
-  res <- chisq.test(tulip, p = c(1/2, 1/2))
-  c("stat"=res$stat,"p"=res$p.value)
-})
-
-res <- as.data.frame(do.call(rbind,l))
-res$logp <- -log10(res$p)
-
-```
-
-Now we can make a plot of the relationship of the -log10(p-values) and the
-test statistics.
-
-```{r,plot1}
-
-plot(res$logp,res$`stat.X-squared`)
-
-```
-
-## Testing the bounds
-
-Now we'll test plus and minus 50% the of the test statistic value.
-
-```{r,bound1}
-
-bound=0.5
-
-res$lower <- res$`stat.X-squared` - ( res$`stat.X-squared` * bound )
-res$upper <- res$`stat.X-squared` + ( res$`stat.X-squared` * bound )
-
-res
-
-```
-
-That looks good but the bounds can be quite wide, approximately two orders of
-magnitude in either direction.
-
-```{r,bound2}
-
-bound=0.25
-
-res$lower <- res$`stat.X-squared` - ( res$`stat.X-squared` * bound )
-res$upper <- res$`stat.X-squared` + ( res$`stat.X-squared` * bound )
-
-res
-
-```
-
-Now it is 1.3 orders of magnitude.
-
-```{r,bound2}
-
-bound=0.15
-
[…]

### 2024-06-04 misc/pval_dist.Rmd
make p smaller

@@
 
-l <- lapply(0:80, function(x) {
+l <- lapply(0:150, function(x) {
   tulip <- c(200+x, 200-x)
@@
 
+## 20% window
+
+This is the range we will use for determining the similarity of p-values.
+
+The below table will be used to judge whether the p-values are reproducible.
+
 ```{r,bound2}

### 2025-08-29 example_10m.Rmd
making an example dataset for the 10 common mistakes article

--- 2023-04-26 example.Rmd
@@
 ---
-title: "An example of extremely reproducible enrichment analysis"
+title: "Supporting 10 common mistakes article"
 author: "Mark Ziemann & Anusuiya Bora"
@@
 
-This guide is a Rmarkdown script that conducts differential expression and enrichment analysis,
-which are very popular workflows for transcriptome data.
+```{r,libs}
 
-This is meant to be a boilerplate template, which you can remix and modify to suit your analytical
-needs.
-
-In the code chunk below called `libs`, you can add and remove required R library dependancies.
-Check that the libraries listed here match the Dockerfile, otherwise you might get errors.
-
-```{r,libs}
+RhpcBLASctl::blas_set_num_threads(1)
 
@@
   library("clusterProfiler")
-  library("fgsea")
   library("eulerr")
   library("gplots")
+  library("mitch")
 })
@@
 
-## Enrichment using over-representation test
-
-I've compiled a reporting checklist:
-
-| Reporting criteria | Method/resource used |
-| --- | --- |
-| Origin of gene sets | Reactome (2023-03-06) |
-| Tool used | ClusterProfiler (check version at foot of report) |
-| Statistical test used | hypergeometric test |
-| P-value correction for multiple comparisons | FDR method |
-| Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
-| ORA directionality | Separate tests for up- and down-regulation |
-| Data availability | via DEE2 at accession SRP038101 (human) |
-| Other parameters | Min gene set size of 10 |
-
-Get the gene sets loaded in R.
+## Get gene sets
 
@@
 # from https://reactome.org/download/current/ReactomePathways.gmt.zip
-genesets2 <- read.gmt("ReactomePathways_2023-03-06.gmt")
-gsets <- gmtPathways("ReactomePathways_2023-03-06.gmt")
-
-```
-
-Now filter the gene names into three lists, up-regulated, down-regulated and
-background.
-Background is simply all genes that were detected.
-
-```{r,subset1}
-
-defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
-defup <- unique(sapply(strsplit(defup," "),"[[",2))
-
-defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
-defdn <- unique(sapply(strsplit(defdn," "),"[[",2))
-
-bg <- rownames(def)
-bg <- unique(sapply(strsplit(bg," "),"[[",2))
-
-```
-
-Enrichment firstly with upregulated genes
-
-```{r,oraup}
-
-ora_up <- as.data.frame(enricher(gene = defup ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
-
-ora_up$geneID <- NULL
-ora_up <- subset(ora_up,p.adjust<0.05 & Count >=10)
-ora_ups <- rownames(ora_up)
-
-gr <- as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_up$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_up$BgRatio,"/"),"[[",2))
-
-ora_up$es <- gr/br
-ora_up <- ora_up[order(-ora_up$es),]
-ora_up$Description=NULL
-
-head(ora_up) %>%
-  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topup2 <- rev(head(ora_up$es,10))
-names(topup2) <- rev(head(ora_up$ID,10))
-
-```
-
-Now with the downregulated genes
-
-```{r,oradn}
-
-n_pw=10
-
-ora_dn <- as.data.frame(enricher(gene = defdn ,
-  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
-  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))
-
-ora_dn$geneID <- NULL
-ora_dn <- subset(ora_dn,p.adjust<0.05 & Count >=10)
-ora_dns <- rownames(ora_dn)
-
-gr <- as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_dn$GeneRatio,"/"),"[[",2))
-
-br <- as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",1)) /
-  as.numeric(sapply(strsplit(ora_dn$BgRatio,"/"),"[[",2))
-
-ora_dn$es <- gr/br
-ora_dn <- ora_dn[order(-ora_dn$es),]
-ora_dn$Description=NULL
-
-head(ora_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment") %>%
-  kable_paper("hover", full_width = F)
-
-topdn2 <- head(ora_dn$es,n_pw)
-names(topdn2) <- head(ora_dn$ID,n_pw)
-
-```
-
-Make a barplot
-
-```{r,topbarplot2}
-
-par(mar=c(5,20,5,1))
-
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
-
-barplot(c(topdn2,topup2),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE Reactomes",
-  xlab="ES",
-  cex.main=0.9)
-
-mtext("ORA test")
+genesets2 <- read.gmt("ref/ReactomePathways_2023-03-06.gmt")
+gsets <- gmtPathways("ref/ReactomePathways_2023-03-06.gmt")
 
@@
 
-def2 <- def
-def2$genename <- sapply(strsplit(rownames(def2)," "),"[[",2)
-def2 <- def2[,c("stat","genename")]
-def2 <- aggregate(. ~ genename, def2, sum)
-stat <- def2$stat
-names(stat) <- def2$genename
-fgseaRes <- fgsea(pathways=gsets, stats=stat, minSize=10, nPermSimple = 10000)
-fgseaRes <- fgseaRes[order(fgseaRes$pval),]
+gt <- def
+gt$genename <- sapply(strsplit(rownames(gt)," "),"[[",2)
+gt <- gt[,"genename",drop=FALSE]
+gt$id <- rownames(gt)
+m <- mitch_import(x=def,DEtype="deseq2",geneTable=gt)
+mres <- mitch_calc(x=m,genesets=gsets,minsetsize=5,cores=4,priority="significance")
 
-```
+mres_sig_up <- head(subset(mres$enrichment_result,s.dist>0),10)
+mres_sig_dn <- head(subset(mres$enrichment_result,s.dist<0),10)
 
-```{r,fgsea_sig}
-
-fgsea_up <- subset(fgseaRes,padj<0.05 & ES>0)
-fgsea_ups <- fgsea_up$pathway
-fgsea_dn <- subset(fgseaRes,padj<0.05 & ES<0)
-fgsea_dns <- fgsea_dn$pathway
-
-fgsea_up <-  data.frame(fgsea_up[order(-fgsea_up$ES),])
-
-fgsea_dn <-  data.frame(fgsea_dn[order(fgsea_dn$ES),])
-
-head(fgsea_up,n_pw) %>%
-  kbl(row.names = FALSE, caption="FGSEA:Top upregulated pathways in Aza treatment") %>%
+mres_sig_up %>%
+  kbl(caption="Top upregulated pathways when prioritised by FDR") %>%
   kable_paper("hover", full_width = F)
 
-head(fgsea_dn,n_pw) %>%
-  kbl(row.names = FALSE, caption="FGSEA:Top downregulated pathways in Aza treatment") %>%
+mres_sig_dn %>%
+  kbl(caption="Top downregulated pathways when prioritised by FDR") %>%
   kable_paper("hover", full_width = F)
 
-fgsea_up <- head(fgsea_up,n_pw)
-fgsea_up_vec <- fgsea_up$ES
-names(fgsea_up_vec) <- fgsea_up$pathway
-fgsea_up_vec <- rev(fgsea_up_vec)
+mres <- mitch_calc(x=m,genesets=gsets,minsetsize=5,cores=4,priority="effect")
 
-fgsea_dn <- head(fgsea_dn,n_pw)
-fgsea_dn_vec <- fgsea_dn$ES * -1
-names(fgsea_dn_vec) <- fgsea_dn$pathway
+mres_es_up <- head(subset(mres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05),10)
+mres_es_dn <- head(subset(mres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05),10)
+
+mres_es_up %>%
+  kbl(caption="Top upregulated pathways when prioritised by ES (discarding any with FDR>0.05)") %>%
+  kable_paper("hover", full_width = F)
+
+mres_es_dn %>%
+  kbl(caption="Top upregulated pathways when prioritised by ES (discarding any with FDR>0.05)") %>%
+  kable_paper("hover", full_width = F)
 
@@
 
-## Compare pathway results
-
-```{r,venn1,fig.width=6}
-
-v1 <- list("ORA up"=ora_ups, "FCS up"=fgsea_ups)
-v2 <- list("ORA dn"=ora_dns, "FCS dn"=fgsea_dns)
-par(mar=c(10,10,10,10))
-par(mfrow=c(2,1))
-plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
-plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))
-
-```
-
-Jaccard index
-
-```{r,jaccard}
-
-jaccard <- function(a, b) {
-    intersection = length(intersect(a, b))
-    union = length(a) + length(b) - intersection
-    return (intersection/union)
-}
-
-ora <- c(ora_ups,ora_dns)
-fcs <- c(fgsea_ups,fgsea_dns)
-
-jaccard(ora,fcs)
-
-```
-
-## Drill in to a specific pathway
-
-In this case the top upregulated and downregulated pathways
-
-```{r,eplot}
-
-upname <- head(fgsea_up,1)$pathway
-plotEnrichment(gsets[[upname]], stat, gseaParam = 1, ticksSize = 0.2)
-
-dnname <- head(fgsea_dn,1)$pathway
-plotEnrichment(gsets[[dnname]], stat, gseaParam = 1, ticksSize = 0.2)
-
-```
-
-Now a heatmap of these gene sets.
-
-```{r,heat}
-
-# reads per million normalisation
-rpm <- apply(mxf,2, function(x) { x/sum(x) * 1000000} )
-colnames(rpm) <- c("UNTR1","UNTR2","UNTR3","TRT1","TRT2","TRT3")
-gnames_up <- gsets[[which(names(gsets) == upname)]]
-gnames_dn <- gsets[[which(names(gsets) == dnname)]]
-gene_ids <- rownames(mxf)
-gene_names <- sapply(strsplit(gene_ids," "),"[[",2)
-rpm_up <- rpm[which(gene_names %in% gnames_up),]
-rownames(rpm_up) <- sapply(strsplit(rownames(rpm_up)," "),"[[",2)
-rpm_dn <- rpm[which(gene_names %in% gnames_dn),]
-rownames(rpm_dn) <- sapply(strsplit(rownames(rpm_dn)," "),"[[",2)
-colsidecols <- c("blue","blue","blue","red","red","red")
-heatmap.2(rpm_up,scale="row",trace="none",margins=c(10,15),main=upname,ColSideColors=colsidecols)
-heatmap.2(rpm_dn,scale="row",trace="none",margins=c(10,15),main=dnname,ColSideColors=colsidecols)
-
-```
-
 ## Session information
@@
 
+save.image("example_10m.Rdata")
 sessionInfo()

### 2025-08-29 example.Rmd
thread control

@@
   library("gplots")
+  library("RhpcBLASctl")
 })
+
+RhpcBLASctl::blas_set_num_threads(1)
 
@@
 
+save.image("example.Rdata")
 sessionInfo()

### 2025-08-29 example_10m.Rmd
thread control

@@
 
-RhpcBLASctl::blas_set_num_threads(1)
-
 suppressPackageStartupMessages({
@@
   library("mitch")
+  library("RhpcBLASctl")
 })
+
+RhpcBLASctl::blas_set_num_threads(1)
 

### 2025-08-29 example_10m.Rmd
using data to support the 10 mistakes article

--- 2025-08-29 example_10m.Rmd
@@
   library("mitch")
+  library("fgsea")
+  library("RhpcBLASctl")
 })
+
+RhpcBLASctl::blas_set_num_threads(1)
 
@@
 
-## Enrichment analysis with functional class scoring
+## Mistake 1 - Raw p-values
+
+## Mistake 2 - Background
+
+## Mistake 4 - prioritisation by p-values
 
@@
 | Origin of gene sets | Reactome (2023-03-06) |
-| Tool used | FGSEA (check version at foot of report) |
-| Statistical test used | Kolmogorov-Smirnov test |
+| Tool used | FORA (check fgsea version at foot of report) |
+| Statistical test used | hypergeometric |
 | P-value correction for multiple comparisons | FDR method |
 | Background list | Genes with >=10 reads per sample on average across all samples |
-| Gene Selection Criteria | DESeq2 FDR<0.05 |
-| ORA directionality | Separate tests for up- and down-regulation |
+| Gene Selection | DESeq2 FDR<0.05 and log2FC>0 or log2FC<0 |
 | Data availability | via DEE2 at accession SRP038101 (human) |
-| Other parameters | Min gene set size of 10 |
+| Other parameters | Min gene set size of 5 |
 
-```{r,fgsea}
+```{r,mistake2}
+
+up <- unique(sapply(strsplit(rownames(subset(def, log2FoldChange>0 & padj<0.05))," "),"[[",2))
+
+dn <- unique(sapply(strsplit(rownames(subset(def, log2FoldChange<0 & padj<0.05))," "),"[[",2))
+
+bg <- unique(sapply(strsplit(rownames(def)," "),"[[",2))
+
+gt <- read.table("https://dee2.io/data/hsapiens/hsa_gene_info.tsv",header=TRUE)
+head(gt)
+
+all <- unique(gt$GeneSymbol)
+
+lapply(list(up,dn,bg,all),length)
+
+fres_up_bg <- fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf)
+fres_up_bg <- subset(fres_up_bg,padj<0.05)$pathway
+length(fres_up_bg)
+
+fres_up_all <- fora(pathways=gsets, genes=up, universe=all, minSize = 5, maxSize = Inf)
+fres_up_all <- subset(fres_up_all,padj<0.05)$pathway
+length(fres_up_all)
+
+fres_dn_bg <- fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
+fres_dn_bg <- subset(fres_dn_bg,padj<0.05)$pathway
+length(fres_dn_bg)
+
+fres_dn_all <- fora(pathways=gsets, genes=dn, universe=all, minSize = 5, maxSize = Inf)
+fres_dn_all <- subset(fres_dn_all,padj<0.05)$pathway
+length(fres_dn_all)
+
+v1 <- list("up"=fres_up_bg, "up*"=fres_up_all, "dn"=fres_dn_bg, "dn*"=fres_dn_all)
+
+par(mar=c(1,1,1,1))
+par(mfrow=c(1,1))
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
+
+v2 <- list("up"=fres_up_bg, "up*"=fres_up_all)
+plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))
+
+par(mar=c(5.1,4.1,4.1,2.1))
+
+```
+
+## Mistake 4 - prioritisation by p-values
+
+| Reporting criteria | Method/resource used |
+| --- | --- |
+| Origin of gene sets | Reactome (2023-03-06) |
+| Tool used | mitch (check version at foot of report) |
+| Statistical test used | Rank ANOVA (Kruskal Wallis) |
+| P-value correction for multiple comparisons | FDR method |
+| Background list | Genes with >=10 reads per sample on average across all samples |
+| Gene Scoring | DESeq2 Wald stat |
+| Data availability | via DEE2 at accession SRP038101 (human) |
+| Other parameters | Min gene set size of 5 |
+
+```{r,mistake4a}
 
@@
 
-```{r,topbarplot3}
+```{r,mistake4b,fig.height=6,fig.width=6}
 
-par(mar=c(5,20,5,1))
+par(mar=c(2,20,2,0))
 
-cols <- c(rep("blue",n_pw),rep("red",n_pw))
+mres_sig_up_vals <- -log(mres_sig_up$pANOVA)
+names(mres_sig_up_vals) <- mres_sig_up$set
 
-barplot(c(fgsea_dn_vec,fgsea_up_vec),
-  horiz=TRUE,las=1,cex.names=0.65,col=cols,
-  main="top DE Reactomes",
+mres_sig_dn_vals <- -log(mres_sig_dn$pANOVA)
+names(mres_sig_dn_vals) <- mres_sig_dn$set
+
+mres_es_up_vals <- mres_es_up$s.dist
+names(mres_es_up_vals) <- mres_es_up$set
+
+mres_es_dn_vals <- -1 * mres_es_dn$s.dist
+names(mres_es_dn_vals) <- mres_es_dn$set
+
+par(mfrow=c(2,1))
+
+barplot( mres_sig_up_vals,
+  horiz=TRUE,las=1,cex.names=0.65,col="gray",
+  main="Up Sig (-log10 p-value)",
+  xlab="-log10 p-value",
+  cex.main=0.9)
+
+barplot( mres_es_up_vals,
+  horiz=TRUE,las=1,cex.names=0.65,col="gray",
+  main="Up ES",
   xlab="ES",
@@
 
-mtext("FCS test")
+barplot( mres_sig_dn_vals,
+  horiz=TRUE,las=1,cex.names=0.65,col="gray",
+  main="Down Sig (-log10 p-value)",
+  xlab="-log10 p-value",
+  cex.main=0.9)
+
+barplot( mres_es_dn_vals,
+  horiz=TRUE,las=1,cex.names=0.65,col="gray",
+  main="Down ES",
+  xlab="-ES",
+  cex.main=0.9)
+
+par(mfrow=c(1,1))
+par(mar=c(5.1,4.1,4.1,2.1))
+
+```
+
+## Mistake 5 - gene set list size too small or long
+
+```{r,mistake5}
+
+bg <- unique(sapply(strsplit(rownames(def)," "),"[[",2))
+
+nrow(subset(def,padj<0.05 & log2FoldChange>0))
+nrow(subset(def,padj<0.05 & log2FoldChange<0))
+nrow(def)
+
+sizes <- c(50,100,200,300,400,500,750,1000,1500,2000,2500,3000,4000,5000,6000,7000)
+
+fup <- nrow(subset(fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf),padj<0.05))
+fdn <- nrow(subset(fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf),padj<0.05))
+
+nups <- unlist(lapply(sizes, function(i) {
+  genes <- unique(sapply(strsplit(rownames(head(subset(def, log2FoldChange>0),i))," "),"[[",2))
+  fres <- fora(pathways=gsets, genes=genes, universe=bg, minSize = 5, maxSize = Inf)
+  nrow(subset(fres,padj<0.05))
+}))
+
+ndns <- unlist(lapply(sizes, function(i) {
+  genes <- unique(sapply(strsplit(rownames(head(subset(def, log2FoldChange<0),i))," "),"[[",2))
+  fres <- fora(pathways=gsets, genes=genes, universe=bg, minSize = 5, maxSize = Inf)
+  nrow(subset(fres,padj<0.05))
+}))
+
+df <- data.frame("up"=nups,"dn"=ndns,row.names=sizes)
+df
+plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length")
+lines(sizes,df$dn,type="b",col="blue")
+
+plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",xlim=c(0,500))
+lines(sizes,df$dn,type="b",col="blue")
 

### 2025-08-30 example_10m.Rmd
mistake 6

@@
 
+Will try FORA and mitch.
+
 | Reporting criteria | Method/resource used |
@@
 ```
-
-## Mistake 4 - prioritisation by p-values
 
@@
 df <- data.frame("up"=nups,"dn"=ndns,row.names=sizes)
+df$tot <- df$up + df$dn
 df
+
 plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length")
@@
 
-plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",xlim=c(0,500))
+plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",
+  xlim=c(0,500), ylim=c(0,125))
 lines(sizes,df$dn,type="b",col="blue")
+
+```
+
+## Mistake 6 - combining up and downregulated genes into the same test.
+
+```{r,mistake6a}
+
+updn <- unique(c(up,dn))
+
+fres_updn_bg <- fora(pathways=gsets, genes=updn, universe=bg, minSize = 5, maxSize = Inf)
+fres_updn_bg <- subset(fres_updn_bg,padj<0.05)$pathway
+length(fres_updn_bg)
+
+length(fres_up_bg)
+length(fres_dn_bg)
+length(unique(c(fres_up_bg,fres_dn_bg)))
+
+v1 <- list("up"=fres_up_bg,"dn"=fres_dn_bg,"both"=fres_updn_bg)
+plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))
 

### 2025-09-01 example_10m.Rmd
shallow gene annotations

@@
   library("RhpcBLASctl")
+  library("msigdb")
+  library("ExperimentHub")
 })
@@
 
+## Mistake 8 - using shallow gene annotations
+
+```{r,m8a}
+
+msdb <- gmtPathways("msigdb.v2025.1.Hs.symbols.gmt")
+kegg <- msdb[grep("KEGG",names(msdb))]
+kegg_med <- msdb[grep("KEGG_MED",names(msdb))]
+kegg <- kegg[grep("KEGG_MED",names(kegg),invert=TRUE)] #separate KEGG LEG from KEGG MED
+gobp <- msdb[grep("GOBP",names(msdb))]
+reactome <- msdb[grep("REACTOME",names(msdb))]
+gslist <- list("KEGG"=kegg,"KEGGM"=kegg_med,"Reactome"=reactome,"GOBP"=gobp,"MSigDB"=msdb)
+
+# no. gene sets
+d1 <- unlist( lapply(gslist,length) )
+
+# total number of annotations
+d2 <- unlist( lapply(gslist,function(i) { length(unlist(i)) } ) )
+
+# median gene set size
+d3 <- unlist(lapply(gslist,function(i) { median(unlist(lapply(i,length))) } ))
+
+# number of genes with one or more functions
+d4 <- unlist( lapply(gslist,function(i) { length(unique(unlist(i))) } ))
+
+d5 <- unlist(lapply(gslist,function(i) { median(unlist(lapply(i,length))) } ))
+
+data.frame(d1,d2,d3,d4) %>%
+  kbl(caption="Gene set metrics") %>%
+  kable_paper("hover", full_width = F)
+
+kegg_up <- fora(pathways=kegg, genes=up, universe=bg, minSize = 5, maxSize = Inf)
+kegg_dn <- fora(pathways=kegg, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
+nrow(subset(kegg_up,padj<0.05))
+nrow(subset(kegg_dn,padj<0.05))
+sum(nrow(subset(kegg_up,padj<0.05)), nrow(subset(kegg_dn,padj<0.05) ))
+
+kegg_med_up <- fora(pathways=kegg_med, genes=up, universe=bg, minSize = 5, maxSize = Inf)
+kegg_med_dn <- fora(pathways=kegg_med, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
+nrow(subset(kegg_med_up,padj<0.05))
+nrow(subset(kegg_med_dn,padj<0.05))
+sum(nrow(subset(kegg_med_up,padj<0.05)), nrow(subset(kegg_med_dn,padj<0.05) ))
+
+reactome_up <- fora(pathways=reactome, genes=up, universe=bg, minSize = 5, maxSize = Inf)
+reactome_dn <- fora(pathways=reactome, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
+nrow(subset(reactome_up,padj<0.05))
+nrow(subset(reactome_dn,padj<0.05))
+sum(nrow(subset(reactome_up,padj<0.05)), nrow(subset(reactome_dn,padj<0.05) ))
+
+gobp_up <- fora(pathways=gobp, genes=up, universe=bg, minSize = 5, maxSize = Inf)
+gobp_dn <- fora(pathways=gobp, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
+nrow(subset(gobp_up,padj<0.05))
+nrow(subset(gobp_dn,padj<0.05))
+sum(nrow(subset(gobp_up,padj<0.05)), nrow(subset(gobp_dn,padj<0.05) ))
+
+msdb_up <- fora(pathways=msdb, genes=up, universe=bg, minSize = 5, maxSize = Inf)
+msdb_dn <- fora(pathways=msdb, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
+nrow(subset(msdb_up,padj<0.05))
+nrow(subset(msdb_dn,padj<0.05))
+sum(nrow(subset(msdb_up,padj<0.05)), nrow(subset(msdb_dn,padj<0.05) ))
+
+```
+
 ## Session information
